---
title: "Clean-Room TMLE Simulation Study: Results"
subtitle: "Comparing Estimators Under Good and Marginal PS Overlap"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 8
    fig-height: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE
)
library(ggplot2)
```

## Overview

This document presents the results of a Monte Carlo simulation study
evaluating five estimators for the average treatment effect (ATE) on
the risk-difference scale under two overlap scenarios.
The study uses the **cleanTMLE** package's staged clean-room workflow,
including outcome-blind plasmode simulation for TMLE specification
selection.

```{r load-results}
res <- readRDS("results/simulation_results.rds")
cfg <- res$config

cat(sprintf("Sample size per replicate: %d\n", cfg$n_obs))
cat(sprintf("Monte Carlo replicates:    %d\n", cfg$n_reps))
cat(sprintf("Plasmode replicates:       %d\n", cfg$plasmode_reps))
cat(sprintf("SL library:                %s\n", paste(cfg$sl_library, collapse = ", ")))
cat(sprintf("Cross-fitting folds:       %d\n", cfg$n_folds))
cat(sprintf("Total runtime:             %.0f minutes\n", res$runtime_minutes))
```


## Data-Generating Process

The DGP simulates a pharmacoepidemiology setting with 5 baseline
covariates (age, sex, biomarker, comorbidity, CKD), a binary
treatment with confounded assignment, and a binary outcome at 24
months with a true protective treatment effect.

The `overlap_strength` parameter controls the degree of confounding
and PS separation:

- **Scenario A (Good Overlap):** `overlap_strength = 0.5` — moderate
  confounding, PS concentrated in [0.2, 0.8].
- **Scenario B (Marginal Overlap):** `overlap_strength = 1.5` — strong
  confounding, PS extending to [0.04, 0.99].

```{r truths}
for (sc in names(res$truths)) {
  t <- res$truths[[sc]]
  cat(sprintf("%s: Risk(1)=%.4f, Risk(0)=%.4f, True RD=%.5f\n",
              res$scenarios[[sc]]$label, t$risk_1, t$risk_0, t$RD))
}
```


## Plasmode Candidate Selection

Before accessing the real outcome, cleanTMLE evaluates prespecified
TMLE candidates on plasmode-simulated data.  Each candidate specifies
a SuperLearner library and PS truncation threshold.  The candidate
with minimum RMSE is selected and locked.

```{r plasmode-tables}
for (sc in names(res$meta)) {
  meta <- res$meta[[sc]]
  cat(sprintf("\n### %s\n\n", meta$label))
  cat(sprintf("- PS range: [%.3f, %.3f]\n", meta$ps_range[1], meta$ps_range[2]))
  cat(sprintf("- ESS: %.1f%%\n", meta$ess_pct))
  cat(sprintf("- **Selected candidate: `%s`**\n", meta$selected))
  cat(sprintf("- Gate decision: **%s**\n\n", meta$gate))

  plas <- meta$plasmode
  knitr::kable(plas, digits = c(2, NA, 5, 5, 3, 5, 5, 0, 3),
               caption = sprintf("Plasmode metrics: %s", meta$label)) |>
    print()
  cat("\n")
}
```


## Monte Carlo Results

### Summary Tables

```{r summary-tables}
for (sc in names(res$summaries)) {
  truth <- res$truths[[sc]]$RD
  label <- res$scenarios[[sc]]$label
  cat(sprintf("\n#### %s (true RD = %.5f)\n\n", label, truth))
  knitr::kable(res$summaries[[sc]],
               digits = c(NA, 0, 5, 5, 5, 5, 5, 3, 5, 3, 5),
               caption = label) |>
    print()
  cat("\n")
}
```

### Key Findings

```{r findings}
sa <- res$summaries$good_overlap
sb <- res$summaries$marginal_overlap

crude_a_cov <- sa$coverage[sa$method == "Crude"]
tmle_a_cov  <- sa$coverage[sa$method == "TMLE"]
crude_b_cov <- sb$coverage[sb$method == "Crude"]
tmle_b_cov  <- sb$coverage[sb$method == "TMLE"]
tmle_b_bias <- sb$bias[sb$method == "TMLE"]
crude_b_bias <- sb$bias[sb$method == "Crude"]
```

1. **Crude estimation is biased in both scenarios** — coverage
   `r sprintf("%.1f%%", 100*crude_a_cov)` (A) and
   `r sprintf("%.1f%%", 100*crude_b_cov)` (B), far below the nominal 95%.

2. **TMLE achieves near-nominal coverage** in both scenarios:
   `r sprintf("%.1f%%", 100*tmle_a_cov)` (A) and
   `r sprintf("%.1f%%", 100*tmle_b_cov)` (B).

3. **Under marginal overlap, crude bias is
   `r sprintf("%.0f", abs(crude_b_bias/tmle_b_bias))`x larger than TMLE bias.**

4. **SE/SD ratios are near 1.0** for all adjusted methods, confirming
   that influence-curve standard errors are well-calibrated.

5. **Cross-fitted TMLE (TMLE_CF)** performs comparably to standard TMLE,
   with marginally lower bias.

6. **Matched-cohort TMLE (Match_TMLE)** combines the bias reduction of
   matching with the double robustness of TMLE, at the cost of wider
   CIs due to reduced sample size.

7. **Plasmode selection correctly adapted:** `glm_t01` (minimal truncation)
   for good overlap, `glm_t05` (more truncation) for marginal overlap.


## Visualisations

### Bias Comparison

```{r bias-plot, fig.cap = "Bias by method and scenario."}
plot_data <- do.call(rbind, lapply(names(res$summaries), function(sc) {
  df <- res$summaries[[sc]]
  df$scenario <- res$scenarios[[sc]]$label
  df
}))

ggplot(plot_data, aes(x = reorder(method, abs(bias)), y = bias,
                       fill = scenario)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(x = NULL, y = "Bias (estimate - truth)",
       title = "Estimator Bias by Scenario",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Coverage Comparison

```{r coverage-plot, fig.cap = "95% CI coverage by method and scenario."}
ggplot(plot_data, aes(x = reorder(method, coverage), y = coverage,
                       fill = scenario)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "red") +
  geom_hline(yintercept = c(0.936, 0.964), linetype = "dotted",
             colour = "grey50") +
  coord_flip() +
  scale_y_continuous(limits = c(0.3, 1.0), labels = scales::percent) +
  labs(x = NULL, y = "Coverage",
       title = "95% CI Coverage by Scenario",
       subtitle = "Dashed red = nominal 95%; dotted = Monte Carlo simulation interval (depends on rep count)",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### RMSE Comparison

```{r rmse-plot, fig.cap = "RMSE by method and scenario."}
ggplot(plot_data, aes(x = reorder(method, -rmse), y = rmse,
                       fill = scenario)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  labs(x = NULL, y = "Root Mean Squared Error",
       title = "RMSE by Scenario",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### SE Calibration

```{r se-sd-plot, fig.cap = "SE/SD ratio by method and scenario. Values near 1.0 indicate well-calibrated standard errors."}
ggplot(plot_data, aes(x = method, y = se_sd_ratio, colour = scenario,
                       shape = scenario)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 1.0, linetype = "dashed") +
  geom_hline(yintercept = c(0.9, 1.1), linetype = "dotted", colour = "grey50") +
  scale_y_continuous(limits = c(0.8, 1.15)) +
  labs(x = NULL, y = "Mean SE / Empirical SD",
       title = "Standard Error Calibration",
       subtitle = "Values near 1.0 = well-calibrated; <1 = anti-conservative",
       colour = NULL, shape = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Sampling Distribution of Estimates

```{r boxplots, fig.height = 7, fig.cap = "Distribution of point estimates across Monte Carlo replicates (count specified in the configuration table)."}
raw_data <- do.call(rbind, lapply(names(res$results), function(sc) {
  df <- as.data.frame(res$results[[sc]])
  df$scenario <- res$scenarios[[sc]]$label
  df$truth    <- res$truths[[sc]]$RD
  df
}))

ggplot(raw_data, aes(x = method, y = estimate, fill = method)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  geom_hline(aes(yintercept = truth), linetype = "dashed", colour = "red") +
  facet_wrap(~ scenario, scales = "free_y") +
  labs(x = NULL, y = "Point Estimate (Risk Difference)",
       title = "Sampling Distribution of ATE Estimates",
       subtitle = "Dashed red = true RD") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```


## Simulation Configuration

```{r config}
cat("n_obs:          ", cfg$n_obs, "\n")
cat("n_reps:         ", cfg$n_reps, "\n")
cat("n_truth:        ", cfg$n_truth, "\n")
cat("plasmode_reps:  ", cfg$plasmode_reps, "\n")
cat("sl_library:     ", paste(cfg$sl_library, collapse = ", "), "\n")
cat("n_folds:        ", cfg$n_folds, "\n")
cat("seed:           ", cfg$seed, "\n")
cat("runtime:        ", round(res$runtime_minutes, 1), "minutes\n")
```


## Conclusions

The simulation confirms that the cleanTMLE staged workflow produces
valid inference under both good and marginal propensity score overlap:

1. **TMLE is approximately unbiased** with near-nominal 95% coverage
   in both scenarios, while crude estimation suffers severe bias and
   undercoverage when confounding is strong.

2. **Influence-curve standard errors are well-calibrated** (SE/SD
   ratios within 3% of 1.0), supporting the use of Wald-type
   confidence intervals without bootstrap.

3. **Outcome-blind plasmode candidate selection** correctly adapts
   the TMLE specification to the data structure, choosing minimal
   truncation under good overlap and increased truncation under
   marginal overlap.

4. **Cross-fitted TMLE** provides comparable or slightly improved
   performance, guarding against overfitting of nuisance parameters.

5. **Matched-cohort TMLE** combines matching's covariate pruning with
   TMLE's double robustness, offering a middle ground between
   full-cohort TMLE (most efficient) and PS matching (most
   interpretable to clinicians).

6. Under marginal overlap, **PS matching** shows early signs of
   degradation (lower ESS, marginally wider CIs), while **IPTW**
   and **TMLE** remain stable.

```{r session-info}
sessionInfo()
```
