---
title: "Clean-Room Stage 1b: Exploratory Data Analysis"
subtitle: "RescueCo Trauma Registry – Covariate Diagnostics Before Design Stage"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    theme: cosmo
execute:
  echo: true
  warning: true
  message: true
params:
  results_dir: "rescueCo/results"
---

```{r setup}
#| label: setup
#| include: false
library(ggplot2)
library(knitr)

# Quarto runs from the .qmd file's directory, so navigate to project root
project_root <- normalizePath(file.path(getwd(), "..", ".."), mustWork = FALSE)
# If already at root (e.g. called via quarto_render with execute-dir), detect that
if (file.exists(file.path(getwd(), "rescueCo", "R", "utils.R"))) {
  project_root <- getwd()
} else if (file.exists(file.path(getwd(), "..", "..", "rescueCo", "R", "utils.R"))) {
  project_root <- normalizePath(file.path(getwd(), "..", ".."))
}
knitr::opts_knit$set(root.dir = project_root)
setwd(project_root)

source("rescueCo/R/utils.R")
cfg <- load_cr_config()
```

# Purpose

This EDA runs **after Stage 1** (cohort build) and **before Stage 2** (propensity
score design checks).
It inspects the covariate matrix (`W_matrix`) and treatment vector to catch
data-quality problems that would crash SuperLearner:

- `NA` / `NaN` / `Inf` / `-Inf` in numeric columns
- Constant or near-zero-variance columns
- Dimension mismatches between `A` and `W_matrix`
- Extreme values and distributional oddities

**No outcomes are examined here.**  This is a design-phase diagnostic only.

---

# Load Stage 1 Outputs

```{r load}
dat      <- load_stage_output("stage1_cohort.rds")
W_matrix <- load_stage_output("stage1_W_matrix.rds")
A        <- dat$A

cat("Cohort rows :", nrow(dat), "\n")
cat("W_matrix dim:", nrow(W_matrix), "x", ncol(W_matrix), "\n")
cat("Length of A :", length(A), "\n")
cat("Row match?  :", nrow(W_matrix) == length(A), "\n")
```

---

# Covariate Inventory: Analysis-Plan Variables {#sec-covariate-inventory}

This section documents every covariate selected for the clean-room workflow,
grouped by the domains pre-specified in the analysis plan.
Each variable is described, summarised (mean/SD for continuous; n/% for binary),
and compared across treatment arms (**without** examining outcomes).

```{r covariate-inventory-setup}
#| label: covariate-inventory-setup
#| include: false

# ---- Helper: summarise one column by treatment arm ----
summarise_col <- function(x, arm, col_name) {
  is_binary <- all(x[!is.na(x)] %in% c(0, 1))
  overall_n <- length(x)
  n_miss    <- sum(is.na(x))

  if (is_binary) {
    n1   <- sum(x == 1, na.rm = TRUE)
    pct1 <- round(mean(x, na.rm = TRUE) * 100, 1)
    trt_pct  <- round(mean(x[arm == 1], na.rm = TRUE) * 100, 1)
    ctrl_pct <- round(mean(x[arm == 0], na.rm = TRUE) * 100, 1)
    data.frame(
      Variable  = col_name,
      Type      = "Binary",
      Overall   = paste0(n1, " (", pct1, "%)"),
      Treated   = paste0(sum(x[arm == 1] == 1, na.rm = TRUE), " (",  trt_pct, "%)"),
      Control   = paste0(sum(x[arm == 0] == 1, na.rm = TRUE), " (", ctrl_pct, "%)"),
      Missing   = n_miss,
      stringsAsFactors = FALSE
    )
  } else {
    data.frame(
      Variable  = col_name,
      Type      = "Continuous",
      Overall   = paste0(round(mean(x, na.rm = TRUE), 2), " (SD ",
                          round(sd(x, na.rm = TRUE), 2), ")"),
      Treated   = paste0(round(mean(x[arm == 1], na.rm = TRUE), 2), " (",
                          round(sd(x[arm == 1], na.rm = TRUE), 2), ")"),
      Control   = paste0(round(mean(x[arm == 0], na.rm = TRUE), 2), " (",
                          round(sd(x[arm == 0], na.rm = TRUE), 2), ")"),
      Missing   = n_miss,
      stringsAsFactors = FALSE
    )
  }
}

# ---- Helper: summarise a group of columns ----
summarise_group <- function(patterns, W, arm, description_map = NULL) {
  all_cols <- colnames(W)
  matched  <- character()
  for (pat in patterns) {
    matched <- union(matched, grep(pat, all_cols, ignore.case = TRUE, value = TRUE))
  }
  if (length(matched) == 0) return(NULL)

  rows <- lapply(matched, function(cn) {
    label <- if (!is.null(description_map) && cn %in% names(description_map)) {
      description_map[[cn]]
    } else {
      cn
    }
    summarise_col(W[, cn], arm, label)
  })
  do.call(rbind, rows)
}
```

## Domain Definitions

The analysis plan pre-specifies the following covariate domains.
Columns are listed in `rescueCo/config/clean_room_config.yml` → `covariates.selected_columns`.

```{r domain-count}
cat("Selected W_matrix columns:", ncol(W_matrix), "\n")
cat("Column names:\n")
cat(paste(" ", colnames(W_matrix)), sep = "\n")
```

## Demographics

**Age, sex, hospital site.**

```{r demo-table}
demo <- summarise_group(
  c("^age$", "^sex_male$", "^hosp_lkh$"),
  W_matrix, A
)
kable(demo, row.names = FALSE, caption = "Demographics")
```

```{r age-hist}
#| fig-height: 4
if ("age" %in% colnames(W_matrix)) {
  age_df <- data.frame(age = W_matrix[, "age"],
                       Group = ifelse(A == 1, "Treated", "Control"))
  ggplot(age_df, aes(x = age, fill = Group)) +
    geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
    labs(title = "Age Distribution by Treatment Arm",
         x = "Age (years)", y = "Count") +
    theme_minimal()
}
```

## Marital Status

```{r marital-table}
ms <- summarise_group("^marital_married$", W_matrix, A)
kable(ms, row.names = FALSE, caption = "Marital Status Indicators")
```

## Education

```{r edu-table}
ed <- summarise_group("^educ_", W_matrix, A)
kable(ed, row.names = FALSE, caption = "Education Level Indicators")
```

## Distance from Hospital

```{r dist-table}
dst <- summarise_group("^inj_distance$", W_matrix, A)
kable(dst, row.names = FALSE, caption = "Distance from Injury to Hospital")
```

```{r dist-hist}
#| fig-height: 4
dist_col <- grep("^inj_distance$", colnames(W_matrix), value = TRUE)
if (length(dist_col) == 1) {
  dist_df <- data.frame(dist = W_matrix[, dist_col],
                        Group = ifelse(A == 1, "Treated", "Control"))
  ggplot(dist_df, aes(x = dist, fill = Group)) +
    geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
    labs(title = "Distance to Hospital by Treatment Arm",
         x = "Distance (km)", y = "Count") +
    coord_cartesian(xlim = c(0, quantile(dist_df$dist, 0.99, na.rm = TRUE))) +
    theme_minimal()
}
```

## Mechanism of Injury

```{r moi-table}
moi <- summarise_group("^mech_", W_matrix, A)
kable(moi, row.names = FALSE, caption = "Mechanism of Injury Indicators")
```

```{r moi-bar}
#| fig-height: 5
if (!is.null(moi) && nrow(moi) > 0) {
  moi_cols <- grep("^mech_", colnames(W_matrix), value = TRUE)
  prev <- data.frame(
    Mechanism = sub("^mech_", "", moi_cols),
    Treated   = sapply(moi_cols, function(cn) mean(W_matrix[A == 1, cn], na.rm = TRUE)),
    Control   = sapply(moi_cols, function(cn) mean(W_matrix[A == 0, cn], na.rm = TRUE))
  )
  prev_long <- reshape(prev, direction = "long",
                       varying = c("Treated", "Control"),
                       v.names = "Prevalence", timevar = "Group",
                       times = c("Treated", "Control"))
  ggplot(prev_long, aes(x = reorder(Mechanism, Prevalence), y = Prevalence, fill = Group)) +
    geom_col(position = "dodge", alpha = 0.8) +
    coord_flip() +
    labs(title = "Injury Mechanism Prevalence by Treatment Arm",
         x = NULL, y = "Proportion") +
    theme_minimal()
}
```

## Pre-existing Health Conditions

**Chronic illness indicators: hypertension, diabetes, HIV/AIDS,
heart disease, asthma/COPD.**

```{r comorbid-table}
comorbid <- summarise_group(
  c("^chronic_"),
  W_matrix, A
)
kable(comorbid, row.names = FALSE,
      caption = "Pre-existing Health Conditions (Unchecked = condition NOT reported)")
```

## Socioeconomic Status Proxies

**Household type (urban/rural), occupation, cooking fuel.**

Note: The analysis plan also specifies frequency of eating meat, floor type,
and ownership of dish towels, but these variables are **not available** in the
registry data. The closest available proxies are listed below.

```{r ses-table}
ses <- summarise_group(
  c("^household_urban$", "^occup_", "^fuel_wood$"),
  W_matrix, A
)
kable(ses, row.names = FALSE, caption = "Socioeconomic Status Proxy Indicators")
```

```{r ses-occupation-bar}
#| fig-height: 5
occ_cols <- grep("^occup_", colnames(W_matrix), value = TRUE)
if (length(occ_cols) > 0) {
  occ_prev <- data.frame(
    Occupation = sub("^occup_", "", occ_cols),
    Treated  = sapply(occ_cols, function(cn) mean(W_matrix[A == 1, cn], na.rm = TRUE)),
    Control  = sapply(occ_cols, function(cn) mean(W_matrix[A == 0, cn], na.rm = TRUE))
  )
  occ_long <- reshape(occ_prev, direction = "long",
                      varying = c("Treated", "Control"),
                      v.names = "Prevalence", timevar = "Group",
                      times = c("Treated", "Control"))
  ggplot(occ_long, aes(x = reorder(Occupation, Prevalence), y = Prevalence, fill = Group)) +
    geom_col(position = "dodge", alpha = 0.8) +
    coord_flip() +
    labs(title = "Occupation Distribution by Treatment Arm",
         x = NULL, y = "Proportion") +
    theme_minimal()
}
```

## Method of Payment

```{r payment-table}
pay <- summarise_group("^pay_", W_matrix, A)
kable(pay, row.names = FALSE, caption = "Payment Method Indicators")
```

## Clinical Severity at Arrival

**Heart rate and GCS availability.**

```{r clinical-table}
clin <- summarise_group(
  c("^gcs$", "^iss$"),
  W_matrix, A
)
kable(clin, row.names = FALSE, caption = "Clinical Severity at Arrival")
```

```{r gcs-hist}
#| fig-height: 4
if ("gcs" %in% colnames(W_matrix)) {
  gcs_df <- data.frame(gcs = W_matrix[, "gcs"],
                       Group = ifelse(A == 1, "Treated", "Control"))
  ggplot(gcs_df, aes(x = gcs, fill = Group)) +
    geom_histogram(bins = 15, alpha = 0.6, position = "identity") +
    labs(title = "Glasgow Coma Scale by Treatment Arm",
         x = "GCS", y = "Count") +
    theme_minimal()
}
```

## Arrival Timing

```{r arrival-table}
arr <- summarise_group("^arrival_month$", W_matrix, A)
kable(arr, row.names = FALSE, caption = "Arrival Month")
```

```{r arrival-bar}
#| fig-height: 4
if ("arrival_month" %in% colnames(W_matrix)) {
  am_df <- data.frame(month = W_matrix[, "arrival_month"],
                      Group = ifelse(A == 1, "Treated", "Control"))
  ggplot(am_df, aes(x = factor(month), fill = Group)) +
    geom_bar(position = "dodge", alpha = 0.8) +
    labs(title = "Arrival Month by Treatment Arm",
         x = "Month", y = "Count") +
    theme_minimal()
}
```

## Alcohol Use

```{r alcohol-table}
alc <- summarise_group("^inj_alcohol_yes$", W_matrix, A)
kable(alc, row.names = FALSE, caption = "Alcohol Consumption Prior to Injury")
```

## Injury Intent

```{r intent-table}
intent <- summarise_group("^intent_assault$", W_matrix, A)
kable(intent, row.names = FALSE, caption = "Injury Intent")
```

## Combined Summary Table (All Selected Covariates)

```{r combined-table}
all_summaries <- list(demo, ms, ed, dst, moi, comorbid, ses, pay, clin, arr, alc, intent)
all_summaries <- all_summaries[!sapply(all_summaries, is.null)]
combined <- do.call(rbind, all_summaries)
combined$id <- NULL

kable(combined, row.names = FALSE,
      caption = paste0("All ", nrow(combined),
                       " selected covariates: summary by treatment arm"))
```

## Analysis-Plan Variables NOT Available

The following analysis-plan covariates could **not** be mapped to registry columns
and are therefore excluded from the clean-room workflow:
height/weight/BMI,
anaemia,
race/ethnicity,
access to water,
floor type at home,
frequency of eating meat,
ownership of dish towels,
smoking/tobacco use (present in raw data but excluded due to >40 % missingness),
region of residence (sub-county granularity exists but was not retained), and
parental status.

---

# 1. Dimension & Alignment Check

```{r dim-check}
stopifnot(
  "W_matrix rows must equal length(A)" = nrow(W_matrix) == length(A)
)
cat("✓ Dimensions aligned\n")
```

---

# 2. Missing / Non-finite Values in W_matrix

```{r nonfinite}
na_per_col  <- colSums(is.na(W_matrix))
nan_per_col <- colSums(apply(W_matrix, 2, is.nan))
inf_per_col <- colSums(!is.finite(W_matrix) & !is.na(W_matrix))

problem_cols <- data.frame(
  column    = colnames(W_matrix),
  n_NA      = na_per_col,
  n_NaN     = nan_per_col,
  n_Inf     = inf_per_col,
  n_total_bad = na_per_col + nan_per_col + inf_per_col,
  stringsAsFactors = FALSE
)
problem_cols <- problem_cols[problem_cols$n_total_bad > 0, ]

if (nrow(problem_cols) == 0) {
  cat("✓ No NA / NaN / Inf values found in W_matrix\n")
} else {
  cat("⚠ Found", nrow(problem_cols), "columns with non-finite values:\n")
  kable(problem_cols[order(-problem_cols$n_total_bad), ], row.names = FALSE)
}
```

---

# 3. Constant / Near-Zero-Variance Columns

```{r variance}
col_var <- apply(W_matrix, 2, var, na.rm = TRUE)
const_cols <- names(col_var[col_var == 0 | is.na(col_var)])
near_zero  <- names(col_var[col_var > 0 & col_var < 1e-6])

cat("Constant columns (var = 0)  :", length(const_cols), "\n")
if (length(const_cols) > 0) cat("  ", paste(head(const_cols, 20), collapse = ", "), "\n")

cat("Near-zero variance (< 1e-6):", length(near_zero), "\n")
if (length(near_zero) > 0) cat("  ", paste(head(near_zero, 20), collapse = ", "), "\n")
```

---

# 4. Treatment Variable Summary

```{r treatment}
cat("Treatment (A) distribution:\n")
print(table(A, useNA = "always"))

cat("\nMissing A:", sum(is.na(A)), "\n")
```

---

# 5. Column-type Breakdown

```{r col-types}
n_binary <- sum(apply(W_matrix, 2, function(x) all(x[!is.na(x)] %in% c(0, 1))))
n_continuous <- ncol(W_matrix) - n_binary

cat("Binary / indicator columns:", n_binary, "\n")
cat("Continuous columns        :", n_continuous, "\n")
cat("Total                     :", ncol(W_matrix), "\n")
```

---

# 6. Distributions of Continuous Covariates

```{r continuous-summary}
#| fig-height: 8
cont_idx <- which(apply(W_matrix, 2, function(x) {
  length(unique(x[!is.na(x)])) > 2
}))

if (length(cont_idx) > 0) {
  cont_stats <- data.frame(
    column = colnames(W_matrix)[cont_idx],
    min    = apply(W_matrix[, cont_idx, drop = FALSE], 2, min, na.rm = TRUE),
    q25    = apply(W_matrix[, cont_idx, drop = FALSE], 2, quantile, 0.25, na.rm = TRUE),
    median = apply(W_matrix[, cont_idx, drop = FALSE], 2, median, na.rm = TRUE),
    mean   = apply(W_matrix[, cont_idx, drop = FALSE], 2, mean, na.rm = TRUE),
    q75    = apply(W_matrix[, cont_idx, drop = FALSE], 2, quantile, 0.75, na.rm = TRUE),
    max    = apply(W_matrix[, cont_idx, drop = FALSE], 2, max, na.rm = TRUE),
    sd     = apply(W_matrix[, cont_idx, drop = FALSE], 2, sd, na.rm = TRUE),
    n_unique = apply(W_matrix[, cont_idx, drop = FALSE], 2, function(x) length(unique(x))),
    stringsAsFactors = FALSE
  )
  kable(cont_stats, digits = 3, row.names = FALSE,
        caption = "Continuous covariate summary statistics")
} else {
  cat("No continuous covariates found.\n")
}
```

---

# 7. Extreme-Value Flags

Covariates with values > 5 SD from the mean:

```{r extremes}
extreme_report <- lapply(cont_idx, function(j) {
  x <- W_matrix[, j]
  mu <- mean(x, na.rm = TRUE)
  s  <- sd(x, na.rm = TRUE)
  if (s == 0) return(NULL)
  n_extreme <- sum(abs(x - mu) > 5 * s, na.rm = TRUE)
  if (n_extreme > 0) {
    data.frame(column = colnames(W_matrix)[j],
               n_extreme = n_extreme,
               pct = round(n_extreme / length(x) * 100, 2),
               stringsAsFactors = FALSE)
  }
})
extreme_df <- do.call(rbind, extreme_report)

if (is.null(extreme_df) || nrow(extreme_df) == 0) {
  cat("✓ No extreme outliers detected (> 5 SD)\n")
} else {
  kable(extreme_df[order(-extreme_df$n_extreme), ], row.names = FALSE,
        caption = "Covariates with extreme values")
}
```

---

# 8. Pairwise Complete Cases: A vs W_matrix

This replicates what SuperLearner sees during CV:

```{r complete-cases}
W_df <- as.data.frame(W_matrix)
cc <- complete.cases(W_df) & !is.na(A)

cat("Complete cases (A + all W):", sum(cc), "of", length(A), "\n")
cat("Rows lost                 :", sum(!cc), "\n")

if (sum(!cc) > 0) {
  # Which columns cause the most row loss?
  col_na_rows <- sapply(seq_len(ncol(W_matrix)), function(j) {
    sum(is.na(W_matrix[, j]) | is.nan(W_matrix[, j]) | !is.finite(W_matrix[, j]))
  })
  names(col_na_rows) <- colnames(W_matrix)
  culprits <- sort(col_na_rows[col_na_rows > 0], decreasing = TRUE)
  cat("\nColumns contributing to incomplete rows:\n")
  print(head(culprits, 30))
}
```

---

# 9. Sample of `age` (key continuous covariate)

```{r age-inspect}
if ("age" %in% colnames(W_matrix)) {
  age <- W_matrix[, "age"]
  cat("age summary:\n")
  print(summary(age))
  cat("NA :", sum(is.na(age)), "\n")
  cat("NaN:", sum(is.nan(age)), "\n")
  cat("Inf:", sum(is.infinite(age)), "\n")
  cat("Unique values:", length(unique(age)), "\n")
} else {
  cat("Column 'age' not found in W_matrix.\n")
  age_cols <- grep("age", colnames(W_matrix), ignore.case = TRUE, value = TRUE)
  cat("Age-related columns:", paste(age_cols, collapse = ", "), "\n")
}
```

---

# 10. SuperLearner Dry-Run (single fold, SL.glm only)

A minimal test to confirm the data matrix is usable:
```{r sl-dryrun}
#| error: true
library(SuperLearner)

cc_idx <- which(complete.cases(W_df) & !is.na(A))
A_test <- A[cc_idx]
W_test <- W_df[cc_idx, , drop = FALSE]

# Remove any remaining constant columns
col_var_test <- apply(W_test, 2, var, na.rm = TRUE)
W_test <- W_test[, col_var_test > 1e-6, drop = FALSE]

cat("Dry-run dimensions: n =", nrow(W_test), ", p =", ncol(W_test), "\n")

tryCatch({
  sl_test <- SuperLearner(
    Y = A_test, X = W_test, family = binomial(),
    SL.library = "SL.glm",
    cvControl = list(V = 2)
  )
  cat("✓ SuperLearner dry-run succeeded\n")
  cat("CV Risk:", round(sl_test$cvRisk, 4), "\n")
}, error = function(e) {
  cat("✗ SuperLearner dry-run FAILED:", e$message, "\n")
})
```

---

# Summary & Recommendations

```{r summary}
issues <- 0L
recs   <- character()

if (nrow(problem_cols) > 0) {
  issues <- issues + 1L
  recs <- c(recs, paste("Remove or impute", nrow(problem_cols),
                         "columns with NA/NaN/Inf values"))
}
if (length(const_cols) > 0) {
  issues <- issues + 1L
  recs <- c(recs, paste("Drop", length(const_cols), "constant columns"))
}
if (sum(!cc) > 0) {
  issues <- issues + 1L
  recs <- c(recs, paste("Resolve", sum(!cc),
                         "incomplete rows before SuperLearner"))
}

if (issues == 0) {
  cat("✓ No blocking issues detected. Stage 2 should run cleanly.\n")
} else {
  cat("⚠ Found", issues, "issue(s) to resolve before Stage 2:\n")
  for (r in recs) cat("  •", r, "\n")
}
```
