---
title: "cleanTMLE for Applied Analysts"
subtitle: "A plain-language introduction to staged, outcome-blind causal analysis"
author: "Andrew Mertens"
date: today
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    self-contained: true
    code-fold: true
  docx:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

# What this report is, and who it is for

This report is written for an applied analyst (an epidemiologist, biostatistician, data scientist, or health-services researcher) who works with observational health data and has not previously used targeted-learning methods or read the staging-and-clean-room methods literature. It is intended to be readable end-to-end in about an hour. By the end the reader should understand the problem the cleanTMLE R package is designed to address, the two ideas it borrows from the methods literature (targeted learning and the outcome-blind staged analysis), how the workflow runs step by step, what the simulation study and the worked case study show, and when the package is and is not the right tool. No prior exposure to TMLE, propensity-score methods beyond the basics, or staged-analysis methodology is assumed.

# The problem

## What we are trying to do in observational research

Suppose you have data on a group of patients. Some received a treatment or intervention; others received a comparator. You want to estimate what the treatment does: what the difference in some health outcome would be if everyone had received the treatment versus if everyone had received the comparator. This is a *causal* question, and it is what every observational comparative-effectiveness or pharmacoepidemiology study is asking.

In a randomised trial, the question is answered by the randomisation. In observational data, it has to be answered by *adjustment*: we measure as many baseline characteristics as we can, and we use statistical methods to mimic, as well as we are able, the balance that randomisation would have produced.

## Why this is harder than it looks

Three difficulties tend to arise at once. First, analytic decisions accumulate during execution. Even with a written protocol, the analyst runs into sparse events, missing values, model-convergence problems, and outliers, and makes choices to deal with each one. Many of those choices are made after the analyst has at least glanced at the comparison being estimated, so a reviewer reading the final paper cannot tell which choices were prespecified and which were made under outcome pressure. Second, standard diagnostics describe the data, not the estimator: propensity-score overlap, balance tables, and weight distributions tell you whether the design is feasible, but none of them tells you whether the specific statistical method you plan to use will give a reliable answer on your dataset. Third, data-quality concerns are usually documented qualitatively rather than quantitatively. A typical fit-for-purpose review states that outcome ascertainment may have sensitivity around 0.90, and a reviewer cannot tell from that sentence whether the planned estimator is robust to it.

## What we want a workflow to deliver

We want a single document, written before the analyst has seen the comparative result, that lets a reviewer answer five questions: whether the cohort, treatment, outcome, and time horizon were defined before the comparative analysis; whether the statistical method was chosen before the comparative analysis; whether the method was checked against plausible data-quality problems for this dataset; whether the diagnostics are adequate or instead show warning signs that should change the analysis plan; and whether every decision is recorded with a timestamp, a rationale, and a named author. cleanTMLE is software that produces this document. It does not replace the careful design work that has to happen before any code runs; it makes that work auditable.

# The two ideas behind the package

## The outcome-blind staged analysis

The first idea is that the analysis should be split into stages, and the early stages should be completed before anyone has seen the comparative result. The analyst can look at baseline characteristics, propensity scores, weights, and even simulations that use a synthetic version of the outcome, but the observed treatment-outcome comparison (the comparison of the drugs or interventions of interest) is not opened until the prespecified checks have been passed. This restriction is less limiting than it sounds. Most of the work in a careful observational study, including defining the cohort, choosing covariates, fitting the propensity score, and examining balance, does not require seeing the comparative result. The analysis lock simply enforces that the analytic plan is finalised before the comparative analysis runs. The same principle is familiar from randomised trials, where the analysis plan is fixed before the treatment codes are unblinded; here the "outcome" rather than the "treatment assignment" is what is held back.

## Clean-room governance

The second idea, the clean room, is the institutional counterpart of staged analysis. The term is borrowed by @muntner2024staging from semiconductor manufacturing, where a clean room is a controlled environment that keeps contaminants away from production; the analogous "contaminant" here is investigator bias such as the desire to find a large effect. In a clean room the team running the analysis is separated from access to the comparative results, and protocol deviations are decided by a documented review team that is masked to how its decisions would affect the comparative estimate. The point is not the metaphor but the arrangement: it separates decision-making from outcome-driven temptation, in the same spirit as blinding in a trial. The same philosophy can also be applied less formally, for example by a single analyst who simply commits to finalising the plan before unblinding the outcome.

cleanTMLE implements the software side of this arrangement. It cannot, on its own, separate teams or stop someone from inspecting the data outside the package. What it can do is record everything the analysis does, so that, when the institutional side is also in place, the two combine into a record that an external reviewer or regulator can use to build trust in the result. A prespecified plan can still fail during execution, because real datasets surprise the analyst, and a workflow that refuses to update the plan when reality intervenes is worse than no plan at all. Recording every update with a rationale and a timestamp keeps the flexibility while preserving the credibility.

# The statistical method

## What an estimator does

An estimator is a recipe that turns a dataset into a number, with a confidence interval, that targets the causal effect of interest. The causal effect (the *estimand*) is the same regardless of method; estimators differ in how they recover it from the observed data and in the assumptions under which they succeed. The causal roadmap is explicit about this separation between the causal question, the statistical estimand it implies under identification assumptions, and the estimator used to compute it [@dang2023roadmap; @petersen2014roadmap].

## Conventional adjustment methods

In practice the comparison is almost never left unadjusted, because the treated and untreated groups differ at baseline. The usual starting point is a covariate-adjusted comparison, for example an outcome regression that conditions on the measured baseline characteristics. Two further strategies build on the propensity score, the predicted probability of treatment given baseline characteristics. Propensity-score matching pairs each treated person with an untreated person of similar profile and discards the unmatched, which redefines the estimand to the population of patients who could plausibly have received either treatment. Inverse-probability-of-treatment weighting (IPTW) instead reweights everyone so that the weighted treated and untreated groups are balanced; it retains the whole cohort, but a few large weights can dominate the estimate, and the standard-error formula is conservative when the weights are estimated rather than known.

## Targeted maximum-likelihood estimation (TMLE)

Targeted maximum-likelihood estimation is a doubly robust alternative that combines outcome regression with the propensity score. It first predicts treatment from the baseline characteristics, and predicts the outcome from the baseline characteristics and treatment, using flexible models (a SuperLearner ensemble is typical). It then applies a *targeting* step: the outcome predictions are adjusted by a small amount, in a direction determined by the propensity score, so that the resulting treated-minus-untreated contrast is calibrated for the causal question rather than for prediction accuracy. The estimate and its standard error are read off the targeted predictions. The targeting step is what lets TMLE use machine-learning models for the two prediction steps while still delivering a confidence interval that can be interpreted in the usual Wald sense; the formal justification is the efficient influence function. Most other flexible approaches either lock the analyst into a parametric model or lose interpretable inference once machine learning is used.

This structure is what makes TMLE a natural fit for a staged, outcome-blind analysis, and it maps directly onto the clean-room separation of steps. The prediction steps depend only on the baseline characteristics and treatment, not on the comparative result, so the entire recipe (the SuperLearner library, the propensity-weight truncation rule, the targeting procedure, and the prespecified set of candidate estimators) can be written into the analysis plan before the outcome is unblinded. @gruber2022sap show how to express exactly these choices in a targeted-learning statistical analysis plan. The flexibility of the underlying models therefore introduces no outcome-driven choices, which is the technical reason TMLE is the package's default; matching, IPTW, and augmented IPW are also supported and reported as concordance checks.

# The cleanTMLE workflow, step by step

The workflow runs in numbered stages. Each stage produces a record. The comparative analysis only runs after all the records show that the prespecified checks have been passed.

## Stage 0: Protocol and target trial

Before any code runs, the team writes the question down using the target-trial framework: if this were a randomised trial, who would be in it, what would the treatment and comparison be, what would the outcome be, and when would follow-up start and end. This is external work, not software work, and cleanTMLE assumes it has been done. It corresponds to Stage 1 of @muntner2024staging (study development and protocol registration) and to Steps 1 to 4 of the causal roadmap [@dang2023roadmap].

## Stage 1a: Lock the plan

The analytic plan is bundled into a single "lock" object. This object records the data shape, the covariate list, the SuperLearner library, the seed, and the candidate estimators. The lock is fingerprinted with a cryptographic hash, so any later change to the plan is detectable.

```{r}
#| eval: false
lock <- create_analysis_lock(
  data        = study_data,
  treatment   = "A",
  outcome     = "Y",
  covariates  = baseline_covariates,
  sl_library  = c("SL.glm", "SL.glmnet", "SL.ranger"),
  seed        = 2026
)
lock <- attach_estimand(lock,
  description           = "Effect of A on 24-month event risk",
  population            = "Adults eligible at index",
  treatment_strategies  = c("Treatment", "Reference"),
  outcome_label         = "Event by 24 months",
  followup              = "24 months",
  contrast              = "risk_difference"
)
```

## Stage 1b: Cohort adequacy

Check that the cohort has enough people and enough events to support the question being asked. If you have 50 events in each arm, you cannot estimate a small risk difference with any precision. This is a software-enforced checkpoint, corresponding to Check Point 1 (adequate study size and statistical precision) in @muntner2024staging.

```{r}
#| eval: false
cp1 <- checkpoint_cohort_adequacy(
  lock, min_n_per_arm = 200, min_events = 50
)
```

## Stage 2a: Design diagnostics

Standard observational-study diagnostics fit the propensity score, examine overlap and balance after weighting, and compute the effective sample size. None of this involves the outcome, and it corresponds to Stage 2 and Check Point 2 (covariate balance) in @muntner2024staging. The diagnostics are estimator-agnostic in the sense that overlap and balance bear on any adjustment method, but the propensity-score fit produced here is the same nuisance model TMLE uses in its targeting step, so a poorly overlapping design is a warning specifically about the TMLE candidate as well as about the design. Balance is summarised with standardised mean differences; because univariate SMDs compare only the marginal means of each covariate, they can be satisfied while higher moments or covariate interactions remain imbalanced, so the diagnostics also report variance ratios and the three-way Love plot rather than SMDs alone [@austin2009ps; @zhang2019balance].

```{r}
#| eval: false
ps      <- fit_ps_superlearner(lock)
ps_diag <- compute_ps_diagnostics(ps)
plot(ps_diag)                         # overlap density
love_plot_threeway(lock, ps$weights)  # SMDs before/after weighting
```

## Stage 2b: Baseline plasmode candidate selection

This optional stage is where the package does something a standard pipeline does not. When several candidate methods are in play (different SuperLearner libraries, different truncation levels), rather than guessing which to use, the package generates a synthetic outcome from the real covariate and treatment structure with a known synthetic treatment effect, and compares how each candidate performs on that synthetic outcome. This is a "plasmode" simulation [@franklin2014plasmode; @schreck2024plasmode], and the candidate with the best performance is locked in as the primary specification. An analyst who has already committed to a single estimator can skip the candidate-selection step and still use the data-quality stress test in Stage 2c.

```{r}
#| eval: false
candidates <- list(
  tmle_candidate("glm_t01", "GLM PS, trunc = 0.01", truncation = 0.01),
  tmle_candidate("glm_t05", "GLM PS, trunc = 0.05", truncation = 0.05)
)
plas <- run_plasmode_feasibility(
  lock, tmle_candidates = candidates,
  effect_sizes = c(0.03, 0.05), reps = 200
)
```

Crucially, the *real* outcome has not been touched. The selection is based entirely on synthetic outcomes generated from the analyst's model.

## Stage 2c: Data-quality stress test

The fit-for-purpose review for the study lists specific data-quality concerns: outcome ascertainment may be imperfect, exposure may be misclassified, a key confounder may be unmeasured, or some covariates may be missing. The stress test turns each concern into a quantitative perturbation of the synthetic outcome and re-runs the candidate-selection simulation under each. Four threats are built in. Covariate missingness sets a specified fraction of covariate values to missing-completely-at-random and mean-imputes them; exposure misclassification flips treatment status with a specified sensitivity and specificity; outcome misclassification flips outcome status with a specified sensitivity and specificity, optionally differential by treatment arm; and unmeasured confounding adds a latent variable U that affects both treatment and outcome with specified strength.

```{r}
#| eval: false
dq_results <- run_plasmode_dq_stress(
  lock, tmle_candidates = candidates,
  scenarios = default_dq_scenarios(),
  n_reps = 200
)
selected <- select_tmle_candidate(
  plas, rule = "min_max_rmse", dq_results = dq_results
)
```

The output is a degradation gradient for each candidate: how far its bias, RMSE, and coverage move under increasingly severe versions of each threat. A robust candidate stays inside the prespecified thresholds; a fragile candidate degrades sharply.

## Stage 3: Negative controls (optional)

A negative-control outcome is a variable that the exposure should not affect. If the data contain a suitable negative control, an estimated treatment effect on it that is far from the null signals residual bias. This is a different probe of confounding from what plasmode or stress testing can give, and it corresponds to Stage 3 and Check Point 3 (uncontrolled confounding) in @muntner2024staging.

```{r}
#| eval: false
nc_results <- run_residual_confounding_stage(
  lock,
  negative_controls = c("chronic_hypertension", "household_urban", ...)
)
```

## The pre-outcome decision

All the records from the previous stages are aggregated into a single decision against the prespecified thresholds. The three outcomes correspond to the review-team recommendations described by @muntner2024staging, where a masked review team at each check point recommends proceeding to the next stage, conducting additional analyses to address a concern, or terminating the study before the comparative analysis. cleanTMLE labels these GO (proceed to the locked comparative analysis), FLAG (proceed, but with a documented qualification or a prespecified additional analysis, such as the bootstrap variance or sensitivity analysis the concern calls for), and STOP (do not run the comparative analysis as planned: redesign, narrow the question, or record a documented override).

```{r}
#| eval: false
gate <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
auth <- authorize_outcome_analysis(lock, gate)
```

A GO decision is not a guarantee that the answer is correct. It means that, under the assumptions written down, the prespecified checks did not fail.

## Stage 4: Locked primary analysis

Only now does the comparative analysis run. The primary-analysis function refuses to run on a lock that has not been authorised. In software terms, `authorize_outcome_analysis()` records that the gate passed, and `run_clean_tmle()` checks for that authorisation before it will read the outcome column; without it the function stops with an error. This is a software complement to, not a replacement for, the organisational arrangement @muntner2024staging recommend, in which a data manager controls access to the outcome variable and a masked review team signs off at each check point. The software records and enforces the sequence; the institution provides the personnel separation.

```{r}
#| eval: false
fit_primary <- run_clean_tmle(lock, spec = selected, ps_fit = ps,
                              authorization = auth)
```

## Post-outcome sensitivity

After the comparative analysis runs, prespecified sensitivity analyses are added: a tipping-point sensitivity (how far would the estimate have to shift for the conclusion to change), an E-value, and a truncation-grid sweep. These differ from the pre-outcome DQ stress test, because they ask a question about the realised estimate rather than about a candidate's behaviour under synthetic outcomes.

## Export the document

The full set of records (checkpoints, decision-log entries, the lock, the authorisation, and the final estimates) exports as a single bundle, and this is what the reviewer reads. A worked example of the exported bundle accompanies the Rescue.Co case study in the companion repository.

# The simulation study

## Design

We simulated three scenarios on a small dataset (2,000 people, 200 repetitions each) and ran six estimators in each: a covariate-adjusted (but unmatched) comparison, PS matching, IPTW, TMLE, cross-validated TMLE, and matched-cohort TMLE. Cross-validated TMLE (CV-TMLE) differs from ordinary TMLE only in that it fits the nuisance models on held-out folds before targeting, which removes the overfitting bias that flexible learners can introduce when the same data are used to fit and to evaluate. The three scenarios increase in difficulty: in Scenario A the treatment groups overlap well and there are no hidden confounders; in Scenario B they overlap only partially; in Scenario C an unmeasured confounder affects both treatment and outcome. Each scenario was also re-run under the four data-quality stresses.

## What we found

In Scenario A all five adjusted estimators recover the truth and only the unadjusted comparison is biased, so the prespecified decision is GO. This is the benchmark check that the machinery is correctly wired. In Scenario B, TMLE and CV-TMLE keep bias small, but IPTW's confidence intervals come out too wide, because its influence-function standard-error formula treats the propensity score as if it were known when in fact it was estimated from the data; estimating the propensity score removes variance that the formula does not credit back, so the reported standard error is too large and coverage exceeds the nominal level. Matched-TMLE shows the same conservatism more strongly, because its post-match targeting step treats the matched cohort as a fresh independent sample and ignores the extra variability introduced by the matching draw. The decision is therefore GO with a FLAG noting that bootstrap variance is needed if IPTW or matched-TMLE is reported as primary. In Scenario C every estimator is biased in the same direction; TMLE is the least biased but still biased. Baseline plasmode coverage sits at the lower edge of acceptable, and what makes the failure unambiguous is the unmeasured-confounding stress-test row, which pushes bias above the locked threshold and returns STOP. Across scenarios the stress test was most informative when the threat targeted identification (unmeasured confounding) or measurement (outcome misclassification), and least informative when the threat was ordinary sampling variability already captured by the baseline plasmode.

## What this means for method choice

TMLE or CV-TMLE is the recommended primary estimator in most observational settings, because it had the best bias control across all three scenarios while still producing interpretable confidence intervals, and IPTW and PS matching are best reported as concordance checks: when all three agree on direction, there is more to say to a reviewer. If IPTW must be reported as primary, the standard error should be obtained by bootstrap rather than from the conservative influence-function formula, and a matched-TMLE primary should be block-bootstrapped for the same reason. When an unmeasured confounder cannot be measured, no adjustment method is reliable; the honest options are to collect the missing variable, narrow the question, or restrict to bounded sensitivity analysis, which is exactly the situation in which the prespecified STOP decision earns its place.

## Traditional approach versus the outcome-blind workflow

| Scenario | What a traditional analyst would see | What the outcome-blind workflow does |
|---|---|---|
| A. Good overlap | PS diagnostics look fine; pick a method; report. | Plasmode confirms the candidates behave acceptably and selects the one with the best bias and variance. GO. |
| B. Marginal overlap | Overlap is strained; some analysts widen truncation, many do not. | Plasmode quantifies IPTW and matched-TMLE SE conservativeness. GO with FLAG and a specific recommendation to bootstrap. |
| C. Unmeasured confounding | All adjusted estimates agree on a non-null effect. Without an external probe, the analyst likely proceeds and reports. | The stress test re-runs the simulation with a prespecified latent confounder added; the selected candidate's bias then exceeds the locked threshold, which trips the decision rule. STOP. |

The clearest contrast is Scenario C. A careful traditional analyst can pass the standard overlap and balance diagnostics and still proceed with a biased estimator, because those diagnostics never interrogate an unmeasured confounder. The outcome-blind workflow, by inserting a prespecified latent confounder of the severity named in the fit-for-purpose review into the synthetic-outcome simulation, observes the selected candidate's bias cross the locked threshold and flags it as fragile before the real outcome is ever unblinded.

## What the simulation does and does not show

The simulation shows that the candidate-selection machinery is correctly calibrated, that the prespecified stress-test row catches a built-in violation of the identification assumptions (here, the conditional-exchangeability assumption broken by the unmeasured confounder) and triggers STOP, and that the locked decision rule reaches the right verdict in all three scenarios without ever inspecting the realised treatment-outcome association. It does not show that the workflow detects unmeasured confounders on its own: it only scores the threats the analyst specifies, so a threat the analyst does not anticipate will not be caught. Nor does it show that cleanTMLE numerically outperforms a careful traditional analysis on the same dataset; the workflow's value is operational (prespecification, auditability, and gating), not a different point estimate.

# The case study: Rescue.Co Kenya Trauma Registry

## The question

Rescue.Co operates a motorcycle-based emergency ambulance service in Nairobi. We compared 6-month functional outcome (GOSE > 4) in patients arriving by Rescue.Co versus other ambulance services. The analytic cohort is 1,693 trauma patients from 2018 to 2024 who arrived by ambulance direct from scene, split between Rescue.Co (n = 1,013) and other ambulance services (n = 680); the binary outcome is GOSE > 4 at six months, and 24.6 % of outcomes are missing. This is a workflow demonstration on real data, not evidence about whether Rescue.Co transport works.

## What the pre-outcome stage showed

Of the 1,693 eligible patients, 1,277 had a complete outcome and 615 formed a matched subset. The design diagnostics were acceptable: propensity scores stayed within range, balance was achieved after weighting, and the effective sample size retained 78 % of observations. Two baseline plasmode candidates at different truncation levels both passed, and a minimax rule selected the candidate that performed best across the worst-case threats. Of the five negative controls planned in the protocol, three failed an automatic data-cleaning filter and were dropped, with the drops recorded in the audit log, while the two that remained both passed. The prespecified rule returned GO with a documented FLAG about the narrower negative-control panel.

## What the locked analysis showed

| Method | N | Risk difference | 95 % CI |
|---|---:|---:|---|
| Unadjusted | 1,277 | 0.037 | (0.003, 0.071) |
| PS matching | 615 | 0.027 | (−0.020, 0.074) |
| IPTW | 1,693 | 0.037 | (0.002, 0.071) |
| **Full-cohort TMLE (primary)** | **1,693** | **0.031** | **(−0.001, 0.062)** |
| Matched-cohort TMLE | 1,230 | 0.026 | (−0.015, 0.067) |
| Complete-outcome TMLE with IPCW weights | 1,277 | 0.033 | (−0.000, 0.066) |

All six methods agree on direction. The primary estimate just crosses zero. The data are consistent with a small, possibly real, possibly null, effect.

## How a reviewer reads it

Every analytic decision was recorded before the comparative numbers above were produced, and six methods that make different assumptions all point the same way. The primary confidence interval just crosses zero, so the workflow does not confirm the signal under the prespecified diagnostics, and the audit log shows the three negative controls that were dropped and why. A reviewer can reconstruct the entire analysis from the exported bundle and the lock fingerprint.

# How cleanTMLE compares to other tools

Many parts of this workflow already exist as separate tools. cleanTMLE's contribution is to bind them into a single auditable pipeline.

| Feature | TMLE-only packages | causalRisk | Clean-room methods papers | **cleanTMLE** |
|---|:---:|:---:|:---:|:---:|
| Targeted-learning estimation | ✓ | -- | -- | ✓ |
| Cohort-specification grammar | -- | ✓ | -- | ✓ |
| Cumulative-risk reporting | partial | ✓ | -- | ✓ |
| Weight diagnostics | -- | ✓ | -- | ✓ |
| Fingerprinted analysis lock | -- | -- | guidance only | ✓ |
| Audit + decision logs | -- | -- | guidance only | ✓ |
| Outcome-blind plasmode candidate selection | -- | -- | -- | ✓ |
| **Data-quality stress test** | -- | -- | -- | **✓** |
| Prespecified GO / FLAG / STOP rule | -- | -- | guidance only | ✓ |
| Software outcome-access gate | -- | -- | -- | ✓ |

cleanTMLE shares the cohort-specification and cumulative-risk reporting concepts with causalRisk, then adds the prespecification, simulation, stress-test, and gate machinery on top.

# When to use this package

cleanTMLE fits methods development around outcome-blind workflows, where it is the most direct way to run an end-to-end staged analysis in R, and outcome-blind estimator selection, where an analyst unsure which propensity-score truncation or learner library to use can let the baseline plasmode identify which performs best on the study's covariate and treatment structure before unblinding. It also supports protocol and statistical-analysis-plan development, since it can produce a draft document to submit as part of the SAP before any comparative analysis is run, and it suits exploratory or lower-stakes real-world evidence studies in which a structured staged review is useful but a full institutional clean room is not in place.

# When it is not the right tool

cleanTMLE is not sufficient on its own for high-stakes confirmatory or regulatory studies, which need an institutional clean-room arrangement (role separation, credentialed data access, and independent reviewers) that software cannot provide. It also cannot rescue settings where the assumptions clearly fail: if a confounder cannot be measured or a phenotype cannot be validated, the package will at best document the problem rather than solve it. Finally, estimands outside the v0.1 scope are not yet supported; the package is tested for a binary point exposure, a binary outcome, and the marginal risk difference, while time-to-event, time-varying treatment, MAR/MNAR missingness, and native competing-risk estimation are planned but not yet validated.

Several responsibilities remain with the analyst and the institution rather than the software: cleanTMLE does not automate study design, validate source data or phenotypes, guarantee that the identification assumptions hold, replace negative controls or post-outcome sensitivity analyses, or control raw-data access outside the package.

# A short worked example

The full example is in the staged-analysis vignette and the Rescue.Co case study. The skeleton is:

```{r}
#| eval: false
library(cleanTMLE)

# Stage 1a: lock the plan
lock <- create_analysis_lock(
  data = my_data, treatment = "A", outcome = "Y",
  covariates = baseline_covariates,
  sl_library = c("SL.glm", "SL.glmnet"), seed = 2026
)
lock <- attach_estimand(lock,
  description = "Effect of A on 24-month event risk",
  contrast = "risk_difference"
)

# Stage 1b: cohort adequacy
audit <- create_audit_log(lock)
cp1   <- checkpoint_cohort_adequacy(lock,
            min_n_per_arm = 200, min_events = 50)
audit <- record_checkpoint(audit, cp1)

# Stage 2a: PS + balance + weights
ps      <- fit_ps_superlearner(lock)
ps_diag <- compute_ps_diagnostics(ps)
cp2     <- checkpoint_balance(ps_diag, max_smd = 0.10,
                              lock_hash = lock$lock_hash)
audit   <- record_checkpoint(audit, cp2)

# Stage 2b + 2c: plasmode + DQ stress
candidates <- list(
  tmle_candidate("glm_t01", "GLM PS, trunc = 0.01", truncation = 0.01),
  tmle_candidate("glm_t05", "GLM PS, trunc = 0.05", truncation = 0.05)
)
plas <- run_plasmode_feasibility(lock,
            tmle_candidates = candidates, reps = 200)
dq   <- run_plasmode_dq_stress(lock,
            tmle_candidates = candidates,
            scenarios = default_dq_scenarios(), n_reps = 200)
selected <- select_tmle_candidate(plas,
            rule = "min_max_rmse", dq_results = dq)

# Decision log
audit <- log_decision_entry(audit,
  stage = "Stage 2c", decision = "GO",
  rationale = "All candidates within thresholds",
  reviewer = "lead_analyst", y_visible = FALSE
)

# Pre-outcome gate
gate <- gate_all(cp1, cp2, allow_flag = TRUE)
auth <- authorize_outcome_analysis(lock, gate)

# Stage 4: locked primary analysis
fit <- run_clean_tmle(lock, spec = selected, ps_fit = ps,
                     authorization = auth)
```

That is the entire workflow. Each function call produces a record, and the reviewer reads the records in order.

# Glossary

| Term | Plain-language definition |
|---|---|
| **Estimand** | The exact treatment-effect quantity you are trying to estimate (e.g. "the difference in 6-month outcome risk between Rescue.Co and other ambulance services in adults eligible at index"). |
| **Estimator** | A specific statistical recipe for producing a number that estimates the estimand. |
| **Propensity score** | The predicted probability that a given person received treatment, given their baseline characteristics. |
| **Overlap (positivity)** | The degree to which treated and untreated groups have similar baseline profiles. Poor overlap makes adjustment unreliable. |
| **Balance** | How similar the two groups look on baseline variables after weighting or matching. Measured by standardised mean differences (SMDs). |
| **Effective sample size (ESS)** | The "real" sample size after weighting; if a few people have huge weights, the ESS can be much smaller than n. |
| **Plasmode simulation** | A simulation that uses the real covariate and treatment structure but generates a synthetic outcome with a known truth. Used to test how candidate estimators behave on data that looks like your study. |
| **Outcome-blind** | Decisions about cohort, methods, and diagnostics are made without inspecting the observed treatment-outcome comparison. Synthetic outcomes (in plasmode) are allowed. |
| **TMLE** | Targeted maximum-likelihood estimation. An estimator that uses flexible models for prediction and a targeting step to ensure the treatment-effect estimate is properly calibrated. |
| **SuperLearner** | A machine-learning ensemble that combines many candidate prediction models. Used by TMLE for the prediction steps. |
| **IPTW** | Inverse-probability-of-treatment weighting. An adjustment method that re-weights people by the inverse of their propensity score. |
| **IPCW** | Inverse-probability-of-censoring weighting. Same idea applied to missingness instead of treatment. |
| **Negative control** | A pre-treatment outcome that the exposure should not affect. A non-zero estimated effect on a negative control is a sign of residual confounding. |
| **Audit trail / decision log** | A timestamped record of every analytic decision and every checkpoint result, in the sense used by @muntner2024staging. |
| **GO / FLAG / STOP** | The package's labels for the three review-team recommendations of @muntner2024staging: proceed to the comparative analysis, proceed with a documented qualification or additional analysis, or stop before the comparative analysis. |

# Resources

The package lives at <https://github.com/amertens/cleanTMLE>; the manuscript is `reports/manuscript_outcome_blind_dq.qmd`; the worked case study is `rescueCo/reports/rescueco_clean_room_case_study.qmd`; and the package ships a staged-analysis vignette (`cleanTMLE/vignettes/cleanTMLE-staged-analysis.qmd`) and a function-reference vignette (`cleanTMLE/vignettes/cleanTMLE-functions.qmd`). The methods papers referenced throughout are @muntner2024staging on staging and clean-room governance, @mcgrath2021romiplostim on the Romiplostim worked example of staged-analysis decision logs, @dang2023roadmap on the targeted-learning RWE roadmap, @gruber2022sap on the targeted-learning statistical-analysis-plan template, @schuler2017tmle as an applied TMLE introduction, @vdL2011tlbook as the targeted-learning textbook, @franklin2014plasmode and @schreck2024plasmode on plasmode simulation, @austin2009ps and @zhang2019balance on propensity-score balance diagnostics, and @lipsitch2010negative on negative controls.

# Critical evaluation and verification

This section answers four questions about cleanTMLE that an applied analyst might reasonably ask before adopting the package, and documents how the package was independently audited. Both audits were read-only and did not change numerical results.

## Q1. Does cleanTMLE have all the functionality of the closest open-source / proprietary cumulative-risk reporting package?

**Partial.** A function-by-function comparison was run against the comparator cumulative-risk reporting package. cleanTMLE matches it on the core cohort-specification grammar (`specify_models()`, the `identify_*()` family), all four cumulative-risk estimators (IPW, g-computation, AIPW, IPW hazard ratio), and the standard reporting helpers (Table 1, Table 2, weight summary, extreme weights, forest plot, plot.cumrisk). Concrete gaps:

- **Cumulative-count family** (`estimate_ipwcount`, `identify_count`, `update_count`, `plot.cumcount`) is not implemented.
- **`compare_protocols()` analogue** for arbitrary treatment-protocol lists is missing.
- **`subgroup()` as a model-spec verb** is only available indirectly via `subset_idx` arguments.
- **`update_missing`, `update_count`, `update_label`** are not implemented.
- **Public reproducible datasets** (ACTG, WIHS, leukemia) are not shipped with cleanTMLE; the comparator's vignettes can nonetheless be reproduced against public copies of the same trials (see `cleanTMLE_vs_causalRisk_actg.qmd`, which replicates the comparator's ACTG 320 analysis using `mlr3proba::actg`).

The SMR (ATT-targeting) weight scheme, which the comparator selects with `wt_type = 1`, is exposed via `estimate_ipwrisk(weight_type = "smr")`. The remaining gaps are tracked as items in *TODO.md* and are scheduled for cleanTMLE 0.2.

## Q2. Is cleanTMLE an improvement on that package?

**Yes, on governance, estimator depth, and pre-outcome stress testing; equal on cumulative-risk reporting.**

What cleanTMLE adds:

- A **fingerprinted analysis lock, audit trail, and decision log** so the prespecification is itself part of the analysis code rather than a separate paper document.
- **Doubly-robust TMLE** in addition to AIPW (four-step pipeline, modular API, IPCW-TMLE, matched-cohort TMLE).
- **Outcome-blind plasmode candidate selection** under the empirical (W, A) structure before unblinding.
- A **data-quality stress test** that turns qualitative concerns into quantitative reruns of the same simulation.
- A **prespecified GO / FLAG / STOP rule** with software-enforced outcome-access gates on Stage 4 estimators.
- **Negative-control screening**, **E-value / tipping-point** helpers, **three-way Love plot**, **influence-curve diagnostics**.

The comparator package is *just as good* on: IPW cumulative risk, g-computation, AIPW, IPW HR API, Table 1 / Table 2 / weight summary / forest plot, and the model-specification grammar.

## Q3. Is the cleanTMLE vignette as readable as the comparator's vignettes for a new user?

**Not yet.** The comparator's ACTG, WIHS, and leukemia vignettes open with a real dataset and a clinical question, stack incremental models on one risk-curve overlay, and render output after every code chunk. The cleanTMLE staged-analysis vignette currently opens with governance vocabulary, uses a simulated dataset with no clinical anchor, and leaves several teaching units (the four-step TMLE, the DQ stress test) running `eval = FALSE` so the reader never sees the output.

The honest near-term moves:

1. Port one comparator vignette (ACTG or WIHS) to cleanTMLE so the reader gets a real-data, curve-first tutorial on the same problem.
2. Add a concept primer at the head of the staged vignette defining clever covariate, EIC, plasmode, ESS, SMD, MDD, and IPCW in one place.
3. Render the four-step TMLE output between chunks rather than only at the end.
4. Add a stacked-estimators risk-curve overlay (crude → IPW → g-comp → AIPW → TMLE).

These are tracked in *TODO.md* as the highest-priority documentation gap.

The present document (`cleanTMLE_for_applied_analysts.qmd`) is the package's first attempt at a plain-language tutorial for an analyst with no prior TMLE or clean-room background.

## Q4. Is the outcome-blind simulation functionality complete?

**Mostly, with three specific limitations.** The current implementation correctly performs the FIORD five-step procedure (define goals, specify DGP, generate replicates, apply candidate estimators factorially, compare on prespecified metrics) on the hybrid DGP that uses the empirical (W, A) structure with a covariate-only outcome model. The DQ stress test correctly extends this to four prespecified threats (covariate missingness, exposure misclassification, outcome misclassification, unmeasured confounding) and produces degradation gradients on a common scale.

What is *not* yet complete:

1. **The FIORD two-stage selector is now complete.** The point-estimator stage is `select_tmle_candidate(rule = "fiord_two_stage")`, which screens candidates to those achieving nominal oracle coverage (a CI from the Monte Carlo empirical SE) and locks in the lowest-SE survivor; the variance-method stage is `select_variance_method()`, with `bootstrap_rd_variance()` providing the native nonparametric bootstrap that is the principled variance for stabilised IPTW and matched-cohort estimators. The remaining work is convenience: folding the variance stage into a single workflow call rather than a separate step.
2. **The DGP spectrum** beyond the hybrid mode. Fully investigator-specified (structural-equation) and parametric-bootstrap DGPs are documented as planned (see the "Outcome-blind DGP mode" subsection in `cleanTMLE-functions.qmd`).
3. **Nonparametric bootstrap variance** is the FIORD case-study winner for IPTW and matched estimators; cleanTMLE 0.1 supports IF-based and TMLE-robust variance natively and recommends bootstrap externally. A native bootstrap path is on the roadmap.

The simulation we report in the manuscript (Scenarios A / B / C × four DQ threats) exercises the full hybrid-mode pipeline and demonstrates that the workflow returns the correct GO / FLAG / STOP verdict in each scenario before unblinding.

## Using cleanTMLE without the staged-analysis governance layer (minimal mode)

cleanTMLE can be used as a plain outcome-blind RWE estimation package without the staged-analysis governance machinery (audit log, hash, gate, GO / FLAG / STOP rule). For users who want unbiased estimation with the package's design diagnostics but not the full clean-room workflow:

```{r}
#| eval: false
library(cleanTMLE)

# Create a "simple" lock that does NOT activate the software-enforced
# outcome guard, the gate authorisation, or the audit-trail requirement.
# The fingerprint and design diagnostics remain available; outcome
# blindness becomes the analyst's responsibility rather than software-
# enforced.
lock <- create_simple_lock(
  data       = my_data,
  treatment  = "A",
  outcome    = "Y",
  covariates = baseline_covariates,
  sl_library = c("SL.glm", "SL.glmnet"),
  seed       = 2026
)

# Design diagnostics (same as the full workflow)
ps      <- fit_ps_superlearner(lock)
ps_diag <- compute_ps_diagnostics(ps)
plot(ps_diag)
love_plot_threeway(lock, ps$weights)

# Estimation (no authorisation step needed)
fit_tmle  <- run_clean_tmle(lock)
fit_iptw  <- run_iptw_workflow(lock)
fit_gcomp <- estimate_gcomprisk(lock)
fit_aipw  <- estimate_aipwrisk(lock)
```

`create_simple_lock()` returns a lock with `cleanroom_enabled = FALSE`, which short-circuits the internal outcome-access check. Use this when:

- you want to compare estimators on a real-data analysis without buying into the GO / FLAG / STOP framework
- you are running a quick exploratory analysis or a methods-development comparison
- you have an external governance arrangement (institutional clean room, role separation) that already provides the outcome-access controls the software would otherwise enforce

For high-stakes confirmatory studies, use the default `create_analysis_lock()` and the full staged workflow with the gate.

## How the audits were run

Two independent read-only audits were run on the package:

- **Function-usage and TMLE-correctness audit.** Every exported function was cross-referenced against R/, tests/, vignettes/, manuscript, and case study. 99 of ~115 exports were found to be actively used; 14 were flagged as dead exports (advertised but not exercised) and are listed in *TODO.md* with recommended actions. The TMLE implementation was verified against the canonical four-step recipe (initial Q, propensity g, targeting via logistic fluctuation submodel, plug-in psi + IF variance). Both the hand-rolled modular path (`run_tmle_targeting_step()` and friends) and the `tmle::tmle()` delegation path (`estimate_tmle_risk_point()`, `run_ipcw_tmle()`) implement the recipe correctly. Three minor issues were identified and either fixed in this revision (silent targeting-step fallback now emits a warning; IPCW fallback warning now states the loss of double robustness; matched-TMLE EIC carries an in-source note about the paired-design SE caveat) or tracked in *TODO.md* (block-bootstrap variance for matched TMLE).
- **Feature-parity audit** against the comparator cumulative-risk reporting package, summarised in Q1–Q3 above.

A one-paragraph takeaway follows.

# A one-paragraph takeaway

cleanTMLE is software for running an observational study in stages, so that every analytic decision is recorded before anyone sees the comparative result. It uses targeted maximum-likelihood estimation because that method lets the analyst write down the entire recipe in advance while still using flexible machine-learning models. It adds plasmode simulation and a data-quality stress test, so candidate methods can be scored on a quantitative scale before unblinding. A prespecified rule, following the review-team recommendations of @muntner2024staging, then decides whether to proceed (GO), proceed with a documented concern (FLAG), or not run the comparative analysis as planned (STOP). The contribution is not a new estimator or a new theorem; it is the outcome-blind plasmode stress test and an auditable workflow that gives a reviewer everything they need to reconstruct the analysis.

# References {.unnumbered}

::: {#refs}
:::
