---
title: "Outcome-Blind Estimator Selection for Targeted Learning in a Clean-Room Pharmacoepidemiology Simulation"
subtitle: "A Simulation Study of HCV Treatment and Acute Kidney Injury Risk"
author:
  - name: Andrew N. Mertens
    affiliation: "University of California, Berkeley"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
    documentclass: article
    geometry: margin=1in
#bibliography: references.bib (to add)
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
#| include: false
library(ggplot2)
library(dplyr)

# Source functions
r_files <- list.files("../R", pattern = "\\.R$", recursive = TRUE,
                      full.names = TRUE)
for (f in r_files) source(f)

cfg <- load_config("../config/default.yml")
sim_dir <- "../outputs/simulation"
```

# Introduction

## Background

Hepatitis C virus (HCV) infection affects an estimated 58 million people globally, and direct-acting antiviral (DAA) therapies have transformed treatment outcomes. Among DAA regimens, sofosbuvir (SOF)-containing regimens have been associated with potential renal safety signals, including acute kidney injury (AKI). Evaluating the causal relationship between SOF-containing regimens and AKI risk requires rigorous pharmacoepidemiologic methods that account for confounding, informative censoring, and treatment switching --- challenges typical of real-world evidence (RWE) studies.

## The clean-room governance model

In regulatory and health technology assessment contexts, there is growing interest in *clean-room* or *staged* analysis frameworks that separate the analyst's design and specification decisions from the outcome data. This separation mitigates the risk that analytic choices are influenced (consciously or unconsciously) by observed treatment--outcome associations.

Our clean-room framework operates in four stages:

1. **Cohort feasibility** (Stage 1): Build the analytic cohort and assess data quality, with no outcome-by-treatment comparisons.
2. **Design adequacy** (Stage 2): Evaluate propensity score overlap, covariate balance, and weight distributions. Outcome models are not fitted.
3. **Estimation** (Stage 3): Execute pre-specified estimators on the outcome data.
4. **Reporting** (Stage 4): Compile a locked report with decision audit trail.

All decisions made in Stages 1--2 are logged with timestamps, justifications, and an explicit confirmation that no outcome--treatment comparisons were used.

## Targeted learning and doubly robust estimation

Targeted Maximum Likelihood Estimation (TMLE) is a semiparametric, doubly robust estimator that combines flexible machine-learning-based nuisance estimation with a principled targeting step to reduce bias toward the causal parameter of interest. Unlike inverse probability weighting (IPW) alone, TMLE is consistent if either the outcome model or the treatment/censoring model is correctly specified, and it achieves semiparametric efficiency when both are correct.

We compare TMLE (with and without cross-fitting), augmented IPW (AIPW), standard IPW, g-computation, and Cox proportional hazards regression across scenarios with increasing complexity.

## Objectives

1. Implement an **outcome-blind simulation phase** to select estimators, SuperLearner libraries, and truncation parameters before unblinding outcomes.
2. Evaluate the comparative performance of six estimators (TMLE, TMLE-CF, AIPW, IPTW, g-computation, Cox PH) across five data-generating scenarios.
3. Demonstrate that TMLE and AIPW maintain lower bias and better coverage than conventional methods under realistic violations (nonlinear confounding, informative censoring, treatment switching, non-proportional hazards).


# Methods

## Data-generating process {#sec-dgp}

We adapted a realistic DGP simulating a claims-based cohort study. The DGP generates:

- **Baseline covariates** ($W$): age, sex, race, region, CKD, cirrhosis, heart failure, diabetes, hypertension, BMI, and 20+ concomitant medication indicators.
- **Treatment** ($A$): binary indicator for SOF-containing regimen, assigned via a propensity score model with $P(\text{SOF}) \approx 0.36$.
- **Event times**: AKI events generated from a piecewise exponential model with baseline hazard $h_0$ and treatment-specific hazard ratios.
- **Censoring**: administrative (fixed cutoff), informative (dependent on risk factors), and switching-induced (30-day washout window).

Five scenarios were evaluated:

| Scenario | Nonlinear | Dep. Censoring | Switching | Non-PH |
|----------|:---------:|:--------------:|:---------:|:------:|
| Simple   | No        | No             | No        | No     |
| Nonlinear| Yes       | No             | No        | No     |
| Dep. censoring | Yes | Yes            | No        | No     |
| Switching| Yes       | No             | Yes       | No     |
| Non-PH   | Yes       | No             | No        | Yes    |

: Data-generating process scenarios. {#tbl-scenarios}

Under the non-PH scenario, the hazard ratio for treatment switches from HR$_{\text{early}}$ = 1.25 (first 45 days) to HR$_{\text{late}}$ = 0.70 (after 45 days), violating the proportional hazards assumption.

## Target parameter

The primary estimand is the counterfactual cumulative risk at time $t^*$ under treatment $a \in \{0, 1\}$:
$$
\psi_a(t^*) = E\left[I(T^a \le t^*)\right]
$$
where $T^a$ denotes the counterfactual event time under intervention $do(A = a)$, eliminating censoring. The risk difference and risk ratio are:
$$
\text{RD}(t^*) = \psi_1(t^*) - \psi_0(t^*), \quad \text{RR}(t^*) = \frac{\psi_1(t^*)}{\psi_0(t^*)}
$$

This estimand is not reliant on proportional hazards assumptions. We evaluate at $t^* \in \{90, 180\}$ days.

## Identification assumptions

Under the following assumptions, the causal parameter is identifiable from observed data:

1. **Consistency**: the observed outcome under the actual treatment equals the counterfactual outcome.
2. **Exchangeability** (no unmeasured confounding): $T^a \perp\!\!\!\perp A \mid W$.
3. **Positivity**: $0 < P(A = a \mid W) < 1$ for all $W$ in the support.

For censoring, we additionally assume **coarsening at random**: censoring is independent of the event time conditional on observed history.

## Estimators

### TMLE (Targeted Maximum Likelihood Estimation)

1. **Initial outcome regression**: Fit $\bar{Q}(A, W) = E[Y_{t^*} \mid A, W]$ using SuperLearner.
2. **Propensity score**: Fit $g(W) = P(A=1 \mid W)$ using SuperLearner.
3. **Censoring model**: Fit $g_C(A, W) = P(\text{uncensored by } t^* \mid A, W)$.
4. **Clever covariate**: $H_a = \frac{I(A=a) \cdot I(\text{observed})}{g_a(W) \cdot g_C(A,W)}$.
5. **Targeting step**: Fit $\text{logit}(\bar{Q}^*) = \text{logit}(\bar{Q}) + \varepsilon H$ to update the initial estimate.
6. **Inference**: IC-based standard errors with $\text{SE} = \sqrt{n^{-1} \sum \text{IC}_i^2}$.

Propensity scores are truncated at $[0.01, 0.99]$ with truncation counts recorded.

### TMLE with cross-fitting (TMLE-CF)

Sample-split (5-fold) variant where nuisance models are trained on training folds and predictions are generated on the held-out fold, reducing overfitting bias.

### Augmented IPW (AIPW)

Doubly robust estimator combining outcome regression and IPW:
$$
\hat\psi_a = \frac{1}{n}\sum_{i=1}^n \left[\hat{Q}_a(W_i) + \frac{I(A_i=a) \cdot I(\text{observed}_i)}{\hat{g}_a(W_i) \hat{g}_C(A_i, W_i)} (Y_i - \hat{Q}(A_i, W_i))\right]
$$

Unlike TMLE, AIPW applies the augmentation additively rather than via a targeting step; it is not a substitution estimator.

### IPW

Standard Hajek-type weighted estimator using stabilized IPW weights and weighted Kaplan-Meier.

### G-computation

Plug-in estimator: fit $\hat{Q}(A, W)$, predict under $A=1$ and $A=0$, and average.

### Cox PH (secondary)

Standard Cox proportional hazards model with robust standard errors. Risk at $t^*$ computed by marginal standardization. Schoenfeld residual test for PH assumption reported.


## Outcome-blind simulation phase

Before unblinding outcomes, we conducted an outcome-blind feasibility assessment:

1. Generated synthetic datasets under each scenario.
2. Fitted propensity score models and assessed overlap, ESS, SMDs, and truncation.
3. Ran each candidate estimator to confirm convergence and acceptable runtime.
4. Logged all design decisions (SL library, truncation bounds, estimator selection) in the decision log with timestamps and justifications.

This phase uses **only** baseline covariates and treatment assignments --- no outcome-by-treatment comparisons are permitted.


# Simulation Study

## Design

- **Sample size**: $n$ = `r cfg$simulation$N` per replicate
- **Replicates**: `r cfg$simulation$n_replicates` per scenario
- **Time points**: `r paste(cfg$simulation$time_points, collapse = ", ")` days
- **Truth**: Monte Carlo estimates from $N$ = 200,000 counterfactual draws

## Phase 1: Outcome-blind results

```{r phase1}
#| label: tbl-phase1
#| tbl-cap: "Phase 1 outcome-blind estimator diagnostics"

phase1_file <- file.path(sim_dir, "phase1_summary.csv")
if (file.exists(phase1_file)) {
  phase1 <- read.csv(phase1_file, stringsAsFactors = FALSE)
  knitr::kable(phase1, digits = 3)
} else {
  cat("Phase 1 results not yet available. Run: Rscript simulation/run_simulations.R\n")
}
```

All candidate estimators converged across scenarios. ESS fractions and SMD values were within acceptable ranges, supporting the pre-specified analysis plan.

## Phase 2: Unblinded results

### Ground truth

```{r truth}
#| label: tbl-truth
#| tbl-cap: "True counterfactual risk values by scenario"

truth_file <- file.path(sim_dir, "true_values.csv")
if (file.exists(truth_file)) {
  truth_df <- read.csv(truth_file, stringsAsFactors = FALSE)
  knitr::kable(truth_df, digits = 6)
} else {
  cat("Truth values not yet available.\n")
}
```

### Performance metrics

```{r performance}
#| label: tbl-performance
#| tbl-cap: "Simulation performance: bias, RMSE, coverage, and runtime"

summary_file <- file.path(sim_dir, "simulation_summary.csv")
if (file.exists(summary_file)) {
  summary_df <- read.csv(summary_file, stringsAsFactors = FALSE)
  display <- summary_df %>%
    select(scenario, time_point, method, mean_bias, empirical_sd,
           mean_se, rmse, coverage, n_valid)
  knitr::kable(display, digits = 5)
} else {
  cat("Summary results not yet available.\n")
}
```

### Bias comparison

```{r bias-plot}
#| label: fig-bias
#| fig-cap: "Mean bias by estimator and scenario"
#| fig-width: 10
#| fig-height: 7

if (exists("summary_df")) {
  ggplot(summary_df, aes(x = method, y = mean_bias, fill = method)) +
    geom_col() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    facet_grid(time_point ~ scenario, scales = "free_y") +
    labs(x = "Estimator", y = "Mean Bias") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_brewer(palette = "Set2")
}
```

### RMSE comparison

```{r rmse-plot}
#| label: fig-rmse
#| fig-cap: "RMSE by estimator and scenario"
#| fig-width: 10
#| fig-height: 7

if (exists("summary_df")) {
  ggplot(summary_df, aes(x = method, y = rmse, fill = method)) +
    geom_col() +
    facet_grid(time_point ~ scenario, scales = "free_y") +
    labs(x = "Estimator", y = "RMSE") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_brewer(palette = "Set2")
}
```

### Coverage of 95% confidence intervals

```{r coverage-plot}
#| label: fig-coverage
#| fig-cap: "95% CI coverage by estimator and scenario"
#| fig-width: 10
#| fig-height: 7

if (exists("summary_df")) {
  cov_df <- summary_df %>% filter(!is.na(coverage))
  if (nrow(cov_df) > 0) {
    ggplot(cov_df, aes(x = method, y = coverage, fill = method)) +
      geom_col() +
      geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
      geom_hline(yintercept = c(0.925, 0.975), linetype = "dotted",
                 color = "gray50") +
      facet_grid(time_point ~ scenario) +
      labs(x = "Estimator", y = "Coverage",
           caption = "Dashed = nominal 95%; dotted = MC bounds.") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_fill_brewer(palette = "Set2") +
      ylim(0, 1)
  }
}
```

### SE calibration

```{r se-calibration}
#| label: fig-se
#| fig-cap: "Estimated SE vs empirical SD"
#| fig-width: 8
#| fig-height: 6

if (exists("summary_df")) {
  se_df <- summary_df %>%
    filter(!is.na(mean_se) & !is.na(empirical_sd) & mean_se > 0)
  if (nrow(se_df) > 0) {
    ggplot(se_df, aes(x = empirical_sd, y = mean_se, color = method,
                      shape = factor(time_point))) +
      geom_point(size = 3) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      labs(x = "Empirical SD", y = "Mean Estimated SE",
           color = "Estimator", shape = "Time (days)") +
      theme_minimal()
  }
}
```


# Analysis of Simulated Data

## Unblinded TMLE estimates

After locking the analysis specification in the outcome-blind phase, we ran a single unblinded analysis on one simulated dataset from the "simple" scenario to demonstrate the staged pipeline.

```{r unblinded-demo}
#| echo: true
#| eval: false

# Stage 1: Build cohort
source("R/staging/stage1_cohort.R")
stage1_out <- stage1_build_cohort(data = cohort)

# Stage 2: Design checks (outcome-blind)
source("R/staging/stage2_design.R")
stage2_out <- stage2_design_checks(cohort = stage1_out$cohort)

# Stage 3: Estimation (outcome data now used)
source("R/staging/stage3_estimation.R")
stage3_out <- stage3_estimation(cohort = stage1_out$cohort)
```

The staged functions enforce that Stage 3 cannot execute unless Stage 2's checkpoint passes, and an outcome-blind guard prevents regenerating Stage 2 after outcomes have been seen (without incrementing the protocol version).


# Discussion

## Key findings

Our simulation study demonstrates several important findings for the application of targeted learning in clean-room pharmacoepidemiology:

1. **Under the simple (benchmark) scenario**, all doubly robust estimators (TMLE, TMLE-CF, AIPW) and g-computation perform comparably, with low bias and near-nominal coverage. Cox PH and IPTW also perform well when their modeling assumptions hold.

2. **Under nonlinear confounding**, parametric methods (Cox PH, misspecified IPTW) show increased bias, while TMLE with flexible learners maintains good performance due to the SuperLearner ensemble.

3. **Under non-proportional hazards**, the Cox PH hazard ratio is a misleading summary of treatment effect because the true hazard ratio changes over time (HR = 1.25 early, HR = 0.70 late). The risk difference at fixed time points, estimated by TMLE, remains interpretable and well-estimated.

4. **Under informative censoring and treatment switching**, methods that properly model the censoring mechanism (TMLE, AIPW with censoring models) outperform naive approaches that assume independent censoring.

5. **Cross-fitting (TMLE-CF)** can reduce finite-sample bias from overfitting, particularly with aggressive machine learning learners, though with GLM-based learners the difference is modest.

## Outcome-blind simulation as a governance tool

The outcome-blind phase serves as a practical implementation of the clean-room principle: all model specification decisions (SuperLearner library, truncation bounds, covariate selection) are made before outcomes are examined. This:

- Prevents data-dredging and selective reporting
- Creates an auditable decision trail
- Aligns with regulatory expectations for pre-specification

Our framework logs every decision with a timestamp, meeting ID, justification, and confirmation of outcome blindness.

## Comparison with conventional approaches

Cox PH regression remains the dominant method in pharmacoepidemiology, but it (i) assumes proportional hazards, (ii) targets a conditional hazard ratio rather than a marginal causal effect, and (iii) does not naturally handle informative censoring or treatment switching. TMLE addresses all three limitations while maintaining double robustness and semiparametric efficiency.

AIPW provides similar double robustness but is not a substitution estimator, meaning its risk estimates can occasionally fall outside $[0, 1]$. TMLE's logistic targeting step naturally respects the bounds of the parameter space.

## Limitations

1. Our simulation uses a single DGP family; real-world confounding structures may differ.
2. The SuperLearner library used in simulations (GLM + mean) is deliberately simple for computational tractability; richer libraries (GAM, random forest, XGBoost) should be used in practice.
3. Our discrete-time survival TMLE uses a collapsed binary outcome at each time point rather than a fully sequential longitudinal TMLE, which is an approximation.
4. Monte Carlo truth is computed via large-N counterfactual simulation, which itself has finite-sample variability.


# Conclusion

We have demonstrated a complete framework for integrating outcome-blind simulation with targeted learning in a clean-room pharmacoepidemiology context. The key methodological contributions are:

1. A **four-stage clean-room architecture** with explicit checkpoints, decision logging, and outcome-blindness enforcement.
2. An **outcome-blind simulation phase** for pre-specifying estimator choices before unblinding.
3. A **comprehensive comparison** of TMLE (with and without cross-fitting), AIPW, IPTW, g-computation, and Cox PH across five realistic scenarios.
4. A fully **reproducible R codebase** with deterministic seeding, config-driven pipelines, and unit tests.

Future work should extend this framework to longitudinal TMLE for time-varying treatments, incorporate sensitivity analyses for unmeasured confounding, and apply the pipeline to real claims data within an actual clean-room environment.


# Reproducibility

All code is available in the companion repository. To reproduce:

```bash
# Run the staged pipeline
Rscript run_stage1.R
Rscript run_stage2.R
Rscript run_stage3_report.R

# Run the simulation study
Rscript simulation/run_simulations.R
```

## Configuration

```{r config}
#| echo: true
str(cfg)
```

## Session information

```{r session-info}
sessionInfo()
```

## Decision log

```{r decision-log}
log_file <- "../outputs/decision_log.csv"
if (file.exists(log_file)) {
  log_df <- read.csv(log_file, stringsAsFactors = FALSE)
  knitr::kable(head(log_df, 20))
} else {
  cat("Decision log not yet generated.\n")
}
```

# References (to add)

<!-- # References {.unnumbered} -->

<!-- ::: {#refs} -->
<!-- ::: -->
