---
title: "cleanTMLE: An R Package for Staged, Outcome-Blind Targeted Learning with Plasmode Data-Quality Stress Testing"
author:
  - name: "Andrew Mertens"
    affiliation: "University of California, Berkeley"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
    fig-width: 8
    fig-height: 5
    number-sections: true
bibliography: references.bib
csl-hanging-indent: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "R> ",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(ggplot2)
library(cleanTMLE)
```

# Abstract {.unnumbered}

Real-world evidence (RWE) for regulatory and reimbursement
decisions has to clear two hurdles. The analysis must use a valid
estimator, and the data must be fit for the research question. The
first hurdle is usually cleared with simulation. The second is
usually cleared with a written template such as the Structured
Process to Identify Fit-for-Purpose Study Design and Data, or
SPIFD2 [@gatto2022spifd; @gatto2023spifd2]. The two checks are
performed separately and live in different documents. We argue
that the second check can be folded into the first, and that the
natural place to do this is the outcome-blind plasmode-simulation
stage of a staged Targeted Maximum Likelihood Estimation (TMLE)
workflow [@vanderLaan2006tmle; @vdL2011tlbook; @dang2023roadmap;
@gruber2022sap].

This paper introduces **cleanTMLE**, an R [@rcoreteam2024] package
that implements such a workflow. The package contributes three new
software primitives. First, `run_plasmode_dq_stress()` extends the
standard outcome-blind plasmode loop with four data-quality (DQ)
degradation mechanisms (covariate missingness, treatment
misclassification, outcome misclassification, and unmeasured
confounding), producing a degradation gradient: a curve of bias,
RMSE, and confidence-interval coverage at increasing levels of each
threat, computed separately for each prespecified TMLE candidate.
Second, `gate_check()` and `gate_all()` consume the same plasmode
metrics together with stage-specific checkpoints
(`checkpoint_cohort_adequacy()`, `checkpoint_balance()`,
`checkpoint_residual_bias()`) to produce a single GO/FLAG/STOP
decision per stage. Third, the `cleanroom_lock` object, the audit
log, and the decision log are persisted as hashed RDS artifacts and
guard the outcome column at run time: every Stage 4 estimator
refuses to run unless `authorize_outcome_analysis()` has signed off.
Together these primitives turn the paper-only prespecification
checklists of the TL-RWE literature [@gruber2022sap] into machine-
readable, hash-verified artifacts.

We illustrate the framework in a Monte Carlo simulation study
comparing six estimators (crude, IPTW, propensity-score matching,
TMLE, cross-validated TMLE, and matched-cohort TMLE) under good
overlap, marginal overlap, and unmeasured confounding. The DQ
stress test correctly identifies scenarios in which estimator
performance degrades and the pipeline should be flagged. cleanTMLE
reuses the cohort-specification and reporting functions of the
**causalRisk** package [@brookhart2024causalrisk] — an MIT-licensed
R package that implements IPW, g-computation, and AIPW estimators
with a clean-room-friendly setup — and adds the lock, gate, and
DQ-stress primitives that `causalRisk` does not provide.

**Keywords:** targeted maximum likelihood estimation, plasmode
simulation, real-world evidence, fit-for-purpose data, outcome-
blind analysis, R, statistical analysis plan, regulatory science.

# Introduction

## Why fit-for-purpose data needs a quantitative bridge

Modern frameworks for RWE study design ask the analyst to write
down, before the outcome is accessed, both an analysis plan and an
argument that the data are fit for the research question
[@dang2023roadmap; @gatto2022spifd; @gatto2023spifd2;
@wang2021startrwe]. In practice these two arguments arrive in
different documents and use different evidence. The analysis plan
is checked with simulation: the analyst runs an outcome-blind
plasmode and reports bias, root-mean-squared error (RMSE), and
confidence-interval (CI) coverage of each candidate estimator
[@franklin2014plasmode; @schuler2017tmle; @dang2023roadmap]. The
fit-for-purpose argument is recorded with a stepwise template such
as SPIFD2 [@gatto2023spifd2], which lists each design element of
the hypothetical target trial, the data-source feature that is
expected to operationalise it, and the rationale for any
deviation. The simulation is numeric. The template is qualitative.

The same outcome-blind stage that hosts the simulation could,
without any new data, also host the data-quality argument. A
plasmode generates synthetic outcomes from a parametric model
fitted to the real covariate and treatment distributions. Because
the data-generating process is under the analyst's control, the
analyst can degrade it: introduce covariate missingness, flip a
fraction of the treatment assignments, apply known sensitivity and
specificity to the outcome, or inject a latent unmeasured
confounder. Each degradation produces a bias, an RMSE, and a
coverage value, indexed by candidate. A robust candidate degrades
gradually as the threat increases; a fragile candidate shows a
cliff. The output is a set of degradation curves, on the same units
as the standard plasmode, that the gate function can consume
alongside the rest of the pre-outcome checks.

This is the bridge we propose. The plasmode DQ stress test sits at
the same workflow stage as the SPIFD2 review, takes its inputs from
the same data feasibility documents, and produces a quantitative
output that the GO/FLAG/STOP gate can score against pre-specified
thresholds. The qualitative SPIFD2 narrative is preserved, but the
analyst no longer has to decide between "documented but
unquantified" data threats and "quantified but undocumented"
estimator performance. They are part of the same gate decision.

## Why TMLE and why a clean room

The DQ stress idea does not require TMLE. It requires only an
outcome-blind plasmode loop that fits a candidate estimator on
synthetic outcomes. We embed it in a TMLE workflow because TMLE
[@vanderLaan2006tmle; @vdL2011tlbook; @schuler2017tmle] is
explicitly designed for prespecification: every analytic decision,
including the SuperLearner library and the truncation rule, can be
listed in advance and the algorithm remains data-adaptive while
still admitting valid Wald-type confidence intervals
[@vdL2007superlearner; @phillips2023sl; @gruber2022sap]. TMLE is
also a natural fit for the staged regulatory frameworks of
@dang2023roadmap and @petersen2014roadmap, which separate causal
identification, statistical estimation, and sensitivity analysis
into prespecified steps.

"Clean room" is the term @muntner2024staging use for the same
access-control behaviour cleanTMLE enforces in code; the staged
construct itself was formalised in that paper and is being adopted
across the field, including in the FDA-funded RCT-DUPLICATE
programme [@franklin2020duplicate; @wang2023rctduplicate], the
Sentinel Innovation Center [@desai2024sentinel], and Amgen's
Center for Observational Research [@kent2026fiord]. cleanTMLE
operationalises that construct in software. The published TL-RWE
literature [@dang2023roadmap; @gruber2022sap; @gruber2023future]
uses the term "outcome-blind." We use both terms but prefer "clean
room" when talking about software, because the package literally
refuses to let Stage 4 estimators run on a lock whose outcome
column has not been authorised. The same way a manufacturing
clean room blocks access to a sterile area until protocols are
followed, cleanTMLE blocks the outcome column until the gate has
signed off. The mechanism is a `mask_outcome()` /
`assert_outcome_authorized()` guard that every estimator function
calls before fitting. This is an implementation pattern, not a
methodological novelty.

## What the paper contributes

The paper makes three connected contributions:

1. **A new use of plasmode simulation.** We extend the outcome-
   blind plasmode loop from candidate-estimator selection
   [@dang2023roadmap; @gruber2022sap] to candidate-estimator
   stress testing under realistic data-quality threats. Each
   threat is mapped to a parameter that can be varied in the
   plasmode; each candidate produces its own degradation gradient.
   This is, to our knowledge, the first published implementation
   of plasmode simulation as a quantitative bridge from SPIFD2
   to a GO/FLAG/STOP gate. We elaborate on this contribution in
   §\ref{sec-bridge}.

2. **An R implementation that turns prespecification into a
   verifiable artifact.** The `cleanroom_lock` carries a hash; the
   audit log records every checkpoint with a timestamp and a
   rationale; the decision log records every protocol deviation;
   `authorize_outcome_analysis()` checks all three before a Stage 4
   estimator can fit. The lock and audit are saved as RDS files via
   `save_lock()` / `save_audit()` and re-validated on load.

3. **Compatibility with an existing IPW-focused R package.** We
   reuse the cohort-specification and reporting functions of
   `causalRisk` [@brookhart2024causalrisk], an MIT-licensed package
   that implements IPW, g-computation, and AIPW estimators with a
   clean-room-friendly setup. cleanTMLE adds the lock, gating, and
   plasmode DQ stress primitives that `causalRisk` does not
   provide; an analyst familiar with `causalRisk` can use cleanTMLE
   without relearning the cohort-specification surface.

## Paper structure

§\ref{sec-estimand} states the estimand and identifiability
assumptions used throughout the simulation. §\ref{sec-framework}
describes the staged workflow and the access-control semantics of
the lock. §\ref{sec-dq} introduces the DQ stress test, with the
parameter mapping from SPIFD2 categories to plasmode parameters and
the algorithm. §\ref{sec-bridge} discusses the contribution as a
quantitative bridge from SPIFD2 to the gate, including its scope
and what it does and does not replace. §\ref{sec-implementation}
shows the user-facing functions and the implementation of
`run_plasmode_dq_stress()`. §\ref{sec-simulation} reports the
Monte Carlo simulation. §\ref{sec-discussion} discusses limitations
and connections to E-values [@vanderWeele2017evalue], quantitative
bias analysis [@lash2014qba], the targeted-learning causal-gap
sensitivity [@diaz2013sensitivity], and negative-control outcomes
[@lipsitch2010negative]. §\ref{sec-summary} summarises.

# Estimand, Causal Model, and Identifiability {#sec-estimand}

## Estimand specification

We follow @ich2019e9r1 and @dang2023roadmap by writing the estimand
along five attributes. For the running example the population is
adults eligible at the index visit. The treatment is a binary point
exposure $A \in \{0, 1\}$ at index. The endpoint is a binary
indicator of the primary event by 24 months, denoted $Y$. The
summary measure is the marginal counterfactual risk difference
$\Psi = E\{Y(1)\} - E\{Y(0)\}$. Intercurrent events are handled
under a treatment-policy strategy; censoring and treatment
switching are not modelled in this illustration. The corresponding
statistical estimand under the identification assumptions in
§\ref{sec-id} is $\Psi = E_W\{E[Y \mid A=1, W] - E[Y \mid A=0, W]\}$,
the marginal g-formula contrast [@robins1986newapproach;
@hernan2020causal].

In the package this specification is attached to the lock with
`attach_estimand()` at Stage 1a and is preserved on the lock object
that downstream stages consume. The estimand fields are not part of
the integrity fingerprint computed by `create_analysis_lock()`;
analysts who want a tamper-evident estimand record should pair
`attach_estimand()` with a `record_decision_log_entry()` call so
that any later modification is visible in the exported decision
log.

## Causal model

The data-generating process has five baseline covariates
$W = (\text{age}, \text{sex}, \text{biomarker}, \text{comorbidity}, \text{CKD})$,
a binary treatment $A$ that depends on $W$, and a binary outcome
$Y$ at 24 months. In the unmeasured-confounding scenario a binary
$U \sim \text{Bern}(p_U)$ also affects both $A$ and $Y$. The DAG
is

```
   W ───► A ──► Y
   │          ▲
   └──────────┘
   (Scenario C only:)
   U ──► A
   U ──► Y
```

with no arrow from $A$ to $W$ (point exposure) and no arrow
between $W$ and $U$.

## Identifiability assumptions {#sec-id}

Under this DAG, $\Psi$ is identified by the g-formula if the
following six assumptions hold [@hernan2020causal]. We list each
assumption together with the DQ stress-test scenario that probes
it; this mapping is one of the main reasons the DQ stress test is
useful.

- *Conditional exchangeability*, $Y(a) \perp\!\!\!\perp A \mid W$.
  This holds in scenarios A and B by construction. It is violated
  by construction in scenario C, where $U$ is not in the analyst's
  adjustment set. **Probed by:** the unmeasured-confounding scenario
  in `run_plasmode_dq_stress()`.
- *Positivity*, $0 < \Pr(A = 1 \mid W) < 1$ for every $W$ in the
  support. Scenario A satisfies this comfortably. Scenario B
  pushes the propensity toward the boundary but does not violate
  it in the limit. **Probed by:** the truncation level of each TMLE
  candidate; near-positivity violations show up as inflated SE/SD
  in the baseline plasmode metrics.
- *Consistency / SUTVA*, $Y = Y(A)$. **Probed by:** the treatment-
  misclassification scenario, which simulates a violation of
  consistency caused by exposure-algorithm error.
- *No measurement error in the outcome*, $\tilde{Y} = Y$. **Probed
  by:** the outcome-misclassification scenario, with optional
  per-arm differential misclassification.
- *Stratified MCAR for the covariates*, $W$ is missing completely
  at random within strata of fully measured variables. **Probed
  by:** the covariate-missingness scenario, which sets a fraction
  of values to NA and imputes with the column median.
- *Correctly specified nuisance models* for $g(W)$ and $E[Y \mid A,
  W]$, including the SuperLearner library and the truncation rule.
  **Probed by:** the standard plasmode candidate-selection metrics,
  which compare candidates that vary the SuperLearner library and
  the truncation rule.

The simulation thus contains one analytic challenge (marginal
overlap, B) and one identifiability violation (unmeasured
confounding, C). We use C to demonstrate what the DQ stress test
detects when an unmeasured confounder is anticipated, and to
contrast the stress-test output with the post-hoc E-value
[@vanderWeele2017evalue] and with the negative-control-outcome
check [@lipsitch2010negative].

# The cleanTMLE Framework {#sec-framework}

## Staged workflow

cleanTMLE adopts the staged design described informally in
@muntner2024staging and formalised in @dang2023roadmap and
@gruber2022sap. Stage names below are package-internal labels;
the stage *content* matches Steps 0–6 of @dang2023roadmap.

| Stage | Purpose | Does the function read the outcome? |
|-------|---------|:------------:|
| 0 | Feasibility assessment, utilisation study, draft protocol/SAP [@muntner2024staging; @kent2026fiord] | No (no data access yet) |
| 1a | Lock estimand, covariates, SuperLearner library, candidate grid | No |
| 1b | Cohort adequacy: $N$ per arm, events, design precision | Events only |
| 2a | Propensity-score diagnostics: overlap, ESS, balance | No |
| 2b | Outcome-blind plasmode candidate selection | No (synthetic outcome) |
| **2b\*** | **Plasmode DQ stress test (proposed)** | **No (synthetic outcome)** |
| 3 | Negative-control outcomes, residual-bias check | NC outcomes only |
| Gate | Pre-outcome authorisation: GO/FLAG/STOP | No |
| 4 | Primary comparative analysis | **Yes** |

Stages 1–3 do not access the real outcome. The plasmode at Stage
2b generates synthetic outcomes from a parametric model fitted to
the real covariates and treatment, which is why it is a natural
host for the DQ stress test we describe in §\ref{sec-dq}.

## Access control in code

The lock is a tagged R object created by `create_analysis_lock()`.
It carries the data, the column names (treatment, outcome,
covariates), the SuperLearner library, the random seed, and a
SHA-256 fingerprint of those fields together with the data shape
and column names (computed via the `digest` package). `save_lock()`
and `load_lock()` persist the lock and re-verify the fingerprint on
load. The estimand (`attach_estimand()`) and the prespecified
candidate (`lock_primary_tmle_spec()`) are attached after the
fingerprint is computed and are protected by the audit log rather
than by the fingerprint itself; any modification to those fields
appears in `export_audit_trail()` as a separate entry. `mask_outcome()` returns a copy of the
lock with the outcome column replaced by `NA` and a flag set on the
lock; `unmask_outcome()` reverses this when the gate has signed
off. Every Stage 4 estimator (`run_iptw_workflow()`,
`run_match_workflow()`, `fit_tmle_outcome_mechanism()`, and so on)
calls an internal `.check_outcome_access()` guard that stops with
an informative error if the lock's outcome flag is still masked.
The same guard checks for an `authorize_outcome_analysis()` token
that is produced only when every required checkpoint has decided
GO or FLAG (not STOP).

The audit log accumulates entries automatically from the checkpoint
functions; the decision log records analyst choices and protocol
deviations. Both can be exported as tidy data frames with
`export_audit_trail()` and `export_decision_log()`, archived as RDS
with `save_audit()`, and reloaded later with `load_audit()`. Every
entry carries the lock hash, so any post-hoc tampering with the
lock invalidates the audit. The exported decision log mirrors the
published Romiplostim worked example [@mcgrath2021romiplostim],
which @muntner2024staging cite as the canonical model for what a
clean-room decision log should look like.

Beyond a single binary outcome-mask, @muntner2024staging and
@kent2026fiord recommend tiered, role-based access in which
analytic advisors see summary outputs by stage and checkpoint
reviewers see summary outputs only by checkpoint. cleanTMLE's
`export_audit_trail()` and `export_decision_log()` are the
software analogue of that arrangement: they produce programmatic
summaries that can be exposed to a reviewer role without granting
the reviewer raw-data access.

## Outcome-blind simulation versus the traditional clean-room

A traditional clean-room design is an organisational arrangement
[@muntner2024staging]. A blinded analyst writes the analysis
plan, fixes the cohort, runs every Stage 1 and Stage 2 diagnostic,
and signs off on the gate. Only after the gate is signed off does
a separate, unblinded analyst (or a different role on the same
team) link the outcome and run the comparative analysis. The
separation is enforced by data access controls: the blinded
analyst's view of the database does not include the outcome
column, and a data manager polices the boundary. Reviewers and
regulators see two artifacts at the end: a frozen analysis plan
with a timestamp, and a final estimate produced under that plan.

Outcome-blind simulation, the approach we use in this paper, is a
software arrangement instead. A single analyst can iterate on the
candidate grid, on the gate thresholds, and on the data-quality
stress scenarios, because the iteration runs against synthetic
outcomes generated from a covariate-only model rather than the
real treatment-outcome data. The lock and the access guard
(`mask_outcome()`, `assert_outcome_authorized()`) replace the
data manager. The audit log replaces the timestamped paper plan.
Reviewers see the same two artifacts as before, except now both
are machine-readable and re-executable.

The two approaches address the same underlying concern — that
analytic decisions made after seeing the treatment-outcome
association are not credible — but they have different costs and
different residual risks. The traditional clean-room design
requires a partition of personnel and an institutional commitment
to the partition. It tolerates more ad-hoc deviation by the
blinded analyst because the only ground truth is the paper plan.
The outcome-blind simulation approach removes the personnel
constraint and makes the deviation log machine-readable, but it
relies on the analyst trusting that the synthetic-outcome
mechanism is faithful enough to the real data-generating process
that the simulation-driven candidate selection generalises. We
therefore see the two as complementary, not competing: a fully
defensible RWE pipeline can use the traditional clean-room
arrangement *and* run cleanTMLE inside it, getting the benefit of
quantitative gating on top of the personnel-based partition.

The advantages of the outcome-blind simulation approach, on its
own, are these:

- *No institutional partition required.* A single statistician can
  use cleanTMLE without a separate unblinded role, and still
  produce a documented prespecification.
- *Iteration without unblinding.* The candidate grid, the
  truncation thresholds, and the gate thresholds can all be
  revised after seeing the plasmode results, because the plasmode
  outcomes are synthetic. The traditional clean-room design has
  to re-freeze the plan if the analyst sees something they want
  to change.
- *Reproducibility.* The lock hash, the audit log, and the
  decision log together let any reviewer re-run the entire
  pre-outcome workflow byte-for-byte.
- *Stress testing.* Because the simulation is under the analyst's
  control, the same loop that picks the candidate can also stress
  the candidate against realistic data threats. The traditional
  clean-room design has no equivalent step; data fitness is
  argued in prose.

The disadvantages are real and worth saying out loud:

- *The plasmode uses the real Y to fit the covariate-only Q0
  model.* The analyst learns the marginal distribution of $Y$
  given $W$, but not the joint $(A, Y) \mid W$ relationship. The
  Dang causal-roadmap reading of "outcome-blind" allows this; a
  stricter clean-room reading would require even the marginal
  $E[Y \mid W]$ to come from external pilot data
  [@dang2023roadmap, p. 7]. cleanTMLE supports both modes: passing
  an external Q0 fit to `run_plasmode_feasibility()` is a
  documented option, and the lock records which mode was used.
- *Parametric DGP for the synthetic outcome.* The plasmode adds a
  fixed risk difference to a logistic Q0; effect modification is
  not modelled. A richer DGP is possible but is not exercised in
  the simulation reported here.
- *Single analyst, single trust boundary.* If the analyst is also
  the one writing the plasmode and stress-test code, there is no
  independent check that the synthetic-outcome mechanism is
  faithful. The traditional clean-room arrangement, with a
  separate data manager, has an extra layer of cross-validation
  that the software-only workflow does not.

We document these trade-offs explicitly in the package's
documentation so that the analyst can choose the level of
formality the study requires. For most pharmacoepidemiology RWE
studies the software-only mode is adequate; for high-stakes
regulatory submissions we recommend running cleanTMLE inside a
traditional clean-room arrangement so that both the institutional
and the computational guarantees apply.

# Data Quality as a Quantifiable Threat {#sec-dq}

## Mapping data-quality concerns to plasmode parameters

SPIFD2 [@gatto2023spifd2] does not list a fixed taxonomy of
"data-quality dimensions"; it is a stepwise template that asks the
analyst to record each design element of the hypothetical target
trial and the data feature that operationalises it. The harmonised
data-quality terminology of @kahn2016data, by contrast, defines
three dimensions (conformance, completeness, plausibility) that
have become standard in OHDSI-style assessments. We map both
SPIFD2 design-element categories and the @kahn2016data dimensions
to plasmode parameters as follows.

```{r spifd2-table}
spifd2_map <- data.frame(
  `SPIFD2 / Kahn category` = c(
    "Covariate measurement / availability (Kahn: completeness)",
    "Treatment exposure definition (Kahn: conformance)",
    "Outcome ascertainment (Kahn: conformance + plausibility)",
    "Unmeasured confounding (SPIFD2 sensitivity step)",
    "Sample-size adequacy (SPIFD2 step 2)"
  ),
  `Plasmode parameter` = c(
    "Fraction of MCAR missingness, pi",
    "(sens, spec) of treatment ascertainment",
    "(sens, spec) of outcome ascertainment, optionally by arm",
    "Latent U with prevalence p_U and (OR_A, OR_Y)",
    "Tested by base plasmode through n"
  ),
  `Source of plausible range` = c(
    "Data profiling, linked validation studies",
    "Chart review, claims-validation literature",
    "Algorithm validation, PPV / sensitivity studies",
    "Expert elicitation, literature, E-value bounds",
    "Known from data source"
  ),
  `Gate criterion (default)` = c(
    "RMSE ratio > 1.5 at observed missingness rate",
    "|Bias| > target at plausible (sens, spec)",
    "|Bias| > target at plausible (sens, spec)",
    "|Bias| > target at plausible U strength",
    "Coverage < min_coverage"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  spifd2_map,
  caption = "Table 1: Mapping data-quality concerns to plasmode-parameter inputs of run_plasmode_dq_stress()."
)
```

The mapping is deliberately one-to-many: a single SPIFD2 design
element (say, outcome ascertainment) can drive several plasmode
parameters (per-arm sensitivity and specificity), and a single
plasmode parameter (say, MCAR fraction) can speak to several SPIFD2
elements (covariate availability, missingness handling).

## The degradation gradient

Rather than testing a single value of each threat, the function
sweeps a vector of severities. The output is a curve of bias,
RMSE, and coverage against the severity, computed separately for
each prespecified TMLE candidate. We call this curve a *degradation
gradient* by analogy with a dose-response curve: it answers the
question, "how bad does the data have to be before this candidate
fails the gate?"

A robust candidate produces a flat gradient that stays inside the
gate thresholds. A fragile candidate shows a cliff: a small
worsening of the data tips bias or coverage past the threshold.
The candidate selection rule (`select_tmle_candidate(rule = "min_rmse")`)
can be augmented to require a maximum allowed degradation, in
which case the gate effectively rules out fragile candidates
before the outcome is accessed.

It is worth saying explicitly that the truncation level inside a
TMLE candidate is not just a tuning knob: trimming or truncating
the propensity score changes the support of the population to
which the estimand applies, and therefore the estimand itself
[@brookhart2024causalrisk; @lash2014qba]. cleanTMLE treats the
truncation level as part of the prespecified candidate, so any
estimand drift induced by truncation is locked into the audit log
and scored on the same RMSE and coverage metrics as everything
else, rather than being chosen post-hoc on the realised data.

# A quantitative bridge from SPIFD2 to GO/FLAG/STOP {#sec-bridge}

This section expands the central contribution of the paper. We
describe how the pre-outcome gate is currently scored in published
TL-RWE pipelines, what changes when the DQ stress test feeds into
the same gate, and what the bridge does and does not replace.

A note on terminology: @muntner2024staging describe three
review-team decisions ("proceed to the next stage", "conduct
additional analyses to address bias", or "terminate the study").
We use the labels GO, FLAG, and STOP throughout this paper for
brevity, and the cleanTMLE software uses the same three labels;
the mapping to the Muntner et al. wording is one-to-one.

## What the gate currently does

In the published TL-RWE literature the pre-outcome gate is a
prespecified set of qualitative checks: the SAP exists, the lock is
in place, the cohort meets minimum size and event criteria, the
propensity-score overlap looks acceptable, and the negative-
control-outcome estimate is consistent with no effect
[@dang2023roadmap; @gruber2022sap; @gruber2024evaluating]. Each
check is binary or ordinal, and each is computed from a different
artifact (a pdf SAP, a balance table, a NCO estimate). When the
data-fitness argument is added, it appears as a separate document,
typically the SPIFD2 or HARPER table [@wang2023harper], and is
adjudicated by reviewers rather than scored by the gate.

## What the bridge does

The DQ stress test reports, for each candidate, the same triple
(bias, RMSE, coverage) at each severity of each threat that the
baseline plasmode reports. Because the units match, the gate can
apply the same thresholds it already uses for the baseline plasmode
— maximum allowed bias, minimum allowed coverage, and an allowed
SE-calibration window — to the degraded runs. The candidate then
has to pass the gate at the ideal data-generating process *and*
stay inside the gate thresholds across the plausible range of each
threat documented in SPIFD2.

Three things change when the gate works this way.

1. **The data-fitness argument becomes numeric, candidate-
   specific, and machine-readable.** Instead of "the outcome
   ascertainment algorithm has reported sensitivity 0.90", the
   gate sees "at sens = 0.90, spec = 0.95 the bias of candidate
   `tmle_glmnet_t05` rises from 0.001 to 0.014, which exceeds the
   prespecified threshold of 0.010". The reviewer no longer has to
   translate between literatures.

2. **Candidate selection picks robust, not just optimal,
   specifications.** The default `select_tmle_candidate()` rule
   minimises RMSE on the baseline plasmode. We augment this with
   `select_tmle_candidate(rule = "min_max_rmse")`, which minimises
   the worst-case RMSE across the prespecified DQ scenarios. A
   candidate that wins on the baseline but collapses under a 5%
   missingness rate loses to a slightly less efficient candidate
   that stays inside the gate thresholds across the entire
   plausible range.

3. **The data-fitness argument is recorded in the same audit log
   as the rest of the prespecification.** A reviewer or a
   regulator can read the audit trail, see the SPIFD2 ranges that
   went in, see the gate thresholds that came out, and reproduce
   the gate decision exactly.

## What the bridge does not replace

The DQ stress test is not a quantitative bias analysis (QBA) of
the kind described by @lash2014qba. QBA adjusts the *final* effect
estimate for a postulated bias and is therefore an unblinding-
stage tool. The plasmode DQ stress test does not adjust the
estimate; it asks whether the candidate estimator can still pass
the gate when the data are degraded. Both belong in a
comprehensive analysis. The DQ stress test goes in the
prespecification; QBA goes in the sensitivity-analysis section
after the outcome is unblinded.

The DQ stress test is also not a substitute for the targeted-
learning causal-gap sensitivity analysis of @diaz2013sensitivity,
which characterises how the estimate changes as a function of an
unmeasured-confounder bound. Causal-gap is an unblinding-stage
tool that complements the E-value [@vanderWeele2017evalue].
Conversely, the DQ stress test reaches threats that neither the
E-value nor causal-gap covers, including covariate missingness and
exposure misclassification.

We therefore position the DQ stress test as the *first* of three
sequential bias analyses: pre-outcome (DQ stress test), at-
outcome (negative-control outcomes [@lipsitch2010negative]), and
post-outcome (E-value, QBA, causal-gap). Each addresses a different
threat, at a different stage, with a different artifact.

## Why "outcome-blind" matters here

The DQ stress test runs on synthetic outcomes generated from the
real covariate distribution and the real treatment. The real
outcome is never read. This is the property that makes the test
admissible inside the prespecification: the analyst can iterate on
the candidate grid, on the degradation parameters, and on the gate
thresholds without learning anything about the treatment-outcome
association in the actual study data [@dang2023roadmap, p. 7].
The lock's hash records the final choices. Once the gate signs
off, the outcome is unmasked and the primary analysis runs once.

# Implementation and `run_plasmode_dq_stress()` {#sec-implementation}

## A worked workflow

A typical cleanTMLE session for a binary point-treatment endpoint
looks like the following. We omit print output for compactness;
each print method is documented in `?print.cleanroom_lock` and the
related help pages.

```{r demo-workflow, eval = FALSE}
library(cleanTMLE)

# Stage 1a: lock the analytic specification
lock <- create_analysis_lock(
  data        = study_data,
  treatment   = "treatment",
  outcome     = "event_24",
  covariates  = c("age", "sex", "biomarker", "comorbidity", "ckd"),
  sl_library  = c("SL.glm", "SL.glmnet", "SL.ranger"),
  seed        = 2026
)
lock <- attach_estimand(lock,
  description           = "Effect of A on 24-month event risk",
  population            = "Adults eligible at index",
  treatment_strategies  = c("Treatment", "Control"),
  outcome_label         = "Primary event by 24 months",
  followup              = "24 months",
  contrast              = "risk_difference"
)

# Stage 1b: cohort adequacy + design precision
audit <- create_audit_log(lock)
cp1   <- checkpoint_cohort_adequacy(lock,
                                    min_n_per_arm = 200,
                                    min_events    = 50)
audit <- record_checkpoint(audit, cp1)

# Stage 2a: propensity-score diagnostics + balance checkpoint
ps_fit  <- fit_ps_superlearner(lock)
ps_diag <- compute_ps_diagnostics(ps_fit)
cp2     <- checkpoint_balance(ps_diag, max_smd = 0.10,
                              lock_hash = lock$lock_hash)
audit   <- record_checkpoint(audit, cp2)

# Stage 2b: plasmode candidate selection + DQ stress test
candidates <- list(
  tmle_candidate("glm_t01",   "GLM PS, trunc=0.01",
                 g_library = "SL.glm",     truncation = 0.01),
  tmle_candidate("glm_t05",   "GLM PS, trunc=0.05",
                 g_library = "SL.glm",     truncation = 0.05),
  tmle_candidate("glmnet_t01","glmnet PS, trunc=0.01",
                 g_library = "SL.glmnet",  truncation = 0.01),
  tmle_candidate("glmnet_t05","glmnet PS, trunc=0.05",
                 g_library = "SL.glmnet",  truncation = 0.05)
)
plas <- run_plasmode_feasibility(lock,
                                 tmle_candidates = candidates,
                                 effect_sizes    = c(0.03, 0.05),
                                 reps            = 200)
dq_results <- run_plasmode_dq_stress(lock,
  tmle_candidates = candidates,
  reps            = 50,
  data_quality_scenarios = list(
    covariate_missingness = list(fractions = c(0.05, 0.10, 0.20)),
    treatment_misclass    = list(sensitivity = c(0.95, 0.90),
                                 specificity = c(0.99, 0.95)),
    outcome_misclass      = list(sensitivity = c(0.95, 0.90),
                                 specificity = c(0.99, 0.95)),
    unmeasured_confounding = list(U_prevalence  = 0.20,
                                  U_treatment_OR = c(1.5, 2.0),
                                  U_outcome_OR   = c(1.5, 2.0))
  )
)
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock     <- lock_primary_tmle_spec(lock, selected)

# Outcome blinding for Stage 3 and the gate.
original_lock <- lock
masked        <- mask_outcome(lock)

# Stage 3: negative-control outcomes
masked <- define_negative_control(masked, name = "nc_outcome")
stage3 <- run_residual_confounding_stage(masked, ps_fit)
cp3    <- stage3$checkpoint
audit  <- record_checkpoint(audit, cp3)

# Pre-outcome gate
gate <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
auth <- authorize_outcome_analysis(audit)
stopifnot(isTRUE(auth$authorized))

# Stage 4: unblind and run the primary analysis
final <- unmask_outcome(masked, original_lock)
g_fit <- fit_tmle_treatment_mechanism(final, ps_fit)
Q_fit <- fit_tmle_outcome_mechanism(final, g_fit)
upd   <- run_tmle_targeting_step(g_fit, Q_fit)
psi   <- extract_tmle_estimate(upd)$estimates$ATE

save_lock(final,  "lock.rds")
save_audit(audit, "audit.rds")

# Reporting tables (function names inherited from causalRisk):
make_table1(final, treatment = "treatment", smd = TRUE)
make_table2(list(`TMLE (primary)` = upd), risk_time = 24)
```

The DQ stress test is the only fundamentally new piece in the
above. The rest is conventional TL-RWE plumbing. The point of the
example is to show that the new piece slots into a single existing
function call (`run_plasmode_dq_stress()`) that returns the same
class of metrics object as the standard plasmode and feeds the
same gate.

## Inside `run_plasmode_dq_stress()`

The function loops over a scenario grid, fits every candidate
under each scenario × severity, and aggregates bias, RMSE, and
coverage. Pseudocode:

```
for each (scenario, severity) in scenario grid:
  for each replicate r:
    set seed deterministically from (lock$seed, r, scenario index)
    generate synthetic outcome from the lock's covariate-only Q0 model
    if scenario == unmeasured_confounding:
      sample latent U; modify treatment via a real PS model;
      modify outcome probabilities via OR_Y * U
    if scenario == treatment_misclass:
      apply (sens, spec) to A
    if scenario == outcome_misclass:
      apply (sens, spec) to Y, optionally by treatment arm
    if scenario == covariate_missingness:
      set MCAR fraction of W to NA, impute with column median
    for each candidate:
      fit g, Q, targeting; record est, SE, CI
  aggregate over replicates: bias, RMSE, coverage, SE/SD
return data frame indexed by (scenario, level, candidate)
```

The implementation is in `R/plasmode_dq.R`. Three details matter
for reproducibility. First, the seed is set deterministically from
`lock$seed + rep_i + sg_i * 10000L` so that re-running the same
lock with the same scenario list produces bit-identical metrics.
Second, the unmeasured-confounding scenario fits a real treatment-
mechanism (propensity) model on the lock data once and uses it as
the baseline propensity that $U$ shifts; earlier prototype versions
incorrectly used the outcome model's prediction here, which we
fix in the version released with this paper. Third, treatment
misclassification accepts asymmetric (sens, spec) pairs rather than
symmetric flips, matching the form in which validation studies
typically report exposure-algorithm performance, and outcome
misclassification accepts optional per-treatment-arm sensitivity
and specificity, allowing the user to probe differential
misclassification, which is the most worrying case in observational
outcome ascertainment.

## Relationship to `causalRisk`

`causalRisk` [@brookhart2024causalrisk] is an MIT-licensed R
package that implements IPW, g-computation, and AIPW cumulative-
risk estimators with a cohort-specification grammar that maps
neatly onto the clean-room idea. cleanTMLE reuses that
cohort-specification grammar and the associated reporting helpers,
acknowledging Brookhart, Breskin, and Eledath as copyright holders
for the reused code in the package's DESCRIPTION file. The lock,
gate, audit log, modular TMLE functions, and plasmode DQ stress
test are unique to cleanTMLE. An analyst who has been using
`causalRisk` for IPW analyses can move into cleanTMLE for staged
TMLE analyses without relearning the cohort-specification grammar.

# Simulation Study {#sec-simulation}

## Design

We run a Monte Carlo simulation with $N = 2000$ observations per
replicate. Six estimators are compared:

1. **Crude** — unadjusted risk difference.
2. **IPTW** — stabilised inverse-probability-of-treatment weighting.
3. **PS Match** — 1:1 nearest-neighbour propensity-score matching.
4. **TMLE** — modular TMLE with the plasmode-selected
   specification, no cross-fitting.
5. **TMLE_CF (CV-TMLE)** — cross-validated TMLE with two folds. We
   use the names "cross-validated TMLE" and "cross-fitted TMLE"
   interchangeably; both describe the sample-split estimator of
   @zheng2011crossvalidated and @chernozhukov2018dml in which the
   nuisance fits are made on data held out from the targeting step.
6. **Match_TMLE** — TMLE on the propensity-matched subset.

Three scenarios are run:

- **A. Good overlap.** `overlap_strength = 0.5`, no unmeasured
  confounding. Propensity scores stay inside $[0.2, 0.8]$.
- **B. Marginal overlap.** `overlap_strength = 1.5`, no unmeasured
  confounding. Propensity scores reach $[0.04, 0.99]$.
- **C. Unmeasured confounding.** `overlap_strength = 0.5`, plus a
  binary $U$ with prevalence 0.20 and odds ratios 2.0 for both $A$
  and $Y$. The analyst's covariate set excludes $U$.

For each scenario the DQ stress test is run on three covariate-
missingness fractions (0.05, 0.10, 0.20); a $2 \times 2$ grid of
treatment-ascertainment sensitivity (0.95, 0.90) by specificity
(0.99, 0.95), giving four asymmetric (sens, spec) cells; the same
$2 \times 2$ grid for outcome ascertainment; and a $2 \times 2$
grid of unmeasured-confounder strengths
($\text{OR}_A \in \{1.5, 2.0\}$ by $\text{OR}_Y \in \{1.5, 2.0\}$,
with $U$ prevalence fixed at 0.20). Across the four threats this
yields fifteen scenario × severity cells per Monte Carlo scenario.

## Monte Carlo results

```{r load-mc-results}
res_path <- file.path("..", "results", "simulation_results.rds")
if (file.exists(res_path)) {
  res <- readRDS(res_path)
  has_mc <- TRUE
} else {
  has_mc <- FALSE
  cat("Monte Carlo results not found. Run run_simulation.R first.\n")
}
```

```{r mc-table, eval = has_mc, results = "asis"}
for (sc in names(res$summaries)) {
  truth <- res$truths[[sc]]$RD
  label <- res$scenarios[[sc]]$label
  cat(sprintf("\n### %s (true RD = %.5f)\n\n", label, truth))
  knitr::kable(res$summaries[[sc]],
               digits = c(NA, 0, 5, 5, 5, 5, 5, 3, 5, 3, 5),
               caption = label) |> print()
  cat("\n")
}
```

### Interpretation

We let the simulation results speak for themselves; the printed
tables include the Monte Carlo standard error of the coverage
estimate so the reader can judge which differences exceed Monte
Carlo noise. Three patterns are robust to noise.

First, under good overlap (Scenario A) all adjusted estimators
recover the truth with bias well under the prespecified threshold
of 0.02 and with confidence intervals that cover at the nominal
rate. The crude estimator has substantial residual confounding and
its confidence intervals do not cover at 95%. This is the textbook
benchmark check.

Second, under marginal overlap (Scenario B) the crude estimator
shifts further from the truth and IPTW shows wider CIs reflecting
the heavier weights, but TMLE and CV-TMLE remain
approximately unbiased. The plasmode-selected specification chooses
a higher truncation level than under Scenario A, illustrating that
the candidate selection adapts to the observed data structure.

Third, under unmeasured confounding (Scenario C) every estimator is
biased in the same direction, because no method that adjusts only
on $W$ can remove the contribution of $U$ to the treatment-outcome
association. The bias is smallest for TMLE and largest for the
crude estimator, but the difference is one of degree, not of kind.
The pre-outcome gate correctly returns STOP under the prespecified
thresholds for Scenario C, demonstrating that the staged workflow
declines to run the primary analysis when the available data
cannot identify the estimand.

A fourth observation worth flagging is the standard-error
calibration of the IPW estimator. The mean-SE-to-empirical-SD
ratio for IPTW is often above one in our scenarios, because the
default IF-based variance estimator treats the propensity score
as known and is therefore conservative when the propensity score
has actually been estimated [@robins1992rotnitzky;
@brookhart2024causalrisk]. Doubly robust estimators (TMLE, AIPW,
CV-TMLE) do not share this conservativeness once the nuisance
fits are properly cross-fit; the SE/SD ratio for TMLE_CF stays
near one across our scenarios. A bootstrap variance estimator
[@xu2020bootstrap] is the principled fix when reporting an
IPW-only estimate; we recommend that users wrapping
`run_iptw_workflow()` for primary inference resample the IPTW
estimator externally rather than rely on the default IF-based SE.

```{r mc-boxplots, eval = has_mc, fig.height = 9, fig.cap = "Sampling distribution of the risk-difference estimate across Monte Carlo replicates. The dashed red line marks the true RD. Under unmeasured confounding (C), all estimators are shifted from the truth."}
raw_data <- do.call(rbind, lapply(names(res$results), function(sc) {
  df <- as.data.frame(res$results[[sc]])
  df$scenario <- res$scenarios[[sc]]$label
  df$truth    <- res$truths[[sc]]$RD
  df
}))

ggplot(raw_data, aes(x = method, y = estimate, fill = method)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  geom_hline(aes(yintercept = truth), linetype = "dashed", colour = "red") +
  facet_wrap(~ scenario, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Risk-difference estimate",
       title = "Sampling distribution by estimator and scenario") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r coverage-bar, eval = has_mc, fig.cap = "95% CI coverage by estimator and scenario. The dashed red line marks the nominal 95%."}
plot_data <- do.call(rbind, lapply(names(res$summaries), function(sc) {
  df <- res$summaries[[sc]]
  df$scenario <- res$scenarios[[sc]]$label
  df
}))

ggplot(plot_data, aes(x = method, y = coverage, fill = scenario)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "red") +
  coord_cartesian(ylim = c(0.3, 1.0)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Coverage",
       title = "95% CI coverage",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r se-calibration, eval = has_mc, fig.cap = "Ratio of mean reported SE to empirical SD. Values near 1.0 indicate well-calibrated standard errors."}
ggplot(plot_data, aes(x = method, y = se_sd_ratio, colour = scenario,
                       shape = scenario)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 1.0, linetype = "dashed") +
  scale_y_continuous(limits = c(0.7, 2.2)) +
  labs(x = NULL, y = "Mean SE / Empirical SD",
       title = "Standard-error calibration",
       colour = NULL, shape = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## Plasmode DQ stress-test results

```{r load-dq-results}
dq_files <- list.files(file.path("..", "results"), pattern = "^dq_stress_",
                        full.names = TRUE)
has_dq <- length(dq_files) > 0

if (has_dq) {
  dq_all <- lapply(dq_files, readRDS)
  names(dq_all) <- sub("dq_stress_(.+)\\.rds", "\\1", basename(dq_files))
}
```

```{r dq-degradation-table, eval = has_dq, results = "asis"}
for (sc in names(dq_all)) {
  dq <- dq_all[[sc]]
  sc_label <- if (has_mc && sc %in% names(res$scenarios))
    res$scenarios[[sc]]$label else sc

  cat(sprintf("\n### %s\n\n", sc_label))

  deg <- summarize_dq_degradation(dq)
  if (nrow(deg) > 0) {
    sel_cand <- if (has_mc && sc %in% names(res$meta))
      res$meta[[sc]]$selected else deg$candidate[1]
    sel_deg <- deg[deg$candidate == sel_cand, ]
    if (nrow(sel_deg) > 0) {
      cat(sprintf("Selected candidate: `%s`\n\n", sel_cand))
      knitr::kable(
        sel_deg[, c("scenario", "level", "bias_baseline", "bias_degraded",
                     "rmse_ratio", "cov_baseline", "cov_degraded", "cov_drop")],
        digits = c(NA, NA, 5, 5, 3, 3, 3, 3),
        caption = sprintf("DQ degradation: %s (candidate: %s)", sc_label, sel_cand)
      ) |> print()
    }
  }
  cat("\n")
}
```

### Reading the degradation gradient

The four degradation mechanisms produce qualitatively different
profiles. Covariate missingness produces a flat gradient in both
the good-overlap and marginal-overlap scenarios: median imputation
on top of the locked covariate set is enough to absorb a 20% MCAR
loss. Treatment misclassification attenuates the estimate toward
the null: bias rises as (sens, spec) worsen, but coverage stays
near nominal because the estimator is targeting an attenuated
truth and the SE shrinks accordingly. Outcome misclassification
under (0.90, 0.95) produces the largest bias of the three
"measurement" threats, particularly when applied differentially by
treatment arm. Unmeasured confounding produces the largest
degradation overall and is the only threat that pushes the
selected candidate's RMSE ratio above 1.5 in our default settings.

```{r dq-gradient-plots, eval = has_dq, fig.height = 10, fig.cap = "Degradation gradient. Bias (top) and coverage (bottom) for the selected TMLE candidate at each severity level of each DQ scenario, faceted by Monte Carlo scenario."}
dq_combined <- do.call(rbind, lapply(names(dq_all), function(sc) {
  m <- dq_all[[sc]]$metrics
  m$mc_scenario <- if (has_mc && sc %in% names(res$scenarios))
    res$scenarios[[sc]]$label else sc
  m
}))

dq_plot <- dq_combined[dq_combined$scenario != "none", ]

if (nrow(dq_plot) > 0) {
  if (has_mc) {
    sel_map <- vapply(names(res$meta), function(sc)
      res$meta[[sc]]$selected, character(1))
    sc_label_map <- vapply(names(res$scenarios), function(sc)
      res$scenarios[[sc]]$label, character(1))
    dq_plot <- do.call(rbind, lapply(split(dq_plot, dq_plot$mc_scenario), function(d) {
      sc_key <- names(sc_label_map)[sc_label_map == d$mc_scenario[1]]
      if (length(sc_key) > 0 && sc_key %in% names(sel_map)) {
        d[d$candidate == sel_map[sc_key], ]
      } else d
    }))
  }

  p1 <- ggplot(dq_plot, aes(x = level, y = abs(bias),
                              colour = scenario, group = scenario)) +
    geom_point(size = 2.5) + geom_line(linewidth = 0.8) +
    facet_wrap(~ mc_scenario, scales = "free_x") +
    labs(x = "DQ severity level", y = "|Bias|",
         title = "A. Bias",
         colour = "DQ scenario") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          legend.position = "bottom")

  p2 <- ggplot(dq_plot, aes(x = level, y = coverage,
                              colour = scenario, group = scenario)) +
    geom_point(size = 2.5) + geom_line(linewidth = 0.8) +
    geom_hline(yintercept = 0.95, linetype = "dashed", colour = "red") +
    facet_wrap(~ mc_scenario, scales = "free_x") +
    labs(x = "DQ severity level", y = "Coverage",
         title = "B. Coverage",
         colour = "DQ scenario") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          legend.position = "bottom")

  gridExtra::grid.arrange(p1, p2, ncol = 1)
}
```

## Comparison with E-value-based sensitivity

The E-value [@vanderWeele2017evalue] reports the minimum strength
of unmeasured confounding required to reduce the observed estimate
to the null. It is post-hoc: it is computed from the realised
estimate after the outcome is unblinded. The plasmode DQ stress
test runs in the opposite direction: given a hypothesised confounder
strength, it computes the bias of each candidate. The two are
complementary, and we recommend reporting both.

```{r evalue-comparison, eval = has_mc && has_dq, results = "asis"}
if ("unmeasured_conf" %in% names(res$summaries)) {
  tmle_row <- res$summaries$unmeasured_conf[
    res$summaries$unmeasured_conf$method == "TMLE", ]

  if (nrow(tmle_row) > 0 && !is.na(tmle_row$mean_est)) {
    truth_c <- res$truths$unmeasured_conf$RD
    truth_a <- res$truths$good_overlap$RD
    r0 <- res$truths$good_overlap$risk_0

    tmle_a <- res$summaries$good_overlap[
      res$summaries$good_overlap$method == "TMLE", ]
    rr_hat <- abs((r0 + tmle_a$mean_est) / r0)
    ev <- compute_evalue(rr_hat, ci_bound = 1.0)

    dq_u <- dq_all$good_overlap$metrics
    dq_u_rows <- dq_u[dq_u$scenario == "unmeasured_U", ]

    cat("\n| Quantity | Value |\n|---|---|\n")
    cat(sprintf("| True RD (no U) | %.5f |\n", truth_a))
    cat(sprintf("| True RD (U present, OR = 2.0) | %.5f |\n", truth_c))
    cat(sprintf("| TMLE estimate, Scenario A (no U) | %.5f |\n", tmle_a$mean_est))
    cat(sprintf("| TMLE estimate, Scenario C (U present, not adjusted) | %.5f |\n",
                tmle_row$mean_est))
    cat(sprintf("| MC bias from unmeasured U (Scenario C) | %.5f |\n",
                tmle_row$bias))
    cat(sprintf("| E-value (from Scenario A estimate) | %.2f |\n", ev["e_value"]))

    if (nrow(dq_u_rows) > 0) {
      for (i in seq_len(nrow(dq_u_rows))) {
        cat(sprintf("| DQ stress bias at %s | %.5f |\n",
                    dq_u_rows$level[i], dq_u_rows$bias[i]))
      }
    }

    cat("\n: Unmeasured confounding under two methods: a single E-value threshold and a candidate-specific bias curve.\n\n")
  }
}
```

# Discussion {#sec-discussion}

## What the DQ stress test adds, in one sentence

The standard plasmode answers the question "which candidate has
the lowest RMSE under the locked data-generating process?" The DQ-
augmented plasmode answers the larger question "which candidate
has the lowest RMSE *and* stays inside the gate thresholds across
the plausible range of every threat documented in SPIFD2?"

## Comparison with adjacent tools

E-values [@vanderWeele2017evalue] are post-hoc and not candidate-
specific. They tell the reader how strong an unmeasured confounder
would have to be to nullify the observed effect. The plasmode DQ
stress test is pre-hoc and candidate-specific: it tells the
reviewer how biased each candidate becomes under each hypothesised
strength of confounding, on the same scale as the rest of the
gate. The two are answering different questions and should both be
reported.

Quantitative bias analysis [@lash2014qba] adjusts the final
estimate for postulated biases. Like the E-value, QBA is an
unblinding-stage tool. The DQ stress test does not adjust the
estimate; it asks whether the estimator passes the gate when the
biases are present. We therefore recommend QBA at the analysis
stage and the DQ stress test at the prespecification stage.

The targeted-learning causal-gap sensitivity analysis
[@diaz2013sensitivity] characterises the change in $\Psi$ as the
unmeasured-confounder bound varies. Causal-gap is unblinding-stage
and characterises one threat (unmeasured confounding) in a TL-
specific way. The DQ stress test is broader (it covers four
threats) but cruder (it does not adjust the estimate). Both are
useful and complementary.

Negative-control outcomes [@lipsitch2010negative] test for
residual confounding in the realised data. Because they use real
data, they are at-outcome rather than pre-outcome. The DQ stress
test cannot detect residual confounding; the NCO check cannot
detect covariate missingness or exposure misclassification. They
target different threats. @muntner2024staging recommend selecting
NCOs across four domains — confounding by indication, functional
status, health-seeking behaviour, and access to healthcare — and
@levintow2023nco give pragmatic guidance for choosing them; the
cleanTMLE workflow records each declared NCO in the lock so the
chosen domains are part of the audit trail.

@kent2026fiord reports a worked applied use of the staging
framework for a comparative-effectiveness study of evolocumab
versus non-treatment for major adverse cardiovascular events in
the Komodo Healthcare Map: NCOs flagged residual bias in the
health-seeking-behaviour domain, the review team allowed the
study to proceed under FLAG with a pre-specified E-value
quantitative-bias-analysis plan, and the final risk difference at
four years was reported as roughly $-4\%$. The example
illustrates both the GO/FLAG/STOP gate and the role of NCO
domains in practice.

## Limitations

The current implementation has six limitations the analyst should
keep in mind.

First, covariate missingness is missing-completely-at-random
(MCAR) only. Missing-at-random (MAR) and missing-not-at-random
(MNAR) mechanisms require a missingness model and are not yet
supported. A more principled alternative is inverse-probability-
of-missingness weighting (`identify_missing()` in `causalRisk`,
[@brookhart2024causalrisk]), which trades MCAR for the weaker
stratified-MCAR assumption; we plan to expose the same option in
a future cleanTMLE release.

Second, the plasmode data-generating process is parametric: the
outcome is generated from a logistic regression of $Y$ on $W$,
fitted to the lock data, and the synthetic effect is constant on
the risk-difference scale ($p_1^{\text{sim}} = p_0^{\text{sim}} +
\Delta$). Effect modification, time-varying treatment, and
competing risks would require a richer DGP. The package architecture
permits these extensions; the simulation in this paper does not
exercise them.

Third, the plausible degradation ranges have to come from
external evidence. The software cannot decide whether 5% or 20%
covariate missingness is realistic for a given data source; the
analyst has to cite a profiling or validation study. This is
consistent with the SPIFD2 stepwise template, which makes the
same demand qualitatively, but it is worth saying explicitly.

Fourth, the DQ stress test is computationally expensive. Each
scenario × severity adds a full set of plasmode replicates. With
the configuration of §\ref{sec-simulation} — 50 replicates per
cell, fifteen scenario × severity cells (3 missingness fractions,
$2 \times 2 = 4$ treatment-misclassification cells, 4 outcome-
misclassification cells, and 4 unmeasured-confounding cells), and
four candidates — the test fits 3,000 TMLE models per Monte Carlo
scenario. We recommend running fewer replicates per severity than
the baseline plasmode (50 versus 200, by default) and parallelising
the outer loop over scenarios.

Fifth, the running estimand in this paper is the marginal risk
difference for a binary point exposure. Counterfactual hazard
ratios — available in `causalRisk` via `estimate_ipwhr()` and in
`survtmle` for survival TMLE [@survtmlepackage2018] — are not
yet wrapped into the cleanTMLE staged workflow. Sites that need
hazard-ratio estimation can call those packages directly and feed
the output through the same lock and audit trail.

Sixth, `causalRisk` supports several estimand families that
cleanTMLE does not: counterfactual cumulative-incidence curves
with competing risks, cumulative-count IPW
(`estimate_ipwcount()`), multinomial treatments, and several
weight types (SMR, overlap, normalised, stabilised). The
cleanTMLE staged workflow narrows scope to binary-treatment
risk-difference TMLE in this release; analysts who need the
broader estimand surface can use `causalRisk` directly with the
same cohort-specification grammar.

## Recommendations for practice

We recommend a four-step pre-outcome workflow when the analyst is
preparing a TL-RWE study.

1. Document the SPIFD2 template, citing validation studies or data
   profiling for each plausible range.
2. Lock the candidate grid and the gate thresholds before any
   simulation runs.
3. Run `run_plasmode_feasibility()` and `run_plasmode_dq_stress()`
   in the same prespecified Stage 2b. Use the SPIFD2-derived ranges
   as the inputs to `data_quality_scenarios`.
4. Score the gate. If a candidate is robust to the SPIFD2
   plausible ranges, lock it. If no candidate is robust, the
   correct decision is STOP, and the analyst should reconsider
   either the data source, the estimand, or the gate thresholds
   before unblinding.

The audit log preserves all of the above. The final paper or
regulatory submission can be reconstructed from the lock and the
audit, including the SPIFD2 ranges, the plasmode and DQ stress
metrics, the gate decision, and the post-outcome estimate.

# Summary {#sec-summary}

cleanTMLE is an R package for staged, outcome-blind targeted
learning with reproducible prespecification artifacts. It extends
the outcome-blind plasmode loop of @dang2023roadmap and
@gruber2022sap with four data-quality degradation mechanisms,
producing per-candidate degradation gradients that the
GO/FLAG/STOP gate can consume on the same scale as the baseline
plasmode metrics. The plasmode DQ stress test is the package's
distinguishing contribution: it gives the analyst a quantitative
bridge from the qualitative SPIFD2 data-fitness template to the
prespecified gate, in code, with a single function call. The lock,
audit log, and decision log persist the prespecification as
machine-readable artifacts that downstream estimation functions
verify before they are allowed to run. The cohort-specification
and reporting functions follow `causalRisk`
[@brookhart2024causalrisk] under MIT-licence attribution, so an
analyst can move between the two packages without relearning the
function names or argument order. The package is
available at <https://github.com/amertens/cleanTMLE>; the
simulation code that produced the results in §\ref{sec-simulation}
is in the companion repository <https://github.com/amertens/clean-room-sim>.

# Acknowledgements {.unnumbered}

We thank the authors of `causalRisk` (M. Alan Brookhart, Alexander
Breskin, and Bassim Eledath) for the MIT-licensed code and
function naming that cleanTMLE inherits, and the targeted-learning
RWE roadmap authors (Dang, Gruber, Phillips, Lee, Ho, van der Laan, Petersen,
and colleagues) whose published frameworks cleanTMLE
operationalises in software. Any errors are ours.

# References {.unnumbered}

::: {#refs}
:::

```{r session-info}
sessionInfo()
```
