---
title: "cleanTMLE: An R Package for Staged, Outcome-Blind Targeted Learning with Plasmode Data-Quality Stress Testing"
author:
  - name: "Andrew Mertens"
    affiliation: "University of California, Berkeley"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
    fig-width: 8
    fig-height: 5
    number-sections: true
bibliography: references.bib
csl-hanging-indent: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "R> ",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(ggplot2)
library(cleanTMLE)
```

# Abstract {.unnumbered}

Real-world evidence studies must justify both the causal analysis
strategy and the fitness of the underlying data source for the
research question. In current practice these two arguments are
usually developed in parallel but reported separately:
estimator specification is supported with outcome-blind simulation
[@franklin2014plasmode; @schuler2017tmle; @dang2023roadmap;
@gruber2022sap], whereas data fitness is documented qualitatively
through target-trial articulation [@hernan2016targettrial;
@hernan2016target], reporting templates such as TARGET
[@cashin2024target], and fit-for-purpose checklists such as SPIFD2
[@gatto2022spifd; @gatto2023spifd2].

We present **cleanTMLE**, an R [@rcoreteam2024] package that
provides a prespecified quantitative stress test of estimator
performance under data-quality degradation scenarios derived from
a prior fit-for-purpose assessment, situated inside the causal
roadmap [@dang2023roadmap; @petersen2014roadmap]. The package does
*not* itself determine whether a data source is fit for purpose;
that judgement remains a study-level activity supported by
target-trial protocols, validation studies, and source-data
review. Once investigators have specified plausible data-quality
threats and severity ranges from external evidence, cleanTMLE
records the prespecified analytic choices and audit trail and
extends the standard outcome-blind plasmode estimator-selection
loop with a pre-outcome stress test that evaluates candidate
estimators under those user-specified threats (covariate
missingness, exposure misclassification, outcome misclassification,
and unmeasured confounding). We illustrate the workflow in a Monte
Carlo simulation and in a trauma-registry case study (n = 1,693
ambulance patients, 24.6 % outcome missingness) where six
estimators implemented in cleanTMLE following the staged-analysis
vignette agree on direction (risk difference 0.026 to 0.037 in
favour of Rescue.Co transport, with the prespecified primary
full-cohort TMLE interval just crossing zero). The case study is
presented as a workflow demonstration on a real registry rather
than as evidence that the framework establishes causal validity.

The contribution is operational rather than foundational. The
package provides reproducible analysis records and
candidate-specific robustness screening before the study outcome
is unblinded. It does not replace target-trial design,
variable-validation studies, source-data relevance and reliability
review, independent oversight, or post-unblinding sensitivity
analyses; the explicit list and corresponding regulatory anchors
appear in the regulatory-alignment table (§\ref{sec-roadmap}) and
in the discussion. cleanTMLE reuses the cohort-specification
grammar and reporting helpers of the **causalRisk** package
[@brookhart2024causalrisk] under MIT licence and adds the
prespecification, audit, and DQ stress-test functions that
`causalRisk` does not provide.

**Keywords:** targeted maximum likelihood estimation, plasmode
simulation, real-world evidence, fit-for-purpose data, outcome-
blind analysis, R, statistical analysis plan, regulatory science.

# Introduction

## Prespecification and fit-for-purpose data in RWE

Outcome blinding and prespecification are governance tools, not
substitutes for causal design. In a real-world evidence (RWE) study
the analyst still has to define the target trial
[@hernan2016targettrial; @hernan2016target], specify a statistical
estimand consistent with the ICH E9(R1) framework [@ich2019e9r1],
justify identification from the available data, and defend the
relevance and reliability of the data source for the regulatory or
scientific question.

Regulatory bodies have converged on this list. The FDA RWE
framework and its source-specific guidances cover
real-world data and evidence [@fda2018rwe; @fda2024considerations],
EHR and claims data [@fda2024ehrclaims], registries
[@fda2021registries], and non-interventional studies
[@fda2024noninterventional]. The EMA materials add registry-based
studies [@ema2021registry], a regulator-led RWE guide
[@ema2024guide], and an RWD data-quality framework
[@ema2026rwd]. ICH M14 (March 2026) sets harmonised principles
for non-interventional studies that use real-world data for safety
assessment of medicines [@ich2024m14].

Within this broader workflow, two arguments are typically
developed in parallel but reported separately. The
*estimator-selection* argument is checked with outcome-blind
plasmode simulation [@franklin2014plasmode; @schuler2017tmle;
@dang2023roadmap; @gruber2022sap; @schreck2024plasmode]. The
*fit-for-purpose* argument is documented qualitatively through
templates such as SPIFD2 [@gatto2022spifd; @gatto2023spifd2] or
HARPER [@wang2023harper] and through reporting tools such as
TARGET [@cashin2024target] and START-RWE [@wang2021startrwe]. The
simulation is numeric; the fit-for-purpose argument is qualitative.

These two arguments are connected: a candidate estimator that
performs well on a simulation derived from real covariate and
treatment distributions may still fail under realistic measurement
error, missingness, or unmeasured confounding. The plasmode loop is
under the analyst's control, so it can be extended to ask not only
"which candidate has the lowest RMSE under the locked
data-generating process?" but also "which candidate maintains
acceptable bias, RMSE, and coverage across the plausible range of
each threat documented in the fit-for-purpose review?" We refer to
this extension as a *data-quality (DQ) stress test*. It is a
pre-outcome robustness screen rather than a substitute for
variable validation or for post-unblinding sensitivity analysis.

Throughout, we use "data-quality stress testing" in a restricted
sense. The procedure does not assess data provenance, phenotype
validity, linkage accuracy, representativeness, or regulatory
acceptability directly; those remain study-level and
data-source-level evaluations. The role of cleanTMLE is narrower:
once investigators have specified plausible data-quality threats
and severity ranges from external validation studies, data
profiling, or expert elicitation, the package evaluates how
candidate estimators behave under those prespecified perturbations
before the observed treatment-outcome association is examined.

In this manuscript, *outcome-blind* means that decisions about
cohort construction, candidate estimators, simulation scenarios,
decision thresholds, and primary-estimator selection are made
without inspecting the observed treatment-outcome association.
Analysts may use baseline covariates, treatment assignment,
missingness indicators, and synthetic outcomes generated for
plasmode simulations; the marginal mean of the observed outcome on
covariates is also used internally to fit the plasmode generator.
The observed treatment-outcome relationship is not used for
estimator selection or design revision before the pre-outcome
decision point.

## What the package adds, and what it does not

This paper introduces **cleanTMLE**, an R [@rcoreteam2024] package
that records prespecified analytic decisions and supports the
estimator-selection and stress-testing steps of the causal roadmap
[@dang2023roadmap; @petersen2014roadmap; @gruber2022sap]. cleanTMLE
provides an analysis-lock object, a serialisable audit trail and
decision log [@muntner2024staging], a pre-outcome decision rule
that classifies the analysis as proceed, review, or stop, an
outcome-blind plasmode candidate selector, and the DQ stress test
described above. Targeted Maximum Likelihood Estimation (TMLE)
[@vanderLaan2006tmle; @vdL2011tlbook; @schuler2017tmle;
@zheng2011crossvalidated; @chernozhukov2018dml] is the natural
setting because every analytic decision (the SuperLearner library
[@vdL2007superlearner; @phillips2023sl], the truncation rule, the
prespecified set of candidate estimators) can be listed in advance,
and the algorithm remains data-adaptive while still admitting valid
Wald-type confidence intervals.

The contribution is operational. The software-mediated outcome
guard is, as @muntner2024staging put it, an implementation pattern
rather than a methodological novelty. Its advantage is
transparency: the analysis lock, audit trail, and decision log
turn paper-only prespecification checklists into a structured
record of analytic decisions, supporting reproducible
reconstruction by an external reviewer. cleanTMLE
provides a prespecified quantitative stress test of estimator
performance under data-quality degradation scenarios derived from
a prior fit-for-purpose assessment; it does not itself determine
whether the data source is fit for purpose. Section
§\ref{sec-roadmap} situates that contribution inside the causal
roadmap and Table \ref{tbl-reg-align} maps each component to the
corresponding FDA, EMA, and ICH expectation; we defer the explicit
list of activities the package does *not* replace to those
locations and to the discussion.

Throughout, we use "clean room" interchangeably with "outcome-blind"
when discussing software, because the workflow prevents
outcome-analysis functions from proceeding until the prespecified
review process has been completed, much as a manufacturing clean
room blocks access to a sterile area until protocols have been
followed. The software-mediated workflow partially substitutes for
personnel-based outcome separation by restricting access to the
observed outcome until prespecified review steps have been
completed; it complements, but does not replace, personnel-based
access controls and independent governance. Protocol review, role
separation, source-data audit, and independent checkpoint review
remain the broader institutional process inside which the software
arrangement sits.

::: {.callout-note}
## Current scope of cleanTMLE

The simulation, vignette, and case study in this paper run on a
**binary point exposure**, a **binary outcome**, the **marginal
risk-difference** estimand, and **MCAR** outcome missingness handled
by IPCW. Hazard ratios, time-varying treatments, competing risks,
multinomial exposures, and MAR / MNAR missingness models are out
of scope for the v0.1 release; analysts who need those estimands
should compose cleanTMLE's prespecification and audit functions with `survtmle`
[@survtmlepackage2018], `lmtp` [@lmtppackage2021], or `causalRisk`
[@brookhart2024causalrisk] until those features are added.
:::

## What the paper contributes

The contributions are operational. First, we extend the
outcome-blind plasmode loop from candidate-estimator selection
[@dang2023roadmap; @gruber2022sap; @schreck2024plasmode] to
candidate-estimator robustness *screening* under prespecified
threat ranges. Each candidate estimator produces a degradation
gradient on the same bias / RMSE / coverage scale as the baseline
plasmode, so the prespecified decision rule can be applied
uniformly across baseline and stressed scenarios. We position this screen as a quantitative *supplement*
to fit-for-purpose review (§\ref{sec-bridge}) rather than as a
substitute for variable-validation studies, source-data
assessment, or post-outcome quantitative bias analysis. Second, we
provide an R implementation that turns prespecification into
reproducible analysis records: a fingerprinted analysis-lock
object, an audit log, a decision log, and a pre-outcome decision
point that the prespecified estimator is checked against before
outcome analysis. Third, we reuse the cohort-specification grammar
and reporting helpers of `causalRisk` [@brookhart2024causalrisk]
under MIT licence and add the prespecification, audit, and DQ
stress-test functions that `causalRisk` does not provide. The next
section (§\ref{sec-roadmap}) places these contributions inside the
causal roadmap; subsequent sections follow that roadmap order.

# The causal roadmap and the role of cleanTMLE {#sec-roadmap}

We position cleanTMLE inside the causal roadmap of @petersen2014roadmap
and @dang2023roadmap, which separates a real-world evidence study
into six logically ordered activities: (1) defining the causal
question and target trial; (2) characterising the observed data and
their fitness for that target trial; (3) stating identification
assumptions; (4) defining a statistical estimand; (5) selecting and
locking estimators; and (6) interpreting results under at-outcome
and post-outcome sensitivity analyses. The package operates inside
steps (5) and the pre-outcome portion of (6); steps (1)--(4)
remain external study-level work, supported by target-trial
templates [@hernan2016targettrial; @cashin2024target], fit-for-
purpose checklists [@gatto2023spifd2; @wang2023harper], and
source-specific data-relevance and reliability assessments
[@fda2024considerations; @fda2024ehrclaims; @fda2021registries;
@ema2024guide; @ema2026rwd].

To support readers who are new to the causal-roadmap vocabulary,
Table \ref{tbl-layers} distinguishes the conceptual layers that
the workflow keeps separate. The same table also serves as a
reference for which decisions are addressed at the design stage
versus at the estimation stage.

```{r layers-table}
#| label: tbl-layers
layers <- data.frame(
  Layer = c(
    "Causal question",
    "Causal estimand",
    "Identification assumptions",
    "Statistical estimand",
    "Estimator",
    "Pre-outcome diagnostic metric",
    "Governance decision"),
  Description = c(
    "Clinical or policy question framed as a target trial or causal contrast.",
    "Counterfactual treatment effect of interest (e.g. marginal risk difference).",
    "Conditions under which the causal estimand can be expressed as an observed-data parameter (exchangeability, positivity, consistency, measurement validity, missingness assumptions).",
    "Observed-data parameter identified under the assumptions above.",
    "Algorithm used to estimate the statistical estimand, for example TMLE, IPTW, or matching.",
    "Pre-outcome quantity used for design checks: simulated bias, RMSE, coverage, effective sample size, balance, negative-control performance.",
    "Whether prespecified criteria support proceeding, require additional review, or require stopping the analysis before the outcome is examined."),
  stringsAsFactors = FALSE,
  check.names = FALSE)
knitr::kable(layers,
             caption = "Conceptual layers in the staged workflow. The same vocabulary is used throughout the manuscript to keep estimand, estimator, simulation, and decision rule distinct.")
```

```{mermaid}
%%| fig-cap: "cleanTMLE inside the causal roadmap. Boxes shaded grey indicate work that the package can operationalise; the remaining boxes are external study-level activities."
flowchart TD
    A[1. Causal question and target trial] --> B[2. Observed data and source characterisation]
    B --> C[3. Identification assumptions]
    C --> D[4. Statistical estimand]
    D --> E["5. Lock candidate estimators and gate thresholds<br/><i>(cleanTMLE: create_analysis_lock, lock_primary_tmle_spec)</i>"]
    E --> F["Pre-outcome checks<br/><i>(cleanTMLE: checkpoints, plasmode, DQ stress)</i>"]
    F --> G["Stage gate<br/><i>(cleanTMLE: gate_check, authorize_outcome_analysis)</i>"]
    G -->|GO or documented FLAG| H["Unblind and run locked primary analysis<br/><i>(cleanTMLE: TMLE step-by-step, IPCW-TMLE)</i>"]
    H --> I[At-outcome diagnostics: negative controls, balance]
    I --> J[Post-outcome sensitivity: E-value, QBA, causal-gap]
    J --> K[Interpretation and transportability]

    style E fill:#e6e6e6,stroke:#333,color:#000
    style F fill:#e6e6e6,stroke:#333,color:#000
    style G fill:#e6e6e6,stroke:#333,color:#000
    style H fill:#e6e6e6,stroke:#333,color:#000
```

A complementary view emphasises that the software-level
arrangement is one layer in a broader governance stack. External
validation studies, independent oversight, role-based data access,
and data standards remain necessary inputs to the pre-outcome
decision, regardless of how the prespecified records and audit
trail are implemented:

```{mermaid}
%%| fig-cap: "Governance layers. The software-level arrangement is necessary but not sufficient for regulatory-grade RWE."
flowchart LR
    A[Protocol and estimand] --> B[Analysis-lock object]
    B --> C[Audit trail]
    C --> D[Decision log]
    D --> E[Pre-outcome decision point]
    E --> F[Primary analysis]
    X[External validation studies] --> E
    Y[Independent review and role governance] --> E
    Z[Data standards and provenance] --> E
```

## Regulatory alignment

Table \ref{tbl-reg-align} maps each roadmap component to the
corresponding manuscript section, the relevant FDA / EMA / ICH
expectation, what cleanTMLE can operationalise in software, and
what still requires external study-level justification. The table
is intentionally explicit about the *what cleanTMLE does not
replace* column; that is the row that distinguishes a software
clean room from a regulatory clean room.

```{r reg-align-table}
#| label: tbl-reg-align
reg_align <- data.frame(
  `Roadmap component` = c(
    "Causal question / target trial",
    "Observed data and source characterisation",
    "Identification assumptions",
    "Statistical estimand",
    "Estimator selection and locking",
    "Pre-outcome data-quality stress",
    "Stage gate (GO / FLAG / STOP)",
    "Outcome analysis",
    "At-outcome diagnostics",
    "Post-outcome sensitivity analyses"
  ),
  `Manuscript section` = c(
    "§\\ref{sec-estimand}",
    "§\\ref{sec-estimand}, §\\ref{sec-casestudy}",
    "§\\ref{sec-id}",
    "§\\ref{sec-estimand}",
    "§\\ref{sec-framework}, §\\ref{sec-implementation}",
    "§\\ref{sec-dq}",
    "§\\ref{sec-bridge}",
    "§\\ref{sec-implementation}",
    "§\\ref{sec-discussion}",
    "§\\ref{sec-discussion}"
  ),
  `FDA / EMA / ICH anchor` = c(
    "Hernán target-trial framework; FDA RWE Framework; ICH M14 (March 2026)",
    "FDA EHR/claims, registries; EMA registry guideline; EMA RWD quality framework",
    "ICH E9(R1)",
    "ICH E9(R1); FDA Considerations for RWD/RWE",
    "FDA non-interventional studies; ICH M14 (March 2026)",
    "Source-specific reliability (FDA/EMA); validation evidence",
    "Internal protocol governance; FDA traceability",
    "FDA Considerations for RWD/RWE",
    "EMA RWE Guide; FDA non-interventional studies",
    "FDA non-interventional studies; ICH E9(R1) sensitivity"
  ),
  `What cleanTMLE operationalises` = c(
    "(none)",
    "Lock fingerprint of column names + data shape",
    "Records assumptions in attach_estimand and audit log",
    "attach_estimand, lock fingerprint",
    "tmle_candidate, lock_primary_tmle_spec, save_lock",
    "run_plasmode_dq_stress, default_dq_scenarios",
    "gate_check, gate_all, authorize_outcome_analysis",
    "fit_tmle_treatment/outcome_mechanism, run_ipcw_tmle",
    "run_negative_control(_tmle), checkpoint_residual_bias",
    "compute_evalue, tipping_point_sensitivity, sensitivity_truncation"
  ),
  `External study-level work still required` = c(
    "Full target-trial protocol",
    "Phenotype validation, data lineage, source audit",
    "Subject-matter justification of exchangeability and positivity",
    "Source-specific operationalisation of intercurrent events",
    "Justification of SuperLearner library and severity grids",
    "External validation evidence anchoring DQ severity ranges",
    "Independent review of gate decision",
    "Role separation, oversight, deviation memos",
    "Choice of negative-control outcomes and domains",
    "Quantitative bias analysis tailored to study threats"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(reg_align,
             caption = "Mapping of causal-roadmap components to manuscript sections, regulatory anchors, what cleanTMLE operationalises, and what still requires external study-level justification.")
```

The remaining methods sections follow the roadmap order. §\ref{sec-estimand}
states the target-trial estimand and identification assumptions
used in the simulation. §\ref{sec-framework} describes the lock,
audit, and gate functions. §\ref{sec-dq} introduces the DQ stress
test, including a threat-to-assumption matrix that is explicit
about what each scenario probes and what it does not probe, and
discusses calibration of severity ranges from external evidence.
§\ref{sec-bridge} reframes the DQ stress test as a quantitative
*supplement* to fit-for-purpose review rather than a replacement
for it. §\ref{sec-implementation}--§\ref{sec-casestudy} report
the implementation, the simulation, and the case study;
§\ref{sec-discussion} discusses limitations and the role of
post-outcome sensitivity analyses.

# Estimand, Causal Model, and Identifiability {#sec-estimand}

## Estimand specification

We follow @ich2019e9r1 and @dang2023roadmap by writing the estimand
along five attributes. For the running example the population is
adults eligible at the index visit. The treatment is a binary point
exposure $A \in \{0, 1\}$ at index. The endpoint is a binary
indicator of the primary event by 24 months, denoted $Y$. The
summary measure is the marginal counterfactual risk difference
$\Psi = E\{Y(1)\} - E\{Y(0)\}$. Intercurrent events are handled
under a treatment-policy strategy; censoring and treatment
switching are not modelled in this illustration. The corresponding
statistical estimand under the identification assumptions in
§\ref{sec-id} is $\Psi = E_W\{E[Y \mid A=1, W] - E[Y \mid A=0, W]\}$,
the marginal g-formula contrast [@robins1986newapproach;
@hernan2020causal].

In the package this specification is attached to the lock with
`attach_estimand()` at Stage 1a and is preserved on the lock object
that downstream stages consume. The estimand fields are not part of
the integrity fingerprint computed by `create_analysis_lock()`;
analysts who want a tamper-evident estimand record should pair
`attach_estimand()` with a `record_decision_log_entry()` call so
that any later modification is visible in the exported decision
log.

## Causal model

The data-generating process has five baseline covariates
$W = (\text{age}, \text{sex}, \text{biomarker}, \text{comorbidity}, \text{CKD})$,
a binary treatment $A$ that depends on $W$, and a binary outcome
$Y$ at 24 months. In the unmeasured-confounding scenario a binary
$U \sim \text{Bern}(p_U)$ also affects both $A$ and $Y$. The DAG
is

```
   W ───► A ──► Y
   │          ▲
   └──────────┘
   (Scenario C only:)
   U ──► A
   U ──► Y
```

with no arrow from $A$ to $W$ (point exposure) and no arrow
between $W$ and $U$.

## Identifiability assumptions {#sec-id}

Under this DAG, $\Psi$ is identified by the g-formula if the
following six assumptions hold [@hernan2020causal]. We list each
assumption together with the DQ stress-test scenario that probes
it; this mapping is one of the main reasons the DQ stress test is
useful.

- *Conditional exchangeability*, $Y(a) \perp\!\!\!\perp A \mid W$.
  This holds in scenarios A and B by construction. It is violated
  by construction in scenario C, where $U$ is not in the analyst's
  adjustment set. **Probed by:** the unmeasured-confounding scenario
  in `run_plasmode_dq_stress()`.
- *Positivity*, $0 < \Pr(A = 1 \mid W) < 1$ for every $W$ in the
  support. Scenario A satisfies this comfortably. Scenario B
  pushes the propensity toward the boundary but does not violate
  it in the limit. **Probed by:** the truncation level of each TMLE
  candidate; near-positivity violations show up as inflated SE/SD
  in the baseline plasmode metrics.
- *Consistency / SUTVA*, $Y = Y(A)$. **Probed by:** the treatment-
  misclassification scenario, which simulates a violation of
  consistency caused by exposure-algorithm error.
- *No measurement error in the outcome*, $\tilde{Y} = Y$. **Probed
  by:** the outcome-misclassification scenario, with optional
  per-arm differential misclassification.
- *Stratified MCAR for the covariates*, $W$ is missing completely
  at random within strata of fully measured variables. **Probed
  by:** the covariate-missingness scenario, which sets a fraction
  of values to NA and imputes with the column median.
- *Correctly specified nuisance models* for $g(W)$ and $E[Y \mid A,
  W]$, including the SuperLearner library and the truncation rule.
  **Probed by:** the standard plasmode candidate-selection metrics,
  which compare candidates that vary the SuperLearner library and
  the truncation rule.

The simulation thus contains one analytic challenge (marginal
overlap, B) and one identifiability violation (unmeasured
confounding, C). We use C to demonstrate what the DQ stress test
detects when an unmeasured confounder is anticipated, and to
contrast the stress-test output with the post-hoc E-value
[@vanderWeele2017evalue] and with the negative-control-outcome
check [@lipsitch2010negative].

# The cleanTMLE Framework {#sec-framework}

## Staged workflow

cleanTMLE adopts the staged design described informally in
@muntner2024staging and formalised in @dang2023roadmap and
@gruber2022sap. Stage names below are package-internal labels;
the stage *content* matches Steps 0–6 of @dang2023roadmap.

| Stage | Purpose | Does the function read the outcome? |
|-------|---------|:------------:|
| 0 | Feasibility assessment, utilisation study, draft protocol/SAP [@muntner2024staging; @kent2026fiord] | No (no data access yet) |
| 1a | Lock estimand, covariates, SuperLearner library, candidate grid | No |
| 1b | Cohort adequacy: $N$ per arm, events, design precision | Events only |
| 2a | Propensity-score diagnostics: overlap, ESS, balance | No |
| 2b | Outcome-blind plasmode candidate selection | No (synthetic outcome) |
| **2b\*** | **Plasmode DQ stress test (proposed)** | **No (synthetic outcome)** |
| 3 | Negative-control outcomes, residual-bias check | NC outcomes only |
| Gate | Pre-outcome authorisation: GO/FLAG/STOP | No |
| 4 | Primary comparative analysis | **Yes** |

Stages 1–3 do not access the real outcome. The plasmode at Stage
2b generates synthetic outcomes from a parametric model fitted to
the real covariates and treatment, which is why it is a natural
host for the DQ stress test we describe in §\ref{sec-dq}.

## How the workflow records and protects analytic decisions

The analysis lock is the central object on which all later
diagnostics depend. It is created at the start of the workflow and
records the prespecified analytic specification: the analytic
dataset, the treatment, outcome, and covariate column names, the
nuisance-learning library, and the random seed. A fingerprint of
those fields is recorded so that the same lock can be reconciled
later by an external reviewer; the estimand and the prespecified
choice of estimator are attached after the lock is created and are
captured by the audit trail rather than by the fingerprint, so any
change to those fields appears as a separate entry in the audit.
While the analysis is outcome-blind, the lock is held in a state
in which the outcome column is masked; the workflow restores the
outcome only after the prespecified diagnostics and checkpoint
review have been completed. Outcome-analysis functions verify that
this review has been completed before allowing access to the
observed outcome.

Two complementary records track analytic decisions through the
workflow. The audit trail accumulates entries automatically from
the checkpoint diagnostics, recording what was evaluated, when,
and against which prespecified criteria. The decision log captures
analyst choices and any deviations from the protocol, with a brief
rationale for each. Both can be exported as tidy data frames and
archived alongside the lock; the audit trail is linked to the
prespecified analysis object, so deviations from the recorded
workflow can be detected during review. The exported decision log
follows the structure of the Romiplostim worked example
[@mcgrath2021romiplostim], which @muntner2024staging take as the
reference template for a staged clean-room decision record.

Beyond a single outcome-masking step, @muntner2024staging and
@kent2026fiord recommend tiered, role-based access in which
analytic advisors see summary outputs by stage and checkpoint
reviewers see summary outputs only by checkpoint. The exported
audit and decision logs support that arrangement at the workflow
level, providing reviewer-facing summaries that do not require
raw-data access.

## Personnel-based and simulation-based outcome blinding

Outcome blinding can be enforced in two complementary ways. The
first is personnel-based [@muntner2024staging]. A blinded analyst
writes the analysis plan, defines the cohort, performs the design
diagnostics, and reaches the pre-outcome decision point; only then
does a separate analyst, or another role on the same team, link
the outcome and run the comparative analysis. Separation is
maintained by an institutional partition of duties and a
documented audit by a data manager who controls access to the
outcome variable. Reviewers and regulators evaluate two artefacts
at the end: a timestamped analysis plan and a final estimate
produced under that plan.

The second arrangement, used here, is simulation-based. A single
analyst can iterate on the prespecified set of candidate
estimators, the decision thresholds, and the data-quality stress
scenarios because the iteration is conducted against synthetic
outcomes generated from a covariate-only model rather than against
the observed treatment–outcome association. The software-mediated
workflow partially substitutes for personnel-based separation by
restricting outcome access until prespecified diagnostics and
decision criteria have been completed; it complements, but does
not replace, an institutional partition of duties or independent
review of the analysis plan. The audit and decision logs likewise
complement, rather than replace, the timestamped paper plan and
the data manager's role. Reviewers retain access to a transparent
reconstruction of every analytic decision in addition to the
existing institutional documentation.

The two arrangements address the same underlying concern, that
analytic decisions made after seeing the treatment–outcome
association are not credible, but they differ in their residual
risks. Personnel-based blinding requires an institutional
commitment to the partition of duties and tolerates ad-hoc
deviation by the blinded analyst, since the only ground truth is
the timestamped plan. Simulation-based blinding reduces the
personnel constraint but introduces a different residual risk: the
investigator must defend the assumption that the synthetic-outcome
mechanism is faithful enough to the real data-generating process
that estimator selection on simulations generalises to the
realised data. The arrangements are complementary, and a
defensible RWE pipeline can use both, applying the simulation-based
workflow inside an institutional partition of duties.

### Methodological tradeoffs

Compared with personnel-mediated outcome blinding, the
simulation-mediated approach has the following tradeoffs.

*In favour:* a single statistician can produce a documented
prespecification without a separate unblinded role, with the
caveat that submission-grade studies should retain a personnel
partition; the prespecified set of candidate estimators, the
truncation thresholds, and the decision thresholds can be revised
after viewing simulation results because those results use
synthetic outcomes, so the analytic plan need not be re-frozen
each time a tuning choice is reconsidered; the design records
support transparent reconstruction of every prespecified decision
by an external reviewer; and the same simulation loop that
selects the candidate can be used to evaluate its robustness to
prespecified data-quality threats, an evaluation that has no
direct counterpart in personnel-based blinding (data fitness is
argued in prose).

*Against:* the plasmode generator uses the realised outcome
$Y$ to fit a covariate-only mean model, so the analyst learns the
marginal distribution of $Y$ given $W$, although not the joint
$(A, Y) \mid W$ relationship. The Dang causal-roadmap reading of
*outcome-blind* permits this; a stricter clean-room reading would
require $E[Y \mid W]$ to come from external pilot data
[@dang2023roadmap, p. 7], and the package supports both modes,
recording which mode was used. The synthetic outcome is generated
from a parametric model with a fixed risk difference, so effect
modification is not represented; a richer data-generating process
is possible but is not exercised here. Finally, when the analyst
also writes the plasmode and stress-test specifications, there is
no independent check that the synthetic-outcome mechanism is
faithful, whereas the personnel-based arrangement provides an
additional layer of cross-validation through the data manager.

We document these trade-offs explicitly in the package's
documentation so that the analyst can choose the level of
formality the study requires. For most pharmacoepidemiology RWE
studies the software-only mode is adequate; for high-stakes
regulatory submissions we recommend running cleanTMLE inside a
traditional clean-room arrangement so that both the institutional
and the computational guarantees apply.

# Data Quality as a Quantifiable Threat {#sec-dq}

## Mapping data-quality concerns to plasmode parameters

SPIFD2 [@gatto2023spifd2] does not list a fixed taxonomy of
"data-quality dimensions"; it is a stepwise template that asks the
analyst to record each design element of the hypothetical target
trial and the data feature that operationalises it. The harmonised
data-quality terminology of @kahn2016data, by contrast, defines
three dimensions (conformance, completeness, plausibility) that
have become standard in OHDSI-style assessments. We map both
SPIFD2 design-element categories and the @kahn2016data dimensions
to plasmode parameters as follows.

```{r spifd2-table}
spifd2_map <- data.frame(
  `SPIFD2 / Kahn category` = c(
    "Covariate measurement / availability (Kahn: completeness)",
    "Treatment exposure definition (Kahn: conformance)",
    "Outcome ascertainment (Kahn: conformance + plausibility)",
    "Unmeasured confounding (SPIFD2 sensitivity step)",
    "Sample-size adequacy (SPIFD2 step 2)"
  ),
  `Plasmode parameter` = c(
    "Fraction of MCAR missingness, pi",
    "(sens, spec) of treatment ascertainment",
    "(sens, spec) of outcome ascertainment, optionally by arm",
    "Latent U with prevalence p_U and (OR_A, OR_Y)",
    "Tested by base plasmode through n"
  ),
  `Source of plausible range` = c(
    "Data profiling, linked validation studies",
    "Chart review, claims-validation literature",
    "Algorithm validation, PPV / sensitivity studies",
    "Expert elicitation, literature, E-value bounds",
    "Known from data source"
  ),
  `Gate criterion (default)` = c(
    "RMSE ratio > 1.5 at observed missingness rate",
    "|Bias| > target at plausible (sens, spec)",
    "|Bias| > target at plausible (sens, spec)",
    "|Bias| > target at plausible U strength",
    "Coverage < min_coverage"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  spifd2_map,
  caption = "Table 1: Mapping data-quality concerns to plasmode-parameter inputs of run_plasmode_dq_stress()."
)
```

The mapping is deliberately one-to-many: a single SPIFD2 design
element (say, outcome ascertainment) can drive several plasmode
parameters (per-arm sensitivity and specificity), and a single
plasmode parameter (say, MCAR fraction) can speak to several SPIFD2
elements (covariate availability, missingness handling).

## The degradation gradient

Rather than testing a single value of each threat, the function
sweeps a vector of severities. The output is a curve of bias,
RMSE, and coverage against the severity, computed separately for
each prespecified TMLE candidate. We call this curve a *degradation
gradient* by analogy with a dose-response curve: it answers the
question, "how severe must the data degradation be before this
candidate fails the prespecified criteria?"

A robust candidate produces a flat gradient that remains inside
the prespecified thresholds. A fragile candidate shows a cliff: a
small worsening of the data tips bias or coverage past the
threshold. The candidate-selection rule
(`select_tmle_candidate(rule = "min_rmse")`) can be augmented to
require a maximum allowed degradation, in which case the
prespecified rule excludes fragile candidates
before the outcome is accessed.

The truncation level inside a TMLE candidate is more than a tuning
knob. Trimming or truncating the propensity score changes the
support of the population to which the estimand applies, and
therefore the estimand itself [@brookhart2024causalrisk;
@lash2014qba]. cleanTMLE treats the truncation level as part of
the prespecified candidate, so any estimand drift induced by
truncation is locked into the audit log and scored on the same
RMSE and coverage metrics as everything else rather than chosen
post-hoc on the realised data.

## Threat-to-assumption matrix: what each scenario probes (and what it does not) {#sec-threat-matrix}

Each DQ scenario is best understood as an *engineering
approximation* to a class of threats rather than an
assumption-specific diagnostic. Table \ref{tbl-threat-matrix}
makes that distinction explicit: for each scenario it lists the
identification or measurement assumption nominally probed, the
scenario operationalisation, and the threats the scenario *does
not* probe. The reader should treat the table as a guide to
interpreting `run_plasmode_dq_stress()` output rather than as a
claim that the four scenarios constitute a complete
threat taxonomy.

```{r threat-matrix-table}
#| label: tbl-threat-matrix
threat_mat <- data.frame(
  Scenario = c("Covariate missingness",
               "Treatment misclassification",
               "Outcome misclassification",
               "Unmeasured confounding"),
  `Nominally probes` = c(
    "Stratified MCAR for $W$; robustness of nuisance fits to imputation",
    "Consistency / SUTVA on $A$ when the treatment phenotype is imperfect",
    "Differential or non-differential outcome measurement error",
    "Conditional exchangeability when one or more confounders are unmeasured"
  ),
  Operationalisation = c(
    "MCAR mask of fraction $\\pi$ of $W$, median imputation",
    "Sensitivity / specificity flips of $A$, optionally asymmetric",
    "Sensitivity / specificity flips of $Y$, optionally per arm",
    "Latent $U \\sim \\text{Bern}(p_U)$ with $\\text{OR}_A$, $\\text{OR}_Y$"
  ),
  `Does NOT probe` = c(
    "MAR / MNAR; non-MCAR mechanisms; missingness in $A$ or $Y$",
    "Time-varying treatment versions; per-arm differential exposure error",
    "Outcome measurement processes that depend on covariates other than $A$",
    "Confounding by variables not exchangeable with the chosen latent $U$"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(threat_mat,
             caption = "Threat-to-assumption matrix. For each scenario, the assumption nominally probed, the operationalisation, and the threats the scenario does NOT probe. Severity ranges must come from external validation evidence (see §\\ref{sec-calibration}).")
```

A consequence is that the `run_plasmode_dq_stress()` output is
useful for *robustness screening* of candidate estimators against
prespecified threat ranges, not for *validating* the data source.
Variable validation, source-data audit, and post-outcome
quantitative bias analysis [@lash2014qba] remain separate
activities anchored in the validation literature
[@fda2024ehrclaims; @fda2021registries; @ema2026rwd] rather than
in the simulation engine.

## Calibration of severity ranges from external evidence {#sec-calibration}

The DQ stress test is only as informative as the severity grids
the analyst supplies. Each grid should be anchored in external
evidence, not chosen for convenience:

- **Covariate missingness fractions** should be calibrated to the
  observed missingness in the data source under study, plus
  realistic alternative scenarios from data-profiling studies
  [@kahn2016data; @fda2024ehrclaims; @ema2026rwd].
- **Treatment-ascertainment (sensitivity, specificity)** ranges
  should come from claims-validation studies for the relevant
  exposure phenotype or from independent chart review
  [@fda2024considerations; @fda2024ehrclaims].
- **Outcome-ascertainment (sensitivity, specificity)** ranges
  should come from algorithm-validation studies or PPV / sensitivity
  literature for the outcome phenotype, with optional per-arm
  differential misclassification when the validation literature
  supports it [@fda2024considerations].
- **Unmeasured-confounder strengths** ($p_U$, $\text{OR}_A$,
  $\text{OR}_Y$) should be elicited from subject-matter experts,
  prior-literature E-value bounds [@vanderWeele2017evalue], or
  causal-gap sensitivity ranges [@diaz2013sensitivity].

We recommend that the analyst record, in the decision log, a brief
citation for each severity range alongside the lock fingerprint, so
that a reviewer can check both the values and their provenance. The
boundaries of what this screen does *not* establish (data-source
validation, causal identification, post-outcome sensitivity, and
independent governance) are listed concisely in the regulatory
alignment table (§\ref{sec-roadmap}, Table \ref{tbl-reg-align}) and
in the limitations paragraph in the Discussion (§\ref{sec-discussion}).

# A quantitative supplement to fit-for-purpose review {#sec-bridge}

This section describes how the pre-outcome gate is currently scored
in published TL-RWE pipelines and what changes when the DQ stress
test feeds into the same gate. The test scores robustness of
candidate estimators against prespecified threat ranges; the
validity of those ranges, and of the data source itself, remains
an external study-level argument anchored in source-specific
assessment guidance [@fda2024considerations; @fda2024ehrclaims;
@ema2026rwd] (see Table \ref{tbl-reg-align}).

A note on terminology: @muntner2024staging describe three
review-team decisions ("proceed to the next stage", "conduct
additional analyses to address bias", or "terminate the study"),
which we shorten in figures and tables to *proceed*, *review*,
and *stop* (also written as GO / FLAG / STOP where compactness is
helpful). In prose we prefer the longer wording. The cleanTMLE
software uses the three labels for the same purpose, and the
mapping to the Muntner et al. wording is one-to-one.

## What the pre-outcome decision currently does

In the published TL-RWE literature the pre-outcome gate is a
prespecified set of qualitative checks: the SAP exists, the lock is
in place, the cohort meets minimum size and event criteria, the
propensity-score overlap looks acceptable, and the negative-
control-outcome estimate is consistent with no effect
[@dang2023roadmap; @gruber2022sap; @gruber2024evaluating]. Each
check is binary or ordinal, and each is computed from a different
source (a PDF SAP, a balance table, a negative-control estimate).
When the data-fitness argument is added, it usually appears as a
separate document (the SPIFD2 or HARPER table
[@wang2023harper]) and is adjudicated by reviewers rather than
scored against the prespecified decision rule.

## What the DQ supplement adds

The DQ stress test reports, for each candidate, the same triple
(bias, RMSE, coverage) at each severity of each threat that the
baseline plasmode reports. Because the units match, the same
prespecified thresholds (maximum allowed bias, minimum allowed
coverage, and an allowed SE-calibration window) can be applied
uniformly to the baseline plasmode and to the degraded runs. The
prespecified candidate is required to meet those thresholds both
under the locked data-generating process and across the plausible
range of each threat documented in the fit-for-purpose review.

This shifts the pre-outcome decision in three ways. First, the
data-fitness argument becomes numeric and candidate-specific:
instead of a prose claim that "the outcome ascertainment algorithm
has reported sensitivity 0.90," the analyst sees that at
sensitivity = 0.90 and specificity = 0.95 the bias of candidate
`tmle_glmnet_t05` rises from 0.001 to 0.014, exceeding the
prespecified threshold of 0.010. Second, the default
`select_tmle_candidate()` rule, which picks the RMSE-minimiser on
the baseline plasmode, gains the option `rule = "min_max_rmse"`,
which selects the candidate with the lowest worst-case RMSE across
the DQ scenarios; a candidate that wins on the baseline but
collapses under a 5 % missingness rate loses to a slightly less
efficient one that maintains acceptable performance across the
entire plausible range. Third, because the DQ output is recorded
in the same audit log as the baseline plasmode, a reviewer can
read the audit trail, see the externally derived ranges that went
in, see the prespecified thresholds that came out, and reproduce
the decision.

## Position relative to other bias analyses

The DQ stress test is distinct from quantitative bias analysis (QBA)
[@lash2014qba] and from the targeted-learning causal-gap sensitivity
analysis of @diaz2013sensitivity. QBA adjusts the *final* effect
estimate for a postulated bias and is an unblinding-stage tool;
causal-gap characterises how the estimate changes as a function of
an unmeasured-confounder bound and complements the E-value
[@vanderWeele2017evalue]. The DQ stress test, by contrast, asks
whether candidate estimators still pass the pre-outcome gate when
the data are degraded under prespecified threats, and reaches
covariate missingness and exposure misclassification in particular.

We therefore position the DQ stress test as the *first* of three
sequential bias analyses: pre-outcome (DQ stress test), at-outcome
(negative-control outcomes [@lipsitch2010negative]), and
post-outcome (E-value, QBA, causal-gap). Each addresses a different
threat, at a different stage, with a different artifact.

## Why "outcome-blind" matters here

The DQ stress test runs on synthetic outcomes generated from the
real covariate distribution and the real treatment. The real
outcome is never read. This is the property that makes the test
admissible inside the prespecification: the analyst can iterate on
the prespecified set of candidate estimators, on the degradation
parameters, and on the decision thresholds without learning
anything about the treatment-outcome association in the actual
study data [@dang2023roadmap, p. 7]. The analysis-lock fingerprint
records the final choices. Once the prespecified workflow has been
completed, the outcome is unmasked and the primary analysis is
run.

# Implementation and `run_plasmode_dq_stress()` {#sec-implementation}

## A worked workflow

A typical cleanTMLE session for a binary point-treatment endpoint
looks like the following. We omit print output for compactness;
each print method is documented in `?print.cleanroom_lock` and the
related help pages.

```{r demo-workflow, eval = FALSE}
library(cleanTMLE)

# Stage 1a: lock the analytic specification
lock <- create_analysis_lock(
  data        = study_data,
  treatment   = "treatment",
  outcome     = "event_24",
  covariates  = c("age", "sex", "biomarker", "comorbidity", "ckd"),
  sl_library  = c("SL.glm", "SL.glmnet", "SL.ranger"),
  seed        = 2026
)
lock <- attach_estimand(lock,
  description           = "Effect of A on 24-month event risk",
  population            = "Adults eligible at index",
  treatment_strategies  = c("Treatment", "Control"),
  outcome_label         = "Primary event by 24 months",
  followup              = "24 months",
  contrast              = "risk_difference"
)

# Stage 1b: cohort adequacy + design precision
audit <- create_audit_log(lock)
cp1   <- checkpoint_cohort_adequacy(lock,
                                    min_n_per_arm = 200,
                                    min_events    = 50)
audit <- record_checkpoint(audit, cp1)

# Stage 2a: propensity-score diagnostics + balance checkpoint
ps_fit  <- fit_ps_superlearner(lock)
ps_diag <- compute_ps_diagnostics(ps_fit)
cp2     <- checkpoint_balance(ps_diag, max_smd = 0.10,
                              lock_hash = lock$lock_hash)
audit   <- record_checkpoint(audit, cp2)

# Stage 2b: plasmode candidate selection + DQ stress test
candidates <- list(
  tmle_candidate("glm_t01",   "GLM PS, trunc=0.01",
                 g_library = "SL.glm",     truncation = 0.01),
  tmle_candidate("glm_t05",   "GLM PS, trunc=0.05",
                 g_library = "SL.glm",     truncation = 0.05),
  tmle_candidate("glmnet_t01","glmnet PS, trunc=0.01",
                 g_library = "SL.glmnet",  truncation = 0.01),
  tmle_candidate("glmnet_t05","glmnet PS, trunc=0.05",
                 g_library = "SL.glmnet",  truncation = 0.05)
)
plas <- run_plasmode_feasibility(lock,
                                 tmle_candidates = candidates,
                                 effect_sizes    = c(0.03, 0.05),
                                 reps            = 200)
dq_results <- run_plasmode_dq_stress(lock,
  tmle_candidates = candidates,
  reps            = 50,
  data_quality_scenarios = list(
    covariate_missingness = list(fractions = c(0.05, 0.10, 0.20)),
    treatment_misclass    = list(sensitivity = c(0.95, 0.90),
                                 specificity = c(0.99, 0.95)),
    outcome_misclass      = list(sensitivity = c(0.95, 0.90),
                                 specificity = c(0.99, 0.95)),
    unmeasured_confounding = list(U_prevalence  = 0.20,
                                  U_treatment_OR = c(1.5, 2.0),
                                  U_outcome_OR   = c(1.5, 2.0))
  )
)
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock     <- lock_primary_tmle_spec(lock, selected)

# Outcome blinding for Stage 3 and the gate.
original_lock <- lock
masked        <- mask_outcome(lock)

# Stage 3: negative-control outcomes
masked <- define_negative_control(masked, name = "nc_outcome")
stage3 <- run_residual_confounding_stage(masked, ps_fit)
cp3    <- stage3$checkpoint
audit  <- record_checkpoint(audit, cp3)

# Pre-outcome gate
gate <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
auth <- authorize_outcome_analysis(audit)
stopifnot(isTRUE(auth$authorized))

# Stage 4: unblind and run the primary analysis
final <- unmask_outcome(masked, original_lock)
g_fit <- fit_tmle_treatment_mechanism(final, ps_fit)
Q_fit <- fit_tmle_outcome_mechanism(final, g_fit)
upd   <- run_tmle_targeting_step(g_fit, Q_fit)
psi   <- extract_tmle_estimate(upd)$estimates$ATE

save_lock(final,  "lock.rds")
save_audit(audit, "audit.rds")

# Reporting tables (function names inherited from causalRisk):
make_table1(final, treatment = "treatment", smd = TRUE)
make_table2(list(`TMLE (primary)` = upd), risk_time = 24)
```

The DQ stress test is the only fundamentally new piece in the
above; the rest is conventional TL-RWE plumbing.
`run_plasmode_dq_stress()` returns a metrics object of the same
class as the standard plasmode and feeds the same gate, so adding
the data-quality argument to the workflow does not require new
infrastructure on the analyst's side.

## Inside `run_plasmode_dq_stress()`

The function loops over a scenario grid, fits every candidate
under each scenario × severity, and aggregates bias, RMSE, and
coverage. Pseudocode:

```
for each (scenario, severity) in scenario grid:
  for each replicate r:
    set seed deterministically from (lock$seed, r, scenario index)
    generate synthetic outcome from the lock's covariate-only Q0 model
    if scenario == unmeasured_confounding:
      sample latent U; modify treatment via a real PS model;
      modify outcome probabilities via OR_Y * U
    if scenario == treatment_misclass:
      apply (sens, spec) to A
    if scenario == outcome_misclass:
      apply (sens, spec) to Y, optionally by treatment arm
    if scenario == covariate_missingness:
      set MCAR fraction of W to NA, impute with column median
    for each candidate:
      fit g, Q, targeting; record est, SE, CI
  aggregate over replicates: bias, RMSE, coverage, SE/SD
return data frame indexed by (scenario, level, candidate)
```

The implementation is in `R/plasmode_dq.R`. Three details matter
for reproducibility. First, the seed is set deterministically from
`lock$seed + rep_i + sg_i * 10000L` so that re-running the same
lock with the same scenario list produces bit-identical metrics.
Second, the unmeasured-confounding scenario fits a real treatment-
mechanism (propensity) model on the lock data once and uses it as
the baseline propensity that $U$ shifts; earlier prototype versions
incorrectly used the outcome model's prediction here, which we
fix in the version released with this paper. Third, treatment
misclassification accepts asymmetric (sens, spec) pairs rather than
symmetric flips, matching the form in which validation studies
typically report exposure-algorithm performance, and outcome
misclassification accepts optional per-treatment-arm sensitivity
and specificity, allowing the user to probe differential
misclassification, which is the most worrying case in observational
outcome ascertainment.

## Relationship to `causalRisk`

`causalRisk` [@brookhart2024causalrisk] is an MIT-licensed R
package that implements IPW, g-computation, and AIPW cumulative-
risk estimators with a cohort-specification grammar that maps
neatly onto the clean-room idea. cleanTMLE reuses that grammar
and the associated reporting helpers (Brookhart, Breskin, and
Eledath are credited as copyright holders for the reused code in
the package's `DESCRIPTION` file) and adds the lock, gate, audit
log, step-by-step TMLE functions, and plasmode DQ stress test as
cleanTMLE-original components. An analyst already using
`causalRisk` for IPW analyses can move into cleanTMLE without
relearning the cohort-specification surface.

# Simulation Study {#sec-simulation}

## Design

We run a Monte Carlo simulation with $N = 2000$ observations per
replicate. Six estimators are compared:

1. **Crude:** unadjusted risk difference.
2. **IPTW:** stabilised inverse-probability-of-treatment weighting.
3. **PS Match:** 1:1 nearest-neighbour propensity-score matching.
4. **TMLE:** TMLE with the plasmode-selected specification, no
   cross-fitting.
5. **TMLE_CF (CV-TMLE):** cross-validated TMLE with two folds. We
   use the names "cross-validated TMLE" and "cross-fitted TMLE"
   interchangeably; both describe the sample-split estimator of
   @zheng2011crossvalidated and @chernozhukov2018dml in which the
   nuisance fits are made on data held out from the targeting step.
6. **Match_TMLE:** TMLE on the propensity-matched subset.

Three scenarios are run:

- **A. Good overlap.** `overlap_strength = 0.5`, no unmeasured
  confounding. Propensity scores stay inside $[0.2, 0.8]$.
- **B. Marginal overlap.** `overlap_strength = 1.5`, no unmeasured
  confounding. Propensity scores reach $[0.04, 0.99]$.
- **C. Unmeasured confounding.** `overlap_strength = 0.5`, plus a
  binary $U$ with prevalence 0.20 and odds ratios 2.0 for both $A$
  and $Y$. The analyst's covariate set excludes $U$.

For each scenario the DQ stress test is run on three covariate-
missingness fractions (0.05, 0.10, 0.20); a $2 \times 2$ grid of
treatment-ascertainment sensitivity (0.95, 0.90) by specificity
(0.99, 0.95), giving four asymmetric (sens, spec) cells; the same
$2 \times 2$ grid for outcome ascertainment; and a $2 \times 2$
grid of unmeasured-confounder strengths
($\text{OR}_A \in \{1.5, 2.0\}$ by $\text{OR}_Y \in \{1.5, 2.0\}$,
with $U$ prevalence fixed at 0.20). Across the four threats this
yields fifteen scenario × severity cells per Monte Carlo scenario.

## Monte Carlo results

```{r load-mc-results}
res_path <- file.path("..", "results", "simulation_results.rds")
if (file.exists(res_path)) {
  res <- readRDS(res_path)
  has_mc <- TRUE
} else {
  has_mc <- FALSE
  cat("Monte Carlo results not found. Run run_simulation.R first.\n")
}
```

```{r mc-table, eval = has_mc, results = "asis"}
for (sc in names(res$summaries)) {
  truth <- res$truths[[sc]]$RD
  label <- res$scenarios[[sc]]$label
  cat(sprintf("\n### %s (true RD = %.5f)\n\n", label, truth))
  knitr::kable(res$summaries[[sc]],
               digits = c(NA, 0, 5, 5, 5, 5, 5, 3, 5, 3, 5),
               caption = label) |> print()
  cat("\n")
}
```

### Interpretation

We let the simulation results speak for themselves; the printed
tables include the Monte Carlo standard error of the coverage
estimate so the reader can judge which differences exceed Monte
Carlo noise. Three patterns are robust to noise.

First, under good overlap (Scenario A) all adjusted estimators
recover the truth with bias well under the prespecified threshold
of 0.02 and with confidence intervals that cover at the nominal
rate. The crude estimator has substantial residual confounding and
its confidence intervals do not cover at 95%. This is the textbook
benchmark check.

Second, under marginal overlap (Scenario B) the crude estimator
shifts further from the truth and IPTW shows wider CIs reflecting
the heavier weights, but TMLE and CV-TMLE remain
approximately unbiased. The plasmode-selected specification chooses
a higher truncation level than under Scenario A, illustrating that
the candidate selection adapts to the observed data structure.

Third, under unmeasured confounding (Scenario C) every estimator is
biased in the same direction, because no method that adjusts only
on $W$ can remove the contribution of $U$ to the treatment-outcome
association. The bias is smallest for TMLE and largest for the
crude estimator, but the difference is one of degree, not of kind.
The pre-outcome gate correctly returns STOP under the prespecified
thresholds for Scenario C, demonstrating that the staged workflow
declines to run the primary analysis when the available data
cannot identify the estimand.

A fourth pattern concerns standard-error calibration for the IPW
estimator and is worth surfacing because it routinely surprises
applied users. The mean-SE-to-empirical-SD
ratio for IPTW is often above one in our scenarios, because the
default IF-based variance estimator treats the propensity score
as known and is therefore conservative when the propensity score
has actually been estimated [@robins1992rotnitzky;
@brookhart2024causalrisk]. Doubly robust estimators (TMLE, AIPW,
CV-TMLE) do not share this conservativeness once the nuisance
fits are properly cross-fit; the SE/SD ratio for TMLE_CF stays
near one across our scenarios. A bootstrap variance estimator
[@xu2020bootstrap] is the principled fix when reporting an
IPW-only estimate; we recommend that users wrapping
`run_iptw_workflow()` for primary inference resample the IPTW
estimator externally rather than rely on the default IF-based SE.

```{r mc-boxplots, eval = has_mc, fig.height = 9, fig.cap = "Sampling distribution of the risk-difference estimate across Monte Carlo replicates. The dashed red line marks the true RD. Under unmeasured confounding (C), all estimators are shifted from the truth."}
raw_data <- do.call(rbind, lapply(names(res$results), function(sc) {
  df <- as.data.frame(res$results[[sc]])
  df$scenario <- res$scenarios[[sc]]$label
  df$truth    <- res$truths[[sc]]$RD
  df
}))

ggplot(raw_data, aes(x = method, y = estimate, fill = method)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  geom_hline(aes(yintercept = truth), linetype = "dashed", colour = "red") +
  facet_wrap(~ scenario, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Risk-difference estimate",
       title = "Sampling distribution by estimator and scenario") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r coverage-bar, eval = has_mc, fig.cap = "95% CI coverage by estimator and scenario. The dashed red line marks the nominal 95%."}
plot_data <- do.call(rbind, lapply(names(res$summaries), function(sc) {
  df <- res$summaries[[sc]]
  df$scenario <- res$scenarios[[sc]]$label
  df
}))

ggplot(plot_data, aes(x = method, y = coverage, fill = scenario)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "red") +
  coord_cartesian(ylim = c(0.3, 1.0)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Coverage",
       title = "95% CI coverage",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r se-calibration, eval = has_mc, fig.cap = "Ratio of mean reported SE to empirical SD. Values near 1.0 indicate well-calibrated standard errors."}
ggplot(plot_data, aes(x = method, y = se_sd_ratio, colour = scenario,
                       shape = scenario)) +
  geom_point(size = 4) +
  geom_hline(yintercept = 1.0, linetype = "dashed") +
  scale_y_continuous(limits = c(0.7, 2.2)) +
  labs(x = NULL, y = "Mean SE / Empirical SD",
       title = "Standard-error calibration",
       colour = NULL, shape = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## Plasmode DQ stress-test results

```{r load-dq-results}
dq_files <- list.files(file.path("..", "results"), pattern = "^dq_stress_",
                        full.names = TRUE)
has_dq <- length(dq_files) > 0

if (has_dq) {
  dq_all <- lapply(dq_files, readRDS)
  names(dq_all) <- sub("dq_stress_(.+)\\.rds", "\\1", basename(dq_files))
}
```

```{r dq-degradation-table, eval = has_dq, results = "asis"}
for (sc in names(dq_all)) {
  dq <- dq_all[[sc]]
  sc_label <- if (has_mc && sc %in% names(res$scenarios))
    res$scenarios[[sc]]$label else sc

  cat(sprintf("\n### %s\n\n", sc_label))

  deg <- summarize_dq_degradation(dq)
  if (nrow(deg) > 0) {
    sel_cand <- if (has_mc && sc %in% names(res$meta))
      res$meta[[sc]]$selected else deg$candidate[1]
    sel_deg <- deg[deg$candidate == sel_cand, ]
    if (nrow(sel_deg) > 0) {
      cat(sprintf("Selected candidate: `%s`\n\n", sel_cand))
      knitr::kable(
        sel_deg[, c("scenario", "level", "bias_baseline", "bias_degraded",
                     "rmse_ratio", "cov_baseline", "cov_degraded", "cov_drop")],
        digits = c(NA, NA, 5, 5, 3, 3, 3, 3),
        caption = sprintf("DQ degradation: %s (candidate: %s)", sc_label, sel_cand)
      ) |> print()
    }
  }
  cat("\n")
}
```

### Reading the degradation gradient

The four degradation mechanisms produce qualitatively different
profiles. Covariate missingness produces a flat gradient in both
the good-overlap and marginal-overlap scenarios: median imputation
on top of the locked covariate set is enough to absorb a 20% MCAR
loss. Treatment misclassification attenuates the estimate toward
the null: bias rises as (sens, spec) worsen, but coverage stays
near nominal because the estimator is targeting an attenuated
truth and the SE shrinks accordingly. Outcome misclassification
under (0.90, 0.95) produces the largest bias of the three
"measurement" threats, particularly when applied differentially by
treatment arm. Unmeasured confounding produces the largest
degradation overall and is the only threat that pushes the
selected candidate's RMSE ratio above 1.5 in our default settings.

```{r dq-gradient-plots, eval = has_dq, fig.height = 10, fig.cap = "Degradation gradient. Bias (top) and coverage (bottom) for the selected TMLE candidate at each severity level of each DQ scenario, faceted by Monte Carlo scenario."}
dq_combined <- do.call(rbind, lapply(names(dq_all), function(sc) {
  m <- dq_all[[sc]]$metrics
  m$mc_scenario <- if (has_mc && sc %in% names(res$scenarios))
    res$scenarios[[sc]]$label else sc
  m
}))

dq_plot <- dq_combined[dq_combined$scenario != "none", ]

if (nrow(dq_plot) > 0) {
  if (has_mc) {
    sel_map <- vapply(names(res$meta), function(sc)
      res$meta[[sc]]$selected, character(1))
    sc_label_map <- vapply(names(res$scenarios), function(sc)
      res$scenarios[[sc]]$label, character(1))
    dq_plot <- do.call(rbind, lapply(split(dq_plot, dq_plot$mc_scenario), function(d) {
      sc_key <- names(sc_label_map)[sc_label_map == d$mc_scenario[1]]
      if (length(sc_key) > 0 && sc_key %in% names(sel_map)) {
        d[d$candidate == sel_map[sc_key], ]
      } else d
    }))
  }

  p1 <- ggplot(dq_plot, aes(x = level, y = abs(bias),
                              colour = scenario, group = scenario)) +
    geom_point(size = 2.5) + geom_line(linewidth = 0.8) +
    facet_wrap(~ mc_scenario, scales = "free_x") +
    labs(x = "DQ severity level", y = "|Bias|",
         title = "A. Bias",
         colour = "DQ scenario") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          legend.position = "bottom")

  p2 <- ggplot(dq_plot, aes(x = level, y = coverage,
                              colour = scenario, group = scenario)) +
    geom_point(size = 2.5) + geom_line(linewidth = 0.8) +
    geom_hline(yintercept = 0.95, linetype = "dashed", colour = "red") +
    facet_wrap(~ mc_scenario, scales = "free_x") +
    labs(x = "DQ severity level", y = "Coverage",
         title = "B. Coverage",
         colour = "DQ scenario") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          legend.position = "bottom")

  gridExtra::grid.arrange(p1, p2, ncol = 1)
}
```

## Comparison with E-value-based sensitivity

The E-value [@vanderWeele2017evalue] reports the minimum strength
of unmeasured confounding required to reduce the observed estimate
to the null. It is post-hoc: it is computed from the realised
estimate after the outcome is unblinded. The plasmode DQ stress
test runs in the opposite direction: given a hypothesised confounder
strength, it computes the bias of each candidate. The two are
complementary, and we recommend reporting both.

```{r evalue-comparison, eval = has_mc && has_dq, results = "asis"}
if ("unmeasured_conf" %in% names(res$summaries)) {
  tmle_row <- res$summaries$unmeasured_conf[
    res$summaries$unmeasured_conf$method == "TMLE", ]

  if (nrow(tmle_row) > 0 && !is.na(tmle_row$mean_est)) {
    truth_c <- res$truths$unmeasured_conf$RD
    truth_a <- res$truths$good_overlap$RD
    r0 <- res$truths$good_overlap$risk_0

    tmle_a <- res$summaries$good_overlap[
      res$summaries$good_overlap$method == "TMLE", ]
    rr_hat <- abs((r0 + tmle_a$mean_est) / r0)
    ev <- compute_evalue(rr_hat, ci_bound = 1.0)

    dq_u <- dq_all$good_overlap$metrics
    dq_u_rows <- dq_u[dq_u$scenario == "unmeasured_U", ]

    cat("\n| Quantity | Value |\n|---|---|\n")
    cat(sprintf("| True RD (no U) | %.5f |\n", truth_a))
    cat(sprintf("| True RD (U present, OR = 2.0) | %.5f |\n", truth_c))
    cat(sprintf("| TMLE estimate, Scenario A (no U) | %.5f |\n", tmle_a$mean_est))
    cat(sprintf("| TMLE estimate, Scenario C (U present, not adjusted) | %.5f |\n",
                tmle_row$mean_est))
    cat(sprintf("| MC bias from unmeasured U (Scenario C) | %.5f |\n",
                tmle_row$bias))
    cat(sprintf("| E-value (from Scenario A estimate) | %.2f |\n", ev["e_value"]))

    if (nrow(dq_u_rows) > 0) {
      for (i in seq_len(nrow(dq_u_rows))) {
        cat(sprintf("| DQ stress bias at %s | %.5f |\n",
                    dq_u_rows$level[i], dq_u_rows$bias[i]))
      }
    }

    cat("\n: Unmeasured confounding under two methods: a single E-value threshold and a candidate-specific bias curve.\n\n")
  }
}
```

# Case Study: Rescue.Co Kenya Trauma Registry {#sec-casestudy}

## Why a real-data workflow demonstration

We use the Rescue.Co Kenya Trauma Registry as a **workflow
demonstration** and software stress test on a real registry
structure, not as evidence that the framework improves causal
validity. The simulation in §\ref{sec-simulation} is small (5
covariates, n = 2,000, fully observed outcome) by design, because
we wanted analytic control over each threat. A working framework
also has to execute on real observational data: covariates that
come from multiple instruments, outcomes that are missing for
non-trivial fractions of the cohort, and confounders the protocol
forgot. The Rescue.Co Kenya Trauma Registry is an applied
comparative-effectiveness study at three Nairobi hospitals (KNH,
Mama Lucy, Mbagathi) comparing prehospital transport by a
specialised emergency-medical-services provider (Rescue.Co) with
traditional ambulance care, on functional outcome six months
post-injury. All five stages were run on cleanTMLE 0.1.1; the
released `manuscript_artifacts/case_study_metadata.json` records
the version and lock hash. The example is fully reproducible from
the bundled `rescueCo/` directory in the companion repository.
The DQ stress test in this run was executed at low replication
(see §\ref{sec-casestudy-dq}), so the Rescue.Co example is not
used as primary evidence of DQ-stress validity; it is presented
as a real-registry execution example.

The example is interesting for four reasons that the simulation
does not exercise: (a) the primary outcome (Glasgow Outcome Scale
Extended, GOSE 1--8) has 24.6 % missing values from loss to
6-month follow-up; (b) 492 ambulance arrivals were inter-facility
transfers and were excluded from the primary cohort, leaving the
direct-from-scene 1,693; (c) an 8-week healthcare-worker strike
(15 March 2024 to 8 May 2024) acts as a quasi-natural-experiment
confounder that is not part of either SPIFD2 or the four threats
encoded by `run_plasmode_dq_stress()`; and (d) two prespecified
socioeconomic negative controls (urban household, primary cooking
fuel) screen for residual SES confounding.

## SuperLearner library specification

We follow @phillips2023sl on SuperLearner library composition. With
$n_{\text{eff}} \approx 1{,}277$ complete-outcome observations, the
recommended cross-validation count is $V = 10$, and the recommended
library is a screened mixture of a parametric base learner
(`SL.glm`), a regularised learner (`SL.glmnet`), and a generalised-
additive learner (`SL.gam`). The cleanTMLE helper
`validate_superlearner_spec(n_eff, library = ...)` records this
recommendation in the lock and warns when the user-supplied library
is degenerate. The rescueCo lock uses the same recommended library.

## Cohort, estimand, and pre-specification

The lock fixes the binary version of GOSE (`gose_good = 1` if GOSE
> 4) as the primary outcome, treating ambulance arrival of any kind
as in-scope and excluding inter-facility transfers. After
exclusions the analytic cohort is **n = 1,693 (1,013 Rescue.Co; 680
other ambulance)**; n = 1,277 have non-missing 6-month GOSE.
Twenty-one covariates enter the propensity-score model after a
near-zero-variance filter drops nine of the originally selected
thirty (combined prevalence below 1 % in the analytic cohort): the
retained set covers demographics (age, sex, marital, education),
distance, mechanism of injury (a subset of indicators), the
chronic-illness panel (the higher-prevalence conditions), SES
proxies (urban household, occupation), payment, GCS, ISS, arrival
month, alcohol use at injury, and assault intent.

The lock declares two prespecified sensitivity plans (a 4-point PS
truncation grid and a transfer-inclusion rerun) and five negative
controls, each tagged with its @muntner2024staging confounding
domain: `chronic_hypertension`, `chronic_diabetes_insulin`, and
`chronic_hiv_art` cover *confounding by indication* and
*functional status*; `household_urban` and `fuel_wood` cover
*health-seeking behaviour* and *access to healthcare*. The
`define_negative_control(..., domain = ...)` argument added in
cleanTMLE 0.1.1 records the domain in the lock and the audit log so
a reviewer can verify that the NC set spans the four domains rather
than concentrating in one. Three of the five NCs
(`chronic_diabetes_insulin`, `chronic_hiv_art`, `fuel_wood`) are
dropped before fitting by the same near-zero-variance filter that
trims the PS-covariate set, and the bundle's Stage 1 script logs the
drops to the pipeline log (the cleanTMLE 0.1.1 audit-log capture for
NC failures applies at fit time, not at sanitisation time, so these
drops surface in the run log rather than in the structured audit).
The remaining two NCs are fit at Stage 3. The estimand is attached via `attach_estimand()` and the lock is
fingerprinted with SHA-256. Stage 2b runs the baseline
`run_plasmode_feasibility()` on two GLM-PS truncation candidates
(`glm_t01` with $t = 0.01$, `glm_t05` with $t = 0.05$) for 50
replicates each, and Stage 3 runs the full
`run_plasmode_dq_stress()` over the four threats on a wider
candidate set (`glm_t01`, `glmnet_t01`, `ensemble_t01`,
`ensemble_t05`). The locked candidate is selected via
`select_tmle_candidate(rule = "min_max_rmse", dq_results = ...)`,
the manuscript's recommended minimax rule.

## Pre-outcome gate and plasmode results {#sec-casestudy-dq}

The Stage 2b feasibility plasmode metrics on the locked DGP are:

| Candidate | Bias    | RMSE   | Coverage | SE/SD |
|-----------|--------:|-------:|---------:|------:|
| `glm_t01` | -0.0006 | 0.0120 |     0.94 | 1.00  |
| `glm_t05` | -0.0006 | 0.0120 |     0.94 | 1.00  |

: Stage 2b feasibility plasmode metrics (rescueCo lock, 50
replicates per candidate, additive risk-difference effect of 0.05).

Both candidates pass the prespecified gate targets
(`max_abs_bias = 0.01`, `min_coverage = 0.90`,
`se_sd_window = (0.8, 1.2)`); the two truncation thresholds
produce indistinguishable performance because the propensity
scores stay well inside the [0.01, 0.99] bounds (range
0.21--0.81). The Stage 3 DQ stress test sweeps the four threats
over four candidates and 3 replicates per cell (the bundled
`config/clean_room_config.yml` sets `simulation.n_sims = 3`, low
enough that coverage estimates are noisy; we recommend re-running
with at least 50 replicates for inference); the minimax rule
locks `glm_t01` as the primary specification. Negative controls
run at Stage 3 on the locked PS:

| Negative control       | Estimate | SE      | p-value | Pass |
|------------------------|---------:|--------:|--------:|:----:|
| chronic hypertension   |  0.00248 | 0.0072  |   0.73  |  ✓   |
| household urban        | −0.00077 | 0.0071  |   0.91  |  ✓   |

: Stage 3 negative-control TMLE results.

At the pre-outcome decision point (`authorize_outcome_analysis()`),
the two passing negative controls and the plasmode metrics support
a *proceed* classification. The audit log records the three NC
drops alongside the two passing checkpoints, so a reviewer can see
at a glance that the negative-control coverage is partial. The
analysis is then unmasked and the outcome-analysis functions run.

## Effect estimates

```{r casestudy-table}
case_eff <- data.frame(
  Method = c("Crude (unadjusted)",  "PS matching",
             "IPTW (stabilised)",   "Full-cohort TMLE",
             "Matched-cohort TMLE", "IPCW-weighted TMLE"),
  N      = c(1277L, 615L, 1693L, 1693L, 1230L, 1277L),
  RD     = c(0.0370, 0.0272, 0.0366, 0.0306, 0.0259, 0.0326),
  CI_lo  = c( 0.0031, -0.0198,  0.0017, -0.0012, -0.0150, -0.0004),
  CI_hi  = c( 0.0708,  0.0741,  0.0714,  0.0623,  0.0668,  0.0656),
  stringsAsFactors = FALSE
)
knitr::kable(case_eff, digits = 4,
             caption = "rescueCo binary-outcome estimates (good functional outcome at 6 months, GOSE > 4). All six estimates produced by cleanTMLE 0.1.1 following the staged-analysis vignette.")
```

The six estimators agree on direction. The crude unadjusted
contrast (RD = 0.037) and the stabilised IPTW (RD = 0.037) sit at
the upper end of the range, with 95 % confidence intervals that
exclude zero. The simulation-locked full-cohort TMLE (the
prespecified primary estimator) returns RD = 0.031 with a 95 % CI
that just crosses zero (−0.001, 0.062). The IPCW-weighted TMLE,
which uses inverse-probability-of-censoring weights to address the
24.6 % outcome missingness, returns RD = 0.033 with a CI that just
crosses zero (−0.0004, 0.066). The two matching-based estimators
give the smallest point estimates (RD = 0.027 for matched RD
without further adjustment; RD = 0.026 for matched-cohort TMLE) and
the widest CIs, reflecting the smaller matched-pair samples
(n = 615 and n = 1230). The full-cohort TMLE remains the
prespecified primary because it is doubly robust and uses the
locked propensity-score truncation chosen at Stage 2b on the full
cohort; the IPCW-TMLE is the prespecified missing-outcome
sensitivity row.

Following @gruber2022sap (Step 5) and @diaz2013sensitivity, we
report a tipping-point bound on the simulation-locked full-cohort
TMLE estimate via `tipping_point_sensitivity()`. The lower CI bound
sits 0.0012 below zero, so the qualitative conclusion is already
that the data are consistent with no effect; the function returns
0.0012 as the distance from zero. The IPTW estimate, whose lower
bound is 0.0017 above zero, would be flipped through zero by a
residual bias of 0.0017 (4.6 % of the IPTW point estimate). The
E-value [@vanderWeele2017evalue] for the full-cohort TMLE is 1.22
(CI bound 1.00), small in either case. The qualitative reading is
"Rescue.Co transport is consistent with a small favourable shift
in functional outcome that the prespecified primary estimator
cannot distinguish from null at $\alpha = 0.05$." An unmeasured
confounder of the strength suggested by the healthcare-worker
strike, or by the inter-facility transfer imbalance (40.7 % of
non-Rescue.Co versus 2.5 % of Rescue.Co ambulance arrivals were
transfers, all excluded from the analytic cohort), would push any
of the six estimates further toward null.

## What the case study tells us about the workflow

### Prespecification, audit, and the pre-outcome decision

The decision log captured every prespecified analytic choice
(transfer exclusion, GOSE construction, truncation thresholds)
with timestamps. Outcome-analysis functions did not run until the
prespecified workflow had been completed, and the analysis-lock
fingerprint reconciled on every reload. One real gap surfaced: the
audit trail did not capture covariates dropped for near-zero
variance during data sanitisation, because those drops occurred
upstream of the negative-control fits rather than at fit time.
The cleanTMLE 0.1.1 fix only partially closes that gap. With that
caveat, the prespecification and audit functions performed as
described in §\ref{sec-framework}.

### Outcome-blind plasmode

Stage 2b feasibility coverage of 0.94 with bias below 0.001 across
50 replicates is the kind of validation a reviewer would want.
Stage 3 then runs the full DQ stress test (`run_plasmode_dq_stress()`)
over the four threats; the bundled configuration uses 3 replicates
per scenario × severity cell, which is low enough that the coverage
estimates are noisy (the baseline coverage rounds to 1.00 because
all 3 CIs happened to cover), and we treat the DQ run as a
structural smoke test rather than a calibration measurement until
it is rerun with more replicates. The four threats encoded by
`run_plasmode_dq_stress()` (covariate MCAR, exposure misclassification,
outcome misclassification, latent $U$) do not cover two design
challenges that mattered in the rescueCo study, the strike-window
confounder and the inter-facility transfer imbalance, which
motivates the time-window and MAR / MNAR extensions planned for
cleanTMLE 0.2.

### Estimator functions and real-data robustness

Applying cleanTMLE 0.1.0 to the rescueCo lock surfaced more than a
dozen distinct issues, six of them blocking, which the companion
0.1.1 release addresses. The marquee problem was that
`run_plasmode_dq_stress()` failed with
`"mu must be a nonempty numeric vector"` on a lock with NA in the
outcome (now guarded). The step-by-step TMLE functions also refused
to fit on NA outcomes (now subset to complete cases with an MCAR
warning); `fit_ps_superlearner()` crashed on stale parallel
backends (now defaults to plain sequential evaluation);
`make_table1()` rejected a lock object (now accepts it); and
`run_residual_confounding_stage()` silently dropped failed NC fits
(now records each failure in the audit log). We document these
candidly because the working theory of the framework is that the
prespecification and audit functions should make problems visible,
including problems with the package itself, rather than absorb
them into a hand-edited final table.

The full release notes are in `cleanTMLE/NEWS.md`, and the
rescueCo companion repository's `clean_room/results/manuscript_artifacts/`
directory contains the data behind the tables above.

# Discussion {#sec-discussion}

## What is genuinely novel

The novelty in this paper is not that cleanTMLE solves causal
identification. The roadmap and target-trial literature
[@dang2023roadmap; @petersen2014roadmap; @hernan2016targettrial]
already provides that frame, and the package operates inside its
estimation and pre-outcome diagnostics steps rather than around
them. The novelty is that staged prespecification, estimator
locking, and the pre-outcome decision rule become explicit in
code: an analysis-lock object, an audit trail, a decision log, and
a proceed / review / stop classification of the analysis before
the outcome is examined. The DQ stress test then adds a
candidate-specific pre-outcome robustness screen on top of the
standard plasmode loop. Together, these features support
transparent reconstruction of paper-only prespecification
checklists [@gruber2022sap; @cashin2024target] by an external
reviewer.

## Why the contribution is mainly operational

A useful test is to ask what kind of evidence the package can and
cannot produce. cleanTMLE can document that the analyst
prespecified the estimand, the set of candidate estimators, the
decision thresholds, and the severity ranges; that the pre-outcome
decision was made before unblinding; that the locked candidate
was the one that ran on the real outcome; and that an audit trail
with a fingerprint records the sequence. cleanTMLE *cannot*
demonstrate that the data source is fit for the regulatory
question, that the exposure or outcome phenotypes have been
validated, that the target population is the regulatory decision
population, or that the unmeasured-confounder bound is
defensible. Those remain study-level activities, anchored in
target-trial protocols [@hernan2016targettrial; @cashin2024target],
source-specific guidance [@fda2024considerations; @fda2024ehrclaims;
@fda2021registries; @ema2024guide; @ema2026rwd], and post-outcome
sensitivity analyses [@vanderWeele2017evalue; @lash2014qba;
@diaz2013sensitivity].

## How the workflow aligns with regulatory expectations

For regulatory-grade RWE, cleanTMLE is best used as one layer in a
broader governed workflow. The lock and audit machinery track the
prespecification and traceability that FDA RWE program guidance
expects [@fda2018rwe; @fda2024considerations;
@fda2024noninterventional] and match the data-quality dimensions
in the EMA RWD framework [@ema2026rwd] and ICH M14 (March 2026)
[@ich2024m14], as summarised in Table \ref{tbl-reg-align}. The
software-mediated workflow can complement, but does not replace,
personnel-based access controls and independent oversight.
Documented source assessment, role-based access where feasible,
and structured protocol records remain the regulatory backbone,
with the software arrangement sitting inside that backbone rather
than substituting for it.

## What the DQ stress test adds

The baseline plasmode asks which candidate has the lowest RMSE
under the locked data-generating process. The DQ-augmented
plasmode asks the larger question: which candidate has the lowest
RMSE and also stays inside the prespecified gate thresholds across
the plausible range of each threat documented in the
fit-for-purpose review.

## Comparison with adjacent tools

E-values [@vanderWeele2017evalue] are post-hoc and not candidate-
specific. They tell the reader how strong an unmeasured confounder
would have to be to nullify the observed effect. The plasmode DQ
stress test is pre-hoc and candidate-specific: it tells the
reviewer how biased each candidate becomes under each hypothesised
strength of confounding, on the same bias / RMSE / coverage scale
as the other pre-outcome diagnostics. The two are answering
different questions and should both be reported.

Quantitative bias analysis [@lash2014qba] adjusts the final
estimate for postulated biases. Like the E-value, QBA is an
unblinding-stage tool. The DQ stress test does not adjust the
estimate; it asks whether the estimator continues to meet the
prespecified performance criteria when the biases are present. We
therefore recommend QBA at the analysis stage and the DQ stress
test at the prespecification stage.

The targeted-learning causal-gap sensitivity analysis
[@diaz2013sensitivity] characterises the change in $\Psi$ as the
unmeasured-confounder bound varies. Causal-gap is unblinding-stage
and characterises one threat (unmeasured confounding) in a TL-
specific way. The DQ stress test is broader (it covers four
threats) but cruder (it does not adjust the estimate). Both are
useful and complementary.

Negative-control outcomes [@lipsitch2010negative] test for
residual confounding in the realised data. Because they use real
data, they are at-outcome rather than pre-outcome. The DQ stress
test cannot detect residual confounding; the NCO check cannot
detect covariate missingness or exposure misclassification. They
target different threats. @muntner2024staging recommend selecting
NCOs across four domains (confounding by indication, functional
status, health-seeking behaviour, and access to healthcare) and
@levintow2023nco give pragmatic guidance for choosing them; the
cleanTMLE workflow records each declared NCO in the lock so the
chosen domains are part of the audit trail.

@kent2026fiord reports a worked applied use of the staging
framework for a comparative-effectiveness study of evolocumab
versus non-treatment for major adverse cardiovascular events in
the Komodo Healthcare Map: negative-control outcomes flagged
residual bias in the health-seeking-behaviour domain, the review
team allowed the study to proceed conditional on a pre-specified
E-value quantitative-bias-analysis plan, and the final risk
difference at four years was reported as roughly $-4\%$. The
example illustrates both the staged decision rule and the role of
negative-control domains in practice.

## Limitations of plasmode data-quality stress testing

A plasmode data-quality stress test is only as informative as the
synthetic outcome model and degradation mechanisms used to
generate it. Passing the pre-outcome decision rule does not
establish identification, remove unmeasured confounding, validate
the data source, or demonstrate that the target population matches
the regulatory decision population. It shows only that, under the
prespecified outcome-generating mechanism and degradation ranges,
a candidate estimator maintained acceptable finite-sample
performance according to the chosen criteria. Severity ranges
should therefore be justified from external validation studies,
data profiling, subject-matter knowledge, expert elicitation, or
formal sensitivity-analysis bounds. When such evidence is weak,
the stress test should be interpreted as exploratory rather than
confirmatory.

The DQ stress test also depends on the same parametric
synthetic-outcome model used by the baseline plasmode. If that
model misrepresents the joint covariate–treatment–outcome
distribution (for example by omitting effect modification or
mis-specifying functional form), the simulated bias and coverage
under stressed scenarios may be optimistic or pessimistic in ways
that no amount of replication can correct.

## Other limitations

The current implementation has six additional limitations the
analyst should keep in mind, grouped into model assumptions,
computational cost, and estimand scope.

### Model assumptions

The covariate-missingness scenario assumes MCAR. MAR and MNAR
mechanisms require an explicit missingness model and are not yet
supported. The principled alternative is inverse-probability-of-
missingness weighting (`identify_missing()` in `causalRisk`
[@brookhart2024causalrisk]), which weakens the assumption to
stratified MCAR; we plan to expose the same option in a future
cleanTMLE release.

The plasmode DGP itself is parametric. The outcome is generated
from a logistic regression of $Y$ on $W$ fitted to the lock data,
and the synthetic effect is constant on the risk-difference scale
($p_1^{\text{sim}} = p_0^{\text{sim}} + \Delta$). Effect
modification, time-varying treatment, and competing risks would
need a richer DGP. The package architecture permits these
extensions; the simulation reported here does not exercise them.

The plausible degradation ranges have to come from external
evidence. The software cannot decide whether 5 % or 20 %
covariate missingness is realistic for a given data source; the
analyst has to cite a profiling or validation study. SPIFD2 makes
the same demand qualitatively, and the DQ stress test inherits it.

### Computational cost

Each scenario $\times$ severity adds a full set of plasmode
replicates. With the configuration of §\ref{sec-simulation} (50
replicates per cell, fifteen scenario $\times$ severity cells, and
four candidates) the test fits 3,000 TMLE models per Monte Carlo
scenario. The default settings already use fewer replicates than
the baseline plasmode (50 versus 200); for larger candidate grids,
parallelising the outer loop over scenarios is the right next step.

### Estimand scope

The running estimand in this paper is the marginal risk difference
for a binary point exposure. Counterfactual hazard ratios are
available in `causalRisk` via `estimate_ipwhr()` and in `survtmle`
[@survtmlepackage2018] but are not yet wrapped into the cleanTMLE
staged workflow. Sites that need hazard-ratio estimation can call
those packages directly and feed the output through the same lock
and audit trail.

Several other estimand families sit outside cleanTMLE's current
scope: counterfactual cumulative-incidence curves with competing
risks, cumulative-count IPW (`estimate_ipwcount()`), multinomial
treatments, and the SMR / overlap / normalised / stabilised weight
families. cleanTMLE in this release narrows to binary-treatment
risk-difference TMLE; analysts who need the broader estimand
surface can use `causalRisk` directly with the same cohort-
specification grammar.

## Future extensions

The most consequential extensions, in priority order, are:

1. **Richer synthetic-outcome models:** effect modification, time-
   varying treatment, and competing risks (the current DGP is a
   logistic regression with a constant additive effect on the
   risk-difference scale).
2. **MAR / MNAR missingness:** `identify_missing()` from `causalRisk`
   [@brookhart2024causalrisk] gives an inverse-probability-of-
   missingness weighting path that could be lifted into cleanTMLE
   without requiring a structural missingness model.
3. **Survival and competing-risk estimands:** wrapping `survtmle`
   [@survtmlepackage2018] and `lmtp` [@lmtppackage2021] inside the
   lock and audit machinery so that hazard-ratio and cumulative-
   incidence estimands inherit the same governance.
4. **Transportability and decision-population shift:** the
   plasmode framework can in principle simulate covariate-
   distribution shift between the analytic cohort and the target
   population, but the package does not yet expose this scenario.
5. **Calibration helpers for severity ranges:** small helpers that
   read validation-literature parameters (PPV, sensitivity, claims-
   accuracy ranges) from a YAML file and translate them into
   `data_quality_scenarios` lists, with the citations attached to
   the audit log.

## When to use cleanTMLE inside a broader governed clean-room process

For routine pharmacoepidemiology and most academic RWE studies,
cleanTMLE used on its own provides an adequate prespecification
record: the analyst prespecifies the analytic plan, runs the
workflow inside a single R session, and archives the analysis
lock together with the audit and decision logs. For
submission-oriented RWE (FDA, EMA, or HTA contexts), the
recommended deployment places cleanTMLE inside a broader governed
process with separate study-level work items: source-data audit
and provenance, exposure and outcome phenotype validation,
role-separated analyst and reviewer access, and an independent
review of the pre-outcome decision by personnel who did not
author the analytic plan. The software-mediated workflow records
that the prespecified contract has been followed; the governed
process establishes that the contract is the right one
[@muntner2024staging; @fda2018rwe; @ema2024guide].

## Recommendations for practice

A fillable SAP template aligned with the prespecification checklist
of @gruber2022sap and the staging construct of @muntner2024staging
is shipped with the package as
`system.file("templates", "sap_template.qmd", package = "cleanTMLE")`.
The template walks the analyst through the eight-section structure
(estimand under ICH E9(R1), identifiability assumptions and DQ
mappings, candidate grid, gate thresholds, NC domain coverage,
sensitivity plan, decision log, reporting checklist) so that the
prespecification is filed in the same form that the audit log will
later reference.

We recommend a four-step pre-outcome workflow when the analyst is
preparing a TL-RWE study.

1. Document the SPIFD2 template, citing validation studies or data
   profiling for each plausible range.
2. Lock the candidate grid and the gate thresholds before any
   simulation runs.
3. Run `run_plasmode_feasibility()` and `run_plasmode_dq_stress()`
   in the same prespecified Stage 2b. Use the SPIFD2-derived ranges
   as the inputs to `data_quality_scenarios`.
4. Score the gate. If a candidate is robust to the SPIFD2
   plausible ranges, lock it. If no candidate is robust, the
   correct decision is STOP, and the analyst should reconsider
   either the data source, the estimand, or the gate thresholds
   before unblinding.

The audit log preserves all of the above. The final paper or
regulatory submission can be reconstructed from the lock and the
audit, including the SPIFD2 ranges, the plasmode and DQ stress
metrics, the pre-outcome decision, and the post-outcome estimate.

# Summary {#sec-summary}

cleanTMLE is an R package for staged, outcome-blind targeted
learning that produces reproducible prespecification records. It
extends the outcome-blind plasmode loop of @dang2023roadmap and
@gruber2022sap with four data-quality degradation mechanisms,
producing per-candidate degradation gradients on the same bias /
RMSE / coverage scale as the baseline plasmode so the prespecified
decision rule can be applied uniformly to both. cleanTMLE
operationalises a quantitative bridge from qualitative
data-fitness concerns to prespecified outcome-blind plasmode
stress-test scenarios; it does not itself determine whether a data
source is fit for purpose, and the severity ranges entered into the
stress test must come from prior validation studies, data
profiling, or expert elicitation. The analysis-lock object, audit
log, and decision log persist the prespecification as reproducible
records that downstream estimation functions check before
running. Cohort-specification and reporting functions follow
`causalRisk` [@brookhart2024causalrisk] under MIT-licence
attribution, so an analyst can move between the two packages
without relearning the function names or argument order. The
package is available at <https://github.com/amertens/cleanTMLE>;
the simulation code for the results in §\ref{sec-simulation} is in
the companion repository <https://github.com/amertens/clean-room-sim>.

# Acknowledgements {.unnumbered}

We thank the authors of `causalRisk` (M. Alan Brookhart, Alexander
Breskin, and Bassim Eledath) for the MIT-licensed code and
function naming that cleanTMLE inherits, and the targeted-learning
RWE roadmap authors (Dang, Gruber, Phillips, Lee, Ho, van der Laan, Petersen,
and colleagues) whose published frameworks cleanTMLE
operationalises in software. Any errors are ours.

# References {.unnumbered}

::: {#refs}
:::

```{r session-info}
sessionInfo()
```
