---
title: "Which estimands can this design deliver? The Rescue.Co case study"
subtitle: "A cleanTMLE clean-room analysis of five transport contrasts"
author: "cleanTMLE case study"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
OUT <- file.path("..", "results", "multiarm")
rd <- function(f) {
  p <- file.path(OUT, f)
  if (file.exists(p)) read.csv(p, stringsAsFactors = FALSE) else NULL
}
locks    <- rd("lock_summary.csv")
support  <- rd("support_by_contrast.csv")
feas     <- rd("estimand_feasibility.csv")
nd       <- rd("near_deterministic_covariates.csv")
vr       <- rd("violation_regions.csv")
who      <- rd("who_is_unsupported.csv")
slw      <- rd("sl_propensity_weights.csv")
ncl      <- rd("nc_ladder.csv")
chk      <- rd("collider_check.csv")
smap     <- rd("support_map.csv")
dcmp     <- rd("plasmode_design_comparison.csv")
est      <- rd("ladder_estimates.csv")
dlog     <- rd("ladder_design_log.csv")
recon    <- {
  p <- file.path("..", "results", "reconciliation_table.csv")
  if (file.exists(p)) read.csv(p, stringsAsFactors = FALSE) else NULL
}
fmt <- function(x, d = 4) formatC(x, format = "f", digits = d)
ci <- function(e, lo, hi, d = 4)
  sprintf("%s [%s, %s]", fmt(e, d), fmt(lo, d), fmt(hi, d))
```

## The question

The study protocol names Rescue.Co against all other transport as the
primary comparison. Whether that comparison is estimable is not a given: an
average treatment effect requires every kind of patient to have a real
probability of appearing in both arms, and most patients who reach a Nairobi
trauma registry without an ambulance are not patients Rescue.Co serves. This
case study runs the cleanTMLE design stage, with no outcome access, on five
locks: the four multi-arm contrasts on the known-transport cohort and the
restricted primary cohort. The design stage grades each contrast's support,
maps which estimands each can deliver, and pre-registers the fallback rule;
only then are outcomes joined and the declared ladder estimated. Every
number below is computed from the pipeline's CSV outputs at render time.

## Cohorts and locks

```{r locks}
if (!is.null(locks)) kable(locks, col.names = c(
  "Lock", "Contrast", "n", "Treated", "Control", "Design columns"))
```

The design matrix is rebuilt from the raw registry to the main pipeline's
construction rules and matches its 90 design columns exactly (see
`design_columns.csv`); prevalence and variance filters are anchored on the
primary cohort, and the five negative controls are registered as cohort
columns before any filter, so no screen can drop them.

## Support by contrast

```{r support}
if (!is.null(support)) {
  s <- support[, c("contrast", "n", "n_treated", "n_control",
                   "pct_outside_band", "max_iptw_weight", "p99_iptw_weight",
                   "ess_treated", "ess_control", "c_statistic", "verdict",
                   "escalated")]
  kable(s, col.names = c("Lock", "n", "Treated", "Control",
                         "% outside [0.05, 0.95]", "Max weight",
                         "99th pct weight", "ESS treated", "ESS control",
                         "c-statistic", "Verdict", "Escalated"),
        digits = c(0, 0, 0, 0, 2, 1, 1, 0, 0, 3, 0, 0))
}
```

```{r support-narrative, results='asis'}
if (!is.null(support)) {
  sev <- support$contrast[support$verdict %in% c("SEVERE", "FAIL")]
  ok  <- support$contrast[support$verdict == "PASS"]
  cat(sprintf(paste0(
    "The gate grades %s as SEVERE or worse and %s as PASS. ",
    "The verdict string travels with every estimate downstream.\n"),
    paste(sev, collapse = ", "),
    if (length(ok)) paste(ok, collapse = ", ") else "none"))
}
```

Mirrored propensity histograms per lock (band shaded):

```{r support-figs, results='asis'}
for (cn in c("C1", "C2", "C3", "C4", "PRIMARY")) {
  f <- file.path(OUT, sprintf("fig_support_%s.png", cn))
  if (file.exists(f)) cat(sprintf("![](%s)\n\n", f))
}
```

### Near-deterministic strata and tree-identified regions

```{r nd}
if (!is.null(nd)) kable(nd, digits = 4)
```

```{r vr}
if (!is.null(vr)) kable(vr, digits = 4)
```

## Which estimands does each design support?

```{r feas}
if (!is.null(feas)) {
  f2 <- feas[, c("contrast", "estimand", "max_weight", "ess_treated",
                 "ess_control", "n_removed_treated", "n_removed_control",
                 "verdict", "feasible")]
  kable(f2, col.names = c("Lock", "Estimand", "Max weight", "ESS treated",
                          "ESS control", "Removed treated",
                          "Removed control", "Verdict", "Feasible"),
        digits = c(0, 0, 1, 0, 0, 0, 0, 0, 0))
}
```

The rows tell the study's story: on the protocol contrast the ATE's weights
explode, the ATT's control weights stay bounded by max g/(1-g), and the
ATO's weights are bounded by one by construction, so the effect among the
treated and the overlap-weighted effect remain estimable where the ATE is
not.

### Who is unsupported

Standardised differences of the patients outside the band against those
inside it, on interpretable variables. Trimming changes the estimand; this
is who the trimmed analysis is no longer about.

```{r who}
if (!is.null(who)) {
  w10 <- do.call(rbind, lapply(split(who, who$contrast), function(d) {
    d <- d[d$population == "all", ]
    head(d[order(-abs(d$smd)), ], 8)
  }))
  kable(w10, row.names = FALSE, digits = 3)
}
```

## The outcome-blind support map

`simulate_support()` generates synthetic data under the generate-treatment
plasmode design across a prespecified family of outcome surfaces
(confounding along the propensity direction, effect modification along it),
computes each estimand's truth on every replicate, and reports where each
estimator stays within the bias tolerance at near-nominal coverage.

```{r smap}
if (!is.null(smap)) {
  m <- smap[, c("contrast", "confounding", "modification", "estimand",
                "truth_mean", "bias", "coverage", "se_cal", "pass")]
  kable(m, digits = c(0, 1, 1, 0, 4, 4, 3, 2, 0))
}
```

```{r smap-figs, results='asis'}
for (cn in c("PRIMARY", "C1", "C2", "C3", "C4")) {
  f <- file.path(OUT, sprintf("fig_supportmap_%s.png", cn))
  if (file.exists(f)) cat(sprintf("![](%s)\n\n", f))
}
```

### The plasmode design comparison

Keeping each subject's observed treatment and simulating only the outcome
induces a positivity violation by construction (Shaw et al. 2025). The table
compares the two designs on this study's own cohorts at the same replicate
count.

```{r dcmp}
if (!is.null(dcmp)) {
  d <- dcmp[dcmp$estimand == "ATE",
            c("contrast", "design", "confounding", "bias", "coverage",
              "se_cal", "emp_sd", "mean_se")]
  kable(d[order(d$contrast, d$confounding, d$design), ],
        row.names = FALSE, digits = c(0, 0, 1, 4, 3, 2, 4, 4))
}
```

## Negative controls along the restriction ladder

```{r ncl}
if (!is.null(ncl)) {
  n2 <- ncl[, c("contrast", "cohort", "negative_control", "n", "estimate",
                "ci_lower", "ci_upper", "p_value", "flagged", "status")]
  kable(n2, digits = c(0, 0, 0, 0, 4, 4, 4, 4, 0, 0))
}
```

A control that fails on the full ambulance cohort and turns null after the
transfer and prior-care restrictions is evidence for the restrictions:
the confounding was removed by design, not by adjustment, and a
negative-control step that runs on one cohort cannot see that.

## Care-process collider check

```{r chk}
if (!is.null(chk)) kable(chk, digits = 4)
```

## The fitted propensity ensembles

A propensity fit that puts substantial weight on the marginal-mean learner
is itself a diagnostic. The fitted SuperLearner weights per lock:

```{r slw}
if (!is.null(slw)) {
  s3 <- reshape(slw, idvar = "learner", timevar = "contrast",
                direction = "wide")
  names(s3) <- sub("^weight\\.", "", names(s3))
  kable(s3, row.names = FALSE, digits = 3)
}
```

## Estimation down the declared ladder

Every lock declared, before outcome access: primary ATE, fallbacks trimmed
ATE, then ATT, then ATO, trigger SEVERE, with a pre-registered
bias-to-null floor of 0.35 percentage points on mortality. Outcomes were
joined only at this stage. Every row carries its estimand, the support
verdict, and the implausibility flags; on C1 the ATE row exists only as a
labelled override for reconciliation against the main pipeline.

```{r est}
if (!is.null(est)) {
  e2 <- est[, c("contrast", "outcome", "estimand", "is_primary", "estimate",
                "ci_lower", "ci_upper", "p_value", "n", "support_verdict",
                "implausible")]
  kable(e2[order(e2$contrast, e2$outcome, e2$estimand), ],
        row.names = FALSE,
        col.names = c("Lock", "Outcome", "Estimand", "Primary", "Estimate",
                      "CI low", "CI high", "p", "n", "Verdict",
                      "Implausible"),
        digits = c(0, 0, 0, 0, 4, 4, 4, 4, 0, 0, 0))
}
```

### Gate path and logged decisions

```{r dlog}
if (!is.null(dlog)) {
  d2 <- unique(dlog[dlog$type %in% c("estimand_switch", "override",
                                     "outcome_access"),
                    c("contrast", "outcome", "type", "note")])
  kable(d2, row.names = FALSE)
}
```

There is no STOP followed by estimation with nothing between: an infeasible
primary is either absent from the estimates or present as an override row
whose reason appears in the log above.

## Reconciliation against the main pipeline

The reconciliation table (`results/reconciliation_table.csv`, regenerated by
`scripts/build_reconciliation_table.R`) compares every clean-room estimate
with the main pipeline's published value.

```{r recon}
if (!is.null(recon)) {
  kable(as.data.frame(table(recon$verdict)),
        col.names = c("Verdict", "Rows"))
  key <- recon[recon$outcome %in% c("death_by_6mo", "time_to_arrival") &
                 is.finite(recon$cr_est) & is.finite(recon$main_est),
               c("contrast", "outcome", "estimand", "cr_est", "main_est",
                 "abs_diff", "verdict")]
  if (nrow(key)) kable(key, row.names = FALSE, digits = 4)
}
```

## Limitations

The support verdicts and the feasibility table are computed from the fitted
propensity, so a PASS is necessary, not sufficient: a propensity model that
misses the covariate that separates the arms will not warn here, which is
why the near-deterministic stratum check and the tree search run beside the
global statistics, and why the negative-control ladder runs beside all of
them. The support map inherits the synthetic outcome family: it spans
confounding, effect modification and curvature along the direction that
drives non-overlap, but a real outcome surface can misbehave along a
direction the family does not span. The proxy survival analysis of the
earlier case study remains strictly secondary (its event times are coarsened
and partly imputed), and two data amendments are pending in every lock's
design log: 76 treated patients without a Rescue.Co ID under review, and a
possible mid-study improvement in Rescue.Co ascertainment.
