---
title: "6-Month Follow-Up Data Quality Check"
subtitle: "RescueCo Kenya Trauma Registry"
author: "RescueCo Analysis Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    theme: cosmo
    self-contained: true
    number-sections: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(haven)
library(dplyr)
library(tidyr)
library(stringr)
library(knitr)
library(ggplot2)
library(labelled)

# ── Locate project root ──
find_project_root <- function() {
  candidates <- c(
    getwd(),
    normalizePath(file.path(getwd(), ".."), mustWork = FALSE),
    normalizePath(file.path(getwd(), "..", ".."), mustWork = FALSE)
  )
  for (d in candidates) {
    if (dir.exists(file.path(d, "data"))) return(normalizePath(d))
  }
  stop("Cannot locate project root")
}
project_root <- find_project_root()

# ── Load 6-month follow-up .dta ──
dta_path <- file.path(
  project_root, "data",
  "6mo_followup_03102026_labeled_notcleaned.dta"
)
stopifnot(file.exists(dta_path))
raw <- read_dta(dta_path)

# ── Convenience: convert haven_labelled to factors for display ──
raw_labelled <- raw  # keep original for label lookups
raw <- raw %>% mutate(across(where(is.labelled), ~ as_factor(.x)))

n_total <- nrow(raw)

# ── Column name vectors ──
# Patient ID
id_col <- "patients_id_gose"

# Hospital
hospital_col <- "hospital"

# Contact attempt date columns (phone + SMS)
phone_date_cols <- c("date_attempt1", "date_attempt2", "date_attempt4",
                     "date_attempt5", "date_attempt7", "date_attempt8",
                     "date_attempt10")
sms_date_cols   <- c("date_attempt_sms", "date_attempt6_sms", "date_attempt9_sms")
# Rename for ordering: attempts 3, 6, 9 are SMS
all_date_cols_named <- c(
  "1" = "date_attempt1",   "2" = "date_attempt2",
  "3" = "date_attempt_sms", "4" = "date_attempt4",
  "5" = "date_attempt5",   "6" = "date_attempt6_sms",
  "7" = "date_attempt7",   "8" = "date_attempt8",
  "9" = "date_attempt9_sms", "10" = "date_attempt10"
)
# Only keep columns that actually exist
all_date_cols_named <- all_date_cols_named[all_date_cols_named %in% names(raw)]
date_col_order <- as.integer(names(all_date_cols_named))

# Contact attempt result columns
phone_result_cols <- c("attempt1", "attempt2", "attempt4", "attempt5",
                       "attempt7", "attempt8", "attempt10")
sms_result_cols   <- c("sms_attempt", "sms_attempt_6", "sms_attempt_9")
all_result_cols_named <- c(
  "1" = "attempt1",    "2" = "attempt2",
  "3" = "sms_attempt", "4" = "attempt4",
  "5" = "attempt5",    "6" = "sms_attempt_6",
  "7" = "attempt7",    "8" = "attempt8",
  "9" = "sms_attempt_9", "10" = "attempt10"
)
all_result_cols_named <- all_result_cols_named[all_result_cols_named %in% names(raw)]

# GOSE / outcome columns
gose_item_cols <- c("consciousness", "independence_home", "need_assistance",
                    "need_assistance_new", "shop", "shop_beforeinj",
                    "travel_locally", "travel_locally_beforeinj",
                    "independence_out_ped", "independence_out_ped_bfinj",
                    "work", "majorchange_work", "work_quantity",
                    "work_hours_lost", "work_money_lost", "money_ksh",
                    "social_activities", "social_restriction",
                    "social_restrictions_new", "change_personality",
                    "freq_change_personality", "change_personality_new",
                    "dailylife", "dailylife_new", "dailylife_factor")
gose_item_cols <- gose_item_cols[gose_item_cols %in% names(raw)]

# Derive hospital prefix from patient ID for cross-check
raw$id_hospital <- str_extract(raw[[id_col]], "^[A-Z]+")

# Ensure date columns are Date class
for (dc in all_date_cols_named) {
  if (!inherits(raw[[dc]], "Date")) {
    raw[[dc]] <- as.Date(raw[[dc]])
  }
}
if ("dod" %in% names(raw) && !inherits(raw[["dod"]], "Date")) {
  raw[["dod"]] <- as.Date(raw[["dod"]])
}

extract_date <- as.Date("2026-03-10")  # date of the .dta export
```

# Overview

This report performs automated data quality checks on the **6-month follow-up** dataset
(`r basename(dta_path)`), exported from REDCap on `r extract_date`.

```{r overview-table}
overview <- data.frame(
  Metric = c("Total records", "Columns", "Unique patient IDs",
             "Hospital sites", "Date columns detected", "GOSE item columns"),
  Value = c(
    format(n_total, big.mark = ","),
    ncol(raw),
    length(unique(raw[[id_col]])),
    paste(sort(unique(na.omit(raw[[hospital_col]]))), collapse = ", "),
    length(all_date_cols_named),
    length(gose_item_cols)
  )
)
kable(overview, caption = "Dataset Overview")
```

```{r site-breakdown}
site_tbl <- raw %>%
  count(!!sym(hospital_col), name = "n") %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))
kable(site_tbl, caption = "Records by Hospital Site")
```

---

# Previously Identified Issues (Re-verification)

Your colleague flagged several categories of issues in the earlier CSV export. This section
**re-verifies** each category against the updated `.dta` file.

## Contact Date Ordering Violations {.tabset}

Sequential contact attempts must have non-decreasing dates. We flag any record where
attempt *N* has a date **after** attempt *N+1*.

```{r date-order-checks}
# Build a tidy date matrix
date_matrix <- raw[, all_date_cols_named, drop = FALSE]
names(date_matrix) <- paste0("attempt_", names(all_date_cols_named))
date_matrix$patient_id <- raw[[id_col]]

# Check consecutive pairs
date_issues <- list()
for (idx in seq_len(length(all_date_cols_named) - 1)) {
  i <- date_col_order[idx]
  j <- date_col_order[idx + 1]
  col_a <- paste0("attempt_", i)
  col_b <- paste0("attempt_", j)
  bad <- date_matrix %>%
    filter(!is.na(.data[[col_a]]) & !is.na(.data[[col_b]]) &
           .data[[col_a]] > .data[[col_b]]) %>%
    select(patient_id, earlier = all_of(col_a), later = all_of(col_b))
  if (nrow(bad) > 0) {
    bad$issue <- paste0("Attempt ", i, " date > Attempt ", j, " date")
    bad$gap_days <- as.numeric(bad$earlier - bad$later)
    date_issues[[paste0(i, "_", j)]] <- bad
  }
}

if (length(date_issues) > 0) {
  all_date_issues <- bind_rows(date_issues)
  kable(
    all_date_issues %>%
      select(patient_id, issue, earlier, later, gap_days) %>%
      arrange(issue, patient_id),
    caption = "Contact dates out of chronological order",
    col.names = c("Patient ID", "Issue", "Earlier Attempt Date",
                  "Later Attempt Date", "Gap (days)")
  )
} else {
  cat("No date ordering issues found.")
}
```

```{r date-order-summary}
if (length(date_issues) > 0) {
  issue_summary <- all_date_issues %>%
    count(issue) %>%
    arrange(issue)
  kable(issue_summary, caption = "Summary: date ordering violations by attempt pair",
        col.names = c("Pair", "Count"))
}
```

---

# NEW Automated Data Quality Checks

The following checks go **beyond** the previously flagged issues to identify additional
impossibilities and anomalies.

## Duplicate Patient IDs

```{r dup-ids}
dup_ids <- raw %>%
  count(!!sym(id_col), name = "n") %>%
  filter(n > 1) %>%
  arrange(desc(n))
names(dup_ids)[1] <- "patient_id"

if (nrow(dup_ids) > 0) {
  kable(dup_ids, caption = "Patient IDs appearing more than once",
        col.names = c("Patient ID", "Count"))
} else {
  cat("No duplicate patient IDs found.")
}
```

## Malformed or Missing Patient IDs

```{r bad-ids}
expected_pattern <- "^(KNH|MGH|MLK)\\d{5}$"
bad_ids <- raw %>%
  filter(!grepl(expected_pattern, .data[[id_col]]) |
         is.na(.data[[id_col]]) | .data[[id_col]] == "") %>%
  select(record_id, !!sym(id_col))

if (nrow(bad_ids) > 0) {
  kable(head(bad_ids, 50),
        caption = paste("Malformed / missing patient IDs (showing up to 50 of",
                        nrow(bad_ids), ")"))
} else {
  cat("All patient IDs match expected format (KNH/MGH/MLKxxxxx).")
}
```

## Implausible Contact Dates

### Dates After the Data Extract (`r extract_date`)

```{r future-dates}
future_flags <- list()
for (nm in names(all_date_cols_named)) {
  dc <- all_date_cols_named[nm]
  future <- raw %>%
    filter(!is.na(.data[[dc]]) & .data[[dc]] > extract_date) %>%
    select(patient_id = !!sym(id_col), date = !!sym(dc)) %>%
    mutate(attempt = as.integer(nm))
  if (nrow(future) > 0) future_flags[[nm]] <- future
}

if (length(future_flags) > 0) {
  all_future <- bind_rows(future_flags)
  kable(all_future %>% arrange(date) %>% select(patient_id, attempt, date),
        caption = "Contact dates after the data extract date")
} else {
  cat("No future dates found.")
}
```

### Dates Suspiciously Early (before 2024-01-01)

```{r early-dates}
early_cutoff <- as.Date("2024-01-01")

early_flags <- list()
for (nm in names(all_date_cols_named)) {
  dc <- all_date_cols_named[nm]
  early <- raw %>%
    filter(!is.na(.data[[dc]]) & .data[[dc]] < early_cutoff) %>%
    select(patient_id = !!sym(id_col), date = !!sym(dc)) %>%
    mutate(attempt = as.integer(nm))
  if (nrow(early) > 0) early_flags[[nm]] <- early
}

if (length(early_flags) > 0) {
  all_early <- bind_rows(early_flags)
  kable(all_early %>% arrange(date),
        caption = paste("Contact dates before", early_cutoff, "(suspiciously early)"))
} else {
  cat("No suspiciously early dates found.")
}
```

### Late Date Values (year 2026+)

These may reflect study delays rather than data errors, but should be verified.

```{r dates-2026}
cutoff_2026 <- as.Date("2026-01-01")

flags_2026 <- list()
for (nm in names(all_date_cols_named)) {
  dc <- all_date_cols_named[nm]
  bad <- raw %>%
    filter(!is.na(.data[[dc]]) & .data[[dc]] >= cutoff_2026) %>%
    select(patient_id = !!sym(id_col), date = !!sym(dc)) %>%
    mutate(attempt = as.integer(nm))
  if (nrow(bad) > 0) flags_2026[[nm]] <- bad
}

if (length(flags_2026) > 0) {
  all_2026 <- bind_rows(flags_2026)
  kable(all_2026 %>% arrange(date),
        caption = "Contact dates in 2026 or later (may reflect study delays -- verify)")
} else {
  cat("No dates in 2026 or later found.")
}
```

## Very Large Gaps Between Consecutive Attempts

Consecutive contact attempts more than **90 days** apart may indicate data entry errors.

```{r large-gaps}
gap_flags <- list()
for (idx in seq_len(length(all_date_cols_named) - 1)) {
  i <- date_col_order[idx]
  j <- date_col_order[idx + 1]
  dc_a <- all_date_cols_named[as.character(i)]
  dc_b <- all_date_cols_named[as.character(j)]
  gaps <- raw %>%
    filter(!is.na(.data[[dc_a]]) & !is.na(.data[[dc_b]])) %>%
    mutate(gap = as.numeric(.data[[dc_b]] - .data[[dc_a]])) %>%
    filter(abs(gap) > 90) %>%
    select(patient_id = !!sym(id_col),
           date_a = !!sym(dc_a), date_b = !!sym(dc_b), gap) %>%
    mutate(pair = paste0("Attempt ", i, " -> ", j))
  if (nrow(gaps) > 0) gap_flags[[idx]] <- gaps
}

if (length(gap_flags) > 0) {
  all_gaps <- bind_rows(gap_flags) %>% arrange(desc(abs(gap)))
  kable(head(all_gaps, 50) %>% select(patient_id, pair, date_a, date_b, gap),
        caption = paste("Gaps > 90 days between consecutive attempts",
                        "(top 50 of", nrow(all_gaps), ")"),
        col.names = c("Patient ID", "Attempt Pair", "Date A", "Date B", "Gap (days)"))
} else {
  cat("No large gaps found.")
}
```

## Same Date for All Attempts (Possible Bulk Entry)

Records where 3+ distinct attempts share the exact same date.

```{r same-date}
date_vals <- raw[, unname(all_date_cols_named), drop = FALSE]
date_vals$patient_id <- raw[[id_col]]

same_date_flags <- date_vals %>%
  rowwise() %>%
  mutate(
    dates_filled = list(na.omit(c_across(-patient_id))),
    n_dates = length(dates_filled),
    n_unique = length(unique(dates_filled))
  ) %>%
  ungroup() %>%
  filter(n_dates >= 3 & n_unique == 1) %>%
  select(patient_id, n_dates, n_unique)

if (nrow(same_date_flags) > 0) {
  kable(same_date_flags,
        caption = "Records where 3+ attempts all share the same date",
        col.names = c("Patient ID", "# Attempts with dates", "# Unique dates"))
} else {
  cat("No records with all-same-date attempts found.")
}
```

## Missing Post-Discharge Timepoint

```{r missing-timepoint}
missing_tp <- raw %>%
  filter(is.na(post_discharge_timepoint) |
         trimws(as.character(post_discharge_timepoint)) == "") %>%
  select(patient_id = !!sym(id_col))

if (nrow(missing_tp) > 0) {
  kable(missing_tp,
        caption = paste("Patients missing post-discharge timepoint (",
                        nrow(missing_tp), "records)"),
        col.names = "Patient ID")
} else {
  cat("All records have a post-discharge timepoint.")
}
```

## Missing Contact Attempt Results

Records with a **date** for an attempt but **no result** (or vice versa).

```{r missing-results}
result_issues <- list()
for (nm in names(all_date_cols_named)) {
  dc <- all_date_cols_named[nm]
  rc <- all_result_cols_named[nm]
  if (is.na(rc) || is.null(rc)) next

  has_date_no_result <- raw %>%
    filter(!is.na(.data[[dc]]) &
           (is.na(.data[[rc]]) | trimws(as.character(.data[[rc]])) == "")) %>%
    select(patient_id = !!sym(id_col)) %>%
    mutate(issue = paste0("Attempt ", nm, ": date but no result"))

  has_result_no_date <- raw %>%
    filter(is.na(.data[[dc]]) &
           !is.na(.data[[rc]]) & trimws(as.character(.data[[rc]])) != "") %>%
    select(patient_id = !!sym(id_col)) %>%
    mutate(issue = paste0("Attempt ", nm, ": result but no date"))

  combined <- bind_rows(has_date_no_result, has_result_no_date)
  if (nrow(combined) > 0) result_issues[[nm]] <- combined
}

if (length(result_issues) > 0) {
  all_result <- bind_rows(result_issues)
  summary_result <- all_result %>% count(issue) %>% arrange(issue)
  kable(summary_result,
        caption = "Date / result mismatch summary by attempt",
        col.names = c("Issue", "Count"))

  kable(all_result %>% arrange(issue, patient_id) %>% head(100),
        caption = "Date / result mismatches (up to 100 rows)")
} else {
  cat("All attempts with dates also have results, and vice versa.")
}
```

## GOSE / Functional Outcome Data Quality

### Patient Reported Alive but Has Death Date (or vice versa)

```{r alive-death-conflict}
# Alive = Yes but death date present
alive_but_dead <- raw %>%
  filter(as.character(patient_alive) == "Yes" & !is.na(dod)) %>%
  select(patient_id = !!sym(id_col), patient_alive, dod)

# Dead but no death date
dead_no_date <- raw %>%
  filter(as.character(patient_alive) == "No" & is.na(dod)) %>%
  select(patient_id = !!sym(id_col), patient_alive)

if (nrow(alive_but_dead) > 0) {
  kable(alive_but_dead,
        caption = "Patient marked ALIVE but has a death date recorded")
}
if (nrow(dead_no_date) > 0) {
  kable(dead_no_date,
        caption = "Patient marked NOT ALIVE but missing death date")
}
if (nrow(alive_but_dead) == 0 & nrow(dead_no_date) == 0) {
  cat("No alive/death status conflicts found.")
}
```

### Death Date Implausibility

```{r death-date-check}
death_sub <- raw %>% filter(!is.na(dod))

# Death in future
future_death <- death_sub %>%
  filter(dod > extract_date) %>%
  select(patient_id = !!sym(id_col), dod)

# Death before first contact attempt
first_dc <- all_date_cols_named["1"]
death_before_contact <- death_sub %>%
  filter(!is.na(.data[[first_dc]]) & dod < .data[[first_dc]]) %>%
  select(patient_id = !!sym(id_col), dod, first_contact = !!sym(first_dc))

if (nrow(future_death) > 0)
  kable(future_death, caption = "Death dates after data extract date")
if (nrow(death_before_contact) > 0)
  kable(death_before_contact,
        caption = "Death date recorded before first contact attempt date")
if (nrow(future_death) == 0 & nrow(death_before_contact) == 0)
  cat("No death date implausibilities found.")
```

### Missing GOSE Responses for Successfully Contacted Patients

Patients who were **successfully contacted** (answered by patient or relative) but have
**no GOSE item responses**. We also investigate *why* GOSE may be missing --- possible
reasons include consent refusal, patient reported dead, or notes indicating unsuccessful
follow-up.

```{r missing-gose}
# Determine who was successfully contacted
success_vals <- c("Answered by patient", "Answered by relative",
                  "Answered by caretaker")

phone_results <- all_result_cols_named[all_result_cols_named %in%
                                        c("attempt1","attempt2","attempt4",
                                          "attempt5","attempt7","attempt8","attempt10")]

contacted <- raw %>%
  rowwise() %>%
  mutate(
    was_contacted = any(
      sapply(phone_results, function(rc) {
        as.character(.data[[rc]]) %in% success_vals
      }), na.rm = TRUE
    )
  ) %>%
  ungroup() %>%
  filter(was_contacted)

# Check if all GOSE items are blank / NA
gose_missing <- contacted %>%
  rowwise() %>%
  mutate(
    n_gose_answered = sum(
      sapply(gose_item_cols, function(gc) {
        val <- .data[[gc]]
        !is.na(val) & trimws(as.character(val)) != ""
      })
    )
  ) %>%
  ungroup() %>%
  filter(n_gose_answered == 0) %>%
  select(patient_id = !!sym(id_col))

if (nrow(gose_missing) > 0) {
  kable(gose_missing,
        caption = paste("Successfully contacted but ALL GOSE items blank (",
                        nrow(gose_missing), "patients)"),
        col.names = "Patient ID")
} else {
  cat("All successfully contacted patients have at least some GOSE data.")
}
```

#### Why are GOSE responses missing?

For the patients above, we check consent status, alive/dead status, respondent type,
and the `notes` field to understand whether GOSE was intentionally skipped (e.g.,
patient died, consent refused) or is a true data gap.

```{r gose-missing-reasons}
if (nrow(gose_missing) > 0) {
  gose_miss_detail <- raw %>%
    filter(.data[[id_col]] %in% gose_missing$patient_id) %>%
    mutate(
      consent   = as.character(consent_agose),
      alive     = as.character(patient_alive),
      resp      = as.character(respondent),
      notes_val = if ("notes" %in% names(.)) trimws(notes) else ""
    ) %>%
    select(patient_id = !!sym(id_col), consent, alive, resp, notes_val)

  # Classify reason
  gose_miss_detail <- gose_miss_detail %>%
    mutate(
      likely_reason = case_when(
        alive == "No"                          ~ "Patient dead (GOSE = 1, not survey items)",
        consent == "No"                        ~ "Consent refused",
        grepl("UNSUCCESSFUL|DISCONNECTED|WRONG|DEAD|DECEASED|LOST|UNREACHABLE",
              notes_val, ignore.case = TRUE)   ~ "Notes indicate unsuccessful / unreachable",
        grepl("NOT TRAUMA|NON.TRAUMA|MEDICAL|SICK",
              notes_val, ignore.case = TRUE)   ~ "Not a trauma case (survey not applicable)",
        is.na(consent) | consent == ""         ~ "Consent field blank (survey may not have started)",
        TRUE                                   ~ "Unknown -- needs manual review"
      )
    )

  reason_summary <- gose_miss_detail %>%
    count(likely_reason, name = "n") %>%
    arrange(desc(n))

  kable(reason_summary,
        caption = "Likely reason for missing GOSE among contacted patients",
        col.names = c("Likely Reason", "Count"))

  # Show detail for "Unknown" cases that need review
  unknown_cases <- gose_miss_detail %>%
    filter(likely_reason == "Unknown -- needs manual review")
  if (nrow(unknown_cases) > 0) {
    kable(head(unknown_cases, 50),
          caption = paste("Cases needing manual review (up to 50 of",
                          nrow(unknown_cases), ")"),
          col.names = c("Patient ID", "Consent", "Alive", "Respondent",
                        "Notes", "Likely Reason"))
  }
}
```

## Consent Issues

### Contacted but No Consent Recorded

```{r no-consent}
no_consent <- contacted %>%
  filter(is.na(consent_agose) |
         trimws(as.character(consent_agose)) == "") %>%
  select(patient_id = !!sym(id_col))

if (nrow(no_consent) > 0) {
  kable(no_consent,
        caption = paste("Successfully contacted but consent not recorded (",
                        nrow(no_consent), "patients)"),
        col.names = "Patient ID")
} else {
  cat("All contacted patients have consent recorded.")
}
```

### Consent Refused but GOSE Data Present

```{r refused-but-data}
refused_with_data <- raw %>%
  filter(as.character(consent_agose) == "No") %>%
  rowwise() %>%
  mutate(
    n_gose = sum(
      sapply(gose_item_cols, function(gc) {
        val <- .data[[gc]]
        !is.na(val) & trimws(as.character(val)) != ""
      })
    )
  ) %>%
  ungroup() %>%
  filter(n_gose > 0) %>%
  select(patient_id = !!sym(id_col), n_gose)

if (nrow(refused_with_data) > 0) {
  kable(refused_with_data,
        caption = "Consent REFUSED but GOSE data was entered",
        col.names = c("Patient ID", "# GOSE items answered"))
} else {
  cat("No consent-refused patients have GOSE data.")
}
```

## Numeric / Free-Text Field Anomalies

### Lost Work Days — Extreme Values

```{r lost-workdays}
if ("work_hours_lost" %in% names(raw)) {
  whl <- suppressWarnings(as.numeric(raw$work_hours_lost))
  extreme_wh <- raw %>%
    mutate(.val = suppressWarnings(as.numeric(work_hours_lost))) %>%
    filter(!is.na(.val) & (.val > 180 | .val < 0)) %>%
    select(patient_id = !!sym(id_col), work_hours_lost = .val)

  if (nrow(extreme_wh) > 0) {
    kable(extreme_wh,
          caption = "Extreme values in work_hours_lost (>180 or <0 days in 30-day window)")
  } else {
    cat("No extreme work-hours-lost values found.")
  }
}
```

### Financial Loss — Extreme Values (KSh)

```{r financial-extreme}
if ("money_ksh" %in% names(raw)) {
  extreme_fin <- raw %>%
    mutate(.val = suppressWarnings(as.numeric(gsub("[^0-9.-]", "", money_ksh)))) %>%
    filter(!is.na(.val) & (.val > 1000000 | .val < 0)) %>%
    select(patient_id = !!sym(id_col), money_ksh, parsed = .val) %>%
    arrange(desc(parsed))

  if (nrow(extreme_fin) > 0) {
    kable(head(extreme_fin, 30),
          caption = "Extreme financial values in money_ksh (>1M KSh or negative)")
  } else {
    cat("No extreme financial values found.")
  }
}
```

## Completion / Notes Field Consistency

### Records Marked "Complete" but Missing Alive/Dead Status

```{r complete-but-missing}
complete_records <- raw %>%
  filter(as.character(patient_follow_up_au_v_1) == "Complete")

complete_no_alive <- complete_records %>%
  filter(is.na(patient_alive) |
         trimws(as.character(patient_alive)) == "") %>%
  select(patient_id = !!sym(id_col))

if (nrow(complete_no_alive) > 0) {
  kable(complete_no_alive,
        caption = paste("Marked COMPLETE but missing alive/dead status (",
                        nrow(complete_no_alive), "records)"),
        col.names = "Patient ID")
} else {
  cat("All 'complete' records have alive/dead status recorded.")
}
```

### Notes Field Distribution

```{r notes-dist}
if ("notes" %in% names(raw)) {
  notes_tbl <- raw %>%
    mutate(notes_clean = trimws(notes)) %>%
    filter(notes_clean != "") %>%
    count(notes_clean, name = "n") %>%
    arrange(desc(n)) %>%
    mutate(pct = round(100 * n / sum(n), 1))
  kable(head(notes_tbl, 20),
        caption = "Notes field values (top 20)",
        col.names = c("Notes Value", "Count", "%"))
}
```

### Never Contacted (No Attempt Dates at All)

```{r never-contacted}
date_vals <- raw[, unname(all_date_cols_named), drop = FALSE]
never_contacted <- raw %>%
  mutate(any_date = rowSums(!is.na(date_vals)) > 0) %>%
  filter(!any_date) %>%
  count(!!sym(hospital_col), name = "n_no_attempts")

if (nrow(never_contacted) > 0 && sum(never_contacted$n_no_attempts) > 0) {
  kable(never_contacted,
        caption = "Records with NO contact attempt dates at all")
} else {
  cat("All records have at least one contact attempt date.")
}
```

## Cross-Hospital ID Validation

```{r site-id-mismatch}
# Map hospital column values to expected prefixes
mismatch <- raw %>%
  mutate(
    id_prefix = str_extract(.data[[id_col]], "^[A-Z]+"),
    expected_prefix = case_when(
      grepl("KNH|Kenyatta", .data[[hospital_col]], ignore.case = TRUE) ~ "KNH",
      grepl("MCH|Mbagathi", .data[[hospital_col]], ignore.case = TRUE) ~ "MGH",
      grepl("MKH|Mama Lucy|MLK", .data[[hospital_col]], ignore.case = TRUE) ~ "MLK",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(expected_prefix) & id_prefix != expected_prefix) %>%
  select(patient_id = !!sym(id_col), id_prefix, hospital = !!sym(hospital_col),
         expected_prefix)

if (nrow(mismatch) > 0) {
  kable(mismatch,
        caption = "Patient ID prefix does not match hospital site")
} else {
  cat("All patient ID prefixes match their hospital site.")
}
```

## Multiple Contacts on Same Day Across Different Attempt Slots

Repeated dates across non-consecutive slots may indicate copy-paste errors.

```{r repeat-dates}
date_vals2 <- raw[, unname(all_date_cols_named), drop = FALSE]
date_vals2$patient_id <- raw[[id_col]]

repeat_date_flags <- date_vals2 %>%
  rowwise() %>%
  mutate(
    all_dates = list(na.omit(c_across(-patient_id))),
    n_filled = length(all_dates),
    n_unique = length(unique(all_dates)),
    n_repeated = n_filled - n_unique
  ) %>%
  ungroup() %>%
  filter(n_repeated > 0 & n_filled >= 2) %>%
  select(patient_id, n_filled, n_unique, n_repeated) %>%
  arrange(desc(n_repeated))

if (nrow(repeat_date_flags) > 0) {
  kable(head(repeat_date_flags, 50),
        caption = paste("Repeated dates across attempt slots",
                        "(top 50 of", nrow(repeat_date_flags), ")"),
        col.names = c("Patient ID", "# Dates Filled", "# Unique", "# Repeated"))
} else {
  cat("No repeated dates found across attempt slots.")
}
```

---

# Contact Attempt Pattern Summary

```{r attempt-patterns, fig.width=8, fig.height=4}
raw$n_attempts <- rowSums(!is.na(raw[, unname(all_date_cols_named), drop = FALSE]))

attempt_dist <- raw %>% count(n_attempts) %>% mutate(pct = round(100 * n / sum(n), 1))

ggplot(attempt_dist, aes(x = factor(n_attempts), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), vjust = -0.3, size = 3) +
  labs(x = "Number of Contact Attempts (with dates)",
       y = "Number of Patients",
       title = "Distribution of Contact Attempts per Patient") +
  theme_minimal()
```

```{r attempt-result-dist}
result_long <- raw %>%
  select(patient_id = !!sym(id_col), all_of(unname(all_result_cols_named))) %>%
  pivot_longer(-patient_id, names_to = "attempt", values_to = "result") %>%
  mutate(result = as.character(result)) %>%
  filter(!is.na(result) & trimws(result) != "")

if (nrow(result_long) > 0) {
  result_counts <- result_long %>%
    count(result) %>%
    arrange(desc(n)) %>%
    mutate(pct = round(100 * n / sum(n), 1))
  kable(result_counts,
        caption = "Contact attempt results across all attempts",
        col.names = c("Result", "Count", "%"))
}
```

---

# Overall Missingness Heatmap

```{r missingness-heatmap, fig.width=12, fig.height=5}
key_cols <- c(unname(all_date_cols_named), unname(all_result_cols_named))
key_cols <- key_cols[key_cols %in% names(raw)]

miss_by_site <- raw %>%
  group_by(!!sym(hospital_col)) %>%
  summarise(
    across(all_of(key_cols),
           ~ round(100 * mean(is.na(.) | trimws(as.character(.)) == ""), 1)),
    .groups = "drop"
  )

miss_long <- miss_by_site %>%
  pivot_longer(-!!sym(hospital_col), names_to = "column", values_to = "pct_missing")

ggplot(miss_long, aes(x = column, y = !!sym(hospital_col), fill = pct_missing)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "white", mid = "orange", high = "red",
                       midpoint = 50, limits = c(0, 100), name = "% Missing") +
  labs(x = NULL, y = "Hospital",
       title = "Missingness by Hospital and Contact Attempt Column") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```

---

# Summary of All Flagged Issues

```{r final-summary}
all_flagged <- character(0)
if (length(date_issues) > 0) all_flagged <- c(all_flagged, all_date_issues$patient_id)
if (length(future_flags) > 0) all_flagged <- c(all_flagged, bind_rows(future_flags)$patient_id)
if (length(early_flags) > 0) all_flagged <- c(all_flagged, bind_rows(early_flags)$patient_id)
if (nrow(dup_ids) > 0) all_flagged <- c(all_flagged, dup_ids$patient_id)
if (exists("gose_missing") && nrow(gose_missing) > 0)
  all_flagged <- c(all_flagged, gose_missing$patient_id)
if (exists("no_consent") && nrow(no_consent) > 0)
  all_flagged <- c(all_flagged, no_consent$patient_id)

all_flagged <- sort(unique(all_flagged))

summary_tbl <- data.frame(
  Check = c(
    "Date ordering violations",
    "Future dates (after extract)",
    "Suspiciously early dates (pre-2024)",
    "Dates in 2026+",
    "Duplicate patient IDs",
    "Large gaps (>90 days) between attempts",
    "Contacted but all GOSE blank",
    "Contacted but no consent recorded",
    "Consent refused but GOSE data entered",
    "Records with repeated dates",
    "Alive/death status conflicts",
    "Complete but missing alive/dead",
    "TOTAL unique patients flagged"
  ),
  Count = c(
    if (length(date_issues) > 0) length(unique(all_date_issues$patient_id)) else 0,
    if (length(future_flags) > 0) length(unique(bind_rows(future_flags)$patient_id)) else 0,
    if (length(early_flags) > 0) length(unique(bind_rows(early_flags)$patient_id)) else 0,
    if (length(flags_2026) > 0) length(unique(bind_rows(flags_2026)$patient_id)) else 0,
    nrow(dup_ids),
    if (length(gap_flags) > 0) length(unique(bind_rows(gap_flags)$patient_id)) else 0,
    if (exists("gose_missing")) nrow(gose_missing) else 0,
    if (exists("no_consent")) nrow(no_consent) else 0,
    if (exists("refused_with_data")) nrow(refused_with_data) else 0,
    if (exists("repeat_date_flags")) nrow(repeat_date_flags) else 0,
    nrow(alive_but_dead) + nrow(dead_no_date),
    nrow(complete_no_alive),
    length(all_flagged)
  )
)

kable(summary_tbl, caption = "Summary of Data Quality Issues Found",
      col.names = c("Check Category", "# Patients Flagged"))
```

```{r flagged-by-site}
if (length(all_flagged) > 0) {
  flagged_site <- raw %>%
    filter(.data[[id_col]] %in% all_flagged) %>%
    count(!!sym(hospital_col), name = "n_flagged") %>%
    left_join(site_tbl, by = hospital_col) %>%
    mutate(pct_flagged = round(100 * n_flagged / n, 1)) %>%
    select(!!sym(hospital_col), n_flagged, n_total = n, pct_flagged)
  kable(flagged_site,
        caption = "Flagged patients by hospital",
        col.names = c("Hospital", "# Flagged", "# Total", "% Flagged"))
}
```

---

# Variable Cleaning & Key Decisions

This section documents data cleaning decisions and analytical choices made during cohort construction that were not pre-specified in the original analysis plan. These are recorded here for transparency and to support reproducibility.

## Inter-Facility Transfer Discovery and Exclusion

During exploratory analysis, the injury-to-arrival time variable showed Rescue.Co ambulance patients arriving ~341 minutes faster than other ambulance patients — a difference far too large to be plausible for a transport quality effect.

**Root cause:** 40.7% of non-Rescue.Co ambulance patients were inter-facility transfers, compared to only 2.5% of Rescue.Co patients. Transfer patients have much longer "injury-to-arrival" times because the clock includes time spent at the referring facility, not just transport time. This created a severe compositional confounder.

**Decision:**

- **Primary analysis:** Exclude all inter-facility transfers, restricting to direct-from-scene ambulance patients only. This ensures a fair comparison of ambulance transport quality.
- **Sensitivity analysis:** Retain all ambulance patients with a binary `is_transfer` covariate added to the adjustment set.
- **Extreme outliers:** Patients with injury-to-arrival time > 24 hours (1440 minutes) are excluded as likely data entry errors.

**Config:** `treatment$exclude_transfers: true` (primary); set to `false` for sensitivity.

## GOSE Derivation from Individual Survey Items

The Glasgow Outcome Scale – Extended (GOSE) score is not directly recorded in the 6-month follow-up data. Instead, it is derived algorithmically from individual adapted GOSE (aGOSE) survey items using an 8-level branching algorithm:

| Level | aGOSE Category | Score |
|-------|----------------|-------|
| Dead | Patient confirmed dead | 1 |
| Vegetative State | VS items endorsed | 2 |
| Lower Severe Disability | Requires daily assistance | 3 |
| Upper Severe Disability | Can be left alone 8+ hours | 4 |
| Lower Moderate Disability | Cannot return to work/social | 5 |
| Upper Moderate Disability | Reduced capacity but participates | 6 |
| Lower Good Recovery | Some residual problems | 7 |
| Upper Good Recovery | No problems or minor only | 8 |

**Key assumptions:**

1. Patients confirmed dead at follow-up are assigned GOSE = 1 (not derived from survey items)
2. Patients contacted but missing all GOSE survey items are treated as missing outcomes (not imputed), unless a reason is documented (e.g., non-trauma case, consent refused)
3. Proxy respondents (family members) may introduce measurement error vs. patient self-report
4. The 6-month timing window varies across patients due to study delays

## Missing Outcome Handling

Of ~600 patients successfully contacted at 6-month follow-up, approximately 300+ are missing all GOSE question responses. Investigation of this missingness revealed several documented reasons:

- **Patient dead:** GOSE = 1 assigned directly (not from survey items)
- **Consent refused:** Patient declined participation
- **Unsuccessful contact:** Phone disconnected, wrong number, or patient unreachable despite multiple attempts
- **Not a trauma case:** Some patients were flagged as non-trauma/medical cases
- **Unknown:** ~50-100 patients with no documented reason — flagged for manual review

This loss to follow-up is likely **informative** (sicker patients harder to reach), which may bias the functional outcome analysis. The survival analysis is less affected because death events are captured from hospital disposition records independent of follow-up.

## Survival Time Construction

Survival time is constructed from multiple data sources with the following priority:

1. **Died in hospital:** Time = max(1, days from injury to disposition date); Event = 1
2. **Died after discharge (date of death known):** Time = days from injury to DOD from 6mo follow-up; Event = 1
3. **Died after discharge (date unknown):** Time = midpoint between discharge and follow-up contact; Event = 1
4. **Alive at follow-up:** Time = days from injury to follow-up contact; Event = 0
5. **Lost to follow-up:** Time = 30 days (conservative censoring); Event = 0

**Key assumptions:** Day-0 deaths are set to day 1 (for log-transform compatibility). The 30-day censoring for lost patients is conservative — true follow-up time is unknown.

## Near-Zero-Variance Covariate Filtering

After imputation, `caret::nearZeroVar()` is applied to remove binary covariates with extremely low prevalence (e.g., `chronic_diabetes_insulin` at 0.09%, `chronic_heart_disease` at 0.05%). These near-constant columns cause:

- SuperLearner instability (perfect separation in logistic models)
- Inflated propensity score variance
- Unreliable coefficient estimates

Removed covariates are logged and documented in the covariate dictionary. This was not in the original analysis plan but is standard practice for TMLE/SuperLearner workflows.

## Other Cleaning Decisions

- **Non-ambulance patients excluded:** The analysis compares Rescue.Co EMS vs. other ambulance services only. Patients who did not arrive by ambulance (`rcvamb == NA`) are excluded before cohort construction.
- **Labelled Stata values:** All labelled doubles from .dta files are converted via `haven::as_factor()` before creating indicator variables, ensuring correct category mapping.
- **Duplicate follow-up records:** De-duplicated by patient ID, keeping the first occurrence.
- **Median imputation:** Applied to continuous covariates with missing values; categorical missingness handled by a "Missing" category where applicable (per config: `missingness_threshold: 0.40`).

---

::: {.callout-note}
## Next Steps for Data Collection Team

1. **Date corrections in REDCap**: Fix all chronologically misordered contact dates
2. **Missing results**: Fill in contact attempt outcomes for flagged IDs
3. **GOSE completion**: Ensure all successfully contacted patients have GOSE data entered
4. **Consent documentation**: Record verbal consent status for all contacted patients
5. **Impossible dates**: Verify and correct any dates in 2026 or before 2024
6. **Duplicate IDs**: Resolve records sharing the same patient ID
7. **Alive/death conflicts**: Reconcile alive status vs. death date discrepancies
:::
