---
title: "Clean-Room Staged Analysis: HCV-AKI Case Study"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  warning: false
  message: false
---

## Overview

This vignette demonstrates the end-to-end clean-room staged workflow for
analyzing the causal effect of sofosbuvir-containing (SOF) vs non-SOF DAA
regimens on acute kidney injury (AKI) risk using a simulated cohort.

The analysis proceeds through four stages with explicit checkpoints:

1. **Stage 1**: Cohort feasibility and build (outcome-blind)
2. **Stage 2**: Design adequacy checks (treatment model only; no outcomes)
3. **Stage 3**: Outcome modeling and estimation (TMLE + comparators)
4. **Stage 4**: Reporting and lock

**Estimand**: The primary estimand is the cumulative incidence (risk) at 90 and
180 days under hypothetical treatment vs control, operationalized as binary
endpoints $Y_t = I(T \le t \text{ and event} = 1)$. The risk difference (RD) and
risk ratio (RR) do not rely on proportional hazards assumptions. A secondary
estimand is the Cox PH hazard ratio (which does assume PH). Inference uses
nonparametric bootstrap CIs as the primary method, with IC-based SEs reported
for TMLE and IPTW. Bootstrap is preferred in this vignette because with
realistic event rates (2--4%) and finite samples, IC-based SEs can be unstable.

```{r setup}
#| label: setup

# Source all R modules (for helper functions used in rendering)
r_files <- list.files("../R", pattern = "\\.R$", recursive = TRUE,
                      full.names = TRUE)
for (f in r_files) source(f)

# Load configuration
cfg <- load_config("../config/default.yml")

# --- Check for precomputed outputs ---
rds_files <- c(
  stage1 = "../outputs/stage1_feasibility.rds",
  stage2 = "../outputs/stage2_design_adequacy.rds",
  stage3 = "../outputs/stage3_estimation.rds"
)
missing <- names(rds_files)[!file.exists(rds_files)]
if (length(missing) > 0) {
  stop("Precomputed outputs not found: ",
       paste(rds_files[missing], collapse = ", "),
       "\n\nRun the pipeline first:\n",
       "  Rscript scripts/00_run_all_stages.R\n",
       "or:  make run\n",
       call. = FALSE)
}

stage1_out <- readRDS(rds_files["stage1"])
stage2_out <- readRDS(rds_files["stage2"])
stage3_out <- readRDS(rds_files["stage3"])
```

## Stage 1: Cohort Feasibility and Build

Stage 1 generates the analysis cohort and produces feasibility summaries.
**No outcome-by-treatment comparisons are permitted at this stage.**

Checkpoint criteria: N $\ge$ `r cfg$stage1$min_N`, treated fraction $\ge$
`r cfg$stage1$min_treated_frac`, events $\ge$ `r cfg$stage1$min_events`,
missingness $\le$ `r cfg$stage1$max_missingness`.

```{r stage1-report}
#| label: stage1-report

report <- stage1_out$report
cat(sprintf("Checkpoint 1: %s\n\n", stage1_out$checkpoint))
cat(sprintf("Cohort size: %d\n", report$N_total))
cat(sprintf("Treated (SOF): %d (%.1f%%)\n",
            report$N_treated, report$frac_treated * 100))
cat(sprintf("Control: %d (%.1f%%)\n",
            report$N_control, report$frac_control * 100))
cat(sprintf("Overall event rate: %.4f\n", report$overall_event_rate))
cat(sprintf("Total events: %d\n", report$overall_N_events))
cat(sprintf("Number of covariates: %d\n", report$n_covariates))
```

## Stage 2: Design Adequacy Checks

Stage 2 evaluates propensity score overlap, covariate balance, and weight
distributions. **This stage does NOT fit outcome models or compute treatment
effects.**

Checkpoint criteria: max weighted SMD $\le$ `r cfg$stage2$max_smd`,
ESS fraction $\ge$ `r cfg$stage2$min_ess_frac`,
PS overlap fraction $\ge$ `r cfg$stage2$min_ps_overlap`.

```{r stage2-report}
#| label: stage2-report

cat(sprintf("Checkpoint 2: %s\n", stage2_out$checkpoint))
```

### Propensity Score Overlap

```{r ps-overlap}
#| label: ps-overlap
#| fig-width: 8
#| fig-height: 5

library(ggplot2)

ps_df <- data.frame(
  ps = stage2_out$ps,
  treatment = factor(stage1_out$cohort$treatment,
                     labels = c("Control", "SOF"))
)

ggplot(ps_df, aes(x = ps, fill = treatment)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = c(cfg$tmle$truncation_lower,
                             cfg$tmle$truncation_upper),
             linetype = "dashed", color = "red") +
  labs(title = "Propensity Score Overlap",
       x = "P(SOF | W)", y = "Density", fill = "Treatment") +
  theme_minimal()
```

### Covariate Balance (SMD Love Plot)

```{r smd-love-plot}
#| label: smd-love-plot
#| fig-width: 8
#| fig-height: 7

covar_cols <- stage2_out$covar_cols
A <- stage1_out$cohort$treatment

smd_unw <- sapply(covar_cols, function(v) {
  compute_smd(stage1_out$cohort[[v]], A)
})
smd_wtd <- sapply(covar_cols, function(v) {
  compute_smd(stage1_out$cohort[[v]], A, weights = stage2_out$weights)
})

smd_df <- data.frame(
  variable = rep(covar_cols, 2),
  smd = c(smd_unw, smd_wtd),
  type = rep(c("Unweighted", "Weighted"), each = length(covar_cols))
)

ggplot(smd_df, aes(x = abs(smd), y = reorder(variable, abs(smd)),
                   color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed") +
  geom_vline(xintercept = cfg$stage2$max_smd,
             linetype = "dotted", color = "red") +
  labs(title = "Covariate Balance (SMD Love Plot)",
       x = "|SMD|", y = "Covariate", color = "Weighting") +
  theme_minimal()
```

### Diagnostics Summary

```{r diag-summary}
#| label: diag-summary

diag <- stage2_out$diagnostics
cat(sprintf("Max weighted SMD: %.4f (threshold: %.2f)\n",
            diag$max_smd_weighted, cfg$stage2$max_smd))
cat(sprintf("ESS fraction: %.4f (threshold: %.2f)\n",
            diag$ess_frac, cfg$stage2$min_ess_frac))
cat(sprintf("PS overlap fraction: %.4f\n", diag$ps_overlap_frac))
cat(sprintf("PS truncated: %d lower, %d upper\n",
            diag$n_truncated_lower, diag$n_truncated_upper))
```

## Stage 3: Outcome Modeling and Estimation

Stage 3 runs TMLE and comparator estimators. It only proceeds if Stage 2
checkpoint is PASS.

**Primary estimand**: Cumulative risk at 90 and 180 days (risk difference
and risk ratio). The estimand is $Y_t = I(T \le t \text{ and event}=1)$;
censoring is handled via IPCW in TMLE and IPTW. This estimand does NOT
rely on proportional hazards.

**Secondary estimand**: Cox PH hazard ratio (labeled as such).

**Inference**: Bootstrap CIs (percentile method) are the primary inference
for all three estimators. IC-based SEs are also reported for TMLE and IPTW.

```{r stage3-table}
#| label: stage3-table

knitr::kable(stage3_out$summary, digits = 5,
             caption = "Stage 3: Effect Estimates by Method and Time Point")
```

### Cox PH Secondary Estimand

```{r cox-secondary}
#| label: cox-secondary

cox_hr <- stage3_out$results$cox_hr
if (!is.null(cox_hr) && !is.na(cox_hr$HR)) {
  cat(sprintf("Hazard Ratio: %.4f (95%% CI: %.4f, %.4f)\n",
              cox_hr$HR, cox_hr$ci_HR[1], cox_hr$ci_HR[2]))
  cat(sprintf("PH Test p-value: %.4f\n", cox_hr$ph_test$pvalue))
  cat(sprintf("PH Violated (p < 0.05): %s\n", cox_hr$ph_test$ph_violated))
  cat("\nNote: The HR estimate assumes proportional hazards. ",
      "When PH is violated, interpret with caution.\n")
} else {
  cat("Cox PH estimation did not produce valid results.\n")
}
```

### TMLE Diagnostics (180 days)

```{r tmle-diag}
#| label: tmle-diag

tmle_180 <- stage3_out$results$t180$tmle
diag_180 <- tmle_180$diagnostics
has_full_diag <- !is.null(diag_180) && is.null(diag_180$error) &&
  is.numeric(diag_180$clever_covariate$H1_mean)
if (has_full_diag) {
  cat("Clever covariate H1 mean:", round(diag_180$clever_covariate$H1_mean, 4), "\n")
  cat("IC mean (should be ~0):", round(diag_180$ic$ic_rd_mean, 6), "\n")
  cat("Positivity: min g =", round(diag_180$positivity$min_g1W, 4),
      ", max g =", round(diag_180$positivity$max_g1W, 4), "\n")
  cat("Observations censored:", diag_180$n_censored, "\n")
  cat("Fluctuation eps:", round(as.numeric(diag_180$eps$eps1), 4),
      ",", round(as.numeric(diag_180$eps$eps0), 4), "\n")

  # Positivity warnings
  frac_boundary <- diag_180$positivity$frac_near_boundary
  if (!is.null(frac_boundary) && frac_boundary > 0.05) {
    cat("\nWARNING: ", round(frac_boundary * 100, 1),
        "% of observations have g near 0/1 boundary.\n")
  }
} else {
  msg <- if (!is.null(diag_180$error)) diag_180$error else "diagnostics fields missing or non-numeric"
  cat("TMLE diagnostics unavailable:", msg, "\n")
}
```

## Stage 4: Report Lock

```{r stage4}
#| label: stage4

# Fix relative paths when running from analysis/ subdirectory
cfg_local <- cfg
cfg_local$output$base_dir   <- "../outputs"
cfg_local$output$stage1_dir <- "../outputs/stage1"
cfg_local$output$stage2_dir <- "../outputs/stage2"
cfg_local$output$stage3_dir <- "../outputs/stage3"
stage4_report(cfg = cfg_local, output_dir = "../outputs/report")
```

## Decision Log

```{r decision-log}
#| label: decision-log

if (file.exists("../outputs/decision_log.csv")) {
  log <- read.csv("../outputs/decision_log.csv", stringsAsFactors = FALSE)
  knitr::kable(log[, c("stage", "model_component", "decision",
                        "outcome_blind_confirm")],
               caption = "Decision Log Excerpt")
}
```

## Limitations

This analysis uses simulated data with illustrative parameters. The DGP is
calibrated to produce realistic event rates (~2--4% at 180 days) and
confounding structure, but in real studies additional data-quality checks,
measurement-validity assessment, and sensitivity analyses (e.g., for
unmeasured confounding, informative censoring models, and positivity
violations) are required. The SL library is intentionally minimal
(GLM + mean) for computational speed; production analyses should include
flexible learners (GAM, random forest, elastic net). The staged clean-room
design supports pre-specification of the treatment model and estimand before
any outcome data are examined.

## Appendix

### Configuration Used

```{r config}
#| label: config

str(cfg, max.level = 2)
```

### Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```
