---
title: "Targeted Learning in Clean-Room Design Analyses"
subtitle: "Comparing Conventional PS, Fixed-Specification TMLE, and Simulation-Selected TMLEs"
author: "Andrew Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

## Abstract {.unnumbered}

A clean-room staged workflow — where analytic decisions are locked before
outcome data are examined — can accommodate multiple analysis strategies.
Adapting the staging and clean-room constructs described by Muntner et al.
(2024), who propose sequential preliminary analyses with checkpoint decisions
before the final comparative analysis, this tutorial introduces a clean-room
design with **two alternative blinded feasibility paths** and explicitly
compares three analysis workflows that differ in how the analysis pipeline is
validated before outcome access: **(1) a
conventional PS-diagnostics-gated workflow** using logistic-regression PS for
matching or IPTW, gated primarily by traditional design diagnostics; **(2) a
fixed-specification TMLE workflow** where a single TMLE nuisance-learning
strategy is locked and validated via plasmode simulation; and **(3) a
simulation-selected TMLE workflow** where a prespecified candidate set of TMLE
specifications is evaluated in plasmode simulation and the best-performing
specification is selected using a locked rule — all before real outcomes are
accessed. In **Stage 1**, the analysis team pre-specifies and locks the
estimand, covariate set, nuisance-learning strategies, truncation rules, and
performance thresholds. The workflow then branches into two alternative
pre-unblinding feasibility paths. In **Stage 2a** (traditional feasibility
diagnostics), the team accesses only treatment and covariate data to estimate
propensity scores, evaluate overlap, and assess matching feasibility — without
touching outcomes. The conventional PS workflow is gated here. Alternatively, in
**Stage 2b** (plasmode-simulation feasibility evaluation), the team runs the
analysis pipelines on synthetic data generated from plasmode-style simulations —
where the covariate and treatment structure come from the observed blinded data
but outcomes are simulated under pre-specified mechanisms with known truth —
evaluating bias, variance, and empirical coverage to make a GO/STOP gate
decision. The TMLE workflows are gated here. Only in **Stage 3** are real
outcomes accessed and the outcome regressions fitted for the first time. Using a
simulated hepatitis C antiviral safety example, we demonstrate the workflow with
three simulation scenarios that illustrate distinct clean-room decision states.
The central manuscript demonstration (Scenario 3) uses a DGP identified through
systematic grid search where the traditional Stage 2a feasibility path flags
concerns and the PS-diagnostics-gated workflow fails its performance targets
under overlap stress, while the TMLE workflow validated through Stage 2b
plasmode simulation remains adequate — illustrating that traditional feasibility
diagnostics alone cannot determine whether a valid analysis is possible. The
tutorial clarifies that TMLE does not require SuperLearner or machine
learning — it can use ordinary regression nuisance models — but flexible
nuisance estimation is often recommended because it better supports robust and
efficient estimation when treatment and outcome surfaces are complex.

```{r setup}
library(knitr)
library(ggplot2)
library(glmnet)
set.seed(20260309)

# Utility functions
expit <- function(x) 1 / (1 + exp(-x))
logit <- function(p) log(p / (1 - p))
```

# What This Tutorial Is and Is Not

This is **not** a generic TMLE tutorial or a comprehensive estimator comparison.
For standalone introductions to TMLE, see van der Laan and Rose (2011), the
`tmle` R package vignette, or Schuler and Rose (2017). This tutorial teaches
how **multiple analysis workflows** — differing in how the analysis pipeline
is validated before outcome access — can be **operationalized within a
clean-room staged workflow**: a governance structure that controls *when*
different data elements may be accessed and *what* analytic tasks are permitted
at each stage.

**Elevator pitch.** This tutorial shows how three distinct analysis workflows
can be embedded in a clean-room design with **two alternative blinded
feasibility paths**: **Stage 2a** (traditional feasibility diagnostics —
propensity score overlap, balance, matching) and **Stage 2b**
(plasmode-simulation feasibility evaluation — where covariate and treatment
structure come from observed blinded data but outcomes are simulated under
pre-specified mechanisms with known truth). The conventional PS workflow is
gated by Stage 2a; the TMLE workflows are gated by Stage 2b. Real outcome
regressions and estimation occur only in Stage 3, after the chosen feasibility
path is completed.

## Relationship to the Staging and Clean-Room Framework of Muntner et al.

Muntner et al. (2024) describe a **staging framework** for observational
comparative effectiveness and safety studies in which sequential preliminary
analyses — each with checkpoint decisions to continue, revise, or stop — are
completed before the final comparative analysis. Their framework includes
study development and cohort construction, a precision or sample-size check,
a covariate-balance stage based on propensity scores, a residual-confounding
assessment using tools such as negative control outcomes, and finally the
comparative analysis itself. Clean-room protections restrict which data
elements and analytic results are accessible at each stage, and a review team
documents checkpoint decisions in an audit trail.

This tutorial follows the same basic logic of **staged, masked,
pre-outcome decision-making**. Stage 1 corresponds to Muntner's study
development and protocol-lock stages. Stage 2a — evaluating propensity score
overlap, covariate balance, and matching feasibility — corresponds closely to
their covariate-balance checkpoint. Stage 3 corresponds to the final
comparative analysis.

The main methodological extension in this tutorial is **Stage 2b**: where
Muntner et al. emphasize negative control outcomes as a tool for assessing
residual confounding before the final analysis, this tutorial introduces
**plasmode simulation** as an alternative or additional blinded feasibility
path for validating the full estimator pipeline — especially for TMLE
workflows whose performance depends on both the treatment and outcome models.
Plasmode simulation shifts the focus from residual-bias diagnostics alone to
end-to-end estimator validation under plausible data-generating processes
with known truth, all before real outcomes are accessed. This is an extension
of the staging framework rather than a departure from it.

**Stage mapping (this tutorial ↔ Muntner et al.):**

- **Stage 1** ↔ study development / cohort build / protocol lock
- **Stage 2a** ↔ covariate-balance checkpoint (PS overlap, balance, ESS)
- **Stage 2b** ↔ pre-outcome bias / validity checkpoint (extended here via
  plasmode simulation rather than primarily negative control outcomes)
- **Stage 3** ↔ final comparative analysis

::: {.callout-note}
### Tutorial vs. production clean-room implementation

In a real operational clean-room analysis, the workflow would also include
role separation (e.g., a blinded analyst team and an unblinded data
custodian), audit trails with timestamped artifacts, frozen data extracts,
and versioned analysis scripts archived at each stage transition. This
tutorial simplifies those operational components so the reader can focus on
the statistical logic — the sequence of analysis lock → blinded feasibility
evaluation (via traditional diagnostics or plasmode simulation) → unblinded
estimation. The core sequencing principle is the same in both settings:
analytic decisions are locked before outcome data are accessed.
:::

## TMLE and Nuisance Estimation: Regression vs. Flexible Learning

TMLE is a framework for targeted bias reduction and efficient estimation. It
requires nuisance estimates for two quantities — the treatment mechanism
$g(W) = P(A=1|W)$ and the outcome regression $Q(A,W) = E[Y|A,W]$ — but it
**does not mandate any specific nuisance learner**. One can implement TMLE with
ordinary logistic regression models for both $g$ and $Q$.

In practice, however, flexible nuisance estimation — including ensemble methods
such as SuperLearner — is often recommended for TMLE because it can better
capture nonlinearities and interactions in the treatment and outcome mechanisms.
When the nuisance functions are estimated more accurately, TMLE's double
robustness and efficiency properties are better supported, and the resulting
estimates tend to be less biased and more precise.

In this tutorial, the TMLE workflow uses flexible nuisance estimation (penalized
regression via `glmnet`) to illustrate how a targeted-learning strategy may
differ from more conventional regression-based workflows. This is a pedagogical
choice, not a mathematical requirement: the same TMLE framework would work with
simpler models, though its performance advantages might be smaller.

## Three Clean-Room Analysis Workflows

A clean-room staged design is not tied to a single analysis strategy. This
tutorial compares three workflows that differ in **which blinded feasibility
path they use** before outcome access:

**Workflow 1. Conventional PS-diagnostics-gated workflow** *(uses Stage 2a)*

This mirrors standard pharmacoepidemiology practice.

- *Stage 1:* Define estimand; prespecify PS model (logistic regression);
  prespecify matching or IPTW approach.
- *Stage 2a:* Estimate PS; evaluate overlap, balance, ESS, matching feasibility.
- *Decision rule:* If traditional diagnostics are acceptable → proceed to
  Stage 3. If diagnostics are poor → redesign or stop.
- *Stage 3:* Estimate treatment effect using matching or IPTW on the real
  outcome.

This workflow relies primarily on **traditional feasibility diagnostics** rather
than simulation-based estimator validation.

**Workflow 2. Fixed-specification TMLE workflow** *(uses Stage 2b)*

This represents the standard targeted-learning approach.

- *Stage 1:* Prespecify a single TMLE specification: nuisance learner strategy
  for $g$ and $Q$, truncation rules, simulation performance thresholds.
- *Stage 2b:* Run plasmode simulation using the fixed TMLE specification.
  Plasmode simulation uses the covariate and treatment structure from the
  observed blinded data $(W, A)$ but generates synthetic outcomes under
  pre-specified mechanisms with known truth. Evaluate bias, coverage, variance.
  Apply GO/STOP gate decision.
- *Stage 3:* Fit $Q$ on real outcomes; run the targeting step.

The estimator specification is **locked before simulation**. Stage 2b validates
but does not select.

**Workflow 3. Simulation-selected TMLE workflow** *(uses Stage 2b)*

This is the key conceptual extension. Instead of locking a single TMLE
specification, Stage 1 prespecifies a **candidate set of TMLE nuisance-learning
strategies** and a **selection rule**.

- *Stage 1:* Define a candidate set (e.g., TMLE-GLM, TMLE-glmnet,
  TMLE-SuperLearner). Define a selection rule: e.g., choose the specification
  with the lowest RMSE among those with |bias| < threshold, coverage ≥
  threshold, and acceptable SE calibration. Both the candidate set and the rule
  must be written in the analysis lock before outcome access.
- *Stage 2b:* Run plasmode simulations under plausible DGPs. Evaluate each
  candidate TMLE specification. Select the best-performing specification using
  the locked rule.
- *Stage 3:* Apply the selected TMLE specification to the real dataset.

This selection occurs **entirely before real outcomes are accessed**, so it
remains compatible with the clean-room principle. The rule is prespecified; only
its application to plasmode-simulation results occurs at Stage 2b.

Each workflow maps onto the same clean-room structure (Stage 1 pre-specification
→ blinded feasibility evaluation → Stage 3 unblinded estimation), but they
differ in **which feasibility path** they take: Stage 2a (traditional
diagnostics, Workflow 1) vs. Stage 2b (plasmode simulation, Workflows 2 and 3).
The two paths are alternatives, not sequential steps — a workflow uses one or
the other to decide whether to proceed to Stage 3.

```{r workflow-comparison-table}
wf_tab <- data.frame(
  Workflow = c("1. Conventional PS",
               "2. Fixed-specification TMLE",
               "3. Simulation-selected TMLE"),
  `Feasibility path` = c(
    "Stage 2a: traditional diagnostics (overlap, balance, ESS)",
    "Stage 2b: plasmode-simulation validation of a single specification",
    "Stage 2b: plasmode-simulation model selection from candidate set"
  ),
  `Estimator flexibility` = c(
    "Low — fixed logistic PS",
    "Moderate — fixed nuisance strategy",
    "High — candidate set evaluated in simulation"
  ),
  `Risk of misspecification` = c(
    "Higher — no outcome-model validation",
    "Moderate — single specification may not be optimal",
    "Lower — simulation selects best-performing candidate"
  ),
  check.names = FALSE
)
kable(wf_tab, caption = "Comparison of the three clean-room analysis workflows")
```

### The running example

We use a simulated pharmacoepidemiology study loosely modeled on HCV antiviral
safety surveillance. The causal question: does a sofosbuvir (SOF)-containing
regimen increase the risk of acute kidney injury (AKI) compared to a non-SOF
regimen, among patients with chronic hepatitis C? The example is entirely
simulated — no real patient data are used — but the covariate structure,
treatment assignment mechanism, and outcome model reflect patterns seen in
real-world safety studies.

### The central tension

All three workflows share a common operational constraint: **outcome-dependent
modeling decisions must be locked before outcomes are seen.** The clean-room
stages enforce this by controlling data access:

- **The treatment model $g(W)$ can be estimated early** (Stage 2a), because it
  uses only treatment and covariates — no outcomes required.
- **Outcome models can only be fitted on real data after outcome access**
  (Stage 3). This applies to the TMLE $Q(A,W)$ and any outcome-dependent
  component.
- **In plasmode simulation (Stage 2b)**, analysis pipelines run on synthetic
  data with known truth — where the covariate and treatment structure come from
  the observed blinded data but outcomes are simulated under pre-specified
  mechanisms. For the fixed-specification TMLE workflow, this validates a single
  pipeline. For the simulation-selected TMLE workflow, this also determines
  which candidate specification to carry forward.

### Workflow overview

The following diagram adapts the staged checkpoint logic of Muntner et al.
(2024) to a targeted-learning setting with two alternative feasibility paths.
Stage 2a is analogous to Muntner's covariate-balance checkpoint; Stage 2b
extends the pre-outcome validity assessment through plasmode simulation.

```
┌──────────────────────────────────────────────────────┐
│  Stage 1. Pre-specification & Analysis Lock          │
│  Lock estimand, covariates, nuisance strategies,     │
│  truncation rules, gate thresholds.                  │
│  If simulation-selected: lock candidate set          │
│  and selection rule.                                 │
└──────────────────────┬───────────────────────────────┘
                       │
          ┌────────────┴────────────┐
          ↓                         ↓
┌─────────────────────┐  ┌─────────────────────────────┐
│  Stage 2a.          │  │  Stage 2b.                   │
│  Traditional        │  │  Plasmode-Simulation         │
│  Feasibility        │  │  Feasibility Evaluation      │
│  Diagnostics        │  │                              │
│                     │  │  Run analysis pipelines on   │
│  Access (W,A) only. │  │  plasmode-simulated data:    │
│  Fit g(W). Evaluate │  │  observed (W,A) structure +  │
│  overlap, balance,  │  │  synthetic outcomes under    │
│  ESS, matching      │  │  pre-specified DGPs with     │
│  feasibility.       │  │  known truth. Evaluate bias, │
│                     │  │  coverage, SE calibration.   │
│  → Conventional PS  │  │  Fixed TMLE: validate spec.  │
│    workflow gated   │  │  Sim-selected: select best.  │
│    here             │  │  Make GO / STOP decision.    │
│                     │  │                              │
│                     │  │  → TMLE workflows gated here │
└─────────┬───────────┘  └──────────────┬──────────────┘
          └────────────┬────────────────┘
                       ↓
┌──────────────────────────────────────────────────────┐
│  Stage 3. Unblinded Estimation                       │
│  Access Y. Fit outcome model for the first time      │
│  using the locked (or simulation-selected) spec.     │
│  Run TMLE targeting step. Report results.            │
└──────────────────────────────────────────────────────┘
```

*Muntner-style mapping: Stage 1 ↔ study development / protocol lock;
Stage 2a ↔ covariate-balance checkpoint; Stage 2b ↔ pre-outcome validity
checkpoint (via plasmode simulation); Stage 3 ↔ final comparative analysis.*


## Quick Reference: Allowed vs Not Allowed Before Outcome Access

```{r permissions-table}
perms <- data.frame(
  Stage = c(
    "Stage 1: Pre-specification",
    "Stage 2a: Traditional Feasibility Diagnostics",
    "Stage 2b: Plasmode-Simulation Feasibility",
    "Stage 3: Unblinded Estimation"
  ),
  `Data accessible` = c(
    "None (metadata, variable definitions only)",
    "W and A from the real dataset (not Y)",
    "Real (W, A) + synthetic Y from plasmode simulation",
    "Full dataset: W, A, Y"
  ),
  `Allowed tasks` = c(
    paste("Define causal question and estimand;",
          "specify covariates and treatment;",
          "lock nuisance strategies (fixed or candidate set);",
          "set truncation rules;",
          "define diagnostic thresholds, gate criteria,",
          "and selection rule (if simulation-selected)"),
    paste("Estimate g(W) on real data;",
          "PS overlap diagnostics;",
          "IPTW weight distributions and ESS;",
          "covariate balance (SMD);",
          "matching feasibility assessment.",
          "Conventional PS workflow gated here"),
    paste("Run analysis pipelines on plasmode-simulated data;",
          "evaluate bias, variance, and coverage;",
          "validate fixed specification or select among",
          "candidates using locked rule;",
          "test nuisance specifications and truncation.",
          "TMLE workflows gated here"),
    paste("Fit Q(A,W) on real data using locked or",
          "simulation-selected specification;",
          "run TMLE targeting step;",
          "compute estimates and CIs;",
          "report results")
  ),
  `Not allowed` = c(
    paste("Estimating g or Q on real data;",
          "any estimator involving real outcomes"),
    paste("Fitting outcome models on real Y;",
          "computing treatment effects on real data;",
          "running any workflow on the real outcome"),
    paste("Using real outcomes;",
          "tuning estimator based on real Y"),
    paste("Changing estimand, nuisance strategies, or truncation",
          "without a documented protocol amendment")
  ),
  check.names = FALSE
)
kable(perms, caption = "Clean-room permissions by stage (adapted from Muntner et al. staging framework)",
      col.names = c("Stage", "Data accessible", "Allowed tasks", "Not allowed"),
      align = c("l", "l", "l", "l")) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE, font_size = 13) |>
  kableExtra::column_spec(1, bold = TRUE, width = "14em") |>
  kableExtra::column_spec(2, width = "14em") |>
  kableExtra::column_spec(3, width = "20em") |>
  kableExtra::column_spec(4, width = "14em")
```


# Stage 1. Pre-specification and Analysis Lock {#stage-1}

## What happens at this stage

Before any real study data are accessed, the analysis team writes a statistical
analysis plan (SAP) that locks all major analytic decisions. This is analogous
to a clinical trial's SAP being finalized before database lock, and corresponds
to the study development and protocol-lock stages in the Muntner et al. (2024)
framework. Muntner et al. emphasize that protocol development, cohort
construction criteria, prespecified roles and responsibilities, and checkpoint
criteria should all be established before any comparative analysis proceeds.
The analysis lock in this tutorial serves the same function in a
targeted-learning setting: it documents the estimand, nuisance strategies,
truncation rules, and gate thresholds that govern all subsequent checkpoint
decisions. The nuisance strategy is pre-specified here — either as a single
fixed specification (Workflow 2) or as a candidate set with a selection rule
(Workflow 3). The versioned, timestamped analysis lock artifact also serves as
the decision log that Muntner et al. recommend for audit-trail purposes.

For our HCV example, the causal question is:

> Among adults with chronic HCV initiating antiviral therapy, what is the
> average causal effect of a SOF-containing regimen (vs. a non-SOF regimen)
> on the 90-day risk of acute kidney injury?

The estimand is the average treatment effect (ATE) on the risk difference scale:
$$\psi = E[Y^1] - E[Y^0]$$

where $Y^a$ denotes the potential AKI outcome under treatment $a$.

Identification assumptions: consistency ($Y = Y^A$), no unmeasured confounding
($Y^a \perp\!\!\!\perp A \mid W$), and positivity ($0 < P(A=1|W) < 1$).

### The analysis lock object

The pre-specified decisions are encoded as a versioned, timestamped artifact.
This object travels through every subsequent stage unchanged unless a formal
amendment is documented.

```{r analysis-lock}
create_analysis_lock <- function() {
  list(
    # --- Protocol metadata ---
    protocol_id   = "HCV-AKI-2024-001",
    version       = "1.0",
    locked_date   = Sys.Date(),
    locked_by     = "Analysis team (blinded)",
    blinded       = TRUE,  # TRUE until Stage 3 unblinding
    status        = "pre-outcome",

    # --- Causal question ---
    study       = "HCV antiviral safety: SOF vs non-SOF",
    estimand    = "ATE",
    effect_scale = "risk_difference",

    # --- Variable roles ---
    covariates  = c("age", "ckd", "cirrhosis", "diabetes", "nsaid", "renal_risk"),
    treatment   = "A",          # 1 = SOF-containing, 0 = non-SOF
    outcome     = "Y",          # 1 = AKI within 90 days

    # --- Workflow comparison ---
    workflow_comparison = TRUE,

    # --- Workflow 1: Conventional PS approach ---
    conventional_ps_model = "glm",
    conventional_g_formula = "A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk",
    conventional_estimators = c("PS-matched regression", "IPTW"),

    # --- Workflow 2: Fixed-specification TMLE ---
    tmle_nuisance_strategy = "flexible",
    tmle_g_formula = "A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk",
    tmle_g_library = c("SL.mean", "SL.glm", "SL.glmnet"),
    tmle_Q_method  = "cv.glmnet (elastic net, alpha = 0.5)",
    tmle_Q_features = paste("Main effects + age^2, age^3, renal_risk^2 +",
                        "treatment interactions (A:ckd, A:age, A:age^2,",
                        "A:renal_risk, A:cirrhosis, A:diabetes) +",
                        "covariate interactions (age:ckd, age:rr, ckd:rr)"),
    tmle_Q_library = c("SL.mean", "SL.glm", "SL.glmnet", "SL.gam"),

    # --- Workflow 3: Simulation-selected TMLE ---
    sim_selected_candidates = c("TMLE-GLM", "TMLE-glmnet", "TMLE-SuperLearner"),
    sim_selected_rule = paste("Select the candidate with the lowest RMSE",
                               "among those with |bias| < max_abs_bias,",
                               "coverage >= min_coverage, and",
                               "SE/SD ratio within [se_sd_low, se_sd_high]"),

    # --- PS truncation (shared) ---
    ps_truncation = c(lower = 0.025, upper = 0.975),

    # --- Stage 2a diagnostic thresholds ---
    diag_thresholds = list(
      extreme_ps_bound = 0.05,    # PS below this or above 1 - this
      max_extreme_frac = 0.15,    # max fraction of extreme PS
      min_ess_frac     = 0.50,    # minimum ESS / n
      max_iptw_weight  = 40,      # max single IPTW weight
      max_smd_weighted = 0.10     # max weighted SMD
    ),

    # --- Stage 2b simulation targets ---
    sim_targets = list(
      max_abs_bias = 0.01,
      min_coverage = 0.93,
      se_sd_low    = 0.80,
      se_sd_high   = 1.20,
      n_reps       = 200
    ),

    # --- Amendment rule ---
    amendment_rule = paste("If Stage 2b returns STOP, the team may file a",
                           "documented amendment to expand the learner library",
                           "or increase the sample size. The amended plan must",
                           "be re-validated through Stage 2b before proceeding",
                           "to Stage 3.")
  )
}

analysis_lock <- create_analysis_lock()

cat("Analysis lock (", analysis_lock$protocol_id, " v", analysis_lock$version, ")\n", sep = "")
cat("Study:", analysis_lock$study, "\n")
cat("Locked:", as.character(analysis_lock$locked_date), "\n")
cat("Blinded:", analysis_lock$blinded, " | Status:", analysis_lock$status, "\n")
cat("Estimand:", analysis_lock$estimand, "on", analysis_lock$effect_scale, "scale\n")
cat("Covariates:", paste(analysis_lock$covariates, collapse = ", "), "\n")
cat("PS truncation:", analysis_lock$ps_truncation["lower"], "-",
    analysis_lock$ps_truncation["upper"], "\n")
cat("Workflow comparison:", analysis_lock$workflow_comparison, "\n")
cat("  Wf 1 - Conventional PS: logistic regression ->",
    paste(analysis_lock$conventional_estimators, collapse = ", "), "\n")
cat("  Wf 2 - Fixed TMLE: flexible nuisance strategy (glmnet-based Q)\n")
cat("  Wf 3 - Simulation-selected TMLE candidates:",
    paste(analysis_lock$sim_selected_candidates, collapse = ", "), "\n")
cat("  Selection rule:", analysis_lock$sim_selected_rule, "\n")
cat("Amendment rule:", analysis_lock$amendment_rule, "\n")
```

::: {.callout-note}
### Three workflows, two feasibility paths

The conventional PS workflow relies on Stage 2a traditional feasibility
diagnostics (overlap, balance, matching) to decide whether to proceed —
reflecting common applied practice. The fixed-specification TMLE workflow
locks a single nuisance-learning strategy and validates it via Stage 2b
plasmode simulation. The simulation-selected TMLE workflow prespecifies a
candidate set and uses Stage 2b plasmode simulation to choose the
best-performing specification. The three workflows differ in which blinded
feasibility path they use: Stage 2a (traditional diagnostics) or Stage 2b
(plasmode-simulation evaluation).
:::

::: {.callout-note}
### What gets locked vs. what gets re-estimated

What is locked is the **modeling strategy**: the learner library (or candidate
set), truncation bounds, covariate set, selection rule, and diagnostic
thresholds. In practice, the team may either carry the Stage 2a
fitted $\hat{g}$ object forward into Stage 3, or re-run the same locked
procedure on the full $(W, A)$ data at Stage 3. Both are valid — the key
constraint is that the *specification* does not change after outcome access.
Locking a strategy is not the same as requiring reuse of a specific fitted
object.
:::

### Stage 1 summary

At this point, no data have been touched. The team has committed to an estimand
(ATE on the RD scale), identification assumptions, covariate list, nuisance
strategies (fixed specification plus candidate set for simulation-selected
TMLE), truncation rules, selection rule, simulation performance targets, and
the choice of which feasibility path each workflow will use (Stage 2a for the
conventional PS workflow; Stage 2b for the TMLE workflows).


# Stage 2a. Traditional Feasibility Diagnostics {#stage-2a}

## What happens at this stage

The analysis team receives the **treatment and covariate data** $(W, A)$ from
the real study. Outcomes $(Y)$ remain sealed. The purpose is to evaluate
whether the treatment mechanism and covariate structure are compatible with the
planned analyses. Stage 2a is the **primary feasibility gate for the
conventional PS workflow** and provides useful diagnostic information for all
workflows.

Stage 2a is the closest analog to the **covariate-balance stage** in the
Muntner et al. (2024) staging framework. It evaluates propensity score
overlap, covariate balance (weighted SMDs), effective sample size, and
matching feasibility — all before any outcome analysis. The decision to
proceed, revise, or stop at this checkpoint is exactly the kind of
checkpointed review that Muntner et al. recommend as part of a staged
preliminary analysis.

Because $g(W) = P(A=1|W)$ does not involve outcomes, the treatment model can
be estimated here for both the conventional PS and TMLE workflows. The key
operational insight: **the treatment-model side of the analysis is fully
accessible during the outcome-blind stage.**

### What each workflow does at Stage 2a

1. **Conventional PS workflow:** Fit logistic-regression PS. Compute overlap
   and matching diagnostics. **This is the conventional workflow's primary
   feasibility gate.**
2. **Fixed-specification TMLE workflow:** Fit the locked $g(W)$ strategy.
   Compute overlap and weight diagnostics. Note: the TMLE workflow's primary
   gate is Stage 2b (plasmode simulation), not Stage 2a.
3. **Simulation-selected TMLE workflow:** Same as the fixed-specification
   TMLE workflow. The candidate set is locked but specification selection
   does not occur until Stage 2b.

::: {.callout-note collapse="true"}
### Operational note: Stage 2a in practice
In a real clean-room implementation, the analyst would receive a blinded
analytic dataset or restricted data extract with outcomes masked or removed.
All Stage 2a diagnostics would be scripted, versioned, and archived as part of
the audit trail. Any changes to the $g$-strategy triggered by Stage 2a
findings (e.g., adding a covariate interaction to improve balance) would
require documented justification and an updated analysis lock before
proceeding to Stage 2b or Stage 3.
:::

### Generate the HCV tutorial dataset

We simulate a dataset that mimics an HCV antiviral safety study. We generate
the full dataset including $Y$, but the Stage 2a analyst sees only $(W, A)$.

```{r generate-data}
simulate_hcv_scenario <- function(n, seed,
                                   b_rr_g     = 0.7,    # g coeff for renal_risk
                                   beta_rr1   = 0.30,   # Q main coeff for renal_risk
                                   beta_rr2   = 0.08,   # Q coeff for renal_risk^2
                                   beta_rrA   = 0.00,   # A:renal_risk interaction (Scen 2 only)
                                   beta_ckdA  = 0.30,   # A:ckd interaction
                                   beta_ageA  = 0.18,   # A:age_std interaction
                                   beta_age2A = 0.10,   # A:age_std^2 interaction
                                   beta_age3  = 0.12,   # cubic age coefficient
                                   alpha_A    = -0.15,  # treatment intercept
                                   alpha_Y    = -2.20) {
  set.seed(seed)

  # --- Baseline covariates ---
  age_raw    <- rnorm(n, 55, 10)
  age        <- (age_raw - 55) / 10           # standardized: age_std = (age_raw - 55) / 10
  ckd        <- rbinom(n, 1, 0.18)            # chronic kidney disease
  cirrhosis  <- rbinom(n, 1, 0.22)
  diabetes   <- rbinom(n, 1, 0.20)
  nsaid      <- rbinom(n, 1, 0.25)            # concurrent NSAID use
  renal_risk <- rnorm(n, 0, 1)               # composite renal-risk score

  # --- Treatment mechanism (g) ---
  # renal_risk drives overlap; b_rr_g = 0.7 (Scen 1), 2.4 (Scen 2), 2.2 (Scen 3)
  lp_g <- alpha_A + 0.25*age + 0.45*ckd + 0.35*cirrhosis +
           0.20*diabetes - 0.15*nsaid + b_rr_g*renal_risk
  g_true <- pmin(pmax(expit(lp_g), 0.005), 0.995)
  A <- rbinom(n, 1, g_true)

  # --- Outcome model (Q): AKI ---
  # Cubic age polynomial (neg. quadratic + pos./weak cubic) + renal_risk^2 +
  # treatment-effect heterogeneity (A:ckd, A:age, A:age^2, A:renal_risk).
  # beta_A = 0.22 fixed across all scenarios; beta_age3 varies.
  # This rich Q structure is learnable by glmnet but not by simple matching.
  lp_Q <- alpha_Y +
           0.22*A +
           0.25*age  - 0.20*age^2  + beta_age3*age^3 +
           0.95*ckd  + 0.55*cirrhosis + 0.40*diabetes + 0.25*nsaid +
           beta_rr1*renal_risk + beta_rr2*renal_risk^2 +
           beta_ckdA*A*ckd + beta_ageA*A*age + beta_age2A*A*age^2 +
           beta_rrA*A*renal_risk
  Y <- rbinom(n, 1, expit(lp_Q))

  data.frame(age, ckd, cirrhosis, diabetes, nsaid, renal_risk, A, Y)
}

# --- Calibration utilities (needed before data generation) ---
# Calibrate treatment-model intercept for target treatment prevalence
calibrate_alpha_A <- function(b_rr_g, target_prev = 0.49,
                               n_cal = 50000, seed = 42) {
  set.seed(seed)
  age        <- rnorm(n_cal, 0, 1)
  ckd        <- rbinom(n_cal, 1, 0.18)
  cirrhosis  <- rbinom(n_cal, 1, 0.22)
  diabetes   <- rbinom(n_cal, 1, 0.20)
  nsaid      <- rbinom(n_cal, 1, 0.25)
  renal_risk <- rnorm(n_cal, 0, 1)
  lp_no_int  <- 0.25*age + 0.45*ckd + 0.35*cirrhosis +
                0.20*diabetes - 0.15*nsaid + b_rr_g*renal_risk

  f <- function(a) mean(expit(a + lp_no_int)) - target_prev
  uniroot(f, c(-5, 5))$root
}

# Calibrate outcome-model intercept for target outcome prevalence
calibrate_alpha_Y <- function(target_prev, beta_A = 0.22,
                               beta_rr1, beta_rr2, beta_rrA,
                               beta_ckdA, beta_ageA, beta_age2A,
                               beta_age3, n_cal = 50000, seed = 43) {
  set.seed(seed)
  age        <- rnorm(n_cal, 0, 1)
  ckd        <- rbinom(n_cal, 1, 0.18)
  cirrhosis  <- rbinom(n_cal, 1, 0.22)
  diabetes   <- rbinom(n_cal, 1, 0.20)
  nsaid      <- rbinom(n_cal, 1, 0.25)
  renal_risk <- rnorm(n_cal, 0, 1)
  A          <- rbinom(n_cal, 1, 0.49)   # approximate treatment prevalence

  lp_no_int <- beta_A*A +
               0.25*age - 0.20*age^2 + beta_age3*age^3 +
               0.95*ckd + 0.55*cirrhosis + 0.40*diabetes + 0.25*nsaid +
               beta_rr1*renal_risk + beta_rr2*renal_risk^2 +
               beta_ckdA*A*ckd + beta_ageA*A*age + beta_age2A*A*age^2 +
               beta_rrA*A*renal_risk

  f <- function(a) mean(expit(a + lp_no_int)) - target_prev
  uniroot(f, c(-8, 2))$root
}

# --- Calibrate Scenario 3 intercepts (manuscript demonstration DGP) ---
scen3_alpha_A <- calibrate_alpha_A(b_rr_g = 2.2, target_prev = 0.49)
scen3_alpha_Y <- calibrate_alpha_Y(
  target_prev = 0.10,
  beta_rr1 = 0.00, beta_rr2 = 0.05, beta_rrA = 0.00,
  beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10,
  beta_age3 = 0.10
)
cat("Scenario 3 calibrated intercepts: alpha_A =", round(scen3_alpha_A, 4),
    ", alpha_Y =", round(scen3_alpha_Y, 4), "\n")

# Generate the "real" study data (n = 1500)
# Using Scenario 3 (manuscript demo) parameters: overlap stress (b_rr=2.2),
# near-zero renal_risk in Q, ~10% outcome prevalence
dat_full <- simulate_hcv_scenario(n = 1500, seed = 42,
                                   b_rr_g = 2.2, beta_rr1 = 0.00,
                                   beta_rr2 = 0.05, beta_rrA = 0.00,
                                   beta_ckdA = 0.30, beta_ageA = 0.18,
                                   beta_age2A = 0.10, beta_age3 = 0.10,
                                   alpha_A = scen3_alpha_A,
                                   alpha_Y = scen3_alpha_Y)

# Stage 2a analyst sees only W and A — Y is sealed
dat_stage2 <- dat_full[, c("age", "ckd", "cirrhosis", "diabetes",
                            "nsaid", "renal_risk", "A")]

cat("Stage 2a data dimensions:", nrow(dat_stage2), "x", ncol(dat_stage2), "\n")
cat("Treatment prevalence:", round(mean(dat_stage2$A), 3), "\n")
cat("Y column accessible:", "Y" %in% names(dat_stage2), "\n")
```

### Fit the treatment model

Using the locked g-formula from the analysis lock. Both the conventional PS
and TMLE workflows use the same logistic-regression PS formula at Stage 2a
(the TMLE workflow may use a richer $g$-learner in Stage 2b/3, but the
Stage 2a diagnostic uses the shared specification):

```{r fit-g}
fit_stage2_g <- function(dat, lock) {
  g_fit <- glm(as.formula(lock$conventional_g_formula), data = dat, family = "binomial")
  g_hat <- fitted(g_fit)
  g_trunc <- pmin(pmax(g_hat, lock$ps_truncation["lower"]),
                  lock$ps_truncation["upper"])
  list(fit = g_fit, g_hat = g_hat, g_trunc = g_trunc)
}

g_stage2 <- fit_stage2_g(dat_stage2, analysis_lock)

cat("PS range (raw):", round(range(g_stage2$g_hat), 4), "\n")
cat("PS range (truncated):", round(range(g_stage2$g_trunc), 4), "\n")
```

### Propensity score diagnostics

These diagnostics use only $(W, A)$ and are fully permitted at Stage 2a.

```{r ps-diagnostics, fig.width=8, fig.height=4}
ps_df <- data.frame(
  ps = g_stage2$g_hat,
  Treatment = factor(dat_stage2$A,
                     labels = c("Non-SOF", "SOF-containing"))
)

ggplot(ps_df, aes(x = ps, fill = Treatment)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = c(analysis_lock$ps_truncation["lower"],
                             analysis_lock$ps_truncation["upper"]),
             linetype = "dashed", color = "red") +
  labs(title = "Stage 2a: Propensity Score Distribution by Treatment",
       subtitle = "Red dashed lines = pre-specified truncation bounds",
       x = "Estimated P(SOF | W)", y = "Count") +
  theme_minimal(base_size = 12)
```

```{r overlap-metrics}
stage2_diagnostics <- function(g_hat, g_trunc, A, dat, lock) {
  thr <- lock$diag_thresholds
  n <- length(A)

  # Fraction with extreme PS
  frac_extreme <- mean(g_hat < thr$extreme_ps_bound |
                       g_hat > (1 - thr$extreme_ps_bound))

  # IPTW weights and ESS
  w_iptw <- A / g_trunc + (1 - A) / (1 - g_trunc)
  ess <- sum(w_iptw)^2 / sum(w_iptw^2)
  ess_frac <- ess / n
  max_w <- max(w_iptw)

  # Weighted SMDs
  compute_smd <- function(x, trt, weights = NULL) {
    if (is.null(weights)) weights <- rep(1, length(x))
    m1 <- weighted.mean(x[trt == 1], weights[trt == 1])
    m0 <- weighted.mean(x[trt == 0], weights[trt == 0])
    s <- sqrt((var(x[trt == 1]) + var(x[trt == 0])) / 2)
    if (s < 1e-10) return(0)
    (m1 - m0) / s
  }

  covars <- lock$covariates
  smd_unw <- sapply(covars, function(v) compute_smd(dat[[v]], A))
  smd_wtd <- sapply(covars, function(v) compute_smd(dat[[v]], A, w_iptw))

  diag_tab <- data.frame(
    Metric = c("Fraction extreme PS",
               "ESS fraction",
               "Max IPTW weight",
               "Max |SMD| (IPTW-weighted)"),
    Value = round(c(frac_extreme, ess_frac, max_w, max(abs(smd_wtd))), 3),
    Threshold = c(paste("<", thr$max_extreme_frac),
                  paste(">", thr$min_ess_frac),
                  paste("<", thr$max_iptw_weight),
                  paste("<", thr$max_smd_weighted)),
    Status = c(ifelse(frac_extreme < thr$max_extreme_frac, "PASS", "FLAG"),
               ifelse(ess_frac > thr$min_ess_frac, "PASS", "FLAG"),
               ifelse(max_w < thr$max_iptw_weight, "PASS", "FLAG"),
               ifelse(max(abs(smd_wtd)) < thr$max_smd_weighted, "PASS", "FLAG"))
  )

  list(diag_tab = diag_tab, smd_unw = smd_unw, smd_wtd = smd_wtd,
       w_iptw = w_iptw, any_flag = any(diag_tab$Status == "FLAG"))
}

s2_diag <- stage2_diagnostics(g_stage2$g_hat, g_stage2$g_trunc,
                               dat_stage2$A, dat_stage2, analysis_lock)
kable(s2_diag$diag_tab, caption = "Stage 2a overlap diagnostics")
```

```{r love-plot, fig.width=7, fig.height=4}
covars <- analysis_lock$covariates
smd_df <- data.frame(
  variable = rep(covars, 2),
  smd = c(s2_diag$smd_unw, s2_diag$smd_wtd),
  type = rep(c("Unweighted", "IPTW-Weighted"), each = length(covars))
)

ggplot(smd_df, aes(x = abs(smd), y = reorder(variable, abs(smd)), color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.10, linetype = "dashed") +
  labs(title = "Stage 2a: Covariate Balance (Love Plot)",
       x = "|Standardized Mean Difference|", y = NULL, color = "Weighting") +
  theme_minimal(base_size = 12)
```

### Stage 2a decision

```{r stage2-decision}
stage2_status <- ifelse(s2_diag$any_flag,
                         "FLAG — traditional diagnostics flagged; proceed to Stage 2b plasmode simulation",
                         "PASS — traditional diagnostics acceptable")
cat("Stage 2a status:", stage2_status, "\n")
if (s2_diag$any_flag) {
  flagged <- s2_diag$diag_tab$Metric[s2_diag$diag_tab$Status == "FLAG"]
  cat("Flagged metrics:", paste(flagged, collapse = "; "), "\n")
}
```

### Why Stage 2a alone is not enough for full validation

Stage 2a evaluates the treatment mechanism $g$ and practical positivity. But
this tells us nothing about whether outcome models can be estimated well,
whether TMLE's targeting step is stable, or whether CI coverage is near
nominal for any workflow.

This matters in both directions:

- A study can have **flagged overlap but excellent TMLE performance** if the
  overlap-driving covariate is weakly related to the outcome — the outcome
  model compensates.
- A study can have **acceptable overlap but poor performance** if the
  outcome surface is difficult to learn at the given sample size.
- The **simulation-selected TMLE workflow** cannot select among candidates
  at Stage 2a, since no plasmode simulation has been run yet.

Stage 2a alone cannot distinguish these cases. **Stage 2b fills the gap** by
testing the workflow pipelines on plasmode-simulated data where the truth is
known — and, for the simulation-selected TMLE workflow, by choosing the best
candidate specification from the pre-registered set. This is why Stage 2a
and Stage 2b are complementary feasibility paths: Stage 2a assesses the
treatment mechanism, while Stage 2b evaluates the full analysis pipeline.

### Stage 2a summary

```{r stage2-summary, results='asis'}
cat("At this point the team has estimated $g(W)$ for the conventional PS and TMLE ")
cat("workflows, evaluated overlap and balance, ")
if (s2_diag$any_flag) {
  cat("and documented that **some Stage 2a diagnostic flags triggered**. ")
  cat("For the conventional PS workflow, these flags raise feasibility concerns. ")
  cat("For the TMLE workflows, Stage 2a flags are informative but not decisive — ")
  cat("the primary feasibility gate is Stage 2b plasmode simulation. ")
} else {
  cat("and found that **all Stage 2a diagnostic thresholds were met**. ")
  cat("For the conventional PS workflow, this is encouraging. ")
  cat("However, passing Stage 2a does not guarantee that any workflow's full ")
  cat("analysis pipeline will perform adequately — as Stage 2b will show. ")
}
cat("The simulation-selected TMLE workflow's candidate set is locked but no ")
cat("plasmode-simulation-based selection has occurred yet. The outcome $Y$ has ")
cat("**not been touched**. All three workflows remain within the clean room.")
```


# Why TMLE Can Survive Poor Overlap {#double-robustness}

Before proceeding to Stage 2b (plasmode simulation), it helps to understand
*why* poor Stage 2a propensity score diagnostics do not automatically
invalidate the TMLE workflow — and why this distinguishes it from the
conventional PS workflow. The answer is **double robustness**.

TMLE relies on two nuisance components:

1. The **outcome regression** $Q(A,W) = E[Y|A,W]$: a model for the expected
   outcome given treatment and covariates.
2. The **treatment mechanism** $g(W) = P(A=1|W)$: a model for the probability
   of receiving treatment given covariates.

The key property is this: **TMLE is consistent if *either* the outcome model
or the treatment model is consistently estimated.** Both do not need to be
correct simultaneously for TMLE to converge to the right answer. If one model
is wrong but the other is right, TMLE uses the correct model to compensate for
the incorrect one.

When *both* models are consistently estimated, TMLE achieves an additional
benefit: it is **efficient**, meaning it reaches the smallest possible variance
among all regular estimators.

What does this mean practically?

- If the overlap-driving covariate barely affects the outcome (weak
  confounding through that variable), then the outcome model $Q$ can be
  estimated well even though $g$ is difficult. TMLE leans on $Q$ to get the
  right answer. (This is the mechanism behind Scenario 3 — the manuscript
  demonstration — in Stage 2b.)
- If the overlap-driving covariate strongly affects both treatment and outcome,
  then *both* $g$ and $Q$ are hard to estimate in the regions of poor overlap.
  TMLE has nothing to fall back on. (This is what happens in Scenario 2.)
- **IPTW** and **PS-matched regression**, by contrast, rely entirely on $g$.
  Poor overlap always hurts these methods, regardless of the outcome surface.

This is why overlap concerns — whether flagged at Stage 2a or not — do not
automatically determine whether TMLE will perform well or poorly. They mean
"the treatment model alone may be under stress — we need Stage 2b plasmode
simulation to check whether the full TMLE pipeline, including the outcome
model, can compensate." Flexible nuisance learning combined with
simulation-based specification selection (as in the simulation-selected TMLE
workflow) further strengthens this ability to adapt, because the
plasmode simulation can identify which nuisance-learning strategy best
exploits the available outcome structure before any real outcomes are seen.

::: {.callout-important}
Double robustness is an asymptotic property that holds under regularity
conditions, including adequate positivity and sufficiently good nuisance
estimation — it is not a finite-sample guarantee. In practice, extreme weights
from a poorly estimated $g$ can still destabilize the targeting step, inflate
variance, and degrade CI coverage even when the outcome model is well-specified.
This is exactly why plasmode-simulation validation (Stage 2b) is necessary:
it checks whether double robustness actually helps at the study's sample size
and covariate distribution, rather than relying on asymptotic theory alone.
:::

### How TMLE uses nuisance estimates across the clean-room stages

In the TMLE workflow, nuisance estimation is distributed across stages:

- **Stage 2a:** The treatment mechanism $g(W)$ can be estimated and diagnosed
  on real $(W,A)$ data. Diagnostics are computed and the $g$-strategy is locked.
- **Stage 2b:** The same locked nuisance-learning strategy for both $g$ and $Q$
  is validated on plasmode-simulated data with known truth. The full TMLE
  pipeline runs end-to-end.
- **Stage 3:** The real-data outcome regression $Q(A,W)$ is fit for the first
  time. The targeted update combines $g$ and $Q$ after unblinding.

One may estimate $g$ using SuperLearner, `glmnet`, or simpler models — the
framework is the same. This tutorial uses flexible nuisance estimation to
illustrate a targeted-learning default strategy, not because TMLE
mathematically requires it.


# Stage 2b. Plasmode-Simulation Feasibility Evaluation {#stage-2b}

## What happens at this stage

Stage 2a told us about the treatment mechanism. Stage 2b answers the harder
question: **do the analysis workflows perform adequately under conditions
that resemble this study?**

The team runs a **plasmode simulation** — a simulation approach where the
covariate and treatment structure come from the observed blinded data $(W, A)$
but outcomes are generated synthetically under pre-specified mechanisms with
known truth. ("Plasmode" means the simulation is grounded in real data
structure rather than fully artificial.) Each DGP is calibrated to match the
observed $(W, A)$ distribution but generates **synthetic outcomes** — not the
real $Y$. Because the truth is known in the plasmode simulation, the team can
evaluate bias, variance, and CI coverage for each workflow.

Stage 2b is the **primary feasibility gate for the TMLE workflows**. It is an
alternative to Stage 2a, not a sequential step — the conventional PS workflow
uses Stage 2a, while the TMLE workflows use Stage 2b.

In the Muntner et al. (2024) staging framework, a later checkpoint assesses
the plausibility of remaining bias — for example, using negative control
outcomes and review-team evaluation — before the final comparative analysis.
Stage 2b serves a related checkpoint function, but uses plasmode simulation to
evaluate the full estimator pipeline under known truth before real outcome
access. Where Muntner et al. emphasize negative control outcomes as a way to
assess residual confounding before the final comparative analysis, this
tutorial instead emphasizes plasmode simulation as a pre-outcome method for
validating estimator performance under plausible data-generating processes.
This is an extension of the staging framework's checkpoint logic rather than a
departure from it: both approaches aim to provide evidence about analysis
validity before unblinding, but plasmode simulation additionally evaluates
bias, variance, and CI coverage of the full estimation pipeline — information
that negative control outcomes alone do not directly provide.

::: {.callout-note collapse="true"}
### Operational note: Stage 2b in practice
In a production clean-room analysis, Stage 2b plasmode simulations would be
pre-run from versioned scripts and their results archived as part of the audit
trail. The candidate DGPs are specified before outcome access, based on
clinical domain knowledge and the observed covariate/treatment distributions
from Stage 2a. The analyst does not select a DGP post hoc to justify
proceeding — the scenarios are locked as part of the Stage 1 analysis plan or
a documented Stage 2a amendment. The purpose is to stress-test the locked
estimator pipeline under plausible candidate outcome surfaces before
unblinding.
:::

We run **three scenarios** to demonstrate three distinct clean-room decision
states, using four estimators that span the workflows: crude RD (benchmark),
PS-matched regression and IPTW (conventional PS workflow), and TMLE (TMLE
workflows). In a simulation-selected TMLE workflow, Stage 2b would run each
candidate specification and apply the locked selection rule; here, we
illustrate with the fixed-specification TMLE pipeline.

### What Stage 2b validates that Stage 2a cannot

| Question | Stage 2a | Stage 2b |
|----------|----------|----------|
| Is $\hat{g}$ well-estimated? | Yes | Yes |
| Are PS values extreme? | Yes | Yes |
| Can adequate matching be achieved on the observed data? | Yes | Partially |
| Does the conventional PS workflow perform adequately? | **No** | Yes |
| Which simulation-selected TMLE candidate performs best? | **No** | Yes |
| Is $\hat{Q}$ adequate for TMLE? | **No** | Yes (on plasmode data) |
| Does the TMLE targeting step work? | **No** | Yes |
| Is IC-based SE reliable? | **No** | Yes |
| Does CI coverage reach nominal for each workflow? | **No** | Yes |
| Which workflow performs best under these conditions? | **No** | Yes |


### Three scenarios for the HCV example

| Scenario | Overlap / matching | Outcome structure | Expected gate | Main lesson (Stage 2a vs 2b) |
|----------|--------------------|-------------------|---------------|------------------------------|
| **1** | Acceptable (`b_rr`=0.7) | Moderate `renal_risk` ($\beta_{rr1}=0.30$), cubic age, $A \times \text{ckd}$, $A \times \text{age}$, $A \times \text{age}^2$ | GO | Stage 2a passes; Stage 2b confirms all workflows viable |
| **2** | Overlap stress (`b_rr`=2.4) | Strong `renal_risk` ($\beta_{rr1}=1.30$, $\beta_{rr2}=0.25$), $A \times \text{renal\_risk}=0.60$, weaker Q elsewhere | STOP | Stage 2a flags concerns; Stage 2b confirms no workflow adequate |
| **3** *(manuscript demo)* | Overlap stress (`b_rr`=2.2, n=1500) | Near-zero `renal_risk` ($\beta_{rr1}=0.00$, $\beta_{rr2}=0.05$), $A \times \text{rr}=0$, cubic age, $A \times \text{ckd}$, $A \times \text{age}$, $A \times \text{age}^2$ | GO | Stage 2a flags concerns; Stage 2b shows TMLE adequate while PS fails |

Scenarios 2 and 3 both create overlap stress through strong `renal_risk`
effects in the treatment model (`b_rr` = 2.4 and 2.2, respectively), so their
Stage 2a overlap and matching diagnostics are similarly challenged. The key
difference is whether `renal_risk` matters for the outcome. All three
scenarios share a baseline outcome structure — cubic age polynomial, binary
comorbidities, and treatment-by-subgroup heterogeneity ($A \times \text{ckd}$,
$A \times \text{age}$, $A \times \text{age}^2$) — giving the DGP learnable
structure that the `glmnet`-based $Q$ model can exploit. The scenarios differ
in the role of `renal_risk` in the outcome: moderate in Scenario 1, dominant
with a strong $A \times \text{renal\_risk}$ interaction in Scenario 2, and
negligible in Scenario 3. Scenario 3 — the manuscript demonstration — uses
parameters identified through a systematic grid search: `b_rr` = 2.2 creates
overlap stress while `renal_risk` has near-zero coefficients in the outcome
model ($\beta_{rr1}=0.00$, $\beta_{rr2}=0.05$, $\beta_{rrA}=0.00$). At
n = 1500 with ~10% outcome prevalence, the traditional PS-diagnostics-gated
workflow fails its performance targets while the TMLE workflow remains adequate.
PS-matched regression (conventional PS workflow) uses a simple paired-difference
estimator and cannot exploit the nonlinear or interaction structure in the
outcome model.

### Define the plasmode-simulation DGP

The DGP must be specified *before* real outcomes are seen. It should reflect
plausible outcome models given clinical domain knowledge. In a plasmode
simulation, the covariate and treatment structure come from the observed
blinded data, while outcomes are generated synthetically under these
pre-specified mechanisms.

```{r sim-dgp}
# Compute true ATE by Monte Carlo marginalization over the covariate distribution.
# Must match simulate_hcv_scenario exactly (same age polynomial, same coefficients).
compute_truth_hcv <- function(n_mc = 500000, seed = 1,
                               beta_rr1   = 0.30,
                               beta_rr2   = 0.08,
                               beta_rrA   = 0.00,
                               beta_ckdA  = 0.30,
                               beta_ageA  = 0.18,
                               beta_age2A = 0.10,
                               beta_age3  = 0.12,
                               alpha_Y    = -2.20) {
  set.seed(seed)
  age        <- rnorm(n_mc, 0, 1)            # already standardized (mean 0, SD 1)
  ckd        <- rbinom(n_mc, 1, 0.18)
  cirrhosis  <- rbinom(n_mc, 1, 0.22)
  diabetes   <- rbinom(n_mc, 1, 0.20)
  nsaid      <- rbinom(n_mc, 1, 0.25)
  renal_risk <- rnorm(n_mc, 0, 1)

  lp_base <- alpha_Y +
             0.25*age - 0.20*age^2 + beta_age3*age^3 +
             0.95*ckd + 0.55*cirrhosis + 0.40*diabetes + 0.25*nsaid +
             beta_rr1*renal_risk + beta_rr2*renal_risk^2
  EY1 <- mean(expit(lp_base + 0.22 + beta_ckdA*ckd +
                     beta_ageA*age + beta_age2A*age^2 + beta_rrA*renal_risk))
  EY0 <- mean(expit(lp_base))

  c(EY1 = EY1, EY0 = EY0, RD = EY1 - EY0)
}

# Scenario 1: Acceptable overlap, rich Q structure (A:ckd, A:age, A:age^2)
truth_1 <- compute_truth_hcv(beta_rr1 = 0.30, beta_rr2 = 0.08, beta_rrA = 0.00,
                              beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10,
                              alpha_Y = -2.20)
# Scenario 2: Overlap stress, strong renal_risk + A:renal_risk interaction, TMLE STOP
truth_2 <- compute_truth_hcv(beta_rr1 = 1.30, beta_rr2 = 0.25, beta_rrA = 0.60,
                              beta_ckdA = 0.20, beta_ageA = 0.10, beta_age2A = 0.05,
                              alpha_Y = -2.00)
# Scenario 3: Manuscript demonstration — overlap stress (b_rr = 2.2),
# near-zero renal_risk in Q, TMLE GO while traditional PS workflow fails
truth_3 <- compute_truth_hcv(beta_rr1 = 0.00, beta_rr2 = 0.05, beta_rrA = 0.00,
                              beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10,
                              beta_age3 = 0.10, alpha_Y = scen3_alpha_Y)

cat("True ATE by scenario:\n")
cat("  Scenario 1 (acceptable overlap):  ", round(truth_1["RD"], 5), "\n")
cat("  Scenario 2 (overlap stress, STOP):", round(truth_2["RD"], 5), "\n")
cat("  Scenario 3 (manuscript demo, GO): ", round(truth_3["RD"], 5), "\n")
```

### Define the estimators

This function runs one TMLE analysis using the pre-specified pipeline. The $Q$
model uses `glmnet` (penalized logistic regression) with a rich design matrix
including polynomial terms, pairwise interactions, and treatment-by-covariate
interactions. This is a modest but meaningful upgrade over a plain GLM: `glmnet`
automatically selects among the candidate features via cross-validated
regularization, approximating what a small SuperLearner library would do but
at tutorial speed. The $g$ model remains a simple logistic GLM.

```{r tmle-function}
# Build the rich design matrix for Q estimation
build_Q_design <- function(dat) {
  A <- dat$A; age <- dat$age; ckd <- dat$ckd; cirrhosis <- dat$cirrhosis
  diabetes <- dat$diabetes; nsaid <- dat$nsaid; rr <- dat$renal_risk

cbind(
    A         = A,
    age       = age,
    age2      = age^2,
    age3      = age^3,
    ckd       = ckd,
    cirrhosis = cirrhosis,
    diabetes  = diabetes,
    nsaid     = nsaid,
    rr        = rr,
    rr2       = rr^2,
    # Treatment-effect heterogeneity terms
    A_ckd     = A * ckd,
    A_age     = A * age,
    A_age2    = A * age^2,
    A_rr      = A * rr,
    A_cirr    = A * cirrhosis,
    A_dm      = A * diabetes,
    A_age_ckd = A * age * ckd,
    # Pairwise covariate interactions
    age_ckd   = age * ckd,
    age_rr    = age * rr,
    ckd_rr    = ckd * rr
  )
}

run_tmle_once <- function(dat, ps_clip = c(0.025, 0.975)) {
  n <- nrow(dat)

  # --- Fit g (treatment model) — simple logistic GLM ---
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])

  # --- Fit Q (outcome model) — glmnet with rich design matrix ---
  X_Q <- build_Q_design(dat)
  Q_fit <- cv.glmnet(X_Q, dat$Y, family = "binomial", alpha = 0.5,
                      nfolds = 5, type.measure = "deviance")
  Q_A <- as.numeric(predict(Q_fit, X_Q, s = "lambda.min", type = "response"))

  dat1 <- dat; dat1$A <- 1
  dat0 <- dat; dat0$A <- 0
  Q_1 <- as.numeric(predict(Q_fit, build_Q_design(dat1),
                             s = "lambda.min", type = "response"))
  Q_0 <- as.numeric(predict(Q_fit, build_Q_design(dat0),
                             s = "lambda.min", type = "response"))

  # Bound Q away from 0/1
  Q_A <- pmin(pmax(Q_A, 1e-5), 1 - 1e-5)
  Q_1 <- pmin(pmax(Q_1, 1e-5), 1 - 1e-5)
  Q_0 <- pmin(pmax(Q_0, 1e-5), 1 - 1e-5)

  # --- Targeting step ---
  H  <- dat$A / g_hat - (1 - dat$A) / (1 - g_hat)
  H1 <- 1 / g_hat
  H0 <- -1 / (1 - g_hat)

  eps_fit <- glm(dat$Y ~ -1 + H, offset = logit(Q_A), family = "binomial")
  eps <- coef(eps_fit)

  Q_1_star <- expit(logit(Q_1) + eps * H1)
  Q_0_star <- expit(logit(Q_0) + eps * H0)
  Q_A_star <- expit(logit(Q_A) + eps * H)

  # --- Point estimate ---
  psi <- mean(Q_1_star - Q_0_star)

  # --- IC-based inference ---
  eif <- H * (dat$Y - Q_A_star) + (Q_1_star - Q_0_star) - psi
  se  <- sqrt(var(eif) / n)
  ci  <- psi + c(-1.96, 1.96) * se

  list(estimate = psi, se = se, ci_low = ci[1], ci_high = ci[2])
}
```

We also define comparators for the conventional PS workflow. IPTW and
PS-matched regression rely primarily on design-stage adjustment through the
treatment model.

```{r comparator-functions}
run_crude <- function(dat) {
  p1 <- mean(dat$Y[dat$A == 1]); p0 <- mean(dat$Y[dat$A == 0])
  rd <- p1 - p0
  n1 <- sum(dat$A == 1); n0 <- sum(dat$A == 0)
  se <- sqrt(p1*(1 - p1)/n1 + p0*(1 - p0)/n0)
  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}

run_iptw <- function(dat, ps_clip = c(0.025, 0.975)) {
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])
  A <- dat$A; Y <- dat$Y; n <- length(Y)
  w  <- A / g_hat + (1 - A) / (1 - g_hat)
  d1 <- sum(w * A); d0 <- sum(w * (1 - A))
  mu1 <- sum(w * A * Y) / d1
  mu0 <- sum(w * (1 - A) * Y) / d0
  rd <- mu1 - mu0
  # Sandwich SE
  phi1 <- w * A * (Y - mu1) / d1 * n
  phi0 <- w * (1 - A) * (Y - mu0) / d0 * n
  se <- sqrt(sum((phi1 - phi0)^2) / n^2)
  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}

run_ps_matched <- function(dat, ps_clip = c(0.025, 0.975)) {
  # Fit propensity score
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  ps <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])
  lps <- logit(ps)

  # 1:1 nearest-neighbor matching without replacement, caliper = 0.2 * SD(logit PS)
  caliper <- 0.2 * sd(lps)
  tidx <- which(dat$A == 1)
  cidx <- which(dat$A == 0)

  # Greedy matching in original order
  matched_t <- matched_c <- integer(0)
  available <- rep(TRUE, length(cidx))

  for (ii in seq_along(tidx)) {
    i <- tidx[ii]
    if (!any(available)) break
    ac <- cidx[available]
    d_lps <- abs(lps[i] - lps[ac])
    j <- which.min(d_lps)
    if (d_lps[j] <= caliper) {
      matched_t <- c(matched_t, i)
      matched_c <- c(matched_c, ac[j])
      available[which(cidx == ac[j])] <- FALSE
    }
  }

  np <- length(matched_t)
  if (np < 20) {
    return(list(estimate = NA, se = NA, ci_low = NA, ci_high = NA,
                n_pairs = np, matched_frac = np / length(tidx), max_smd = NA))
  }

  # Paired difference estimator
  rd <- mean(dat$Y[matched_t] - dat$Y[matched_c])
  se <- sd(dat$Y[matched_t] - dat$Y[matched_c]) / sqrt(np)

  # Post-match balance
  covars <- c("age", "ckd", "cirrhosis", "diabetes", "nsaid", "renal_risk")
  smds <- sapply(covars, function(v) {
    xt <- dat[[v]][matched_t]; xc <- dat[[v]][matched_c]
    s <- sqrt((var(xt) + var(xc)) / 2)
    if (s < 1e-10) 0 else (mean(xt) - mean(xc)) / s
  })

  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se,
       n_pairs = np, matched_frac = np / length(tidx),
       max_smd = max(abs(smds)))
}

```

### Run the three-scenario plasmode simulation

```{r run-sim, cache=TRUE}
run_stage3_simulation <- function(reps, n, b_rr_g,
                                   beta_rr1 = 0.30, beta_rr2 = 0.08,
                                   beta_rrA = 0.00,
                                   beta_ckdA = 0.30, beta_ageA = 0.18,
                                   beta_age2A = 0.10, beta_age3 = 0.12,
                                   alpha_A = -0.15, alpha_Y = -2.20,
                                   true_rd, seed_offset = 0,
                                   ps_clip = c(0.025, 0.975)) {
  methods <- c("Crude", "IPTW", "PS-Matched", "TMLE")
  results <- vector("list", reps * length(methods))
  k <- 0

  for (r in seq_len(reps)) {
    d <- simulate_hcv_scenario(n, seed = seed_offset + r,
                                b_rr_g = b_rr_g, beta_rr1 = beta_rr1,
                                beta_rr2 = beta_rr2, beta_rrA = beta_rrA,
                                beta_ckdA = beta_ckdA, beta_ageA = beta_ageA,
                                beta_age2A = beta_age2A, beta_age3 = beta_age3,
                                alpha_A = alpha_A, alpha_Y = alpha_Y)

    for (meth in methods) {
      res <- tryCatch(
        switch(meth,
               Crude          = run_crude(d),
               IPTW           = run_iptw(d, ps_clip),
               "PS-Matched"   = run_ps_matched(d, ps_clip),
               TMLE           = run_tmle_once(d, ps_clip)),
        error = function(e) list(estimate = NA, se = NA,
                                  ci_low = NA, ci_high = NA))
      k <- k + 1
      results[[k]] <- data.frame(
        rep = r, method = meth,
        est = res$estimate, se = res$se,
        ci_lo = res$ci_low, ci_hi = res$ci_high)
    }
  }

  results <- do.call(rbind, results)

  # Compute performance metrics per method
  metrics <- do.call(rbind, lapply(split(results, results$method), function(sub) {
    ok     <- !is.na(sub$est)
    has_ci <- ok & !is.na(sub$ci_lo)
    e <- sub$est[ok]
    data.frame(
      method   = sub$method[1],
      n_ok     = sum(ok),
      bias     = mean(e) - true_rd,
      rmse     = sqrt(mean((e - true_rd)^2)),
      emp_sd   = sd(e),
      mean_se  = mean(sub$se[ok], na.rm = TRUE),
      coverage = mean(
        (sub$ci_lo[has_ci] <= true_rd) & (true_rd <= sub$ci_hi[has_ci])),
      n_ci     = sum(has_ci))
  }))
  row.names(metrics) <- NULL

  list(results = results, metrics = metrics)
}

n_reps <- 200   # number of replicates
# Per-scenario sample sizes:
#   Scenario 1 (acceptable overlap):   n = 3000
#   Scenario 2 (overlap stress, STOP): n = 3000
#   Scenario 3 (manuscript demo, GO):  n = 1500 — grid-search-identified DGP

# --- Scenario 1: Acceptable overlap, rich Q structure (A:ckd, A:age, A:age^2) ---
sim_1 <- run_stage3_simulation(
  reps = n_reps, n = 3000,
  b_rr_g = 0.7, beta_rr1 = 0.30, beta_rr2 = 0.08, beta_rrA = 0.00,
  beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10,
  alpha_Y = -2.20,
  true_rd = truth_1["RD"], seed_offset = 1000,
  ps_clip = analysis_lock$ps_truncation)

# --- Scenario 2: Overlap stress, strong renal_risk + A:renal_risk, TMLE STOP ---
sim_2 <- run_stage3_simulation(
  reps = n_reps, n = 3000,
  b_rr_g = 2.4, beta_rr1 = 1.30, beta_rr2 = 0.25, beta_rrA = 0.60,
  beta_ckdA = 0.20, beta_ageA = 0.10, beta_age2A = 0.05,
  alpha_Y = -2.00,
  true_rd = truth_2["RD"], seed_offset = 2000,
  ps_clip = analysis_lock$ps_truncation)

# --- Scenario 3: Manuscript demonstration — overlap stress, near-zero renal_risk,
#     TMLE GO while traditional PS workflow fails (n = 1500) ---
sim_3 <- run_stage3_simulation(
  reps = n_reps, n = 3000,
  b_rr_g = 2.2, beta_rr1 = 0.00, beta_rr2 = 0.05, beta_rrA = 0.00,
  beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10, beta_age3 = 0.10,
  alpha_A = scen3_alpha_A, alpha_Y = scen3_alpha_Y,
  true_rd = truth_3["RD"], seed_offset = 3000,
  ps_clip = analysis_lock$ps_truncation)
```

### Overlap and matching diagnostics by scenario

Before examining estimator performance, we compute Stage 2a-style diagnostics
for each scenario using a single representative dataset.

```{r scenario-diagnostics}
compute_scenario_diagnostics <- function(n = 5000, seed = 99,
                                          b_rr_g,
                                          beta_rr1 = 0.30, beta_rr2 = 0.08,
                                          beta_rrA = 0.00,
                                          beta_ckdA = 0.30, beta_ageA = 0.18,
                                          beta_age2A = 0.10, beta_age3 = 0.12,
                                          alpha_A = -0.15, alpha_Y = -2.20,
                                          ps_clip = c(0.025, 0.975)) {
  d <- simulate_hcv_scenario(n, seed = seed,
                              b_rr_g = b_rr_g, beta_rr1 = beta_rr1,
                              beta_rr2 = beta_rr2, beta_rrA = beta_rrA,
                              beta_ckdA = beta_ckdA, beta_ageA = beta_ageA,
                              beta_age2A = beta_age2A, beta_age3 = beta_age3,
                              alpha_A = alpha_A, alpha_Y = alpha_Y)

  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = d, family = "binomial")
  ps <- fitted(g_fit)
  ps_trunc <- pmin(pmax(ps, ps_clip[1]), ps_clip[2])

  frac_extreme <- mean(ps < 0.05 | ps > 0.95)
  w <- d$A / ps_trunc + (1 - d$A) / (1 - ps_trunc)
  ess <- sum(w)^2 / sum(w^2)
  ess_frac <- ess / n

  # Run matching on this dataset
  m <- run_ps_matched(d, ps_clip)

  data.frame(
    frac_extreme = round(frac_extreme, 3),
    ess_frac     = round(ess_frac, 3),
    max_weight   = round(max(w), 1),
    matched_frac = round(ifelse(is.null(m$matched_frac), NA, m$matched_frac), 3),
    max_smd_match = round(ifelse(is.null(m$max_smd), NA, m$max_smd), 3),
    trt_prev     = round(mean(d$A), 3),
    aki_prev     = round(mean(d$Y), 3)
  )
}

diag_1 <- compute_scenario_diagnostics(b_rr_g = 0.7,
                                        beta_rr1 = 0.30, beta_rr2 = 0.08, beta_rrA = 0.00,
                                        beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10,
                                        alpha_Y = -2.20)
diag_2 <- compute_scenario_diagnostics(b_rr_g = 2.4,
                                        beta_rr1 = 1.30, beta_rr2 = 0.25, beta_rrA = 0.60,
                                        beta_ckdA = 0.20, beta_ageA = 0.10, beta_age2A = 0.05,
                                        alpha_Y = -2.00)
diag_3 <- compute_scenario_diagnostics(b_rr_g = 2.2,
                                        beta_rr1 = 0.00, beta_rr2 = 0.05, beta_rrA = 0.00,
                                        beta_ckdA = 0.30, beta_ageA = 0.18, beta_age2A = 0.10,
                                        beta_age3 = 0.10,
                                        alpha_A = scen3_alpha_A, alpha_Y = scen3_alpha_Y)

diag_tab <- rbind(
  cbind(Scenario = "1: Acceptable overlap", diag_1),
  cbind(Scenario = "2: Overlap stress, strong Q role", diag_2),
  cbind(Scenario = "3: Manuscript demo (overlap stress, near-zero rr)", diag_3)
)
# Add qualitative label
diag_tab$quality <- ifelse(diag_tab$frac_extreme < 0.15 & diag_tab$matched_frac > 0.70,
                            "Acceptable", "Poor")
kable(diag_tab, row.names = FALSE,
      caption = "Stage 2a-style diagnostics by scenario (n = 5000 representative datasets)")
```

### What the gate tests

The Stage 2b gate asks: does the TMLE pipeline meet pre-specified performance
targets?

- **GO** means TMLE bias is below the threshold, CI coverage is at or above
  the minimum, and SE calibration is acceptable — the pipeline is adequate.
- **FLAG** means bias and coverage are acceptable but SE calibration is
  borderline — proceed with caution.
- **STOP** means TMLE does not meet the primary targets — the pipeline is
  inadequate as specified.

The interpretation below follows the actual results from this tutorial run.

```{r gate-function}
gate_check <- function(metrics, scenario_name, targets) {
  tmle <- metrics[metrics$method == "TMLE", ]

  bias_ok <- abs(tmle$bias) < targets$max_abs_bias
  cov_ok  <- tmle$coverage >= targets$min_coverage
  se_sd_ratio <- tmle$mean_se / tmle$emp_sd
  se_cal  <- se_sd_ratio > targets$se_sd_low & se_sd_ratio < targets$se_sd_high

  gate_tab <- data.frame(
    Metric = c("TMLE |bias|", "TMLE coverage", "SE/SD ratio"),
    Target = c(paste("<", targets$max_abs_bias),
               paste(">=", targets$min_coverage),
               paste(targets$se_sd_low, "-", targets$se_sd_high)),
    Observed = round(c(abs(tmle$bias), tmle$coverage, se_sd_ratio), 4),
    Status = c(ifelse(bias_ok, "GO", "STOP"),
               ifelse(cov_ok, "GO", "STOP"),
               ifelse(se_cal, "GO", "FLAG"))
  )

  if (bias_ok & cov_ok & se_cal) {
    decision <- "GO"
  } else if (bias_ok & cov_ok) {
    decision <- "FLAG"
  } else {
    decision <- "STOP"
  }

  list(decision = decision, table = gate_tab, scenario = scenario_name)
}
```

### Scenario 1: Acceptable overlap, all workflows viable

```{r scenario1-results, results='asis'}
cat("**True RD:**", round(truth_1["RD"], 5), "\n\n")
kable(sim_1$metrics[, c("method", "n_ok", "bias", "rmse", "emp_sd", "mean_se", "coverage")],
      digits = 4, row.names = FALSE,
      caption = "Scenario 1: Acceptable overlap (b_rr = 0.7), cubic age + A:ckd, A:age, A:age^2 heterogeneity")

gate_1 <- gate_check(sim_1$metrics, "Scenario 1", analysis_lock$sim_targets)
cat("\n")
kable(gate_1$table, row.names = FALSE, caption = "Scenario 1 gate assessment")
cat("\n**Gate decision:**", gate_1$decision, "\n\n")

# Decision summary
tmle_1 <- sim_1$metrics[sim_1$metrics$method == "TMLE", ]
psm_1  <- sim_1$metrics[sim_1$metrics$method == "PS-Matched", ]
iptw_1 <- sim_1$metrics[sim_1$metrics$method == "IPTW", ]
cat("**Workflow comparison:** With acceptable overlap, the conventional PS workflow ",
    "(matching: RMSE = ", round(psm_1$rmse, 4),
    "; IPTW: RMSE = ", round(iptw_1$rmse, 4), ") performed well. ",
    "The TMLE workflow (RMSE = ", round(tmle_1$rmse, 4),
    ") benefited from modeling outcome structure — cubic age effects, ",
    "treatment-by-CKD, treatment-by-age, and treatment-by-age^2 interactions — ",
    "that the conventional PS workflow cannot exploit. ",
    "Under acceptable overlap, both the conventional PS (gated by Stage 2a) and TMLE ",
    "(gated by Stage 2b) workflows are viable. ",
    "A simulation-selected TMLE workflow would confirm that all candidates meet the ",
    "gate and select the best-performing specification.\n", sep = "")
```

### Scenario 2: Overlap stress, all workflows fail → STOP

```{r scenario2-results, results='asis'}
cat("**True RD:**", round(truth_2["RD"], 5), "\n\n")
kable(sim_2$metrics[, c("method", "n_ok", "bias", "rmse", "emp_sd", "mean_se", "coverage")],
      digits = 4, row.names = FALSE,
      caption = "Scenario 2: Overlap stress (b_rr = 2.4), strong renal_risk (beta_rr1=1.30) + A:renal_risk interaction")

gate_2 <- gate_check(sim_2$metrics, "Scenario 2", analysis_lock$sim_targets)
cat("\n")
kable(gate_2$table, row.names = FALSE, caption = "Scenario 2 gate assessment")
cat("\n**Gate decision:**", gate_2$decision, "\n\n")

# Decision summary
tmle_2 <- sim_2$metrics[sim_2$metrics$method == "TMLE", ]
cat("**Workflow comparison:** Matching feasibility was reduced (retained only ",
    round(100 * diag_2$matched_frac), "% of treated). ", sep = "")
if (gate_2$decision == "STOP") {
  cat("No workflow met the Stage 2b gate. ",
      "TMLE: |bias| = ", round(abs(tmle_2$bias), 4),
      ", coverage = ", round(tmle_2$coverage, 3),
      ". When `renal_risk` strongly predicts both treatment and outcome, ",
      "neither the conventional PS workflow nor any TMLE specification ",
      "can adequately compensate for the positivity stress. ",
      "A simulation-selected TMLE workflow would likewise find no candidate ",
      "meeting the gate thresholds.\n", sep = "")
} else {
  cat("TMLE gate returned ", gate_2$decision, " with |bias| = ",
      round(abs(tmle_2$bias), 4), " and coverage = ", round(tmle_2$coverage, 3),
      ". Consult the gate criteria to determine whether this meets the pre-specified ",
      "performance targets.\n", sep = "")
}
```

### Scenario 3: Manuscript demonstration — traditional PS fails, TMLE adequate → GO

```{r scenario3-results, results='asis'}
cat("**True RD:**", round(truth_3["RD"], 5), "\n\n")
kable(sim_3$metrics[, c("method", "n_ok", "bias", "rmse", "emp_sd", "mean_se", "coverage")],
      digits = 4, row.names = FALSE,
      caption = "Scenario 3 (manuscript demo): Overlap stress (b_rr = 2.2), near-zero renal_risk (beta_rr1=0.00), n=1500")

gate_3 <- gate_check(sim_3$metrics, "Scenario 3", analysis_lock$sim_targets)
cat("\n")
kable(gate_3$table, row.names = FALSE, caption = "Scenario 3 gate assessment")
cat("\n**Gate decision:**", gate_3$decision, "\n\n")

# Decision summary
tmle_3 <- sim_3$metrics[sim_3$metrics$method == "TMLE", ]
psm_3 <- sim_3$metrics[sim_3$metrics$method == "PS-Matched", ]
iptw_3 <- sim_3$metrics[sim_3$metrics$method == "IPTW", ]
cat("**Workflow comparison (manuscript demonstration):** ",
    "This scenario uses DGP parameters identified through a systematic grid search ",
    "to illustrate the central Stage 2a vs Stage 2b contrast. ",
    "Stage 2a flagged overlap concerns (matching feasibility reduced to only ",
    round(100 * diag_3$matched_frac), "% of treated). ", sep = "")
if (gate_3$decision == "GO" || gate_3$decision == "FLAG") {
  cat("The conventional PS workflow (gated by Stage 2a) failed to meet performance ",
      "targets (PS-Matched bias = ", round(psm_3$bias, 4),
      ", coverage = ", round(psm_3$coverage, 3),
      "; IPTW bias = ", round(iptw_3$bias, 4),
      ", coverage = ", round(iptw_3$coverage, 3), "). ",
      "However, Stage 2b plasmode simulation showed the TMLE workflow ",
      "(bias = ", round(tmle_3$bias, 4),
      ", RMSE = ", round(tmle_3$rmse, 4),
      ", coverage = ", round(tmle_3$coverage, 3),
      ") met all gate criteria. ",
      "Because `renal_risk` has near-zero coefficients in the outcome model ",
      "($\\beta_{rr1}=0.00$, $\\beta_{rr2}=0.05$, $\\beta_{rrA}=0.00$), the ",
      "TMLE workflow can compensate for the stressed treatment model by learning ",
      "the remaining outcome structure — cubic age effects and treatment-effect ",
      "heterogeneity (A:ckd, A:age, A:age$^2$) — through its flexible Q model. ",
      "A simulation-selected TMLE workflow would use these Stage 2b results to ",
      "choose the best-performing candidate specification.\n\n", sep = "")
  cat("**Decision summary for Scenario 3:**\n\n",
      "| Workflow | Feasibility path | Gate | Rationale |\n",
      "|----------|-----------------|------|-----------|\n",
      "| Conventional PS (matching/IPTW) | Stage 2a | **Would not pass** | ",
      "Overlap stress (frac extreme ≈ ", round(diag_3$frac_extreme, 3),
      ", ESS frac ≈ ", round(diag_3$ess_frac, 3),
      ") degrades PS-based estimation; bias and coverage do not meet targets |\n",
      "| TMLE (fixed-specification) | Stage 2b | **GO** | ",
      "Double-robust Q model compensates for g stress; bias, coverage, and SE calibration adequate |\n",
      "| TMLE (simulation-selected) | Stage 2b | **GO** | ",
      "Would confirm fixed-specification finding and select best candidate |\n\n", sep = "")
} else {
  cat("TMLE also failed the Stage 2b gate (|bias| = ", round(abs(tmle_3$bias), 4),
      ", coverage = ", round(tmle_3$coverage, 3),
      "). Even though `renal_risk` barely affects the outcome, finite-sample performance ",
      "did not meet the pre-specified targets at n = 1500",
      ". The team would need to consider a larger sample or richer learners.\n", sep = "")
}
```

### Sampling distributions across scenarios

```{r sim-sampling-dist, fig.width=10, fig.height=9}
plot_dat <- rbind(
  cbind(scenario = "1: Moderate overlap", sim_1$results),
  cbind(scenario = "2: Overlap stress,\nstrong Q effect", sim_2$results),
  cbind(scenario = "3: Manuscript demo\n(overlap stress, weak rr)", sim_3$results)
)
plot_dat <- plot_dat[!is.na(plot_dat$est), ]
plot_dat$method <- factor(plot_dat$method,
                          levels = c("Crude", "PS-Matched", "IPTW", "TMLE"))

truth_lines <- data.frame(
  scenario = c("1: Moderate overlap",
               "2: Overlap stress,\nstrong Q effect",
               "3: Manuscript demo\n(overlap stress, weak rr)"),
  truth = c(truth_1["RD"], truth_2["RD"], truth_3["RD"])
)

ggplot(plot_dat, aes(x = est, fill = method)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(data = truth_lines, aes(xintercept = truth),
             linewidth = 0.8, linetype = "solid") +
  facet_grid(method ~ scenario, scales = "free") +
  labs(title = "Stage 2b: Sampling Distributions Under Three DGP Scenarios",
       subtitle = "Black line = true ATE",
       x = "Estimated Risk Difference", y = "Count") +
  guides(fill = "none") +
  theme_minimal(base_size = 11)
```

### Stage 2b synthesis

```{r stage3-synthesis, results='asis'}
cat("The three scenarios tell a coherent story about how the two feasibility ",
    "paths (Stage 2a vs Stage 2b) — corresponding to the covariate-balance and ",
    "pre-outcome validity checkpoints in the Muntner et al. staging framework — ",
    "lead to different conclusions:\n\n")

cat("- **Scenario 1** (gate: ", gate_1$decision,
    "): With acceptable overlap, the Muntner-style covariate-balance checkpoint ",
    "(Stage 2a) passes, and the plasmode-simulation checkpoint (Stage 2b) ",
    "confirms that both the conventional PS and TMLE workflows perform well. ",
    "The TMLE workflow achieves comparable or better precision because its ",
    "outcome model can capture cubic age effects and treatment-effect ",
    "heterogeneity (A:ckd, A:age, A:age^2) that the conventional PS workflow ",
    "cannot exploit.\n\n",
    sep = "")

cat("- **Scenario 2** (gate: ", gate_2$decision,
    "): When overlap is stressed *and* the overlap-driving covariate ",
    "strongly predicts the outcome, no workflow is adequate. The staged review ",
    "would stop regardless of approach: Stage 2a flags the overlap concerns, ",
    "and Stage 2b confirms that neither the conventional PS workflow nor any ",
    "TMLE specification can rescue the analysis.\n\n", sep = "")

cat("- **Scenario 3** *(manuscript demonstration)* (gate: ", gate_3$decision,
    "): This scenario — identified through a systematic grid search over DGP ",
    "parameters — provides the central pedagogic example. Overlap stress is ",
    "present (`b_rr` = 2.2), but `renal_risk` has near-zero coefficients in ",
    "the outcome model. **Stage 2a flags concerns equally for both workflow ",
    "families** — matching retains only ~",
    round(100 * diag_3$matched_frac), "% of treated. ",
    "But **Stage 2b plasmode simulation reveals the critical difference**: the ",
    "TMLE workflow remains adequate because its flexible Q model can learn the ",
    "outcome structure — cubic age effects, A:ckd, A:age, and A:age$^2$ ",
    "interactions — independently of the stressed treatment model. The ",
    "traditional PS workflow fails its performance targets. This is the core ",
    "contrast: the Muntner-style covariate-balance checkpoint (Stage 2a) alone ",
    "would have flagged both workflows equally, but the plasmode-simulation ",
    "checkpoint (Stage 2b) supplies additional information before the final ",
    "comparative analysis, revealing that TMLE's double robustness provides a ",
    "viable path forward.\n\n",
    sep = "")

cat("Scenarios 2 and 3 both create overlap stress through large `renal_risk` ",
    "coefficients in the treatment model (`b_rr` = 2.4 and 2.2, respectively). ",
    "Their Stage 2a diagnostics are similarly challenged. The critical difference ",
    "is entirely in the outcome surface, which Stage 2a cannot evaluate. ",
    "**Only Stage 2b distinguishes the two cases.**\n\n")

cat("**Dual role of Stage 2b for TMLE workflows.** In a fixed-specification ",
    "TMLE workflow, Stage 2b validates a single pre-committed pipeline. In a ",
    "simulation-selected TMLE workflow, Stage 2b serves a dual role: it both ",
    "validates each candidate specification against the GO/STOP gate *and* ",
    "selects the best-performing candidate using the pre-registered selection ",
    "rule. Because the candidate set, the selection rule, and the plasmode-",
    "simulation DGPs are all locked before outcome access, this selection does ",
    "not violate the clean-room principle — the analyst never sees real outcomes ",
    "during the selection process.\n")
```

### What this tutorial run demonstrates

```{r tutorial-lesson, results='asis'}
all_go  <- (gate_1$decision %in% c("GO", "FLAG")) &&
           (gate_3$decision %in% c("GO", "FLAG"))
s2_stop <- (gate_2$decision == "STOP")

if (all_go && s2_stop) {
  cat("::: {.callout-tip}\n",
      "### Lesson from this run\n\n",
      "This tutorial run produced the intended contrast across feasibility paths:\n\n",
      "1. **Scenario 1 (GO):** Stage 2a diagnostics passed, and Stage 2b confirmed that both ",
      "the conventional PS and TMLE workflows performed well. The TMLE workflow achieved ",
      "precision gains by leveraging outcome-model learning (cubic age, A:ckd, A:age, ",
      "A:age^2) that design-only estimators cannot exploit.\n",
      "2. **Scenario 2 (STOP):** Overlap was stressed and the overlap-driving covariate ",
      "strongly predicted the outcome — no workflow was adequate. Stage 2b correctly identified ",
      "this as a genuine problem for all workflow families.\n",
      "3. **Scenario 3 — Manuscript demonstration (GO):** This is the central pedagogic example. ",
      "Stage 2a flagged overlap concerns for both workflow families equally. But Stage 2b ",
      "plasmode simulation revealed that the TMLE workflow remained adequate because ",
      "`renal_risk` barely affects the outcome, while the conventional PS workflow failed. ",
      "A workflow gated only by Stage 2a diagnostics would have either stopped both workflows ",
      "equally or proceeded with an inadequate PS-based estimator.\n\n",
      "The core lesson: **traditional feasibility diagnostics (Stage 2a) and plasmode-simulation ",
      "feasibility evaluation (Stage 2b) answer different questions.** Stage 2a evaluates ",
      "overlap. Stage 2b validates the workflow pipelines under plausible DGPs — and for a ",
      "simulation-selected TMLE workflow, also selects the best candidate specification. ",
      "Plasmode simulation is what resolves whether the analysis should GO or STOP.\n",
      ":::\n", sep = "")
} else {
  cat("::: {.callout-tip}\n",
      "### Lesson from this run\n\n",
      "The gate decisions (Scenario 1: ", gate_1$decision,
      ", Scenario 2: ", gate_2$decision,
      ", Scenario 3: ", gate_3$decision,
      ") show that Stage 2b provides information beyond Stage 2a. ",
      "The three scenarios distinguish settings where different workflow families ",
      "succeed or fail.\n",
      ":::\n", sep = "")
}
```

::: {.callout-important}
### The core methodological point

Stage 2a evaluates $g$ alone — propensity scores, overlap, matching
feasibility. Stage 2b evaluates the workflow pipelines under known truth via
plasmode simulation — and, for a simulation-selected TMLE workflow, selects
the best candidate specification. Without Stage 2b, the team can only assess
whether the propensity score behaves well — not whether any workflow produces
valid estimates. The manuscript demonstration (Scenario 3) makes this
concrete: Stage 2a flags both workflow families equally, but Stage 2b reveals
that the TMLE workflow remains adequate while the traditional PS workflow
fails — a distinction that Stage 2a diagnostics alone cannot make. Plasmode
simulation is what turns a clean-room design into a credible analysis,
regardless of which workflow family is used.
:::

### When might TMLE be preferred over the other workflows?

TMLE is not uniformly superior to the conventional PS workflow. When the
outcome surface is simple and overlap is adequate, matching and IPTW may
perform comparably. TMLE may be a sensible default when the outcome surface
contains nonlinearities or treatment-effect heterogeneity that can be learned
well by a pre-committed $Q$ model — and when it can exploit richer nuisance
learning while remaining doubly robust. The conventional PS workflow relies on
a simpler design-stage adjustment strategy that cannot exploit outcome
structure.

The advantage of TMLE can persist even when overlap is stressed, *provided*
the overlap-driving covariate is not itself a major outcome driver. In
Scenario 3 (the manuscript demonstration), `renal_risk` drives treatment
strongly (`b_rr` = 2.2) but has near-zero coefficients in the outcome model
($\beta_{rr1}=0.00$, $\beta_{rr2}=0.05$), allowing the TMLE workflow to
compensate via its flexible Q model. The traditional PS workflow — which
relies entirely on the treatment model — fails under this overlap stress.
In Scenario 2, by contrast, `renal_risk` drives both treatment and outcome,
leaving all workflows without a reliable path to recovery. This distinction
cannot be assessed from Stage 2a overlap diagnostics alone. Stage 2b plasmode
simulation is what resolves it.

In clean-room workflows, each workflow's specification is pre-committed at
Stage 1 and validated via its respective feasibility path (Stage 2a or 2b),
so any precision gain is achieved without data-adaptive outcome model
selection. The simulation-selected TMLE workflow extends this by prespecifying
a candidate set and using Stage 2b plasmode simulation to choose the
best-performing specification — still entirely before outcome access.


# A Brief Intuition for the Efficient Influence Function {#eif-intuition}

Before presenting the Stage 3 results and inference formulas, it helps to
have a brief intuition for the **efficient influence function (EIF)**, which
underlies TMLE's standard errors.

The EIF summarizes **how each observation contributes to the estimator**. For
the ATE, the EIF for observation $i$ is:

$$D_i = \frac{A_i}{\hat{g}(W_i)}\bigl(Y_i - \bar{Q}^*(1, W_i)\bigr)
      - \frac{1 - A_i}{1 - \hat{g}(W_i)}\bigl(Y_i - \bar{Q}^*(0, W_i)\bigr)
      + \bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i) - \hat\psi$$

Three key facts:

1. **TMLE's targeting step ensures that the sample mean of the EIF is
   approximately zero.** This is the "solved estimating equation" property —
   it means the estimator is locally efficient.

2. **The variance of the EIF gives the standard error.** Specifically,
   $\widehat{\text{SE}} = \sqrt{n^{-1}\widehat{\text{Var}}(D)}$.
   This is sometimes called the "IC-based" or "sandwich" variance estimator.

3. **The EIF combines information from both models.** The first two terms
   use $g$ through the weights $A/\hat{g}$ and $(1-A)/(1-\hat{g})$, and $Q$
   through the residuals $Y - \bar{Q}^*$. The third term uses $Q$ alone. If
   $Q$ is perfect, the residuals are zero and $g$ drops out. If $g$ is perfect,
   the weights are accurate and $Q$ errors are corrected. This is double
   robustness expressed through the influence function.

This is all you need to know to interpret the Stage 3 inference: the SE comes
from the variance of the EIF across observations, and the CI is
$\hat\psi \pm 1.96 \times \widehat{\text{SE}}$.


# Stage 3. Unblinded Estimation {#stage-3}

## What happens at this stage

The analyst now receives the full dataset including outcomes $Y$. This is the
**first point** at which any outcome model is estimated on real data. Each of
the three precommitted workflows is now carried to completion:

- **Conventional PS workflow:** Uses its logistic-regression PS (validated at
  Stage 2a) for matching and IPTW estimation on the real outcome.
- **Fixed-specification TMLE workflow:** Fits $Q(A,W)$ on real data for the
  first time and runs the targeting step, using the nuisance strategy validated
  at Stage 2b.
- **Simulation-selected TMLE workflow:** Uses the candidate specification
  selected at Stage 2b to fit $Q(A,W)$ and run the targeting step on the real
  data. The selected specification was determined entirely before outcome
  access.

::: {.callout-note collapse="true"}
### Operational note: Stage 3 in practice
In a real clean-room analysis, outcome unblinding would occur only after
Stage 2a or 2b results have been reviewed and the feasibility decision
documented. The Stage 3 analysis script would be pre-specified (or at least
minimally modifiable) — typically the same code validated in Stage 2b (for
TMLE workflows) or the PS-based estimator locked at Stage 2a, applied to the
real dataset. Any deviations from the locked analysis plan triggered by
unexpected findings (e.g., convergence issues, data quality problems) would
need to be documented as formal amendments and reported alongside the primary
results.
:::

### Prerequisites

Stage 3 proceeds only after:

1. Stage 1 analysis lock is documented and versioned
2. Stage 2a diagnostics are reviewed and recorded (conventional PS workflow
   gated here)
3. Stage 2b plasmode-simulation gate is evaluated for TMLE workflows (if GO,
   proceed; if STOP, document the concern and any plan amendment)

### Run TMLE on the real data

```{r stage4-tmle}
run_stage4_tmle <- function(dat, lock) {
  cat("Stage 3: Outcome column now accessible.\n")
  cat("AKI prevalence:", round(mean(dat$Y), 3), "\n")
  cat("Treatment prevalence:", round(mean(dat$A), 3), "\n\n")

  result <- run_tmle_once(dat, ps_clip = lock$ps_truncation)

  cat("=== Stage 3 TMLE Result ===\n")
  cat("ATE estimate (RD):", round(result$estimate, 5), "\n")
  cat("Standard error:", round(result$se, 5), "\n")
  cat("95% CI: [", round(result$ci_low, 5), ",",
      round(result$ci_high, 5), "]\n")

  invisible(result)
}

stage4_result <- run_stage4_tmle(dat_full, analysis_lock)
```

### All three workflows on real data

We run all precommitted workflows on the real data. The results are presented
by workflow family:

```{r stage4-comparators}
stage4_crude   <- run_crude(dat_full)
stage4_iptw    <- run_iptw(dat_full, ps_clip = analysis_lock$ps_truncation)
stage4_matched <- run_ps_matched(dat_full, ps_clip = analysis_lock$ps_truncation)

est_tab <- data.frame(
  Workflow = c("(Benchmark)", "Conventional PS", "Conventional PS",
               "TMLE"),
  Estimator = c("Crude RD", "PS-Matched", "IPTW (Hajek)",
                "TMLE (fixed-specification)"),
  Estimate = round(c(stage4_crude$estimate, stage4_matched$estimate,
                     stage4_iptw$estimate,
                     stage4_result$estimate), 5),
  SE = round(c(stage4_crude$se, stage4_matched$se,
               stage4_iptw$se, stage4_result$se), 5),
  CI_low = round(c(stage4_crude$ci_low, stage4_matched$ci_low,
                   stage4_iptw$ci_low,
                   stage4_result$ci_low), 5),
  CI_high = round(c(stage4_crude$ci_high, stage4_matched$ci_high,
                    stage4_iptw$ci_high,
                    stage4_result$ci_high), 5)
)
kable(est_tab, caption = "Stage 3: Effect estimates from the workflows on the real (simulated) dataset")
```

### What just happened — step by step

Each workflow was carried to completion using its precommitted specification:

**Conventional PS workflow:** The logistic-regression PS from Stage 2a was
used for 1:1 nearest-neighbor matching (PS-Matched) and for inverse
probability weighting (IPTW). These estimators rely entirely on the treatment
model for confounding adjustment.

**TMLE workflow** (fixed-specification; a simulation-selected TMLE workflow
would use the specification chosen at Stage 2b):

1. **$Q$ was estimated on the real data** for the first time: a penalized
   logistic regression (`cv.glmnet`) with a rich feature set was fit to
   $(W, A, Y)$.

2. **$g$ was re-estimated** using the same locked strategy from Stage 2a.

3. **The targeting step** constructed the clever covariate
   $H = A/\hat{g} - (1-A)/(1-\hat{g})$ and fit a logistic fluctuation
   $\text{logit}(\bar{Q}^*) = \text{logit}(\bar{Q}) + \epsilon H$.

4. **The ATE** was computed as
   $\hat\psi = n^{-1}\sum[\bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i)]$.

5. **Inference** used the EIF as described above.

### Interpreting the results in context

```{r stage4-context, results='asis'}
cat("These Stage 3 estimates should be interpreted alongside the evidence from ",
    "the blinded feasibility stages and across all three workflows.\n\n")

if (s2_diag$any_flag) {
  cat("Stage 2a flagged overlap concerns in the propensity score distribution. ")
} else {
  cat("Stage 2a diagnostics did not flag major overlap concerns. ")
}

# The real dataset uses Scenario 3 (manuscript demo) parameters
cat("The real study dataset was generated with the same DGP as the manuscript ",
    "demonstration (Scenario 3): overlap stress (`b_rr` = 2.2), near-zero ",
    "`renal_risk` effect on outcome, n = 1500, ~10% outcome prevalence. ",
    "In practice, the analyst would not know which candidate Stage 2b scenario ",
    "is closest to reality; interpretation should weigh results across all ",
    "plausible DGPs.\n\n")

if (gate_3$decision %in% c("GO", "FLAG")) {
  cat("Stage 2b plasmode simulation (the manuscript demonstration) returned **",
      gate_3$decision,
      "** for the TMLE workflow under Scenario 3, providing pre-registered ",
      "evidence that the TMLE pipeline produces adequate finite-sample ",
      "performance under these conditions — even though Stage 2a flagged ",
      "matching feasibility concerns and the traditional PS workflow failed its ",
      "performance targets. ",
      "The Stage 3 TMLE estimate therefore carries credibility beyond a ",
      "post-hoc analysis, because the analytic decisions were locked before ",
      "outcomes were accessed. The conventional PS estimates are reported ",
      "alongside for comparison, but Stage 2b identified their limitations.\n\n",
      sep = "")
} else {
  cat("Stage 2b plasmode simulation returned **STOP** under Scenario 3, meaning ",
      "the analysis pipelines did not meet pre-specified performance targets. In a ",
      "real study, these Stage 3 estimates would be reported with caveats, and the ",
      "STOP finding documented alongside the results.\n\n")
}

cat("No analytic decisions — estimand, nuisance strategies, truncation bounds, ",
    "covariate set — were changed based on the observed outcomes.\n")
```


<details>
<summary><strong>Full Simulation Results (production-scale)</strong></summary>

# Full Simulation Results {#full-sim}

```{r full-sim-load}
out_dir <- file.path("..", "results", "scenario1_v4")
has_full_sim <- dir.exists(out_dir)
if (has_full_sim) {
  load_mode <- function(mode) {
    d <- file.path(out_dir, mode)
    tryCatch(list(
      truth   = readRDS(file.path(d, "truth.rds")),
      results = readRDS(file.path(d, "replicate_results.rds")),
      metrics = read.csv(file.path(d, "summary_overall.csv")),
      gate    = readRDS(file.path(d, "gate_results.rds")),
      params  = readRDS(file.path(d, "params_used.rds"))
    ), error = function(e) NULL)
  }
  weak   <- load_mode("weak")
  strong <- load_mode("strong")
  gate_combined <- tryCatch(
    readRDS(file.path(out_dir, "gate_combined.rds")),
    error = function(e) NULL)
} else {
  weak <- strong <- gate_combined <- NULL
}
```

The three-scenario simulation above demonstrates the staged workflow using
enriched-GLM specifications for both $g$ and $Q$. The production-scale
simulation uses richer estimators (TMLE-ML with SuperLearner, CV-TMLE,
PS-matched regression), additional sample sizes, and a mixed-overlap regime.
If those results are available, they are shown below as an extended reference.
Note that the production simulation uses a different DGP parameterization
than the tutorial scenarios, so results are not directly comparable.

```{r full-sim-results, results='asis'}
if (!is.null(weak) && !is.null(strong)) {
  cat("### Full simulation results available\n\n")
  cat("These results come from the v4 simulation with N =", weak$params$n_main,
      "and", weak$params$reps, "replicates per mode.\n\n")

  point_methods <- c("Crude", "Adj. Regression (analytic)",
                     "IPTW (analytic)", "PS-Matched (pairs boot)",
                     "TMLE-GLM (IC)", "TMLE-ML (IC)", "CV-TMLE-ML (IC)")
  show_perf <- function(metrics, label) {
    sub <- metrics[metrics$subset == "all" &
                   metrics$method %in% point_methods, ]
    sub$mode <- label
    sub[, c("mode", "method", "bias", "rmse", "coverage", "mcse_coverage")]
  }
  combined <- rbind(show_perf(weak$metrics, "Weak W4"),
                    show_perf(strong$metrics, "Strong W4"))
  kable(combined, digits = 4, row.names = FALSE,
        caption = "Full simulation: estimator performance (all replicates)")

  if (!is.null(gate_combined)) {
    cat("\n### Stop/go gate results\n\n")
    kable(gate_combined, digits = 4, row.names = FALSE,
          caption = "Gate results from full simulation")
  }
} else {
  cat("Full simulation outputs not available. ",
      "Run: `Rscript scripts/sim_scenario1_v4_false_stop_inference.R`\n")
}
```

</details>


# Common Mistakes When Operationalizing a Clean-Room Analysis {#common-mistakes}

::: {.callout-warning}
### Pitfalls to avoid

1. **Fitting any outcome model on real $Y$ before outcome access is
   permitted.** This violates the clean-room principle for all workflows.
   All outcome-model fitting on real data belongs to Stage 3.

2. **Treating Stage 2a propensity score diagnostics as if they validate the
   full analysis pipeline.** Stage 2a evaluates $g$ only. It says nothing
   about whether any workflow's outcome modeling, inference, or CI coverage
   will be adequate. Stage 2b plasmode simulation is needed to assess the
   full pipeline.

3. **Changing the nuisance strategy, truncation rule, or estimand after
   seeing real outcomes without documenting a formal amendment.** Undocumented
   post-hoc changes undermine the entire clean-room rationale for all
   workflows. All revisions must be recorded and justified.

4. **Interpreting acceptable Stage 2a overlap diagnostics as proof that any
   workflow's inference will be valid.** Good overlap is neither necessary nor
   sufficient for valid estimation. The outcome surface, sample size, and
   learner flexibility all matter — and only Stage 2b can evaluate these.

5. **Assuming poor propensity score overlap automatically invalidates TMLE
   without checking whether the outcome model can compensate.** This ignores
   TMLE's double robustness. Stage 2b plasmode simulation is the appropriate
   tool for resolving this question.

6. **Assuming TMLE requires SuperLearner or machine learning.** TMLE can use
   ordinary regression models for its nuisance estimates. Flexible nuisance
   estimation is recommended because it better supports TMLE's efficiency
   properties, but it is not a mathematical requirement.

7. **Confusing a locked modeling strategy with a requirement to reuse the
   exact same fitted nuisance object.** What is locked is the specification
   (formula, learner library, truncation bounds). Re-running the same procedure
   on the same data is acceptable; changing the specification is not.

8. **Selecting among TMLE specifications after gaining access to real
   outcomes.** In a simulation-selected TMLE workflow, the candidate set and
   selection rule must be locked at Stage 1 and the selection itself must occur
   at Stage 2b, entirely before outcome access. Choosing a TMLE specification
   after seeing real $Y$ is post-hoc model selection and violates the
   clean-room principle.

9. **Confusing Stage 2a and Stage 2b as sequential steps rather than
   alternative feasibility paths.** The conventional PS workflow is gated by
   Stage 2a (traditional diagnostics); the TMLE workflows are gated by
   Stage 2b (plasmode simulation). They are not meant to be done in sequence —
   each workflow uses the feasibility path appropriate to its estimator.

10. **Treating plasmode simulation as a replacement for all existing
    checkpoint tools.** This tutorial extends the Muntner et al. (2024) staging
    framework by adding plasmode simulation as a pre-outcome feasibility path,
    but it does not replace traditional diagnostics, negative control outcomes,
    or other residual-confounding assessments. These tools answer related but
    distinct questions and may be used in combination in a production analysis.
:::


# Key Takeaways {#takeaways}

```{r takeaways-text, results='asis'}
cat("1. **A clean-room design can accommodate multiple analysis workflows with\n",
    "   different feasibility paths.**\n",
    "   This tutorial compared three workflow families — conventional PS,\n",
    "   fixed-specification TMLE, and simulation-selected TMLE — each using a\n",
    "   different blinded feasibility path: Stage 2a (traditional diagnostics) for\n",
    "   the conventional PS workflow, or Stage 2b (plasmode simulation) for the\n",
    "   TMLE workflows.\n\n")

cat("2. **TMLE does not require SuperLearner.**\n",
    "   TMLE can use ordinary regression models for its nuisance estimates.\n",
    "   Flexible nuisance estimation is often recommended because it better supports\n",
    "   TMLE's double-robustness and efficiency properties, but it is not a\n",
    "   mathematical requirement.\n\n")

cat("3. **Traditional feasibility diagnostics (Stage 2a) and plasmode-simulation\n",
    "   feasibility evaluation (Stage 2b) answer different questions.**\n",
    "   Stage 2a evaluates overlap and matching feasibility on the observed data.\n",
    "   Stage 2b validates the workflow pipelines under plausible DGPs via\n",
    "   plasmode simulation. Acceptable Stage 2a overlap does not guarantee valid\n",
    "   inference; Stage 2a overlap stress does not necessarily invalidate TMLE\n",
    "   workflows.\n\n")

cat("4. **The conventional PS workflow relies on a simpler design-stage adjustment strategy.**\n",
    "   Scenario 1 (gate: ", gate_1$decision,
    ") showed that both the conventional PS and TMLE\n",
    "   workflows can perform well when overlap is adequate. The TMLE workflow\n",
    "   gained precision by exploiting learnable outcome structure that the\n",
    "   conventional PS workflow cannot capture.\n\n", sep = "")

cat("5. **Overlap stress plus strong confounding through the overlap driver can\n",
    "   defeat all workflows.** Scenario 2 (gate: ", gate_2$decision,
    ") showed that when\n",
    "   `renal_risk` strongly predicts both treatment and outcome, no workflow\n",
    "   family can rescue the analysis.\n\n", sep = "")

cat("6. **Under overlap stress, TMLE workflows can remain viable when the\n",
    "   conventional PS workflow fails.** The manuscript demonstration\n",
    "   (Scenario 3, gate: ", gate_3$decision,
    ") showed that Stage 2a flagged both\n",
    "   workflow families equally, but Stage 2b plasmode simulation revealed that\n",
    "   the TMLE workflow compensated for the stressed treatment model by learning\n",
    "   outcome structure through its flexible Q model, while the conventional PS\n",
    "   workflow failed. Stage 2b is necessary to determine whether this\n",
    "   compensation works at the study's sample size.\n\n", sep = "")

cat("7. **Stage 2b can serve a dual role for TMLE workflows.**\n",
    "   For a fixed-specification TMLE, Stage 2b validates a single pipeline.\n",
    "   For a simulation-selected TMLE, Stage 2b both validates candidates and\n",
    "   selects the best-performing specification — all before outcome access.\n\n")

cat("8. **Stage 3 is where all workflows fit their real outcome models.**\n",
    "   The $g$-strategy is carried forward from Stage 2a, but outcome models are\n",
    "   fit on real data only at this final stage. No analytic decisions change\n",
    "   after outcome access.\n\n")

cat("9. **This tutorial extends the Muntner et al. (2024) staging framework to\n",
    "   targeted learning.**\n",
    "   The staging and clean-room constructs of Muntner et al. — staged\n",
    "   preliminary analyses, checkpoint decisions, masked data access, and audit\n",
    "   trails — provide the organizational backbone. This tutorial generalizes\n",
    "   that logic by showing how targeted-learning workflows can be embedded in\n",
    "   the same masked checkpoint structure, with plasmode simulation used to\n",
    "   evaluate full estimator performance before unblinding.\n")
```


# Practical Checklist for a Clean-Room Analysis {#checklist}

Use this checklist to ensure each stage is properly documented and archived.

### Stage 1: Pre-specification and Analysis Lock

- [ ] Causal question and estimand clearly stated
- [ ] Eligibility criteria, treatment definition, and outcome definition
- [ ] Covariate list with variable roles (confounders, effect modifiers)
- [ ] Nuisance strategies specified for all workflows (PS model, TMLE learner
      libraries / candidate set for simulation-selected TMLE)
- [ ] Candidate TMLE specifications and prespecified selection rule (if using
      simulation-selected TMLE workflow)
- [ ] PS truncation bounds
- [ ] Stage 2a diagnostic thresholds (extreme PS fraction, ESS, max weight, SMD)
- [ ] Stage 2b performance thresholds (max bias, min coverage, SE calibration)
- [ ] Versioned analysis lock file with timestamp and hash

### Stage 2a: Traditional Feasibility Diagnostics

- [ ] Fitted $g$-strategy for PS and TMLE workflows
- [ ] Propensity score distribution summaries and overlap plots
- [ ] ESS and IPTW weight summaries
- [ ] Covariate balance diagnostics (love plot, weighted SMDs)
- [ ] Matching feasibility assessment
- [ ] List of triggered diagnostic flags
- [ ] Documented Stage 2a decision for conventional PS workflow (PASS / FLAG)

### Stage 2b: Plasmode-Simulation Feasibility Evaluation

- [ ] Pre-specified DGP description (before real outcome access)
- [ ] Plasmode simulation seeds and settings (reps, sample size, scenarios)
- [ ] Code used for simulation runs (archived and versioned)
- [ ] Bias, RMSE, empirical SD, mean SE, and coverage for each workflow
- [ ] Cross-workflow comparison (conventional PS, fixed-specification TMLE)
- [ ] Evaluation of candidate TMLE specifications and selection of final
      pipeline using the locked rule (if using simulation-selected TMLE workflow)
- [ ] Gate decision table: metric, threshold, observed value, GO/STOP
- [ ] Documented GO / FLAG / STOP decision with rationale
- [ ] Any plan amendment and its justification

### Stage 3: Unblinded Estimation

- [ ] Final outcome models fit (first time on real outcomes) for all workflows
- [ ] Point estimates and 95% CIs from each workflow
- [ ] Cross-reference to Stage 2a diagnostics and Stage 2b gate result
- [ ] Confirmation that no specifications were changed after outcome access
- [ ] Any deviations from the locked plan and their justification
- [ ] Final reproducibility metadata (session info, package versions, seed)


## Reproducing the Full Simulation

```bash
# Compact tutorial simulation: runs within this QMD (200 reps, 4 estimators)

# Grid search for the manuscript demonstration DGP (Scenario 3):
Rscript scripts/search_manuscript_demo_tmle_vs_psmatch.R

# Production simulation with SuperLearner, extended estimators, mixed overlap:
Rscript scripts/sim_scenario1_v4_false_stop_inference.R fast   # N=400, 120 reps
Rscript scripts/sim_scenario1_v4_false_stop_inference.R eval   # N=600, 300 reps
```


## References

- Muntner P, Colantonio LD, Engel Brolin EB, et al. Staging and clean-room
  analyses in observational comparative effectiveness and safety studies.
  *Journal of Clinical Epidemiology*, 2024.
- van der Laan MJ, Rose S. *Targeted Learning: Causal Inference for
  Observational and Experimental Data*. Springer, 2011.
- van der Laan MJ, Rose S. *Targeted Learning in Data Science*. Springer, 2018.
- Schuler MS, Rose S. Targeted maximum likelihood estimation for causal
  inference in observational studies. *American Journal of Epidemiology*, 2017.
- Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ. Diagnosing and
  responding to violations in the positivity assumption.
  *Statistical Methods in Medical Research*, 2012.
- Gruber S, van der Laan MJ. tmle: An R package for targeted maximum
  likelihood estimation. *Journal of Statistical Software*, 2012.
