---
title: "Simulation Scenario 1 v2: Bad Overlap, TMLE OK"
subtitle: "Point estimation vs inference under near-positivity violations"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 8
    fig-height: 5
  pdf:
    toc: true
    toc-depth: 3
    fig-width: 7
    fig-height: 4.5
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| code-fold: false

library(ggplot2)
theme_set(theme_minimal(base_size = 13))

# Load precomputed v2 simulation outputs
out_dir <- "../results/scenario1_v2"
rds_files <- c(
  params  = file.path(out_dir, "params_used.rds"),
  truth   = file.path(out_dir, "truth.rds"),
  results = file.path(out_dir, "replicate_results.rds"),
  gate    = file.path(out_dir, "gate_results.rds")
)
csv_file <- file.path(out_dir, "summary_metrics.csv")

missing <- names(rds_files)[!file.exists(rds_files)]
if (length(missing) > 0 || !file.exists(csv_file)) {
  stop("Simulation outputs not found in ", out_dir,
       "\n\nRun the simulation first:\n",
       "  Rscript scripts/sim_scenario1_bad_overlap_tmle_ok_v2.R\n",
       call. = FALSE)
}

params  <- readRDS(rds_files["params"])
truth   <- readRDS(rds_files["truth"])
res     <- readRDS(rds_files["results"])
gate    <- readRDS(rds_files["gate"])
summary <- read.csv(csv_file, stringsAsFactors = FALSE)

# Convenience: extract "all" subset metrics
sm <- summary[summary$subset == "all", ]
```

## Motivation

Propensity score--based estimators require *positivity*: every subject must
have a nonzero probability of receiving each treatment level. When this
assumption is nearly violated --- many propensity scores crowd near 0 or 1 ---
inverse probability of treatment weighted (IPTW) estimators become unstable
because a few observations receive extreme weights that dominate the estimate.
Unadjusted comparisons are biased by confounding regardless of overlap quality.

Targeted minimum loss-based estimation (TMLE) is *doubly robust*:
it is consistent if either the outcome model $Q(A,W)$ or the treatment model
$g(W)$ is correctly specified. In this scenario, TMLE point estimates remain
close to the truth despite severe practical overlap violations because the
covariate driving near-positivity has minimal influence on the outcome model.
However, standard influence-curve based Wald intervals may undercover in finite
samples under near-positivity, motivating bootstrap-based inference checks.

## Estimand

The target estimand is the average treatment effect on the risk difference
scale:

$$\text{RD} = E[Y^1] - E[Y^0]$$

where $Y^a$ is the counterfactual binary outcome under intervention $A = a$.

## Data-Generating Process

### Covariates

Four baseline covariates are generated independently:

| Variable | Distribution | Role |
|----------|-------------|------|
| $W_1$ | $N(0, 1)$ | Shared confounder (moderate in both $g$ and $Q$) |
| $W_2$ | Bernoulli(0.5) | Binary confounder |
| $W_3$ | $N(0, 1)$ | Continuous confounder |
| $W_4$ | $N(0, 1)$ | **Overlap driver**: strong in $g$, weak in $Q$ |

### Treatment mechanism (poor overlap by design)

The propensity score is generated as:

$$g(W) = \text{expit}\!\bigl(\alpha_g + s \cdot (
  `r params$coef_W1_g` \cdot W_1 +
  `r params$coef_W2_g` \cdot W_2 +
  (`r params$coef_W3_g`) \cdot W_3 +
  `r params$coef_W4_g` \cdot W_4
)\bigr)$$

where $s = `r params$strength_overlap`$ is the overlap strength multiplier and
$\alpha_g$ is chosen so that $P(A=1) \approx 0.5$. True propensity scores are
clipped to $[`r params$ps_clip_true[1]`,\; `r params$ps_clip_true[2]`]$.

The key structural feature: **$W_4$ has coefficient `r params$coef_W4_g` in the
treatment model but only `r params$coef_W4_Q` in the outcome model.** This means
the PS is extreme primarily in regions of the covariate space where the outcome
is not changing much, so $Q(A,W)$ remains smooth and learnable even where $g$ is
near 0 or 1. The scenario is constructed such that overlap violations are driven
by a covariate weakly related to the outcome.

### Outcome model (learnable)

$$Q(A,W) = \text{expit}\!\bigl(
  `r params$alpha_Q` +
  `r params$effect_A` \cdot A +
  `r params$coef_W1_Q` \cdot W_1 +
  `r params$coef_W2_Q` \cdot W_2 +
  (`r params$coef_W3_Q`) \cdot W_3 +
  `r params$coef_W4_Q` \cdot W_4
  `r if (params$nonlinear_Q) paste0("+ 0.3 \\cdot W_1^2") else ""`
\bigr)$$

The outcome model includes all four covariates with a correctly specified
GLM as one of the learners in the SuperLearner library, ensuring $Q$ is
well-estimated.

### Ground truth

```{r truth-display}
#| label: truth-display
#| code-fold: false

cat(sprintf("True RD:    %.6f\n", truth$RD_true))
cat(sprintf("E[Y^1]:     %.4f\n", truth$EY1))
cat(sprintf("E[Y^0]:     %.4f\n", truth$EY0))
cat(sprintf("Truth N:    %s\n", format(truth$truth_N, big.mark = ",")))
cat(sprintf("Method:     %s\n", truth$note))
```

## Simulation Design

```{r sim-design}
#| label: sim-design

cat(sprintf("Replicates:          %d\n", params$reps))
cat(sprintf("Sample size (N):     %d\n", params$n_main))
cat(sprintf("Overlap strength:    %.1f\n", params$strength_overlap))
cat(sprintf("PS clip (true):      [%.3f, %.3f]\n", params$ps_clip_true[1], params$ps_clip_true[2]))
cat(sprintf("PS clip (estimated): [%.2f, %.2f]\n", params$ps_clip_hat[1], params$ps_clip_hat[2]))
cat(sprintf("Targeted boot B:     %d\n", params$B_tboot))
cat(sprintf("NP boot B:           %d (0 = off)\n", params$B_npboot))
cat(sprintf("IPTW boot B:         %d\n", params$B_iptw))
cat(sprintf("Master seed:         %d\n", params$master_seed))
```

### Estimators

This simulation compares six estimators, spanning unadjusted, outcome-model
only, propensity-score only, doubly robust, and matching approaches:

| Estimator | Description | Inference |
|-----------|-------------|-----------|
| **Crude RD** | Unadjusted difference in sample proportions | Wald CI (binomial SE) |
| **Adj. Regression** | G-computation via logistic regression of $Y$ on $A + W$ with correct specification; no targeting step | Approximate SE from unit-level predictions |
| **IPTW RD** | Hajek estimator with truncated inverse probability weights | Percentile bootstrap CI (B=`r params$B_iptw`) |
| **PS-Matched Reg.** | Nearest-neighbor matching on logit(PS) with caliper, followed by logistic regression on matched sample | Robust SE on matched data |
| **TMLE (IC)** | Doubly robust TMLE with SuperLearner Q and g models | IC-based Wald CI |
| **TMLE (tboot)** | Same TMLE point estimate; variance from EIF-based Bayesian bootstrap (B=`r params$B_tboot`) | Percentile CI from targeted bootstrap |

**Crude adjusted regression** (outcome regression / g-computation) fits a
correctly specified logistic model for $Q(A,W)$ and marginalizes over the
covariate distribution to obtain the ATE. Unlike TMLE, it does not include a
targeting step to solve the efficient influence function estimating equation,
so it is not doubly robust and its variance estimate is approximate.

**PS-matched regression** performs nearest-neighbor matching on the logit of
the estimated propensity score with a caliper of $0.2 \times \text{sd}(\text{logit}(\hat{g}))$.
Under poor overlap, matching may discard many observations or produce poor
balance, limiting both precision and generalizability. Match diagnostics
(standardized mean differences before and after matching, number of
unmatched units) are recorded for each replicate.

**Targeted bootstrap** resamples the efficient influence function
contributions while holding nuisance fits fixed. Specifically, for each
of $B$ draws, Bayesian bootstrap (Dirichlet) weights $w_i^{(b)}$ are
generated and the bootstrap replicate is
$\hat\psi^{(b)} = \hat\psi + \sum_i w_i^{(b)} D_i / \sum_i w_i^{(b)}$
where $D_i$ are the TMLE EIF values. This is computationally light and
provides a complementary check on variance estimation in near-positivity
regimes.

```{r npboot-note, results='asis'}
if (params$B_npboot > 0) {
  cat("As a sensitivity analysis, we also compute a full nonparametric bootstrap ",
      "for TMLE (B = ", params$B_npboot, "), which refits nuisance models within each ",
      "resample, to assess whether resampling the full estimation pipeline materially ",
      "increases estimated uncertainty under extreme overlap.\n")
}
```

## Stage 2: Outcome-Blind Overlap Diagnostics

Before examining any treatment--outcome associations, we assess propensity
score overlap using only treatment assignment and covariates. This mirrors the
clean-room protocol where design adequacy is evaluated before unblinding
outcomes.

Under a propensity-score-only feasibility gate, these diagnostics would
typically trigger redesign or analysis termination; the purpose of this
scenario is to illustrate how outcome-blind operating characteristic simulations
can justify proceeding with a doubly robust estimator when the outcome surface
remains learnable in the regions with poor overlap.

```{r overlap-diagnostics}
#| label: overlap-diagnostics

thr <- params$overlap_flag_thresholds
cat(sprintf("Fraction of replicates flagged 'bad' overlap: %.1f%%\n",
            mean(res$overlap_flag == "bad") * 100))
cat(sprintf("\nAcross %d replicates:\n", nrow(res)))
cat(sprintf("  Mean fraction g_hat < %.2f or > %.2f:  %.3f\n",
            thr$extreme_ps, 1 - thr$extreme_ps, mean(res$fraction_extreme)))
cat(sprintf("  Mean ESS / N:                           %.3f\n",
            mean(res$ess_frac)))
cat(sprintf("  Mean max IPTW weight:                   %.1f\n",
            mean(res$max_w)))
cat(sprintf("  Median max IPTW weight:                 %.1f\n",
            median(res$max_w)))
cat(sprintf("  Mean min g_hat (raw):                   %.4f\n",
            mean(res$min_g_raw)))
cat(sprintf("  Mean max g_hat (raw):                   %.4f\n",
            mean(res$max_g_raw)))
```

### Distribution of Propensity Score Extremity

```{r ps-extreme-hist}
#| label: ps-extreme-hist
#| fig-height: 4
#| fig-cap: "Fraction of observations with estimated PS below 0.05 or above 0.95. Dashed line shows the pre-specified threshold for flagging 'bad' overlap. The scenario is designed so that W4 (weak in the outcome model) drives most of the PS extremity."

ggplot(res, aes(x = fraction_extreme)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = thr$extreme_prop,
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = thr$extreme_prop + 0.02,
           y = Inf, vjust = 2, hjust = 0, color = "red", size = 3.5,
           label = paste0("threshold = ", thr$extreme_prop)) +
  labs(title = "Fraction of observations with extreme PS per replicate",
       x = paste0("Fraction with g_hat < ", thr$extreme_ps,
                   " or g_hat > ", 1 - thr$extreme_ps),
       y = "Count")
```

### Effective Sample Size Under IPTW

```{r ess-hist}
#| label: ess-hist
#| fig-height: 4
#| fig-cap: "Effective sample size fraction (ESS/N) under IPTW ATE weights. Low ESS indicates that a few extreme weights dominate the IPTW estimate."

ggplot(res, aes(x = ess_frac)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  geom_vline(xintercept = thr$ess_frac,
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = thr$ess_frac + 0.01,
           y = Inf, vjust = 2, hjust = 0, color = "red", size = 3.5,
           label = paste0("threshold = ", thr$ess_frac)) +
  labs(title = "Effective sample size fraction (ESS / N) per replicate",
       x = "ESS / N", y = "Count")
```

### Maximum IPTW Weight

```{r max-weight-hist}
#| label: max-weight-hist
#| fig-height: 4
#| fig-cap: "Maximum IPTW weight per replicate, reflecting the most influential single observation under inverse probability weighting."

ggplot(res, aes(x = max_w)) +
  geom_histogram(bins = 30, fill = "firebrick", alpha = 0.7) +
  labs(title = "Maximum IPTW weight per replicate",
       x = "Max weight", y = "Count")
```

The diagnostics confirm that this DGP produces a *practical near-positivity
violation* in most replicates: many estimated propensity scores fall near the
boundaries, the effective sample size is substantially reduced by extreme
weights, and maximum weights are large. In a real analysis, Stage 2 would flag
these issues before any outcome data are examined.

## Stage 2b: Outcome-Blind Operating Characteristics (Stop/Go Gate)

The stop/go gate assesses whether TMLE has acceptable operating characteristics
*despite* the overlap flag, using simulation-based evidence. This is an
outcome-blind assessment: we examine only whether the estimator's bias and
coverage meet pre-specified criteria across the overlap conditions observed.

```{r stopgo-gate}
#| label: stopgo-gate
#| code-fold: false

knitr::kable(gate, digits = 4,
             caption = "Stop/go gate: TMLE operating characteristics by overlap flag status. 'Acceptable' requires |bias| < 0.01 and targeted bootstrap coverage >= 0.93.")
```

```{r gate-interpret}
#| label: gate-interpret

n_bad <- sum(res$overlap_flag == "bad")
pct_bad <- round(100 * n_bad / nrow(res), 1)
cat(sprintf("Replicates flagged 'bad' overlap: %d/%d (%.1f%%)\n\n",
            n_bad, nrow(res), pct_bad))

if (any(!is.na(gate$acceptable))) {
  bad_row <- gate[gate$overlap_flag == "bad", ]
  if (!is.na(bad_row$acceptable) && bad_row$acceptable) {
    cat("Despite the overlap flag, TMLE demonstrates acceptable operating\n")
    cat("characteristics (low bias and adequate targeted bootstrap coverage)\n")
    cat("in the flagged replicates. Under a clean-room protocol, this\n")
    cat("outcome-blind OC evidence would support proceeding with TMLE.\n")
  } else if (!is.na(bad_row$acceptable) && !bad_row$acceptable) {
    cat("TMLE operating characteristics under bad overlap do NOT meet\n")
    cat("pre-specified criteria. A clean-room protocol might require\n")
    cat("redesign, alternative estimand, or additional sensitivity analyses.\n")
  }
}
```

## Stage 3: Estimator Performance

### Summary Table

```{r summary-table}
#| label: summary-table
#| code-fold: false

knitr::kable(summary, digits = 4,
             caption = "Simulation performance metrics by estimator, CI type, and subset. Coverage is for nominal 95% CI. MCSE = Monte Carlo standard error of coverage estimate.")
```

### Point Estimation vs Inference

```{r metrics-interpret}
#| label: metrics-interpret

tmle_ic   <- sm[sm$method == "TMLE (IC)", ]
tmle_tb   <- sm[sm$method == "TMLE (tboot)", ]
crude_row <- sm[sm$method == "Crude", ]
regadj    <- sm[sm$method == "Adj. Regression", ]
iptw_row  <- sm[sm$method == "IPTW", ]
psmatch   <- sm[sm$method == "PS-Matched Reg.", ]

cat("=== TMLE (IC-based inference) ===\n")
if (nrow(tmle_ic) > 0) {
  cat(sprintf("  Bias:      %+.4f\n", tmle_ic$bias))
  cat(sprintf("  RMSE:      %.4f\n", tmle_ic$rmse))
  cat(sprintf("  Emp SD:    %.4f\n", tmle_ic$emp_sd))
  cat(sprintf("  Mean SE:   %.4f\n", tmle_ic$mean_se))
  cat(sprintf("  Coverage:  %.1f%% (MCSE: %.1f%%)\n",
              tmle_ic$coverage * 100, tmle_ic$mcse_coverage * 100))
}

cat("\n=== TMLE (targeted bootstrap) ===\n")
if (nrow(tmle_tb) > 0) {
  cat(sprintf("  Mean SE:   %.4f\n", tmle_tb$mean_se))
  cat(sprintf("  Coverage:  %.1f%% (MCSE: %.1f%%)\n",
              tmle_tb$coverage * 100, tmle_tb$mcse_coverage * 100))
}

cat("\n=== Adjusted Regression (g-computation) ===\n")
if (nrow(regadj) > 0) {
  cat(sprintf("  Bias:      %+.4f\n", regadj$bias))
  cat(sprintf("  RMSE:      %.4f\n", regadj$rmse))
  cat(sprintf("  Coverage:  %.1f%%\n", regadj$coverage * 100))
}

cat("\n=== IPTW ===\n")
if (nrow(iptw_row) > 0) {
  cat(sprintf("  Bias:      %+.4f\n", iptw_row$bias))
  cat(sprintf("  RMSE:      %.4f\n", iptw_row$rmse))
  cat(sprintf("  Coverage:  %.1f%%\n", iptw_row$coverage * 100))
}

cat("\n=== PS-Matched Regression ===\n")
if (nrow(psmatch) > 0) {
  cat(sprintf("  Bias:      %+.4f\n", psmatch$bias))
  cat(sprintf("  RMSE:      %.4f\n", psmatch$rmse))
  cat(sprintf("  Coverage:  %.1f%%\n", psmatch$coverage * 100))
  cat(sprintf("  Median matched N: %d (of %d)\n",
              median(res$psmatch_n_matched, na.rm = TRUE), params$n_main))
}

cat("\n=== Crude RD ===\n")
if (nrow(crude_row) > 0) {
  cat(sprintf("  Bias:      %+.4f\n", crude_row$bias))
  cat(sprintf("  Coverage:  %.1f%%\n", crude_row$coverage * 100))
}
```

We therefore report both the conventional influence-curve standard error and a
targeted bootstrap variance estimate that resamples the efficient influence
function contributions while holding nuisance fits fixed. This targeted
bootstrap is computationally light and provides a complementary check on
variance estimation in near-positivity regimes.

### Sampling Distribution of Estimators

```{r sampling-dist}
#| label: sampling-dist
#| fig-height: 10
#| fig-cap: "Sampling distributions of all RD estimators across simulation replicates. Vertical black line marks the true RD. Distributions centered on truth indicate low bias; wide distributions indicate high variance. TMLE is tightly centered; adjusted regression is also well-centered but may have different variance properties; IPTW shows a heavy-tailed distribution; PS-matching loses observations; crude RD is biased."

est_long <- data.frame(
  method = rep(c("TMLE", "Adj. Regression", "IPTW", "PS-Matched", "Crude"),
               each = nrow(res)),
  estimate = c(res$tmle_est, res$regadj_est, res$iptw_est,
               res$psmatch_est, res$crude_est),
  stringsAsFactors = FALSE
)
est_long <- est_long[!is.na(est_long$estimate), ]
est_long$method <- factor(est_long$method,
  levels = c("TMLE", "Adj. Regression", "PS-Matched", "IPTW", "Crude"))

ggplot(est_long, aes(x = estimate, fill = method)) +
  geom_histogram(bins = 40, alpha = 0.7) +
  geom_vline(xintercept = truth$RD_true, linewidth = 0.8) +
  annotate("text", x = truth$RD_true, y = Inf, vjust = 2, hjust = -0.1,
           label = paste0("True RD = ", round(truth$RD_true, 4)),
           fontface = "bold", size = 3.5) +
  facet_wrap(~ method, scales = "free_y", ncol = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Sampling distributions of RD estimators",
       x = "Estimated RD", y = "Count") +
  guides(fill = "none")
```

### Bias Comparison

```{r bias-plot}
#| label: bias-plot
#| fig-height: 4
#| fig-cap: "Monte Carlo bias (mean estimate minus truth) by estimator. TMLE and adjusted regression show minimal bias; IPTW may show residual bias from truncation; PS-matching is affected by sample restriction; crude RD is dominated by confounding."

methods_present <- sm$method[!is.na(sm$bias)]
bias_df <- sm[sm$method %in% methods_present & !grepl("npboot", sm$method), ]
bias_df <- bias_df[!is.na(bias_df$bias), ]
bias_df$method <- factor(bias_df$method, levels = rev(bias_df$method))

ggplot(bias_df, aes(x = method, y = bias, fill = method)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(title = "Bias by estimator",
       x = NULL, y = "Bias (mean estimate - truth)") +
  guides(fill = "none") +
  scale_fill_brewer(palette = "Set2")
```

### Coverage Comparison: IC-based vs Targeted Bootstrap

```{r coverage-plot}
#| label: coverage-plot
#| fig-height: 5
#| fig-cap: "95% CI coverage by estimator and CI type. Error bars show Monte Carlo uncertainty. Dashed line = 95% nominal level. 'TMLE (IC)' uses the influence-curve Wald CI; 'TMLE (tboot)' uses the targeted bootstrap percentile CI."

cov_df <- sm[!is.na(sm$coverage) & !grepl("npboot", sm$method), ]
cov_df$ci_lo <- cov_df$coverage - 1.96 * cov_df$mcse_coverage
cov_df$ci_hi <- cov_df$coverage + 1.96 * cov_df$mcse_coverage
cov_df$method <- factor(cov_df$method, levels = rev(cov_df$method))

ggplot(cov_df, aes(x = method, y = coverage, fill = method)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = ci_lo, ymax = pmin(ci_hi, 1)), width = 0.2) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.5, y = 0.95, label = "95% nominal",
           hjust = -0.1, vjust = -0.5, color = "grey40", size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  labs(title = "95% CI coverage by estimator",
       subtitle = "Error bars: Monte Carlo uncertainty in coverage estimate",
       x = NULL, y = "Coverage") +
  guides(fill = "none")
```

### SE Calibration: TMLE IC vs Targeted Bootstrap

A well-calibrated SE estimator should produce $\text{mean}(\widehat{\text{SE}})
\approx \text{empirical SD}$ of the point estimates. If the IC-based SE
systematically underestimates variability under near-positivity, coverage will
fall below nominal --- this is precisely the finite-sample inference limitation
that the targeted bootstrap is designed to detect.

```{r se-calibration}
#| label: se-calibration

if (nrow(tmle_ic) > 0 && !is.na(tmle_ic$mean_se) && !is.na(tmle_ic$emp_sd)) {
  ratio_ic <- tmle_ic$mean_se / tmle_ic$emp_sd
  cat(sprintf("TMLE SE calibration (IC-based):\n"))
  cat(sprintf("  Empirical SD of estimates:  %.4f\n", tmle_ic$emp_sd))
  cat(sprintf("  Mean IC-based SE:           %.4f\n", tmle_ic$mean_se))
  cat(sprintf("  Ratio (mean SE / emp SD):   %.3f\n", ratio_ic))
  if (ratio_ic < 0.90) {
    cat("  => IC SE underestimates variability; undercoverage expected\n")
  } else if (ratio_ic > 1.10) {
    cat("  => IC SE overestimates variability; CIs are conservative\n")
  } else {
    cat("  => Well-calibrated (ratio near 1.0)\n")
  }
}

if (nrow(tmle_tb) > 0 && !is.na(tmle_tb$mean_se) && !is.na(tmle_ic$emp_sd)) {
  ratio_tb <- tmle_tb$mean_se / tmle_ic$emp_sd
  cat(sprintf("\nTMLE SE calibration (targeted bootstrap):\n"))
  cat(sprintf("  Mean targeted bootstrap SE: %.4f\n", tmle_tb$mean_se))
  cat(sprintf("  Ratio (mean SE / emp SD):   %.3f\n", ratio_tb))
  if (ratio_tb < 0.90) {
    cat("  => Targeted bootstrap SE also underestimates\n")
  } else if (ratio_tb > 1.10) {
    cat("  => Targeted bootstrap SE is conservative\n")
  } else {
    cat("  => Well-calibrated (ratio near 1.0)\n")
  }
}
```

```{r se-scatter}
#| label: se-scatter
#| fig-height: 4
#| fig-cap: "TMLE estimated SE (IC-based and targeted bootstrap) vs absolute estimation error. Points near the diagonal indicate well-calibrated SEs; points consistently below the diagonal indicate SE underestimates variability."

tmle_ok <- res[!is.na(res$tmle_ok) & res$tmle_ok == TRUE, ]
if (nrow(tmle_ok) > 0) {
  se_df <- data.frame(
    se = c(tmle_ok$tmle_se_ic, tmle_ok$tmle_se_tboot),
    abs_error = rep(abs(tmle_ok$tmle_est - truth$RD_true), 2),
    type = rep(c("IC-based SE", "Targeted bootstrap SE"), each = nrow(tmle_ok))
  )
  se_df <- se_df[!is.na(se_df$se), ]

  ggplot(se_df, aes(x = se, y = abs_error, color = type)) +
    geom_point(alpha = 0.3, size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    facet_wrap(~ type) +
    labs(title = "TMLE: estimated SE vs absolute error",
         subtitle = "Points near the diagonal indicate well-calibrated SEs",
         x = "Estimated SE", y = "|Estimate - Truth|") +
    guides(color = "none") +
    scale_color_manual(values = c("IC-based SE" = "steelblue",
                                   "Targeted bootstrap SE" = "darkorange"))
}
```

### Overlap Severity and Estimator Error

```{r overlap-vs-error}
#| label: overlap-vs-error
#| fig-height: 5
#| fig-cap: "Absolute estimation error as a function of overlap severity (fraction of observations with extreme PS). Loess smooths highlight trends. Crude error is large regardless (confounding). IPTW error increases with overlap severity (weight instability). TMLE error stays low, demonstrating robustness through outcome model learning. The scenario is designed so overlap extremity is driven by W4, which is weak in Q."

if (nrow(tmle_ok) > 10) {
  err_df <- data.frame(
    fraction_extreme = rep(tmle_ok$fraction_extreme, 5),
    abs_error = c(
      abs(tmle_ok$tmle_est - truth$RD_true),
      abs(tmle_ok$regadj_est - truth$RD_true),
      abs(tmle_ok$iptw_est - truth$RD_true),
      abs(tmle_ok$psmatch_est - truth$RD_true),
      abs(tmle_ok$crude_est - truth$RD_true)
    ),
    method = rep(c("TMLE", "Adj. Regression", "IPTW", "PS-Matched", "Crude"),
                 each = nrow(tmle_ok)),
    stringsAsFactors = FALSE
  )
  err_df <- err_df[!is.na(err_df$abs_error), ]
  err_df$method <- factor(err_df$method,
    levels = c("TMLE", "Adj. Regression", "PS-Matched", "IPTW", "Crude"))

  ggplot(err_df, aes(x = fraction_extreme, y = abs_error, color = method)) +
    geom_point(alpha = 0.2, size = 1) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    scale_color_brewer(palette = "Set1") +
    labs(title = "Absolute error vs overlap severity",
         x = "Fraction of observations with extreme PS",
         y = "|Estimate - Truth|",
         color = "Estimator")
}
```

### PS Matching Diagnostics

```{r match-diagnostics}
#| label: match-diagnostics

if (any(!is.na(res$psmatch_n_matched))) {
  cat(sprintf("Matched sample size:\n"))
  cat(sprintf("  Median:  %d / %d (%.0f%%)\n",
              median(res$psmatch_n_matched, na.rm = TRUE),
              params$n_main,
              100 * median(res$psmatch_n_matched, na.rm = TRUE) / params$n_main))
  cat(sprintf("  Min:     %d\n", min(res$psmatch_n_matched, na.rm = TRUE)))
  cat(sprintf("  Max:     %d\n", max(res$psmatch_n_matched, na.rm = TRUE)))
  cat(sprintf("  Failed:  %d replicates\n", sum(is.na(res$psmatch_est))))
}
```

## Discussion

### Why TMLE point estimation succeeds here

TMLE's double robustness means it is consistent if *either* $Q$ or $g$ is
correctly specified. In this DGP:

1. The outcome model $Q(A,W)$ is a logistic regression with a mild
   nonlinearity ($0.3 W_1^2$), which is well-captured by GLM learners
   in the SuperLearner library.

2. The variable driving poor overlap ($W_4$) has a strong coefficient in
   the treatment model ($`r params$coef_W4_g`$) but a near-zero coefficient
   in the outcome model ($`r params$coef_W4_Q`$). This means the outcome
   surface is *flat* in the direction of PS extremity. The targeting step
   in TMLE, which updates the initial $Q$ estimate using the clever
   covariate $H = A/g - (1-A)/(1-g)$, can make small, well-behaved
   corrections even when $g$ is near the boundary.

### Inference under near-positivity

While TMLE point estimation is stable in this scenario, standard
influence-curve based Wald intervals may exhibit finite-sample undercoverage
when propensity scores approach the truncation boundaries. The targeting step
places nonzero weight on the clever covariate $H(A,W) = A/\hat{g}(W) -
(1-A)/(1-\hat{g}(W))$, which can be large for observations with extreme PS.
Even when the resulting bias correction is small (because $Q$ is already close
to truth), the IC-based variance estimate may not fully capture the additional
variability introduced by these observations.

We therefore report both the conventional influence-curve standard error and a
targeted bootstrap variance estimate that resamples the efficient influence
function contributions while holding nuisance fits fixed. This targeted
bootstrap is computationally light and provides a complementary check on
variance estimation in near-positivity regimes.

### Why adjusted regression also performs well (and its limitations)

The correctly specified outcome regression (g-computation) estimates the ATE
with low bias because the outcome model is simple and learnable. However,
unlike TMLE, this estimator is *not* doubly robust: if the outcome model
were misspecified (e.g., missing interactions or nonlinearities), the
g-computation estimator would be inconsistent regardless of the PS model
quality. Additionally, its variance estimation relies on asymptotic
approximations that may be less reliable than TMLE's efficient influence
function approach.

### Why IPTW and PS-matching struggle

**IPTW** relies entirely on the treatment model $g(W)$. When propensity scores
are near 0 or 1, inverse probability weights become very large for a small
number of observations, inflating variance. The effective sample size shrinks
and bootstrap CIs may under-cover because the bootstrap distribution of the
Hajek estimator can be irregular with extreme weights.

**PS-matching** under poor overlap typically discards many observations that
cannot be matched within the caliper, reducing effective sample size and
potentially introducing selection bias. Even when matches are found, the
quality may be poor (large within-pair PS differences), and the regression
adjustment on matched data cannot fully compensate for residual imbalance.

### Why crude RD fails

The unadjusted estimator ignores confounding entirely. Covariates $W_1$
through $W_4$ affect both treatment assignment and (to varying degrees) the
outcome, inducing selection bias. The confounding bias dominates the sampling
variability, producing CIs that never contain the truth.

### Implications for practice

This scenario illustrates a realistic situation in observational studies:

1. **Stage 2 diagnostics flag a problem.** A clean-room analysis detects
   poor overlap before any outcomes are examined, alerting analysts to
   the positivity concern.

2. **Stage 2b operating characteristics justify proceeding.** Outcome-blind
   simulation evidence shows that TMLE point estimation remains near-unbiased
   and inference (particularly with targeted bootstrap) remains adequate,
   supporting a decision to proceed rather than terminate.

3. **TMLE provides a path forward** when the outcome model is well-specified
   and the positivity violation is *practical* (no structural zeros). The
   analyst should invest in a rich, correctly-specified $Q$ library.

4. **Inference requires scrutiny.** Even when TMLE point estimation is
   reliable, standard IC-based inference should be supplemented with
   targeted bootstrap variance estimation under near-positivity.

5. **IPTW and matching alone are insufficient** under near-positivity
   violations. Truncation and calipers help but do not fully address
   instability.

6. **Unadjusted analyses are not a fallback.** Confounding bias makes crude
   comparisons unreliable regardless of overlap quality.

### Limitations

Because the DGP is constructed so that the primary positivity violations arise
from a covariate weakly related to the outcome, these results do not imply that
TMLE will yield reliable inference under all forms of limited overlap,
particularly when outcome prediction also depends strongly on the covariates
inducing near-deterministic treatment.

## Appendix: Replicate-Level Data

```{r replicate-table}
#| label: replicate-table

# Show first 10 replicates
display_cols <- c("rep", "event_rate", "fraction_extreme", "ess_frac",
                  "max_w", "overlap_flag",
                  "tmle_est", "regadj_est", "iptw_est",
                  "psmatch_est", "crude_est")
display_cols <- intersect(display_cols, names(res))
knitr::kable(head(res[, display_cols], 10), digits = 4,
             caption = "First 10 replicates (sample)")
```

### Session Info

```{r session-info}
#| label: session-info

si_file <- file.path(out_dir, "sessionInfo.txt")
if (file.exists(si_file)) {
  cat(readLines(si_file), sep = "\n")
} else {
  sessionInfo()
}
```
