---
title: "Scenario 1 v4: False-Stop Inference Under Near-Positivity Violations"
subtitle: "Clean-Room TMLE Simulation — NP Bootstrap Primary Inference"
author: "Clean-Room Analysis Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(ggplot2)
library(knitr)

# Locate outputs
out_dir <- file.path("..", "results", "scenario1_v4")
if (!dir.exists(out_dir)) {
  stop("Outputs not found at ", out_dir, "\n",
       "Run: Rscript scripts/sim_scenario1_v4_false_stop_inference.R")
}

# Load both modes
load_mode <- function(mode) {
  d <- file.path(out_dir, mode)
  list(
    truth   = readRDS(file.path(d, "truth.rds")),
    results = readRDS(file.path(d, "replicate_results.rds")),
    metrics = read.csv(file.path(d, "summary_overall.csv")),
    gate    = readRDS(file.path(d, "gate_results.rds")),
    params  = readRDS(file.path(d, "params_used.rds"))
  )
}

weak   <- load_mode("weak")
strong <- load_mode("strong")

gate_combined <- if (file.exists(file.path(out_dir, "gate_combined.rds")))
  readRDS(file.path(out_dir, "gate_combined.rds")) else
  rbind(cbind(mode = "weak", weak$gate), cbind(mode = "strong", strong$gate))
```

## 1. Motivation and Study Design

### 1.1 The False-Stop Problem

In observational pharmacoepidemiological studies, overlap diagnostics sometimes
flag propensity score (PS) distributions as concerning even when the estimand
remains identifiable and the chosen estimator performs well. A **false stop**
occurs when a clean-room feasibility gate halts analysis in a setting where a
doubly robust estimator like TMLE would have produced valid point estimates
*and* valid inferential performance (confidence interval coverage near nominal).

This simulation (v4) is designed to investigate whether the overlap flag
represents a genuine barrier to valid inference or a conservative false alarm,
and whether the answer depends on how strongly the overlap-driving covariate
(W4) enters the outcome model.

### 1.2 What Changed from v3

Version 3 used the Rademacher multiplier bootstrap as the primary TMLE
inferential check. The multiplier bootstrap perturbs the efficient influence
function (EIF) without refitting nuisance models, which can produce
under-coverage when nuisance estimation uncertainty is non-negligible — as is
the case at the sample sizes used in these simulations. **V4 replaces this
with a full nonparametric (NP) bootstrap** that resamples the data and refits
the entire TMLE pipeline (both Q and g nuisance models, plus the targeting
step) on each bootstrap resample. The NP bootstrap properly accounts for
nuisance estimation variability and is the primary inferential check for the
stop/go gate.

Additional v4 changes:

- **Mode-specific overlap severity:** The overlap strength multiplier $s_{bad}$
  now differs by mode (weak: `r weak$params$s_bad`, strong:
  `r strong$params$s_bad`), rather than using the same severity for both.
  This allows the weak-W4 mode to have a milder overlap violation, making a
  false-stop demonstration more realistic.
- **Moderated nonlinearity:** The outcome model's nonlinear terms are slightly
  reduced ($h_1 = 0.50 W_1 + 0.25 W_1^2$, $h_3 = 0.35 |W_3|$) to keep the
  misspecification signal within a range where TMLE-ML can achieve good
  performance.
- **Reduced treatment-effect heterogeneity (weak mode):** Interaction terms
  $\tau_{\text{int}1}$ and $\tau_{\text{int}2}$ are reduced in the weak mode,
  making the average treatment effect more homogeneous and easier to estimate.
- **Multiplier bootstrap retained** as a secondary diagnostic for comparison.

### 1.3 Estimand

The estimand is the average treatment effect on the risk-difference (RD) scale:

$$
\psi = E[Y^1] - E[Y^0]
$$

where $Y^a$ denotes the potential outcome under treatment $A = a$.

## 2. Data-Generating Process

### 2.1 Covariates

- $W_1 \sim N(0,1)$ — continuous
- $W_2 \sim \text{Bernoulli}(0.5)$ — binary
- $W_3 \sim N(0,1)$ — continuous
- $W_4 \sim N(0,1)$ — the "overlap driver"

### 2.2 Treatment Model (g)

The propensity score is generated with a treatment-strength multiplier $s_g$
that differs between overlap regimes:

$$
\text{logit}(g_0(W)) = \alpha_0 + s_g \cdot (1.5 W_1 + 0.8 W_2 - 0.5 W_3 + 3.0 W_4)
$$

where $s_g = s_{ok}$ or $s_g = s_{bad}$ with equal probability per replicate.
The intercept $\alpha_0$ is centered so that $P(A=1) \approx 0.35$, ensuring
more controls than treated (important for PS matching).

W4 has the largest coefficient in $g$ (3.0) and drives the overlap problem.

### 2.3 Outcome Model (Q)

$$
\text{logit}(Q_0(A,W)) = \beta_0 + \beta_A \cdot A +
  h_1(W_1) + h_2(W_2) + h_3(W_3) + h_4(W_4)
  + \tau_1 \cdot A \cdot I(W_1 > 0) + \tau_2 \cdot A \cdot W_2 + \tau_3 \cdot A \cdot W_4
$$

where:

- $h_1(W_1) = `r weak$params$h1_lin` \cdot W_1 + `r weak$params$h1_sq` \cdot W_1^2$ (nonlinear)
- $h_2(W_2) = `r weak$params$h2_coef` \cdot W_2$
- $h_3(W_3) = `r weak$params$h3_coef` \cdot |W_3|$ (nonlinear)
- $h_4(W_4) = \beta_{4,Q} \cdot W_4$ (mode-specific)

The key difference between modes:

```{r params-table}
params_tab <- data.frame(
  Parameter = c("s_bad", "beta4_Q", "tau_int1", "tau_int2", "tau_int3"),
  Description = c(
    "Overlap severity (bad regime)",
    "W4 coefficient in Q",
    "A × I(W1>0) interaction",
    "A × W2 interaction",
    "A × W4 interaction"
  ),
  Weak = c(weak$params$s_bad, weak$params$beta4_Q,
           weak$params$tau_int1, weak$params$tau_int2, weak$params$tau_int3),
  Strong = c(strong$params$s_bad, strong$params$beta4_Q,
             strong$params$tau_int1, strong$params$tau_int2, strong$params$tau_int3)
)
kable(params_tab, digits = 3, caption = "Mode-specific DGP parameters")
```

In the **weak mode**, W4 has a large coefficient in $g$ but barely affects
$Y$ ($\beta_{4,Q} = `r weak$params$beta4_Q`$, $\tau_3 = 0$). Overlap
diagnostics flag the PS distribution as bad, but since W4 is nearly irrelevant
to outcomes, TMLE — which adapts through both Q and g models — may still
produce valid inference. This is the candidate **false stop**.

In the **strong mode**, W4 meaningfully affects $Y$ ($\beta_{4,Q} =
`r strong$params$beta4_Q`$, $\tau_3 = `r strong$params$tau_int3`$). When
overlap is bad, the near-positivity violation creates a genuine barrier to
valid estimation and inference. This is the **true stop**.

### 2.4 Truth

```{r truth}
truth_tab <- data.frame(
  Mode = c("Weak W4", "Strong W4"),
  `True RD` = c(weak$truth$RD_true, strong$truth$RD_true),
  EY1 = c(weak$truth$EY1, strong$truth$EY1),
  EY0 = c(weak$truth$EY0, strong$truth$EY0),
  check.names = FALSE
)
kable(truth_tab, digits = 5, caption = "True ATE (RD) by mode")
```

Truth is computed via plug-in over `r format(weak$truth$truth_N, big.mark=",")`
observations. It depends on $Q_0$ and the marginal distribution of $W$, not
on $g_0$; the treatment mechanism affects only the overlap environment, not
the estimand.

## 3. Clean-Room Staged Analysis

### Stage 1: Pre-specification

The estimand ($\psi$), estimators, overlap diagnostic thresholds, and stop/go
criteria were all pre-specified before examining any outcome data.

### Stage 2: Outcome-Blind Diagnostics

PS estimation and overlap diagnostics use only $(W, A)$; $Y$ is not examined.
A replicate is flagged as "bad overlap" if any of:

- Fraction of PS outside $(0.05, 0.95)$ exceeds
  `r weak$params$overlap_flag_thresholds$extreme_prop`
- IPTW effective sample size fraction below
  `r weak$params$overlap_flag_thresholds$ess_frac`
- Maximum IPTW weight exceeds `r weak$params$overlap_flag_thresholds$max_w`

### Stage 2b: Simulation-Based Operating Characteristics

The stop/go gate uses the simulation results to assess whether TMLE-ML retains
acceptable performance in flagged replicates, specifically:

- **|Bias| < 0.01** (point estimation)
- **NP bootstrap 95% CI coverage ≥ 0.93** (inferential validity)

Both criteria must hold for the flagged stratum to "pass." In v4, the gate
criterion uses the **percentile NP bootstrap CI**, not the multiplier bootstrap.

### Stage 3: Estimation

All seven estimators are run. Inference for TMLE-ML uses both the IC-based CI
and the full NP bootstrap CI; the NP bootstrap is considered primary.

## 4. Estimators and Inference Methods

| # | Estimator | Q model | g model | Inference |
|---|-----------|---------|---------|-----------|
| 1 | Crude RD | None | None | Wald (binomial proportions) |
| 2 | Adj. Regression | Misspecified GLM | None | Analytic SE (SD of unit RDs / √n) |
| 3 | IPTW (Hajek) | None | GLM | Analytic IF-based SE |
| 4 | PS-Matched Reg. | Misspecified GLM | GLM (PS) | Pairs bootstrap |
| 5 | TMLE-GLM | GLM (misspecified) | GLM | IC-based (influence curve) |
| 6 | TMLE-ML | SuperLearner | SuperLearner | IC-based + NP bootstrap (primary) |
| 7 | CV-TMLE-ML | Ranger/GLM (V-fold) | Ranger/GLM (V-fold) | IC-based |

**Note on inference methods:**

- The **IC-based** CI uses the efficient influence function (EIF) to estimate
  variance: $\hat{\sigma}^2 = n^{-1}\sum \hat{D}_i^2$. This is
  asymptotically valid when nuisance models are well-estimated but may have
  finite-sample under-coverage.
- The **NP bootstrap** resamples the full dataset with replacement and refits
  the entire TMLE pipeline (Q, g, targeting) on each resample. It captures
  nuisance estimation uncertainty that the IC-based approach ignores in finite
  samples.
- The **multiplier bootstrap** (Rademacher) perturbs the EIF without refitting
  nuisance models: $\psi^*_b = \hat\psi + n^{-1}\sum\xi_i\hat{D}_i$. This
  is fast but inherits the same finite-sample limitations as the IC-based
  approach. It is retained here only as a secondary diagnostic.
- The **analytic SE** for adjusted regression is $\text{SD}(\hat{p}_{1,i} - \hat{p}_{0,i}) / \sqrt{n}$,
  i.e., the standard deviation of unit-level predicted risk differences.
  This does not account for model estimation uncertainty and may be
  anti-conservative.
- The **analytic IF-based SE** for IPTW uses the Hajek influence function.

## 5. Results

### 5.1 Overlap Flag Distribution

```{r overlap-flags}
flag_tab <- data.frame(
  Mode = c("Weak", "Weak", "Strong", "Strong"),
  Flag = rep(c("ok", "bad"), 2),
  N = c(sum(weak$results$overlap_flag == "ok"),
        sum(weak$results$overlap_flag == "bad"),
        sum(strong$results$overlap_flag == "ok"),
        sum(strong$results$overlap_flag == "bad"))
)
flag_tab$Pct <- round(100 * flag_tab$N / c(
  rep(nrow(weak$results), 2), rep(nrow(strong$results), 2)), 1)
kable(flag_tab, caption = "Overlap flag counts by mode")
```

### 5.2 Point Estimation Performance

```{r point-est-table}
# Show point estimators (unique on estimate, not inference method)
point_methods <- c("Crude", "Adj. Regression (analytic)",
                   "IPTW (analytic)", "PS-Matched (pairs boot)",
                   "TMLE-GLM (IC)", "TMLE-ML (IC)", "CV-TMLE-ML (IC)")

show_point <- function(metrics, label) {
  sub <- metrics[metrics$subset == "all" &
                 metrics$method %in% point_methods, ]
  sub$mode <- label
  sub[, c("mode", "method", "n_success", "bias", "rmse", "emp_sd")]
}

point_tab <- rbind(show_point(weak$metrics, "Weak"),
                   show_point(strong$metrics, "Strong"))
kable(point_tab, digits = 4, row.names = FALSE,
      caption = "Point estimation performance (all replicates)")
```

### 5.3 Inferential Performance (Coverage)

This is the core table for v4. Coverage is computed as the fraction of
replicates (with success AND non-NA CI) where the true RD falls within
the 95% CI. The Monte Carlo standard error (MCSE) quantifies simulation
uncertainty in coverage estimates.

```{r coverage-table}
show_cov <- function(metrics, label) {
  sub <- metrics[metrics$subset == "all" &
                 !is.na(metrics$coverage), ]
  sub$mode <- label
  sub[, c("mode", "method", "n_ci", "mean_se", "coverage", "mcse_coverage")]
}

cov_tab <- rbind(show_cov(weak$metrics, "Weak"),
                 show_cov(strong$metrics, "Strong"))
kable(cov_tab, digits = 4, row.names = FALSE,
      caption = "Coverage by inference method (all replicates)")
```

### 5.4 Coverage by Overlap Stratum

```{r coverage-by-overlap}
show_cov_strat <- function(metrics, label) {
  sub <- metrics[metrics$subset %in% c("overlap_ok", "overlap_bad") &
                 !is.na(metrics$coverage), ]
  sub$mode <- label
  sub[, c("mode", "subset", "method", "n_ci", "coverage", "mcse_coverage")]
}

cov_strat <- rbind(show_cov_strat(weak$metrics, "Weak"),
                   show_cov_strat(strong$metrics, "Strong"))

# Focus on TMLE-ML methods
tmle_cov <- cov_strat[grepl("TMLE-ML", cov_strat$method), ]
if (nrow(tmle_cov) > 0) {
  kable(tmle_cov, digits = 4, row.names = FALSE,
        caption = "TMLE-ML coverage by overlap stratum and inference method")
}
```

### 5.5 Stop/Go Gate Results

```{r gate-results}
kable(gate_combined, digits = 4, row.names = FALSE,
      caption = "Stop/go gate results (NP bootstrap-based)")
```

**Interpretation:**

The gate uses NP bootstrap CI coverage as the primary criterion. For the gate
to pass ("acceptable = TRUE"), the flagged stratum must have |bias| < 0.01 and
NP bootstrap coverage ≥ 0.93.

- **Weak-W4, bad-overlap replicates:** If the gate passes here, this
  demonstrates a false stop — the overlap flag was conservative, and TMLE-ML
  inference remained valid despite the PS difficulties.
- **Strong-W4, bad-overlap replicates:** If the gate fails here, this
  demonstrates a true stop — the overlap violation genuinely degraded
  inferential performance.

::: {.callout-note}
Whether these results constitute a convincing false-stop vs. true-stop
demonstration depends on the observed coverage values and their MCSEs. Coverage
estimates within $\pm 2 \times \text{MCSE}$ of 0.95 are consistent with nominal
performance. Values substantially below 0.93 suggest genuine inferential
degradation.
:::

## 6. Figures

### 6.1 Sampling Distributions

::: {.panel-tabset}

#### Weak W4

```{r fig-samp-weak, fig.width=9, fig.height=14}
methods <- c("Crude", "Adj.Reg", "IPTW", "PS-Match",
             "TMLE-GLM", "TMLE-ML", "CV-TMLE")
ests_w <- list(weak$results$crude_est, weak$results$regadj_est,
               weak$results$iptw_est, weak$results$psmatch_est,
               weak$results$tmle_glm_est, weak$results$tmle_ml_est,
               weak$results$cvtmle_est)
est_long_w <- do.call(rbind, lapply(seq_along(methods), function(i) {
  data.frame(method = methods[i], estimate = ests_w[[i]],
             overlap_flag = weak$results$overlap_flag, stringsAsFactors = FALSE)
}))
est_long_w <- est_long_w[!is.na(est_long_w$estimate), ]
est_long_w$method <- factor(est_long_w$method, levels = methods)

ggplot(est_long_w, aes(x = estimate, fill = overlap_flag)) +
  geom_histogram(bins = 35, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = weak$truth$RD_true, linewidth = 0.8) +
  facet_wrap(~ method, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(ok = "steelblue", bad = "firebrick")) +
  labs(title = "Sampling distributions (Weak W4)",
       subtitle = paste0("True RD = ", round(weak$truth$RD_true, 4)),
       x = "Estimated RD", y = "Count", fill = "Overlap") +
  theme_minimal(base_size = 11)
```

#### Strong W4

```{r fig-samp-strong, fig.width=9, fig.height=14}
ests_s <- list(strong$results$crude_est, strong$results$regadj_est,
               strong$results$iptw_est, strong$results$psmatch_est,
               strong$results$tmle_glm_est, strong$results$tmle_ml_est,
               strong$results$cvtmle_est)
est_long_s <- do.call(rbind, lapply(seq_along(methods), function(i) {
  data.frame(method = methods[i], estimate = ests_s[[i]],
             overlap_flag = strong$results$overlap_flag, stringsAsFactors = FALSE)
}))
est_long_s <- est_long_s[!is.na(est_long_s$estimate), ]
est_long_s$method <- factor(est_long_s$method, levels = methods)

ggplot(est_long_s, aes(x = estimate, fill = overlap_flag)) +
  geom_histogram(bins = 35, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = strong$truth$RD_true, linewidth = 0.8) +
  facet_wrap(~ method, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(ok = "steelblue", bad = "firebrick")) +
  labs(title = "Sampling distributions (Strong W4)",
       subtitle = paste0("True RD = ", round(strong$truth$RD_true, 4)),
       x = "Estimated RD", y = "Count", fill = "Overlap") +
  theme_minimal(base_size = 11)
```

:::

### 6.2 Coverage Comparison

```{r fig-coverage, fig.width=10, fig.height=8}
cov_all <- rbind(
  cbind(mode = "Weak W4", weak$metrics[weak$metrics$subset == "all" &
                                        !is.na(weak$metrics$coverage), ]),
  cbind(mode = "Strong W4", strong$metrics[strong$metrics$subset == "all" &
                                            !is.na(strong$metrics$coverage), ])
)

if (nrow(cov_all) > 0) {
  cov_all$ci_lo <- pmax(cov_all$coverage - 1.96 * cov_all$mcse_coverage, 0)
  cov_all$ci_hi <- pmin(cov_all$coverage + 1.96 * cov_all$mcse_coverage, 1)

  ggplot(cov_all, aes(x = method, y = coverage, fill = mode)) +
    geom_col(position = "dodge", alpha = 0.8, width = 0.7) +
    geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                  position = position_dodge(0.7), width = 0.2) +
    geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
    coord_flip() + ylim(0, 1) +
    scale_fill_manual(values = c("Weak W4" = "steelblue",
                                  "Strong W4" = "firebrick")) +
    labs(title = "95% CI Coverage by Inference Method and Mode",
         x = NULL, y = "Coverage", fill = "Mode") +
    theme_minimal(base_size = 11)
}
```

### 6.3 NP Bootstrap vs IC Coverage (TMLE-ML)

```{r fig-np-vs-ic, fig.width=8, fig.height=6}
np_ic <- rbind(
  cbind(mode = "Weak", weak$metrics[grepl("TMLE-ML \\(IC\\)|TMLE-ML \\(NP boot pctl\\)",
                                          weak$metrics$method) &
                                     weak$metrics$subset %in% c("overlap_ok", "overlap_bad") &
                                     !is.na(weak$metrics$coverage), ]),
  cbind(mode = "Strong", strong$metrics[grepl("TMLE-ML \\(IC\\)|TMLE-ML \\(NP boot pctl\\)",
                                              strong$metrics$method) &
                                         strong$metrics$subset %in% c("overlap_ok", "overlap_bad") &
                                         !is.na(strong$metrics$coverage), ])
)

if (nrow(np_ic) > 0) {
  np_ic$ci_lo <- pmax(np_ic$coverage - 1.96 * np_ic$mcse_coverage, 0)
  np_ic$ci_hi <- pmin(np_ic$coverage + 1.96 * np_ic$mcse_coverage, 1)
  np_ic$label <- paste(np_ic$mode, np_ic$subset, sep = " / ")

  ggplot(np_ic, aes(x = label, y = coverage, fill = method)) +
    geom_col(position = "dodge", alpha = 0.8, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                  position = position_dodge(0.6), width = 0.2) +
    geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
    ylim(0, 1) +
    labs(title = "TMLE-ML: NP Bootstrap vs IC Coverage by Overlap Stratum",
         x = NULL, y = "Coverage", fill = "Inference") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## 7. Discussion

### 7.1 Point Estimation

TMLE-ML and CV-TMLE-ML with flexible SuperLearner libraries are expected to
show smaller bias than misspecified estimators (crude, adjusted regression,
IPTW), particularly in the bad-overlap stratum of the strong-W4 mode. The
degree of bias improvement depends on how well the SuperLearner library
approximates the true nonlinear outcome surface.

### 7.2 Inferential Validity

The central question of v4 is whether TMLE-ML CIs maintain nominal coverage
under the flagged overlap conditions:

- **NP bootstrap CIs** should provide more reliable coverage than IC-based CIs
  at these sample sizes, since they account for nuisance estimation uncertainty.
- **IC-based CIs** may show under-coverage, especially in the bad-overlap
  stratum, because the influence-function variance estimator does not capture
  finite-sample variability in nuisance model estimation.
- **Multiplier bootstrap CIs** inherit the same limitation as IC-based CIs and
  are expected to show similar coverage patterns.

### 7.3 False Stop vs True Stop

If the simulation parameters are well-tuned:

- **Weak-W4, bad overlap:** TMLE-ML NP bootstrap coverage should remain
  near 0.95 (± MCSE), demonstrating that the overlap flag was a false alarm.
  The gate should show "acceptable = TRUE."
- **Strong-W4, bad overlap:** TMLE-ML coverage should drop noticeably below
  nominal, demonstrating that the overlap violation genuinely undermines
  inference. The gate should show "acceptable = FALSE."

::: {.callout-warning}
The false-stop result is not guaranteed by the DGP alone — it depends on the
specific combination of overlap severity, sample size, outcome model complexity,
and SuperLearner library richness. If the NP bootstrap coverage in the
weak-W4 / bad-overlap stratum falls below 0.93, the DGP parameters may need
further tuning (e.g., reducing $s_{bad,weak}$ or increasing $N$).
:::

### 7.4 Why the Multiplier Bootstrap Can Under-Perform

The Rademacher multiplier bootstrap constructs pseudo-samples as:
$$\psi^*_b = \hat\psi + \frac{1}{n}\sum_{i=1}^n \xi_i \hat{D}_i, \quad \xi_i \overset{iid}{\sim} \text{Rademacher}$$

This approximates the distribution of $\hat\psi - \psi_0$ using the estimated
EIF $\hat{D}_i$ but treats the nuisance estimates as fixed. When nuisance
estimation contributes non-negligible variability to $\hat\psi$ — as happens
at moderate $n$ with ML learners — the multiplier bootstrap underestimates
the true sampling variability. The NP bootstrap avoids this by re-estimating
nuisance models on each resample.

## 8. Limitations

1. **Finite-sample simulation:** Coverage estimates are subject to Monte Carlo
   error (reported as MCSE). With `r weak$params$reps` replicates, a true
   coverage of 0.95 has MCSE $\approx$
   `r round(sqrt(0.95*0.05/weak$params$reps), 3)`.

2. **SuperLearner library:** The library
   (`r paste(weak$params$Q_SL_library_ml, collapse=", ")`) is deliberately
   kept small for computational feasibility. Richer libraries (GAMs, deeper
   forests, neural nets) might improve or worsen finite-sample performance.

3. **Single scenario family:** Results are specific to this DGP. Generalization
   to other covariate structures, treatment mechanisms, or outcome models
   requires additional scenarios.

4. **NP bootstrap resamples:** With `r weak$params$B_npboot_tmle` resamples,
   the bootstrap CI itself has some Monte Carlo noise. Increasing
   `B_npboot_tmle` would reduce this at the cost of runtime.

5. **Analytic SEs for non-TMLE estimators:** The adjusted regression SE
   (SD of unit RDs / √n) and the IPTW IF-based SE are known to be
   potentially anti-conservative; their coverage should be interpreted
   cautiously.

## 9. Technical Details

```{r technical}
tech_tab <- data.frame(
  Parameter = c("N (per replicate)", "Replicates", "Preset",
                "NP bootstrap B", "Multiplier bootstrap B",
                "Pairs bootstrap B", "CV-TMLE folds",
                "PS clipping (estimation)", "g intercept shift",
                "Q SL library", "g SL library"),
  Value = c(weak$params$n_main, weak$params$reps, weak$params$preset,
            weak$params$B_npboot_tmle, weak$params$B_multboot,
            weak$params$B_matchboot, weak$params$V_cf,
            paste(weak$params$ps_clip_hat, collapse = ", "),
            weak$params$g_intercept_shift,
            paste(weak$params$Q_SL_library_ml, collapse = ", "),
            paste(weak$params$g_SL_library_ml, collapse = ", "))
)
kable(tech_tab, caption = "Simulation parameters")
```
