---
title: "Simulation Study: TMLE vs Traditional Approaches"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  warning: false
  message: false
---

## Overview

This vignette presents the simulation study comparing TMLE with flexible
nuisance estimation against traditional approaches (Cox PH, IPTW,
G-computation). The study evaluates five scenarios with increasing
complexity to demonstrate how TMLE performs under realistic violations
of standard modeling assumptions.

### Scenarios

| Scenario | Description | Key Feature |
|----------|-------------|-------------|
| simple | Linear confounding, independent censoring, PH | Benchmark |
| nonlinear | Nonlinear PS and outcome models | Model misspecification |
| dep_censor | Dependent (informative) censoring | Censoring bias |
| switching | Treatment switching with washout | Intercurrent events |
| np_hazard | Non-proportional hazards (piecewise) | PH violation |

### Estimand

Primary estimand: Risk difference at 90 and 180 days.
$$
\psi(t) = E[I(T^{a=1} \le t)] - E[I(T^{a=0} \le t)]
$$

```{r setup}
#| label: setup

library(ggplot2)
library(dplyr)

sim_dir <- "../outputs/simulation"
```

## Load Results

```{r load-results}
#| label: load-results

if (!file.exists(file.path(sim_dir, "simulation_summary.csv"))) {
  cat("Simulation results not found. Run simulation first:\n")
  cat("  Rscript simulation/run_simulations.R\n")
  knitr::knit_exit()
}

summary_df <- read.csv(file.path(sim_dir, "simulation_summary.csv"),
                       stringsAsFactors = FALSE)
truth_df <- read.csv(file.path(sim_dir, "true_values.csv"),
                     stringsAsFactors = FALSE)
results_df <- read.csv(file.path(sim_dir, "simulation_results.csv"),
                       stringsAsFactors = FALSE)

knitr::kable(truth_df, digits = 6,
             caption = "True Counterfactual Risk Values by Scenario")
```

## Performance Summary

```{r summary-table}
#| label: summary-table

display_cols <- c("scenario", "time_point", "method", "mean_bias",
                  "empirical_sd", "mean_se", "rmse", "coverage",
                  "n_valid")
knitr::kable(
  summary_df[, display_cols],
  digits = 5,
  caption = "Simulation Performance Metrics by Scenario and Method"
)
```

## Bias Comparison

```{r bias-plot}
#| label: bias-plot
#| fig-width: 10
#| fig-height: 6

ggplot(summary_df, aes(x = method, y = mean_bias, fill = method)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(time_point ~ scenario, scales = "free_y") +
  labs(title = "Bias by Method and Scenario",
       x = "Method", y = "Mean Bias (Estimate - Truth)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

## RMSE Comparison

```{r rmse-plot}
#| label: rmse-plot
#| fig-width: 10
#| fig-height: 6

ggplot(summary_df, aes(x = method, y = rmse, fill = method)) +
  geom_col(position = "dodge") +
  facet_grid(time_point ~ scenario, scales = "free_y") +
  labs(title = "RMSE by Method and Scenario",
       x = "Method", y = "Root Mean Squared Error") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```

## Coverage of 95% CIs

```{r coverage-plot}
#| label: coverage-plot
#| fig-width: 10
#| fig-height: 6

coverage_df <- summary_df %>%
  filter(!is.na(coverage))

ggplot(coverage_df, aes(x = method, y = coverage, fill = method)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  geom_hline(yintercept = c(0.925, 0.975), linetype = "dotted",
             color = "gray50") +
  facet_grid(time_point ~ scenario) +
  labs(title = "95% CI Coverage by Method and Scenario",
       x = "Method", y = "Coverage",
       caption = "Red dashed line = nominal 95%. Dotted = Monte Carlo bounds.") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2") +
  ylim(0, 1)
```

## SE Calibration (Empirical vs Estimated)

```{r se-calibration}
#| label: se-calibration
#| fig-width: 8
#| fig-height: 6

se_df <- summary_df %>%
  filter(!is.na(mean_se) & !is.na(empirical_sd) & mean_se > 0)

if (nrow(se_df) > 0) {
  ggplot(se_df, aes(x = empirical_sd, y = mean_se, color = method,
                    shape = factor(time_point))) +
    geom_point(size = 3) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    labs(title = "SE Calibration: Estimated SE vs Empirical SD",
         x = "Empirical SD", y = "Mean Estimated SE",
         color = "Method", shape = "Time Point") +
    theme_minimal()
}
```

## Non-Proportional Hazards Scenario

The np_hazard scenario is designed to demonstrate Cox PH bias when
proportional hazards are violated. The DGP uses a piecewise hazard
with HR = 1.25 for t < 45 days and HR = 0.70 for t >= 45 days.

```{r np-hazard-detail}
#| label: np-hazard-detail

np_df <- summary_df %>%
  filter(scenario == "np_hazard")

if (nrow(np_df) > 0) {
  knitr::kable(
    np_df[, c("time_point", "method", "mean_bias", "rmse", "coverage")],
    digits = 5,
    caption = "Non-PH Scenario: Method Comparison"
  )
}
```

## Runtime Comparison

```{r runtime}
#| label: runtime

if ("mean_runtime" %in% names(summary_df)) {
  rt_df <- summary_df %>%
    select(scenario, time_point, method, mean_runtime) %>%
    filter(!is.na(mean_runtime))

  if (nrow(rt_df) > 0) {
    ggplot(rt_df, aes(x = method, y = mean_runtime, fill = method)) +
      geom_col() +
      facet_wrap(~ scenario, scales = "free_y") +
      labs(title = "Mean Runtime per Replicate (seconds)",
           x = "Method", y = "Seconds") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  }
}
```

## Key Findings

```{r findings}
#| label: findings

# Identify scenario where Cox is most biased vs TMLE
if (nrow(summary_df) > 0) {
  tmle_perf <- summary_df %>%
    filter(method == "TMLE") %>%
    select(scenario, time_point, tmle_bias = mean_bias, tmle_rmse = rmse)

  cox_perf <- summary_df %>%
    filter(method == "Cox PH") %>%
    select(scenario, time_point, cox_bias = mean_bias, cox_rmse = rmse)

  comparison <- merge(tmle_perf, cox_perf, by = c("scenario", "time_point"))
  comparison$tmle_advantage <- abs(comparison$cox_bias) - abs(comparison$tmle_bias)

  cat("Scenarios where TMLE has lower |bias| than Cox PH:\n")
  adv <- comparison[comparison$tmle_advantage > 0, ]
  if (nrow(adv) > 0) {
    for (i in seq_len(nrow(adv))) {
      cat(sprintf("  %s (t=%d): Cox bias=%.5f, TMLE bias=%.5f, advantage=%.5f\n",
                  adv$scenario[i], adv$time_point[i],
                  adv$cox_bias[i], adv$tmle_bias[i],
                  adv$tmle_advantage[i]))
    }
  } else {
    cat("  (Run simulation with more replicates for clearer comparison)\n")
  }
}
```

## Appendix

### Configuration

```{r appendix-config}
#| label: appendix-config

r_files <- list.files("../R", pattern = "\\.R$", recursive = TRUE,
                      full.names = TRUE)
for (f in r_files) source(f)
cfg <- load_config("../config/default.yml")
str(cfg$simulation)
str(cfg$tmle)
```

### Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```
