---
title: "cleanTMLE"
subtitle: |
  Staged, outcome-blind targeted learning
  with plasmode data-quality stress testing
author: "Andrew Mertens"
institute: "University of California, Berkeley"
date: today
format:
  revealjs:
    theme: simple
    slide-number: c/t
    progress: true
    incremental: false
    self-contained: true
    width: 1280
    height: 720
    fig-width: 8
    fig-height: 4.5
    code-line-numbers: false
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 9,
  fig.height = 4.5
)
library(ggplot2)

# ── Monte Carlo results ────────────────────────────────────────
# Prefer the archived snapshot of the prior run as a baseline,
# then splice in any newer per-scenario interim results from the
# in-progress rerun. The aggregated `simulation_results.rds` in
# results/ is only written at the very end of the rerun, so until
# then we read partial progress from `interim_<scenario>.rds`.

archive_path <- file.path("..", "archive", "results_pre_rerun_2026-04-27",
                          "simulation_results.rds")
current_path <- file.path("..", "results", "simulation_results.rds")

res <- if (file.exists(archive_path)) {
  readRDS(archive_path)
} else if (file.exists(current_path)) {
  readRDS(current_path)
} else NULL
has_mc <- !is.null(res)

# Splice in newer interim files if they cover ≥ 50 replicates
# (enough that the boxplot stops jittering visibly).
mc_data_sources <- list()
if (has_mc) {
  for (sc in names(res$results)) {
    interim_path <- file.path("..", "results", sprintf("interim_%s.rds", sc))
    if (file.exists(interim_path)) {
      interim <- readRDS(interim_path)
      n_done  <- interim$n_done %||% 0
      if (is.null(n_done)) n_done <- 0
      if (n_done >= 50 && !is.null(interim$results) &&
          nrow(interim$results) > 0) {
        res$results[[sc]] <- interim$results
        mc_data_sources[[sc]] <- sprintf(
          "rerun in progress (%d / %d reps, %s)",
          n_done, interim$n_total %||% NA,
          format(interim$saved_at, "%H:%M")
        )
      } else {
        mc_data_sources[[sc]] <- "prior run (archived 2026-04-03)"
      }
    } else {
      mc_data_sources[[sc]] <- "prior run (archived 2026-04-03)"
    }
  }
}
`%||%` <- function(a, b) if (is.null(a)) b else a

# ── DQ stress results ──────────────────────────────────────────
# DQ stress is written all-or-nothing per scenario, so we read
# whatever is currently on disk; once a scenario finishes in the
# rerun, its file is overwritten and the slide picks up the new
# gradient automatically.

dq_files <- list.files(
  file.path("..", "results"),
  pattern    = "^dq_stress_",
  full.names = TRUE
)
has_dq <- length(dq_files) > 0
dq_data_sources <- list()
if (has_dq) {
  dq_all <- lapply(dq_files, readRDS)
  names(dq_all) <- sub("dq_stress_(.+)\\.rds", "\\1", basename(dq_files))
  for (sc in names(dq_all)) {
    f <- dq_files[sub("dq_stress_(.+)\\.rds", "\\1",
                       basename(dq_files)) == sc]
    mtime <- file.info(f)$mtime
    dq_data_sources[[sc]] <- if (as.Date(mtime) >= as.Date("2026-04-27")) {
      sprintf("rerun (%s)", format(mtime, "%Y-%m-%d %H:%M"))
    } else {
      sprintf("prior run (%s)", format(mtime, "%Y-%m-%d"))
    }
  }
}
```

## RWE has to clear two hurdles {.smaller}

Before the outcome is unblinded, an RWE study has to show two
things:

1. **the analysis is valid** — the chosen estimator gives the
   right answer under the planned conditions, and
2. **the data are fit for the question** — the data source
   actually has the cohort, the exposure, and the outcome the
   target trial asks for.

These two arguments live in different documents and use different
evidence.

::: {.fragment}
- Analysis validity is checked with **outcome-blind plasmode
  simulation**, reporting bias, RMSE, and CI coverage for each
  candidate estimator [@dang2023roadmap; @gruber2022sap;
  @franklin2014plasmode].
- Data fitness is recorded in **stepwise written templates**
  such as SPIFD2 [@gatto2022spifd; @gatto2023spifd2] or
  STaRT-RWE [@wang2021startrwe].
:::

::: {.fragment .center}
The simulation is numeric. The template is qualitative.
:::

---

## The two checks meet at the gate {.smaller}

@muntner2024staging describe a staged "clean-room" workflow in
which each pre-outcome stage ends with a review-team decision:
*proceed*, *do more bias analyses*, or *stop*. We use the labels
**GO / FLAG / STOP** for these three decisions throughout.

::: {.fragment}
| Stage | What gets done | Source |
|------|----------------|--------|
| 0 | Feasibility, utilisation, draft protocol/SAP | @kent2026fiord |
| 1a | Lock the analysis plan | @muntner2024staging |
| 1b | Cohort adequacy + design precision (CP1) | @muntner2024staging |
| 2a | Propensity-score diagnostics + balance (CP2) | @muntner2024staging |
| **2b** | **Outcome-blind plasmode candidate selection** | **this paper** |
| 3 | Negative-control outcomes (CP3) | @muntner2024staging |
| Gate | Pre-outcome authorisation | @muntner2024staging |
| 4 | Primary comparative analysis | @muntner2024staging |
:::

::: {.fragment}
In the published staging workflow, the **data-fitness argument
arrives as a separate document** and is read by reviewers rather
than scored by the gate.
:::

---

## What we propose

> **Fold the data-fitness check into the same outcome-blind
> simulation that already scores the analysis check.**

::: {.fragment}
- The plasmode is under the analyst's control, so we can degrade
  it on purpose.
- For each candidate estimator, vary the severity of each
  realistic data threat and read off bias, RMSE, and coverage.
- Apply the same gate thresholds to the degraded runs as to the
  baseline plasmode.
:::

::: {.fragment .center}
*This is the bridge from SPIFD2 to GO/FLAG/STOP.*
:::

::: {.fragment}
Implemented in **`cleanTMLE`** (R package, MIT-licensed) as
`run_plasmode_dq_stress()`.
:::

---

## What the package contributes {.smaller}

Three new pieces of software:

::: {.fragment}
1. **`run_plasmode_dq_stress()`** — an outcome-blind plasmode
   loop with four data-quality (DQ) degradation mechanisms. For
   each prespecified TMLE candidate it returns a degradation
   gradient: bias, RMSE, and coverage as the threat gets worse.
:::

::: {.fragment}
2. **`gate_check()` / `gate_all()`** — take the baseline plasmode
   metrics, the DQ-stress metrics, and the stage checkpoints and
   return a single GO / FLAG / STOP decision per stage.
:::

::: {.fragment}
3. **The `cleanroom_lock`** — a hashed, saveable analysis plan
   with an audit log and a decision log. Every Stage 4 estimator
   (`fit_tmle_outcome_mechanism()`, `run_iptw_workflow()`, …)
   refuses to run until `authorize_outcome_analysis()` has
   signed off, so the lock physically enforces the workflow.
:::

::: {.fragment}
The cohort-specification grammar (`specify_models()`,
`identify_treatment()`, `identify_outcome()`, …) and the reporting
helpers (`make_table1()`, `make_table2()`, …) follow `causalRisk`
[@brookhart2024causalrisk] under MIT-licence attribution.
:::

---

## Estimand, DAG, identifiability {.smaller}

Following ICH E9(R1) [@ich2019e9r1] and @dang2023roadmap:

| Attribute | Specification |
|-----------|---------------|
| Population | Adults eligible at index |
| Treatment | Binary point exposure $A \in \{0,1\}$ |
| Endpoint | Event by 24 months ($Y$) |
| Summary | Marginal counterfactual RD, $\Psi = E\{Y(1)\} - E\{Y(0)\}$ |
| Intercurrent events | Treatment-policy strategy |

::: {.fragment}
**Statistical estimand** under exchangeability + positivity +
consistency + correctly specified nuisance models:
$$
\Psi = E_W\{E[Y \mid A=1, W] - E[Y \mid A=0, W]\}
$$
:::

::: {.fragment}
**Each DQ stress scenario probes one identifiability assumption:**
unmeasured-U → exchangeability; misclassified A → consistency;
misclassified Y → outcome measurement; covariate missingness →
stratified MCAR; truncation/library → nuisance specification.
:::

---

## Outcome-blind simulation vs. traditional clean room {.smaller}

::: {.columns}

::: {.column width="50%"}
**Traditional clean room** [@muntner2024staging]

- Two roles: blinded analyst and unblinded analyst.
- The blinded analyst writes the plan, fixes the cohort, and
  runs Stages 1–3.
- A data manager controls who sees the outcome.
- The plan and the final estimate are paper artifacts.
:::

::: {.column width="50%"}
**Outcome-blind simulation** *(this paper)*

- One analyst.
- Iteration runs on synthetic outcomes generated from a
  covariate-only $Q_0$ model — the analyst never sees the
  real treatment-outcome association.
- `mask_outcome()` and `assert_outcome_authorized()` replace
  the data manager.
- The lock and audit log are machine-readable, so the whole
  pre-outcome workflow is re-executable.
:::

:::

::: {.fragment .center}
The two are **complementary**, not competing.
:::

::: {.fragment}
What the software-only mode adds: no need to split the team into
two roles; the analyst can iterate without seeing the outcome;
the lock hash makes the workflow byte-for-byte reproducible; and
the same loop can stress-test data quality.
:::

---

## The DQ stress test — four mechanisms {.smaller}

Mapping data-quality concerns from SPIFD2 [@gatto2023spifd2] and
the harmonised data-quality dimensions of @kahn2016data to
plasmode parameters:

| Data-quality concern | Plasmode parameter |
|------------------------|--------------------|
| Covariate measurement / availability (Kahn: completeness) | Fraction $\pi$ of MCAR missingness |
| Treatment exposure definition (Kahn: conformance) | (sens, spec) of treatment ascertainment |
| Outcome ascertainment (Kahn: conformance + plausibility) | (sens, spec) of outcome ascertainment, optionally per arm |
| Unmeasured confounding (SPIFD2 sensitivity step) | Latent $U$ with prevalence $p_U$ and $(\text{OR}_A, \text{OR}_Y)$ |

::: {.fragment}
The mapping is **one-to-many**: a single SPIFD2 element can drive
several plasmode parameters, and a single plasmode parameter can
address several SPIFD2 elements.
:::

::: {.fragment}
Plausible ranges for each parameter come from data profiling,
validation studies, or expert elicitation. The chosen ranges are
recorded in the lock.
:::

---

## The degradation gradient

For each candidate the test runs a range of severity levels and
reports bias, RMSE, and coverage at each level — on the **same
units** as the baseline plasmode.

::: {.fragment}
- **Robust candidate:** a flat curve that stays inside the gate
  thresholds.
- **Fragile candidate:** a cliff — a small worsening of the data
  pushes bias or coverage past the threshold.
:::

::: {.fragment}
The candidate-selection rule can require a **maximum tolerated
degradation**, which lets the gate rule out fragile candidates
*before* the outcome is accessed.
:::

::: {.fragment .center}
*Same metrics, same gate, candidate-specific answer.*
:::

---

## Worked example — simulation design {.smaller}

- $n = 2{,}000$ per replicate, 200 Monte Carlo replicates.
- Six estimators: **Crude, IPTW, PS Match, TMLE, CV-TMLE,
  Match_TMLE** (CV-TMLE = cross-fitted TMLE in the sense of
  @zheng2011crossvalidated and @chernozhukov2018dml).
- Three scenarios:

| | Overlap | Confounding |
|-|---------|-------------|
| **A** Good overlap | PS in [0.2, 0.8] | only measured |
| **B** Marginal overlap | PS reaches [0.04, 0.99] | only measured |
| **C** Unmeasured confounding | PS in [0.2, 0.8] | $U$, $\text{OR}_A=\text{OR}_Y=2$ |

- DQ stress: covariate missingness {5%, 10%, 20%}, treatment
  misclass {(0.95, 0.99), (0.90, 0.95)}, outcome misclass
  {(0.95, 0.99), (0.90, 0.95)}, unmeasured U with
  $\text{OR} \in \{1.5, 2.0\}$.

---

## Results — Monte Carlo {.smaller}

```{r mc-boxplots, eval = has_mc, fig.cap = "Sampling distribution by estimator and scenario; dashed red line = true RD."}
raw_data <- do.call(rbind, lapply(names(res$results), function(sc) {
  df <- as.data.frame(res$results[[sc]])
  df$scenario <- res$scenarios[[sc]]$label
  df$truth    <- res$truths[[sc]]$RD
  df
}))
ggplot(raw_data, aes(x = method, y = estimate, fill = method)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  geom_hline(aes(yintercept = truth), linetype = "dashed",
             colour = "red") +
  facet_wrap(~ scenario, ncol = 3, scales = "free_y") +
  labs(x = NULL, y = "Risk-difference estimate") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1, size = 9))
```

```{r mc-fallback, eval = !has_mc, results = "asis"}
cat("*Monte Carlo results not yet available; rerun `run_simulation.R` to populate this slide.*")
```

```{r mc-source-note, eval = has_mc, results = "asis"}
parts <- vapply(names(mc_data_sources), function(sc) {
  sprintf("%s: %s", res$scenarios[[sc]]$label, mc_data_sources[[sc]])
}, character(1))
cat("<div style='font-size:60%; color:#555;'>Data sources — ",
    paste(parts, collapse = "; "), ".</div>", sep = "")
```

::: {.fragment}
**Standard-error calibration.** The IPW influence-function SE
treats the propensity score as known and is therefore
conservative when the propensity score has actually been
estimated [@robins1992rotnitzky; @brookhart2024causalrisk]. TMLE
and CV-TMLE do not share this conservativeness once the nuisance
fits are cross-fit. Crude under-covers when there is confounding;
under unmeasured $U$ the gate returns **STOP**.
:::

---

## Results — DQ degradation gradients {.smaller}

```{r dq-gradient-plot, eval = has_dq, fig.height = 4.2, fig.cap = "Bias gradient for the selected candidate at each severity level of each DQ scenario, faceted by Monte Carlo scenario."}
dq_combined <- do.call(rbind, lapply(names(dq_all), function(sc) {
  m <- dq_all[[sc]]$metrics
  m$mc_scenario <- if (has_mc && sc %in% names(res$scenarios))
    res$scenarios[[sc]]$label else sc
  m
}))
dq_plot <- dq_combined[dq_combined$scenario != "none", ]

if (has_mc) {
  sel_map <- vapply(names(res$meta),
                    function(sc) res$meta[[sc]]$selected, character(1))
  sc_label_map <- vapply(names(res$scenarios),
                         function(sc) res$scenarios[[sc]]$label, character(1))
  dq_plot <- do.call(rbind, lapply(split(dq_plot, dq_plot$mc_scenario), function(d) {
    sc_key <- names(sc_label_map)[sc_label_map == d$mc_scenario[1]]
    if (length(sc_key) > 0 && sc_key %in% names(sel_map)) {
      d[d$candidate == sel_map[sc_key], ]
    } else d
  }))
}

ggplot(dq_plot, aes(x = level, y = abs(bias),
                     colour = scenario, group = scenario)) +
  geom_point(size = 2.4) + geom_line(linewidth = 0.8) +
  facet_wrap(~ mc_scenario, scales = "free_x") +
  labs(x = "DQ severity level", y = "|Bias|", colour = "DQ scenario") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 8),
        legend.position = "bottom")
```

```{r dq-fallback, eval = !has_dq, results = "asis"}
cat("*DQ stress results not yet available; rerun `run_simulation.R` to populate this slide.*")
```

```{r dq-source-note, eval = has_dq, results = "asis"}
parts <- vapply(names(dq_data_sources), function(sc) {
  label <- if (!is.null(res) && sc %in% names(res$scenarios))
    res$scenarios[[sc]]$label else sc
  sprintf("%s: %s", label, dq_data_sources[[sc]])
}, character(1))
cat("<div style='font-size:60%; color:#555;'>Data sources — ",
    paste(parts, collapse = "; "), ".</div>", sep = "")
```

::: {.fragment}
The covariate-missingness gradient is flat. Treatment
misclassification attenuates the estimate toward the null.
Outcome misclassification gets worse as (sens, spec) drop.
Unmeasured $U$ produces the largest degradation, and is the
only threat that the standard plasmode alone cannot catch.
:::

---

## What the bridge does **not** replace {.smaller}

::: {.fragment}
**E-values** [@vanderWeele2017evalue]. *Post-hoc and not
candidate-specific.* Reports the minimum confounder strength
that would nullify the observed effect.
:::

::: {.fragment}
**Quantitative bias analysis** [@lash2014qba]. *Adjusts the
final estimate.* The DQ stress test does not adjust the
estimate; it asks whether the estimator can still pass the gate
when the bias is present.
:::

::: {.fragment}
**Causal-gap ($\delta$) sensitivity** [@diaz2013sensitivity]. The
targeted-learning sensitivity tool: it shows how $\Psi$ moves as
a hypothesised unmeasured-confounder bound varies. Used at
unblinding, not pre-outcome.
:::

::: {.fragment}
**Negative-control outcomes** [@lipsitch2010negative;
@levintow2023nco; @muntner2024staging Table 2]. *Used on the
real outcome data.* The DQ stress test cannot detect residual
confounding; the NCO check cannot detect missingness or
misclassification. They target different threats.
:::

::: {.fragment .center}
*Pre-outcome (DQ stress) → at-outcome (NCO) → post-outcome
(E-value, QBA, causal-gap).* Three sequential bias analyses.
:::

---

## Limitations + roadmap {.smaller}

::: {.fragment}
- **MCAR only** for covariate missingness. The principled
  alternative is inverse-probability-of-missingness weighting
  (`identify_missing()` in `causalRisk`); we plan to expose it
  in a future release.
:::

::: {.fragment}
- **Parametric data-generating process.** The plasmode adds a
  fixed risk difference to a logistic $Q_0$ model fitted to the
  lock data; effect modification is not modelled.
:::

::: {.fragment}
- **Single binary unmeasured confounder.** A vector or
  continuous $U$ would require a richer DGP than the current
  package supports.
:::

::: {.fragment}
- **Computational cost.** The number of TMLE fits scales with
  scenarios × severity levels × candidates × replicates. We run
  scenarios in parallel.
:::

::: {.fragment}
- **Estimand scope.** The current release targets the binary
  risk difference. Counterfactual hazard ratios via
  `estimate_ipwhr()` and survival TMLE via `survtmle` are not
  yet wrapped in the staged workflow.
:::

::: {.fragment}
- **Single trust boundary.** A software-only clean room has no
  independent role-based check. For high-stakes regulatory
  work, run cleanTMLE *inside* a traditional clean-room
  arrangement [@muntner2024staging; @kent2026fiord] so both
  the institutional and the computational guarantees apply.
:::

---

## Take-home {.smaller}

::: {.fragment}
- Outcome-blind plasmode simulation can do two jobs at once:
  pick a candidate estimator **and** stress-test it against the
  data threats SPIFD2 lists qualitatively.
:::

::: {.fragment}
- The output — bias, RMSE, and coverage gradients on the same
  units as the baseline plasmode — feeds the same GO / FLAG /
  STOP gate the rest of the pre-outcome workflow already uses.
:::

::: {.fragment}
- The result is a reproducible, hash-verified pre-outcome
  bundle: the lock, the audit log, the decision log, and the
  degradation gradient.
:::

::: {.fragment}
- Implementation: **`cleanTMLE`** (R, MIT-licensed) at
  <https://github.com/amertens/cleanTMLE>; companion simulation
  at <https://github.com/amertens/clean-room-sim>.
:::

::: {.fragment .center}
**Acknowledgements.** The `causalRisk` authors (Brookhart,
Breskin, Eledath) for the MIT-licensed cohort-specification
grammar that cleanTMLE inherits. The TL-RWE roadmap authors
(Dang, Gruber, Phillips, Lee, Ho, van der Laan, Petersen, and
colleagues) whose published frameworks cleanTMLE operationalises.
Kent and the Amgen Center for Observational Research for the
applied staging examples used in the FIORD 2026 talk.
:::

::: footer
References in `references.bib` and the companion manuscript.
:::
