---
title: "Clean-Room Staged Analysis: HCV-AKI Case Study"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  warning: false
  message: false
---

## Overview

This vignette demonstrates the end-to-end clean-room staged workflow for
analyzing the causal effect of sofosbuvir-containing (SOF) vs non-SOF DAA
regimens on acute kidney injury (AKI) risk using a simulated cohort.

The analysis proceeds through four stages with explicit checkpoints:

1. **Stage 1**: Cohort feasibility and build (outcome-blind)
2. **Stage 2**: Design adequacy checks (treatment model only; no outcomes)
3. **Stage 3**: Outcome modeling and estimation (TMLE + comparators)
4. **Stage 4**: Reporting and lock

```{r setup}
#| label: setup

# Source all R modules
r_files <- list.files("../R", pattern = "\\.R$", recursive = TRUE,
                      full.names = TRUE)
for (f in r_files) source(f)

# Load configuration
cfg <- load_config("../config/default.yml")
```

## Stage 1: Cohort Feasibility and Build

Stage 1 generates the analysis cohort and produces feasibility summaries.
**No outcome-by-treatment comparisons are permitted at this stage.**

```{r stage1}
#| label: stage1

# Generate cohort data using simple scenario
set_deterministic_seed(cfg$dgp$seed)
cohort_data <- generate_hcv_data(
  N = cfg$dgp$N,
  h0 = cfg$dgp$h0,
  HR_early = cfg$dgp$HR_early,
  HR_late = cfg$dgp$HR_late,
  tau = cfg$dgp$tau,
  max_follow = cfg$dgp$max_follow,
  np_hazard = FALSE,
  dep_censor = FALSE,
  complexity = FALSE,
  switch_on = FALSE,
  seed = cfg$dgp$seed
)

# Run Stage 1
stage1_out <- stage1_build_cohort(
  data = cohort_data,
  cfg = cfg,
  output_dir = "../outputs/stage1"
)

cat("Checkpoint 1:", stage1_out$checkpoint, "\n")
```

### Stage 1 Report

```{r stage1-report}
#| label: stage1-report

report <- stage1_out$report
cat(sprintf("Cohort size: %d\n", report$N_total))
cat(sprintf("Treated (SOF): %d (%.1f%%)\n",
            report$N_treated, report$frac_treated * 100))
cat(sprintf("Control: %d (%.1f%%)\n",
            report$N_control, report$frac_control * 100))
cat(sprintf("Overall event rate: %.4f\n", report$overall_event_rate))
cat(sprintf("Total events: %d\n", report$overall_N_events))
cat(sprintf("Number of covariates: %d\n", report$n_covariates))
```

## Stage 2: Design Adequacy Checks

Stage 2 evaluates propensity score overlap, covariate balance, and weight
distributions. **This stage does NOT fit outcome models or compute treatment
effects.**

```{r stage2}
#| label: stage2

stage2_out <- stage2_design_checks(
  cohort = stage1_out$cohort,
  cfg = cfg,
  output_dir = "../outputs/stage2"
)

cat("Checkpoint 2:", stage2_out$checkpoint, "\n")
```

### Propensity Score Overlap

```{r ps-overlap}
#| label: ps-overlap
#| fig-width: 8
#| fig-height: 5

library(ggplot2)

ps_df <- data.frame(
  ps = stage2_out$ps,
  treatment = factor(stage1_out$cohort$treatment,
                     labels = c("Control", "SOF"))
)

ggplot(ps_df, aes(x = ps, fill = treatment)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = c(cfg$tmle$truncation_lower,
                             cfg$tmle$truncation_upper),
             linetype = "dashed", color = "red") +
  labs(title = "Propensity Score Overlap",
       x = "P(SOF | W)", y = "Density", fill = "Treatment") +
  theme_minimal()
```

### Covariate Balance (SMD Love Plot)

```{r smd-love-plot}
#| label: smd-love-plot
#| fig-width: 8
#| fig-height: 7

covar_cols <- stage2_out$covar_cols
A <- stage1_out$cohort$treatment

smd_unw <- sapply(covar_cols, function(v) {
  compute_smd(stage1_out$cohort[[v]], A)
})
smd_wtd <- sapply(covar_cols, function(v) {
  compute_smd(stage1_out$cohort[[v]], A, weights = stage2_out$weights)
})

smd_df <- data.frame(
  variable = rep(covar_cols, 2),
  smd = c(smd_unw, smd_wtd),
  type = rep(c("Unweighted", "Weighted"), each = length(covar_cols))
)

ggplot(smd_df, aes(x = abs(smd), y = reorder(variable, abs(smd)),
                   color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed") +
  geom_vline(xintercept = cfg$stage2$max_smd,
             linetype = "dotted", color = "red") +
  labs(title = "Covariate Balance (SMD Love Plot)",
       x = "|SMD|", y = "Covariate", color = "Weighting") +
  theme_minimal()
```

### Diagnostics Summary

```{r diag-summary}
#| label: diag-summary

diag <- stage2_out$diagnostics
cat(sprintf("Max weighted SMD: %.4f (threshold: %.2f)\n",
            diag$max_smd_weighted, cfg$stage2$max_smd))
cat(sprintf("ESS fraction: %.4f (threshold: %.2f)\n",
            diag$ess_frac, cfg$stage2$min_ess_frac))
cat(sprintf("PS overlap fraction: %.4f\n", diag$ps_overlap_frac))
cat(sprintf("PS truncated: %d lower, %d upper\n",
            diag$n_truncated_lower, diag$n_truncated_upper))
```

## Stage 3: Outcome Modeling and Estimation

Stage 3 runs TMLE and comparator estimators. It only proceeds if Stage 2
checkpoint is PASS.

**Primary estimand**: Cumulative risk at 90 and 180 days (risk difference
and risk ratio). This estimand does NOT rely on proportional hazards.

**Secondary estimand**: Cox PH hazard ratio (labeled as such).

```{r stage3}
#| label: stage3

stage3_out <- stage3_estimation(
  cohort = stage1_out$cohort,
  stage2_result = stage2_out,
  cfg = cfg,
  output_dir = "../outputs/stage3"
)

# Display results table
knitr::kable(stage3_out$summary, digits = 5,
             caption = "Stage 3: Effect Estimates by Method and Time Point")
```

### Cox PH Secondary Estimand

```{r cox-secondary}
#| label: cox-secondary

cox_hr <- stage3_out$results$cox_hr
cat(sprintf("Hazard Ratio: %.4f (95%% CI: %.4f, %.4f)\n",
            cox_hr$HR, cox_hr$ci_HR[1], cox_hr$ci_HR[2]))
cat(sprintf("PH Test p-value: %.4f\n", cox_hr$ph_test$pvalue))
cat(sprintf("PH Violated (p < 0.05): %s\n", cox_hr$ph_test$ph_violated))
cat("\nNote: The HR estimate assumes proportional hazards. ",
    "When PH is violated, interpret with caution.\n")
```

### TMLE Diagnostics (180 days)

```{r tmle-diag}
#| label: tmle-diag

tmle_180 <- stage3_out$results$t180$tmle
if (!is.null(tmle_180$diagnostics)) {
  cat("Clever covariate H1 mean:", round(tmle_180$diagnostics$clever_covariate$H1_mean, 4), "\n")
  cat("IC mean (should be ~0):", round(tmle_180$diagnostics$ic$ic_rd_mean, 6), "\n")
  cat("Positivity: min g =", round(tmle_180$diagnostics$positivity$min_g1W, 4),
      ", max g =", round(tmle_180$diagnostics$positivity$max_g1W, 4), "\n")
  cat("Observations censored:", tmle_180$diagnostics$n_censored, "\n")
  cat("Fluctuation eps:", round(tmle_180$diagnostics$eps$eps1, 4),
      ",", round(tmle_180$diagnostics$eps$eps0, 4), "\n")
}
```

## Stage 4: Report Lock

```{r stage4}
#| label: stage4

stage4_report(cfg = cfg, output_dir = "../outputs/report")
```

## Now Run with Complex Scenario

```{r complex-scenario}
#| label: complex-scenario

set_deterministic_seed(cfg$dgp$seed)
cohort_complex <- generate_hcv_data(
  N = cfg$dgp$N,
  h0 = cfg$dgp$h0,
  HR_early = cfg$dgp$HR_early,
  HR_late = cfg$dgp$HR_late,
  tau = cfg$dgp$tau,
  max_follow = cfg$dgp$max_follow,
  np_hazard = TRUE,
  dep_censor = TRUE,
  complexity = TRUE,
  switch_on = TRUE,
  seed = cfg$dgp$seed + 1000
)

# Stage 1 (complex)
s1_cx <- stage1_build_cohort(data = cohort_complex, cfg = cfg,
                              output_dir = "../outputs/stage1")
cat("Complex - Checkpoint 1:", s1_cx$checkpoint, "\n")
cat("Complex - N:", s1_cx$report$N_total,
    ", Events:", s1_cx$report$overall_N_events, "\n")
```

## Decision Log

```{r decision-log}
#| label: decision-log

if (file.exists("../outputs/decision_log.csv")) {
  log <- read.csv("../outputs/decision_log.csv", stringsAsFactors = FALSE)
  knitr::kable(log[, c("stage", "model_component", "decision",
                        "outcome_blind_confirm")],
               caption = "Decision Log Excerpt")
}
```

## Appendix

### Configuration Used

```{r config}
#| label: config

str(cfg, max.level = 2)
```

### Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```
