---
title: "Operationalizing TMLE in a Clean-Room Staged Design"
subtitle: "A Pharmacoepidemiology Tutorial with Simulated HCV Safety Data"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(ggplot2)
set.seed(2024)

# Utility functions
expit <- function(x) 1 / (1 + exp(-x))
logit <- function(p) log(p / (1 - p))
```

# What This Tutorial Is and Is Not

This is **not** a generic TMLE tutorial. For standalone introductions to TMLE,
see van der Laan and Rose (2011), the `tmle` R package vignette, or Schuler and
Rose (2017). This tutorial teaches how TMLE is **operationalized within a
clean-room staged workflow** — a governance structure that controls *when*
different data elements may be accessed and *what* analytic tasks are permitted
at each stage.

**Elevator pitch.** This tutorial shows how TMLE can be embedded in a clean-room
workflow where treatment-mechanism diagnostics occur before outcome access,
simulation is used to validate the full estimator pipeline under known truth,
and the real outcome regression and targeting step occur only after the earlier
stages are completed. The result is an analysis whose credibility does not rest
solely on post-hoc sensitivity checks, because the key analytic decisions were
locked before outcomes were seen.

### The running example

We use a simulated pharmacoepidemiology study loosely modeled on HCV antiviral
safety surveillance. The causal question: does a sofosbuvir (SOF)-containing
regimen increase the risk of acute kidney injury (AKI) compared to a non-SOF
regimen, among patients with chronic hepatitis C? The example is entirely
simulated — no real patient data are used — but the covariate structure,
treatment assignment mechanism, and outcome model reflect patterns seen in
real-world safety studies.

### The central tension

TMLE requires both a treatment model $g(W) = P(A=1|W)$ and an outcome model
$Q(A,W) = E[Y|A,W]$. In a clean-room workflow, these two components live in
different stages:

- **$g$ can be estimated early** (Stage 2), because it uses only treatment and
  covariates — no outcomes required.
- **$Q$ for the real data can only be estimated after outcome access**
  (Stage 4), because fitting an outcome regression requires $Y$.
- **The targeting step** combines $Q$ and $g$ and therefore also belongs to
  Stage 4 for the real analysis.
- **In simulation (Stage 3)**, the full pipeline ($g$, $Q$, targeting) runs on
  synthetic data with known truth, validating the estimator before real
  outcomes are unblinded.

### Workflow overview

```
┌─────────────────────────────────────────────┐
│  Stage 1. Pre-specification & Analysis Lock │
│  Lock estimand, covariates, SL libraries,   │
│  truncation rules, gate thresholds          │
└────────────────────┬────────────────────────┘
                     ↓
┌─────────────────────────────────────────────┐
│  Stage 2. Outcome-Blind Design Diagnostics  │
│  Access (W, A) only. Fit g(W), evaluate     │
│  overlap, balance, ESS. Lock g-strategy.    │
└────────────────────┬────────────────────────┘
                     ↓
┌─────────────────────────────────────────────┐
│  Stage 3. Outcome-Blind Simulation          │
│  Run full TMLE pipeline on synthetic data   │
│  with known truth. Evaluate bias, coverage. │
│  Make GO / STOP gate decision.              │
└────────────────────┬────────────────────────┘
                     ↓
┌─────────────────────────────────────────────┐
│  Stage 4. Unblinded TMLE Estimation         │
│  Access Y. Fit Q(A,W) for the first time.   │
│  Run targeting step. Report results.        │
└─────────────────────────────────────────────┘
```


## Quick Reference: Allowed vs Not Allowed Before Outcome Access

```{r permissions-table}
perms <- data.frame(
  Stage = c(
    "Stage 1:\nPre-specification",
    "Stage 2:\nOutcome-Blind\nDiagnostics",
    "Stage 3:\nSimulation\nValidation",
    "Stage 4:\nUnblinded\nEstimation"
  ),
  `Data accessible` = c(
    "None (metadata, variable definitions only)",
    "W and A from the real dataset (not Y)",
    "Synthetic data with simulated Y",
    "Full dataset: W, A, Y"
  ),
  `Allowed TMLE-related tasks` = c(
    paste("Define causal question and estimand;",
          "specify covariates and treatment;",
          "select SL libraries for g and Q;",
          "set truncation rules;",
          "define diagnostic thresholds and gate criteria"),
    paste("Estimate g(W) on real data;",
          "PS overlap diagnostics;",
          "IPTW weight distributions and ESS;",
          "covariate balance (SMD)"),
    paste("Run the full TMLE pipeline on simulated data;",
          "evaluate bias, variance, and coverage;",
          "compare TMLE with simpler estimators;",
          "test SL specifications and truncation"),
    paste("Estimate Q(A,W) on real data;",
          "combine with locked g-strategy;",
          "run TMLE targeting step;",
          "compute estimates and CIs;",
          "report results")
  ),
  `Not allowed` = c(
    paste("Estimating g or Q on real data;",
          "any estimator involving real outcomes"),
    paste("Fitting Q(A,W) on real outcomes;",
          "computing treatment effects on real data;",
          "running TMLE on the real dataset"),
    paste("Using real outcomes;",
          "tuning estimator based on real Y"),
    paste("Changing estimand, SL library, or truncation",
          "without a documented protocol amendment")
  ),
  check.names = FALSE
)
kable(perms, caption = "Clean-room permissions by stage")
```


# Stage 1. Pre-specification and Analysis Lock {#stage-1}

## What happens at this stage

Before any real study data are accessed, the analysis team writes a statistical
analysis plan (SAP) that locks all major analytic decisions. This is analogous
to a clinical trial's SAP being finalized before database lock.

For our HCV example, the causal question is:

> Among adults with chronic HCV initiating antiviral therapy, what is the
> average causal effect of a SOF-containing regimen (vs. a non-SOF regimen)
> on the 90-day risk of acute kidney injury?

The estimand is the average treatment effect (ATE) on the risk difference scale:
$$\psi = E[Y^1] - E[Y^0]$$

where $Y^a$ denotes the potential AKI outcome under treatment $a$.

Identification assumptions: consistency ($Y = Y^A$), no unmeasured confounding
($Y^a \perp\!\!\!\perp A \mid W$), and positivity ($0 < P(A=1|W) < 1$).

### The analysis lock object

The pre-specified decisions are encoded as a versioned, timestamped artifact.
This object travels through every subsequent stage unchanged unless a formal
amendment is documented.

```{r analysis-lock}
create_analysis_lock <- function() {
  list(
    # --- Causal question ---
    study       = "HCV antiviral safety: SOF vs non-SOF",
    estimand    = "ATE",
    effect_scale = "risk_difference",

    # --- Variable roles ---
    covariates  = c("age", "ckd", "cirrhosis", "diabetes", "nsaid", "renal_risk"),
    treatment   = "A",          # 1 = SOF-containing, 0 = non-SOF
    outcome     = "Y",          # 1 = AKI within 90 days

    # --- Treatment model (g) ---
    g_formula   = "A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk",
    g_SL_library = c("SL.mean", "SL.glmnet", "SL.xgboost"),

    # --- Outcome model (Q) ---
    Q_SL_library = c("SL.mean", "SL.glmnet", "SL.xgboost"),

    # --- PS truncation ---
    ps_truncation = c(lower = 0.025, upper = 0.975),

    # --- TMLE variant ---
    tmle_variant  = "TMLE-ML",

    # --- Stage 2 diagnostic thresholds ---
    diag_thresholds = list(
      extreme_ps_bound = 0.05,    # PS below this or above 1 - this
      max_extreme_frac = 0.15,    # max fraction of extreme PS
      min_ess_frac     = 0.50,    # minimum ESS / n
      max_iptw_weight  = 40,      # max single IPTW weight
      max_smd_weighted = 0.10     # max weighted SMD
    ),

    # --- Stage 3 simulation targets ---
    sim_targets = list(
      max_abs_bias = 0.01,
      min_coverage = 0.93,
      n_reps       = 200
    ),

    # --- Metadata ---
    version     = "1.0",
    locked_date = Sys.Date(),
    locked_by   = "Analysis team"
  )
}

analysis_lock <- create_analysis_lock()

cat("Analysis lock (version", analysis_lock$version, ")\n")
cat("Study:", analysis_lock$study, "\n")
cat("Locked:", as.character(analysis_lock$locked_date), "\n")
cat("Estimand:", analysis_lock$estimand, "on", analysis_lock$effect_scale, "scale\n")
cat("Covariates:", paste(analysis_lock$covariates, collapse = ", "), "\n")
cat("PS truncation:", analysis_lock$ps_truncation["lower"], "-",
    analysis_lock$ps_truncation["upper"], "\n")
```

::: {.callout-note}
### What gets locked vs. what gets re-estimated

What is locked is the **modeling strategy**: the learner library, truncation
bounds, covariate set, and diagnostic thresholds. In practice, the team may
either carry the Stage 2 fitted $\hat{g}$ object forward into Stage 4, or
re-run the same locked procedure on the full $(W, A)$ data at Stage 4. Both
are valid — the key constraint is that the *specification* does not change
after outcome access. Locking a strategy is not the same as requiring reuse of
a specific fitted object.
:::

### Stage 1 summary

At this point, no data have been touched. The team has committed to an estimand
(ATE on the RD scale), identification assumptions, covariate list, SL
libraries, truncation rules, and simulation performance targets.


# Stage 2. Outcome-Blind Design Diagnostics {#stage-2}

## What happens at this stage

The analysis team receives the **treatment and covariate data** $(W, A)$ from
the real study. Outcomes $(Y)$ remain sealed. The purpose is to evaluate
whether the treatment mechanism and covariate structure are compatible with the
planned analysis.

Because $g(W) = P(A=1|W)$ does not involve outcomes, the entire treatment model
can be estimated here. This is the key operational insight: **the $g$-side of
TMLE is fully accessible during the outcome-blind stage.**

### Generate the HCV tutorial dataset

We simulate a dataset that mimics an HCV antiviral safety study. We generate
the full dataset including $Y$, but the Stage 2 analyst sees only $(W, A)$.

```{r generate-data}
simulate_hcv_data <- function(n, seed,
                               beta_renal_Q = 0.05,
                               tau_renal = 0.0) {
  set.seed(seed)

  # --- Baseline covariates ---
  age_raw   <- rnorm(n, 55, 10)
  age       <- (age_raw - 55) / 10           # standardized
  ckd       <- rbinom(n, 1, 0.15)            # chronic kidney disease

  cirrhosis <- rbinom(n, 1, 0.20)
  diabetes  <- rbinom(n, 1, 0.25)
  nsaid     <- rbinom(n, 1, 0.30)            # concurrent NSAID use
  renal_risk <- rnorm(n, 0, 1)               # composite renal-risk score

  # --- Treatment mechanism (g) ---
  # renal_risk is the overlap driver: coefficient = 2.2
  lp_g <- -0.3 + 0.3*age + 0.5*ckd + 0.4*cirrhosis +
           0.2*diabetes - 0.3*nsaid + 2.2*renal_risk
  g_true <- pmin(pmax(expit(lp_g), 0.005), 0.995)
  A <- rbinom(n, 1, g_true)

  # --- Outcome model (Q): AKI ---
  # beta_renal_Q controls how much renal_risk affects AKI directly
  # tau_renal controls treatment-by-renal_risk interaction
  lp_Q <- -3.0 + 0.4*age + 0.3*age^2 + 0.8*ckd + 0.5*cirrhosis +
           0.3*diabetes + 0.4*nsaid + beta_renal_Q * renal_risk +
           0.30*A + 0.20*A*ckd + tau_renal*A*renal_risk
  Y <- rbinom(n, 1, expit(lp_Q))

  data.frame(age, ckd, cirrhosis, diabetes, nsaid, renal_risk, A, Y)
}

# Generate the "real" study data (n = 800)
dat_full <- simulate_hcv_data(n = 800, seed = 42,
                               beta_renal_Q = 0.01, tau_renal = 0.0)

# Stage 2 analyst sees only W and A — Y is sealed
dat_stage2 <- dat_full[, c("age", "ckd", "cirrhosis", "diabetes",
                            "nsaid", "renal_risk", "A")]

cat("Stage 2 data dimensions:", nrow(dat_stage2), "x", ncol(dat_stage2), "\n")
cat("Treatment prevalence:", round(mean(dat_stage2$A), 3), "\n")
cat("Y column accessible:", "Y" %in% names(dat_stage2), "\n")
```

### Fit the treatment model

Using the locked g-formula from the analysis lock:

```{r fit-g}
fit_stage2_g <- function(dat, lock) {
  g_fit <- glm(as.formula(lock$g_formula), data = dat, family = "binomial")
  g_hat <- fitted(g_fit)
  g_trunc <- pmin(pmax(g_hat, lock$ps_truncation["lower"]),
                  lock$ps_truncation["upper"])
  list(fit = g_fit, g_hat = g_hat, g_trunc = g_trunc)
}

g_stage2 <- fit_stage2_g(dat_stage2, analysis_lock)

cat("PS range (raw):", round(range(g_stage2$g_hat), 4), "\n")
cat("PS range (truncated):", round(range(g_stage2$g_trunc), 4), "\n")
```

### Propensity score diagnostics

These diagnostics use only $(W, A)$ and are fully permitted at Stage 2.

```{r ps-diagnostics, fig.width=8, fig.height=4}
ps_df <- data.frame(
  ps = g_stage2$g_hat,
  Treatment = factor(dat_stage2$A,
                     labels = c("Non-SOF", "SOF-containing"))
)

ggplot(ps_df, aes(x = ps, fill = Treatment)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = c(analysis_lock$ps_truncation["lower"],
                             analysis_lock$ps_truncation["upper"]),
             linetype = "dashed", color = "red") +
  labs(title = "Stage 2: Propensity Score Distribution by Treatment",
       subtitle = "Red dashed lines = pre-specified truncation bounds",
       x = "Estimated P(SOF | W)", y = "Count") +
  theme_minimal(base_size = 12)
```

```{r overlap-metrics}
stage2_diagnostics <- function(g_hat, g_trunc, A, dat, lock) {
  thr <- lock$diag_thresholds
  n <- length(A)

  # Fraction with extreme PS
  frac_extreme <- mean(g_hat < thr$extreme_ps_bound |
                       g_hat > (1 - thr$extreme_ps_bound))

  # IPTW weights and ESS
  w_iptw <- A / g_trunc + (1 - A) / (1 - g_trunc)
  ess <- sum(w_iptw)^2 / sum(w_iptw^2)
  ess_frac <- ess / n
  max_w <- max(w_iptw)

  # Weighted SMDs
  compute_smd <- function(x, trt, weights = NULL) {
    if (is.null(weights)) weights <- rep(1, length(x))
    m1 <- weighted.mean(x[trt == 1], weights[trt == 1])
    m0 <- weighted.mean(x[trt == 0], weights[trt == 0])
    s <- sqrt((var(x[trt == 1]) + var(x[trt == 0])) / 2)
    if (s < 1e-10) return(0)
    (m1 - m0) / s
  }

  covars <- lock$covariates
  smd_unw <- sapply(covars, function(v) compute_smd(dat[[v]], A))
  smd_wtd <- sapply(covars, function(v) compute_smd(dat[[v]], A, w_iptw))

  diag_tab <- data.frame(
    Metric = c("Fraction extreme PS",
               "ESS fraction",
               "Max IPTW weight",
               "Max |SMD| (IPTW-weighted)"),
    Value = round(c(frac_extreme, ess_frac, max_w, max(abs(smd_wtd))), 3),
    Threshold = c(paste("<", thr$max_extreme_frac),
                  paste(">", thr$min_ess_frac),
                  paste("<", thr$max_iptw_weight),
                  paste("<", thr$max_smd_weighted)),
    Status = c(ifelse(frac_extreme < thr$max_extreme_frac, "PASS", "FLAG"),
               ifelse(ess_frac > thr$min_ess_frac, "PASS", "FLAG"),
               ifelse(max_w < thr$max_iptw_weight, "PASS", "FLAG"),
               ifelse(max(abs(smd_wtd)) < thr$max_smd_weighted, "PASS", "FLAG"))
  )

  list(diag_tab = diag_tab, smd_unw = smd_unw, smd_wtd = smd_wtd,
       w_iptw = w_iptw, any_flag = any(diag_tab$Status == "FLAG"))
}

s2_diag <- stage2_diagnostics(g_stage2$g_hat, g_stage2$g_trunc,
                               dat_stage2$A, dat_stage2, analysis_lock)
kable(s2_diag$diag_tab, caption = "Stage 2 overlap diagnostics")
```

```{r love-plot, fig.width=7, fig.height=4}
covars <- analysis_lock$covariates
smd_df <- data.frame(
  variable = rep(covars, 2),
  smd = c(s2_diag$smd_unw, s2_diag$smd_wtd),
  type = rep(c("Unweighted", "IPTW-Weighted"), each = length(covars))
)

ggplot(smd_df, aes(x = abs(smd), y = reorder(variable, abs(smd)), color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.10, linetype = "dashed") +
  labs(title = "Stage 2: Covariate Balance (Love Plot)",
       x = "|Standardized Mean Difference|", y = NULL, color = "Weighting") +
  theme_minimal(base_size = 12)
```

### Stage 2 decision

```{r stage2-decision}
stage2_status <- ifelse(s2_diag$any_flag,
                         "FLAG — proceed to Stage 3 simulation",
                         "PASS")
cat("Stage 2 status:", stage2_status, "\n")
if (s2_diag$any_flag) {
  flagged <- s2_diag$diag_tab$Metric[s2_diag$diag_tab$Status == "FLAG"]
  cat("Flagged metrics:", paste(flagged, collapse = "; "), "\n")
}
```

### Why Stage 2 is not enough for TMLE validation

Stage 2 evaluates the treatment mechanism $g$ and practical positivity. But
TMLE depends on *three* components: $g$, $Q$, and the targeting step. Stage 2
tells us nothing about whether the outcome model can be estimated well,
whether the targeting step is stable, whether the variance estimator is
reliable, or whether CI coverage is near nominal.

This matters in both directions:

- A study can have **flagged overlap but excellent TMLE performance** if the
  overlap-driving covariate is weakly related to the outcome — the outcome
  model compensates.
- A study can have **acceptable overlap but poor TMLE performance** if the
  outcome surface is difficult to learn at the given sample size.

Stage 2 alone cannot distinguish these cases. **Stage 3 fills the gap** by
testing the complete TMLE pipeline on synthetic data where the truth is known.

### Stage 2 summary

```{r stage2-summary, results='asis'}
cat("At this point the team has estimated $g(W)$, evaluated overlap and balance, ")
if (s2_diag$any_flag) {
  cat("and documented that **some diagnostic flags triggered**. ")
  cat("These flags indicate potential overlap concerns — but as discussed below, ")
  cat("overlap flags alone do not determine whether TMLE will perform adequately. ")
} else {
  cat("and found that **all diagnostic thresholds were met**. ")
  cat("However, passing Stage 2 does not guarantee that the full TMLE pipeline ")
  cat("will perform adequately — as Stage 3 will show. ")
}
cat("The outcome $Y$ has **not been touched**. The $g$-modeling strategy is now locked.")
```


# Why TMLE Can Survive Poor Overlap {#double-robustness}

Before proceeding to the simulation stage, it helps to understand *why* poor
propensity score diagnostics do not automatically invalidate TMLE. The answer
is **double robustness**.

TMLE relies on two nuisance components:

1. The **outcome regression** $Q(A,W) = E[Y|A,W]$: a model for the expected
   outcome given treatment and covariates.
2. The **treatment mechanism** $g(W) = P(A=1|W)$: a model for the probability
   of receiving treatment given covariates.

The key property is this: **TMLE is consistent if *either* the outcome model
or the treatment model is consistently estimated.** Both do not need to be
correct simultaneously for TMLE to converge to the right answer. If one model
is wrong but the other is right, TMLE uses the correct model to compensate for
the incorrect one.

When *both* models are consistently estimated, TMLE achieves an additional
benefit: it is **efficient**, meaning it reaches the smallest possible variance
among all regular estimators.

What does this mean practically?

- If the overlap-driving covariate barely affects the outcome (weak
  confounding through that variable), then the outcome model $Q$ can be
  estimated well even though $g$ is difficult. TMLE leans on $Q$ to get the
  right answer.
- If the overlap-driving covariate strongly affects both treatment and outcome,
  then *both* $g$ and $Q$ are hard to estimate in the regions of poor overlap.
  TMLE has nothing to fall back on.
- **IPTW**, by contrast, relies entirely on $g$. Poor overlap always hurts
  IPTW, regardless of the outcome surface.

This is why overlap concerns — whether flagged at Stage 2 or not — do not
automatically determine whether TMLE will perform well or poorly. They mean
"the treatment model alone may be under stress — we need Stage 3 to check
whether the full TMLE pipeline, including the outcome model, can compensate."

::: {.callout-important}
Double robustness is an asymptotic property that holds under regularity
conditions, including adequate positivity and sufficiently good nuisance
estimation — it is not a finite-sample guarantee. In practice, extreme weights
from a poorly estimated $g$ can still destabilize the targeting step, inflate
variance, and degrade CI coverage even when the outcome model is well-specified.
This is exactly why simulation validation (Stage 3) is necessary: it checks
whether double robustness actually helps at the study's sample size and
covariate distribution, rather than relying on asymptotic theory alone.
:::


# Stage 3. Outcome-Blind Simulation Validation {#stage-3}

## What happens at this stage

Stage 2 told us about the treatment mechanism. Stage 3 answers the harder
question: **does the entire TMLE pipeline perform adequately under conditions
that resemble this study?**

The team runs a simulation using a pre-specified data-generating process (DGP).
The DGP is calibrated to match the observed $(W, A)$ distribution but generates
**synthetic outcomes** — not the real $Y$. Because the truth is known in the
simulation, the team can evaluate bias, variance, and CI coverage.

### What Stage 3 validates that Stage 2 cannot

| Question | Stage 2 | Stage 3 |
|----------|---------|---------|
| Is $\hat{g}$ well-estimated? | Yes | Yes |
| Are PS values extreme? | Yes | Yes |
| Is $\hat{Q}$ adequate? | **No** | Yes (on synthetic data) |
| Does the targeting step work? | **No** | Yes |
| Is IC-based SE reliable? | **No** | Yes |
| Does CI coverage reach nominal? | **No** | Yes |
| Is TMLE better than alternatives? | **No** | Yes |


### Define the simulation DGP

The DGP must be specified *before* real outcomes are seen. It should reflect
plausible outcome models given clinical domain knowledge. We define two
scenarios:

- **Scenario A (weak renal-risk effect on AKI):** The baseline renal-risk
  score strongly predicts treatment choice (coefficient 2.2 in $g$) but has
  almost no direct effect on AKI (coefficient 0.01 in $Q$). If the outcome
  model can compensate for poor overlap, TMLE should still perform well.

- **Scenario B (strong renal-risk effect on AKI):** The same covariate
  strongly predicts both treatment and outcome (coefficient 1.0 in $Q$, with
  a treatment interaction of 0.5). Both $g$ and $Q$ are difficult to estimate
  in the tails, so TMLE may have nothing to fall back on.

```{r sim-dgp}
# Compute true ATE by marginalizing Q over the covariate distribution
compute_truth_hcv <- function(n_mc = 200000, seed = 1,
                               beta_renal_Q = 0.05, tau_renal = 0.0) {
  set.seed(seed)
  age       <- rnorm(n_mc, 0, 1)
  ckd       <- rbinom(n_mc, 1, 0.15)
  cirrhosis <- rbinom(n_mc, 1, 0.20)
  diabetes  <- rbinom(n_mc, 1, 0.25)
  nsaid     <- rbinom(n_mc, 1, 0.30)
  renal_risk <- rnorm(n_mc, 0, 1)

  lp_base <- -3.0 + 0.4*age + 0.3*age^2 + 0.8*ckd + 0.5*cirrhosis +
             0.3*diabetes + 0.4*nsaid + beta_renal_Q*renal_risk
  lp_Q1 <- lp_base + 0.30 + 0.20*ckd + tau_renal*renal_risk
  lp_Q0 <- lp_base

  c(EY1 = mean(expit(lp_Q1)), EY0 = mean(expit(lp_Q0)),
    RD = mean(expit(lp_Q1)) - mean(expit(lp_Q0)))
}

truth_A <- compute_truth_hcv(beta_renal_Q = 0.01, tau_renal = 0.0)
truth_B <- compute_truth_hcv(beta_renal_Q = 1.00, tau_renal = 0.50)

cat("True ATE, Scenario A (weak renal-risk):", round(truth_A["RD"], 5), "\n")
cat("True ATE, Scenario B (strong renal-risk):", round(truth_B["RD"], 5), "\n")
```

### Define the TMLE estimator

This function runs one TMLE analysis using the pre-specified pipeline. It fits
$Q$ (here via an enriched GLM for tutorial speed), fits $g$, performs the
targeting step, and returns the estimate with IC-based inference.

```{r tmle-function}
run_tmle_once <- function(dat, ps_clip = c(0.025, 0.975)) {
  n <- nrow(dat)

  # --- Fit g (treatment model) ---
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])

  # --- Fit Q (outcome model) — enriched GLM with interactions ---
  Q_fit <- glm(Y ~ A * (age + ckd + cirrhosis + diabetes + nsaid + renal_risk) +
                 I(age^2),
               data = dat, family = "binomial")
  Q_A <- predict(Q_fit, type = "response")
  Q_1 <- predict(Q_fit, newdata = transform(dat, A = 1), type = "response")
  Q_0 <- predict(Q_fit, newdata = transform(dat, A = 0), type = "response")

  # Bound Q away from 0/1
  Q_A <- pmin(pmax(Q_A, 1e-5), 1 - 1e-5)
  Q_1 <- pmin(pmax(Q_1, 1e-5), 1 - 1e-5)
  Q_0 <- pmin(pmax(Q_0, 1e-5), 1 - 1e-5)

  # --- Targeting step ---
  H  <- dat$A / g_hat - (1 - dat$A) / (1 - g_hat)
  H1 <- 1 / g_hat
  H0 <- -1 / (1 - g_hat)

  eps_fit <- glm(dat$Y ~ -1 + H, offset = logit(Q_A), family = "binomial")
  eps <- coef(eps_fit)

  Q_1_star <- expit(logit(Q_1) + eps * H1)
  Q_0_star <- expit(logit(Q_0) + eps * H0)
  Q_A_star <- expit(logit(Q_A) + eps * H)

  # --- Point estimate ---
  psi <- mean(Q_1_star - Q_0_star)

  # --- IC-based inference ---
  eif <- H * (dat$Y - Q_A_star) + (Q_1_star - Q_0_star) - psi
  se  <- sqrt(var(eif) / n)
  ci  <- psi + c(-1.96, 1.96) * se

  list(estimate = psi, se = se, ci_low = ci[1], ci_high = ci[2])
}
```

We also define two simpler comparators:

```{r comparator-functions}
run_crude <- function(dat) {
  p1 <- mean(dat$Y[dat$A == 1]); p0 <- mean(dat$Y[dat$A == 0])
  rd <- p1 - p0
  n1 <- sum(dat$A == 1); n0 <- sum(dat$A == 0)
  se <- sqrt(p1*(1 - p1)/n1 + p0*(1 - p0)/n0)
  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}

run_iptw <- function(dat, ps_clip = c(0.025, 0.975)) {
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])
  A <- dat$A; Y <- dat$Y; n <- length(Y)
  w  <- A / g_hat + (1 - A) / (1 - g_hat)
  d1 <- sum(w * A); d0 <- sum(w * (1 - A))
  mu1 <- sum(w * A * Y) / d1
  mu0 <- sum(w * (1 - A) * Y) / d0
  rd <- mu1 - mu0
  # Sandwich SE
  phi1 <- w * A * (Y - mu1) / d1 * n
  phi0 <- w * (1 - A) * (Y - mu0) / d0 * n
  se <- sqrt(sum((phi1 - phi0)^2) / n^2)
  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}
```

### Run the Stage 3 simulation

```{r run-sim, cache=TRUE}
run_stage3_simulation <- function(reps, n, beta_renal_Q, tau_renal,
                                   true_rd, seed_offset = 0,
                                   ps_clip = c(0.025, 0.975)) {
  results <- vector("list", reps * 3)
  k <- 0

  for (r in seq_len(reps)) {
    d <- simulate_hcv_data(n, seed = seed_offset + r,
                            beta_renal_Q = beta_renal_Q,
                            tau_renal = tau_renal)

    for (meth in c("Crude", "IPTW", "TMLE")) {
      res <- tryCatch(
        switch(meth,
               Crude = run_crude(d),
               IPTW  = run_iptw(d, ps_clip),
               TMLE  = run_tmle_once(d, ps_clip)),
        error = function(e) list(estimate = NA, se = NA,
                                  ci_low = NA, ci_high = NA))
      k <- k + 1
      results[[k]] <- data.frame(
        rep = r, method = meth,
        est = res$estimate, se = res$se,
        ci_lo = res$ci_low, ci_hi = res$ci_high)
    }
  }

  results <- do.call(rbind, results)

  # Compute performance metrics
  metrics <- do.call(rbind, lapply(split(results, results$method), function(sub) {
    ok     <- !is.na(sub$est)
    has_ci <- ok & !is.na(sub$ci_lo)
    e <- sub$est[ok]
    data.frame(
      method   = sub$method[1],
      n_ok     = sum(ok),
      bias     = round(mean(e) - true_rd, 5),
      rmse     = round(sqrt(mean((e - true_rd)^2)), 5),
      emp_sd   = round(sd(e), 5),
      mean_se  = round(mean(sub$se[ok], na.rm = TRUE), 5),
      coverage = round(mean(
        (sub$ci_lo[has_ci] <= true_rd) & (true_rd <= sub$ci_hi[has_ci])), 3),
      n_ci     = sum(has_ci))
  }))
  row.names(metrics) <- NULL

  list(results = results, metrics = metrics)
}

# --- Scenario A: Weak renal-risk effect on AKI ---
sim_A <- run_stage3_simulation(
  reps = 200, n = 800,
  beta_renal_Q = 0.01, tau_renal = 0.0,
  true_rd = truth_A["RD"],
  seed_offset = 1000,
  ps_clip = analysis_lock$ps_truncation)

# --- Scenario B: Strong renal-risk effect on AKI ---
sim_B <- run_stage3_simulation(
  reps = 200, n = 800,
  beta_renal_Q = 1.00, tau_renal = 0.50,
  true_rd = truth_B["RD"],
  seed_offset = 2000,
  ps_clip = analysis_lock$ps_truncation)
```

### Simulation results

```{r sim-results}
cat("=== Scenario A: Weak renal-risk effect on AKI ===\n")
cat("True RD:", round(truth_A["RD"], 5), "\n\n")
kable(sim_A$metrics, digits = 4,
      caption = "Scenario A: renal_risk weakly affects AKI (beta = 0.01)")

cat("\n=== Scenario B: Strong renal-risk effect on AKI ===\n")
cat("True RD:", round(truth_B["RD"], 5), "\n\n")
kable(sim_B$metrics, digits = 4,
      caption = "Scenario B: renal_risk strongly affects AKI (beta = 1.00, tau = 0.50)")
```

```{r sim-sampling-dist, fig.width=9, fig.height=6}
plot_dat <- rbind(
  cbind(scenario = "A: Weak renal-risk\nin outcome", sim_A$results),
  cbind(scenario = "B: Strong renal-risk\nin outcome", sim_B$results)
)
plot_dat <- plot_dat[!is.na(plot_dat$est), ]
plot_dat$method <- factor(plot_dat$method, levels = c("Crude", "IPTW", "TMLE"))

truth_lines <- data.frame(
  scenario = c("A: Weak renal-risk\nin outcome",
               "B: Strong renal-risk\nin outcome"),
  truth = c(truth_A["RD"], truth_B["RD"])
)

ggplot(plot_dat, aes(x = est, fill = method)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(data = truth_lines, aes(xintercept = truth),
             linewidth = 0.8, linetype = "solid") +
  facet_grid(method ~ scenario, scales = "free") +
  labs(title = "Stage 3: Sampling Distributions Under Two DGP Scenarios",
       subtitle = "Black line = true ATE",
       x = "Estimated Risk Difference", y = "Count") +
  guides(fill = "none") +
  theme_minimal(base_size = 11)
```

### What the gate tests

Before examining the results, it is worth being explicit about the decision
framework. The Stage 3 gate asks: does the TMLE pipeline meet the pre-specified
performance targets?

- **GO** means TMLE bias is below the threshold and CI coverage is at or above
  the minimum — the pipeline is adequate for this scenario.
- **STOP** means TMLE does not meet these targets — the pipeline is inadequate
  as specified, and the team must either revise the plan or report the analysis
  with caveats.

The interpretation below follows the actual results from this tutorial run.

### Gate decision

The gate logic compares TMLE performance against pre-specified thresholds. We
also check whether the mean estimated SE is reasonably calibrated to the
empirical SD (SE/SD ratio between 0.8 and 1.2 is a useful heuristic).

```{r sim-gate}
gate_check <- function(metrics, scenario_name, targets) {
  tmle <- metrics[metrics$method == "TMLE", ]

  bias_ok <- abs(tmle$bias) < targets$max_abs_bias
  cov_ok  <- tmle$coverage >= targets$min_coverage
  se_sd_ratio <- tmle$mean_se / tmle$emp_sd
  se_cal  <- se_sd_ratio > 0.8 & se_sd_ratio < 1.2

  gate_tab <- data.frame(
    Metric = c("TMLE |bias|", "TMLE coverage", "SE/SD ratio"),
    Target = c(paste("<", targets$max_abs_bias),
               paste(">=", targets$min_coverage),
               "0.8 - 1.2"),
    Observed = round(c(abs(tmle$bias), tmle$coverage, se_sd_ratio), 4),
    Status = c(ifelse(bias_ok, "GO", "STOP"),
               ifelse(cov_ok, "GO", "STOP"),
               ifelse(se_cal, "GO", "FLAG"))
  )

  decision <- ifelse(bias_ok & cov_ok, "GO", "STOP")

  cat(scenario_name, "\n")
  print(kable(gate_tab))
  cat("Gate decision:", decision, "\n\n")

  invisible(list(decision = decision, table = gate_tab))
}

cat("--- Gate Assessment ---\n\n")
gate_A <- gate_check(sim_A$metrics, "Scenario A (Weak renal-risk)", analysis_lock$sim_targets)
gate_B <- gate_check(sim_B$metrics, "Scenario B (Strong renal-risk)", analysis_lock$sim_targets)
```

### Stage 3 interpretation

```{r stage3-interpret, results='asis'}
tmle_A <- sim_A$metrics[sim_A$metrics$method == "TMLE", ]
tmle_B <- sim_B$metrics[sim_B$metrics$method == "TMLE", ]
iptw_A <- sim_A$metrics[sim_A$metrics$method == "IPTW", ]
iptw_B <- sim_B$metrics[sim_B$metrics$method == "IPTW", ]

s2_flagged <- s2_diag$any_flag

# --- Scenario A ---
cat("**Scenario A** (weak renal-risk effect on AKI): TMLE shows ",
    sprintf("bias of %.5f and coverage of %.1f%%", tmle_A$bias, 100*tmle_A$coverage),
    ". ", sep = "")

if (gate_A$decision == "GO") {
  if (s2_flagged) {
    cat("Despite the Stage 2 overlap flags, the simulation gate says **GO**. ",
        "This is a **false stop averted**: the overlap concern was real for $g$ alone, ",
        "but because `renal_risk` barely affects AKI, the outcome model compensated ",
        "for the difficult treatment model — double robustness in action.\n\n")
  } else {
    cat("The simulation gate says **GO**, confirming that the TMLE pipeline ",
        "performs adequately under this DGP.\n\n")
  }
} else {
  cat("The simulation gate says **STOP**. Even though `renal_risk` barely affects AKI, ",
      "TMLE performance did not meet the pre-specified thresholds at this sample size. ",
      "This illustrates that double robustness is a large-sample property that does not ",
      "always guarantee adequate finite-sample performance.\n\n")
}

# --- Scenario B ---
cat("**Scenario B** (strong renal-risk effect on AKI): ",
    "The same treatment mechanism now creates a harder problem because `renal_risk` ",
    "strongly affects both treatment and outcome. TMLE shows ",
    sprintf("bias of %.5f and coverage of %.1f%%", tmle_B$bias, 100*tmle_B$coverage),
    ". ", sep = "")

if (gate_B$decision == "STOP") {
  cat("The gate says **STOP**: when the overlap-driving covariate also strongly ",
      "predicts the outcome, TMLE cannot compensate.\n\n")
} else {
  cat("The gate says **GO**. Even with strong confounding through ",
      "`renal_risk`, the enriched Q model captures enough of the ",
      "outcome surface to compensate.\n\n")
}

# --- Comparative ---
cat(sprintf("For comparison, IPTW (which depends entirely on $g$) shows coverage of ",
            "%.1f%% in Scenario A and %.1f%% in Scenario B — poor in both cases, ",
            "because IPTW has no outcome model to fall back on.",
            100*iptw_A$coverage, 100*iptw_B$coverage))
cat("\n\n")

# --- Overall lesson ---
if (gate_A$decision == "GO" && gate_B$decision == "STOP") {
  cat("Both scenarios share identical treatment mechanisms and identical Stage 2 ",
      "diagnostics. The difference is visible only through Stage 3 simulation, ",
      "which tests the full estimator pipeline — not just the propensity score. ",
      "Scenario A is a **false stop averted**: Stage 2 overlap flags alone would ",
      "have unnecessarily stopped a valid analysis. Scenario B is a **true stop**: ",
      "the overlap problem genuinely invalidates the TMLE pipeline.\n")
} else if (gate_A$decision == "STOP" && gate_B$decision == "STOP") {
  cat("Both scenarios fail the Stage 3 gate. ",
      "This demonstrates that even when Stage 2 diagnostics ",
      ifelse(s2_flagged, "flag overlap concerns, ", "look acceptable, "),
      "Stage 3 simulation can reveal that the planned TMLE pipeline does not yet ",
      "meet pre-specified performance targets. The tutorial's DGP, sample size, ",
      "and enriched-GLM specification were not sufficient to achieve adequate ",
      "performance under these overlap conditions.\n")
} else {
  cat("The gate decisions show that Stage 3 provides information beyond ",
      "Stage 2 — the two scenarios share identical treatment mechanisms but ",
      "produce different TMLE performance depending on how the overlap-driving ",
      "covariate relates to the outcome.\n")
}
```

### What this tutorial run demonstrates

```{r tutorial-lesson, results='asis'}
if (gate_A$decision == "GO" && gate_B$decision == "STOP") {
  if (s2_flagged) {
    cat("::: {.callout-tip}\n",
        "### Lesson from this run\n\n",
        "Stage 2 flagged overlap concerns that would, under a traditional PS-only ",
        "workflow, lead to stopping the analysis. Stage 3 simulation showed that ",
        "TMLE still performs adequately in Scenario A (where the overlap driver ",
        "barely affects the outcome) but not in Scenario B (where it strongly ",
        "affects the outcome). This is the **false-stop / true-stop contrast**: ",
        "the same overlap pattern is or is not a problem depending on the outcome ",
        "surface, and only Stage 3 can distinguish the two cases.\n",
        ":::\n", sep = "")
  } else {
    cat("::: {.callout-tip}\n",
        "### Lesson from this run\n\n",
        "Stage 2 diagnostics did not flag major overlap concerns, but Stage 3 ",
        "revealed that TMLE performs adequately in Scenario A but not in ",
        "Scenario B. This shows that Stage 3 adds information beyond Stage 2 ",
        "even when Stage 2 looks acceptable.\n",
        ":::\n", sep = "")
  }
} else if (gate_A$decision == "STOP" && gate_B$decision == "STOP") {
  cat("::: {.callout-tip}\n",
      "### Lesson from this run\n\n",
      "Stage 2 diagnostics ", ifelse(s2_flagged, "flagged some overlap concerns",
                                      "did not flag major problems"),
      ", but Stage 3 showed that the planned TMLE pipeline did not meet ",
      "pre-specified performance targets under either scenario. This demonstrates ",
      "that **Stage 2 diagnostics alone are not sufficient to validate a TMLE ",
      "analysis** — outcome-blind simulation can reveal performance problems that ",
      "propensity-score diagnostics miss. In a real study, the team would now ",
      "consider revising the analysis plan (e.g., larger sample, richer Q learners, ",
      "tighter truncation, or a different estimand) before proceeding to Stage 4.\n",
      ":::\n", sep = "")
} else {
  cat("::: {.callout-tip}\n",
      "### Lesson from this run\n\n",
      "Stage 3 simulation provided information that Stage 2 could not: it ",
      "distinguished scenarios where the TMLE pipeline works from those where ",
      "it does not, based on the relationship between the overlap-driving ",
      "covariate and the outcome.\n",
      ":::\n", sep = "")
}
```

::: {.callout-important}
### The core methodological point

Stage 2 evaluates $g$ alone. Stage 3 evaluates the entire TMLE pipeline
($g$, $Q$, targeting, inference). Without Stage 3, the team can only assess
whether the propensity score behaves well — not whether the full estimator
produces valid estimates. Outcome-blind simulation is what turns a generic
TMLE workflow into a credible clean-room analysis.
:::


# A Brief Intuition for the Efficient Influence Function {#eif-intuition}

Before presenting the Stage 4 results and inference formulas, it helps to
have a brief intuition for the **efficient influence function (EIF)**, which
underlies TMLE's standard errors.

The EIF summarizes **how each observation contributes to the estimator**. For
the ATE, the EIF for observation $i$ is:

$$D_i = \frac{A_i}{\hat{g}(W_i)}\bigl(Y_i - \bar{Q}^*(1, W_i)\bigr)
      - \frac{1 - A_i}{1 - \hat{g}(W_i)}\bigl(Y_i - \bar{Q}^*(0, W_i)\bigr)
      + \bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i) - \hat\psi$$

Three key facts:

1. **TMLE's targeting step ensures that the sample mean of the EIF is
   approximately zero.** This is the "solved estimating equation" property —
   it means the estimator is locally efficient.

2. **The variance of the EIF gives the standard error.** Specifically,
   $\widehat{\text{SE}} = \sqrt{n^{-1}\widehat{\text{Var}}(D)}$.
   This is sometimes called the "IC-based" or "sandwich" variance estimator.

3. **The EIF combines information from both models.** The first two terms
   use $g$ through the weights $A/\hat{g}$ and $(1-A)/(1-\hat{g})$, and $Q$
   through the residuals $Y - \bar{Q}^*$. The third term uses $Q$ alone. If
   $Q$ is perfect, the residuals are zero and $g$ drops out. If $g$ is perfect,
   the weights are accurate and $Q$ errors are corrected. This is double
   robustness expressed through the influence function.

This is all you need to know to interpret the Stage 4 inference: the SE comes
from the variance of the EIF across observations, and the CI is
$\hat\psi \pm 1.96 \times \widehat{\text{SE}}$.


# Stage 4. Unblinded TMLE Estimation {#stage-4}

## What happens at this stage

The analyst now receives the full dataset including outcomes $Y$. This is the
**first point** at which the outcome regression $Q(A,W)$ is estimated on real
data. The treatment model strategy was locked in Stage 2; the simulation
validation was completed in Stage 3; only now does the targeting step run on
real data.

### Prerequisites

Stage 4 proceeds only after:

1. Stage 1 analysis lock is documented and versioned
2. Stage 2 diagnostics are reviewed and recorded
3. Stage 3 simulation gate is evaluated (if GO, proceed; if STOP, document
   the concern and any plan amendment)

### Run TMLE on the real data

```{r stage4-tmle}
run_stage4_tmle <- function(dat, lock) {
  cat("Stage 4: Outcome column now accessible.\n")
  cat("AKI prevalence:", round(mean(dat$Y), 3), "\n")
  cat("Treatment prevalence:", round(mean(dat$A), 3), "\n\n")

  result <- run_tmle_once(dat, ps_clip = lock$ps_truncation)

  cat("=== Stage 4 TMLE Result ===\n")
  cat("ATE estimate (RD):", round(result$estimate, 5), "\n")
  cat("Standard error:", round(result$se, 5), "\n")
  cat("95% CI: [", round(result$ci_low, 5), ",",
      round(result$ci_high, 5), "]\n")

  invisible(result)
}

stage4_result <- run_stage4_tmle(dat_full, analysis_lock)
```

### Comparator estimates

For transparency, we also run the pre-specified comparator estimators on the
real data:

```{r stage4-comparators}
stage4_crude <- run_crude(dat_full)
stage4_iptw  <- run_iptw(dat_full, ps_clip = analysis_lock$ps_truncation)

est_tab <- data.frame(
  Estimator = c("Crude RD", "IPTW (Hajek)", "TMLE"),
  Estimate = round(c(stage4_crude$estimate, stage4_iptw$estimate,
                     stage4_result$estimate), 5),
  SE = round(c(stage4_crude$se, stage4_iptw$se, stage4_result$se), 5),
  CI_low = round(c(stage4_crude$ci_low, stage4_iptw$ci_low,
                   stage4_result$ci_low), 5),
  CI_high = round(c(stage4_crude$ci_high, stage4_iptw$ci_high,
                    stage4_result$ci_high), 5)
)
kable(est_tab, caption = "Stage 4: Effect estimates on the real (simulated) dataset")
```

### What just happened — step by step

1. **$Q$ was estimated on the real data** for the first time: an enriched GLM
   with treatment-by-covariate interactions and $\text{age}^2$ was fit to
   $(W, A, Y)$.

2. **$g$ was re-estimated** using the same locked strategy from Stage 2. (In
   a production analysis, one might load the Stage 2 $\hat{g}$ object instead.)

3. **The targeting step** constructed the clever covariate
   $H = A/\hat{g} - (1-A)/(1-\hat{g})$ and fit a logistic fluctuation
   $\text{logit}(\bar{Q}^*) = \text{logit}(\bar{Q}) + \epsilon H$.

4. **The ATE** was computed as
   $\hat\psi = n^{-1}\sum[\bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i)]$.

5. **Inference** used the EIF as described above.

### Interpreting the result in context

```{r stage4-context, results='asis'}
cat("This Stage 4 estimate should be interpreted alongside the evidence from ",
    "earlier stages.\n\n")

if (s2_diag$any_flag) {
  cat("Stage 2 flagged overlap concerns in the propensity score distribution. ")
} else {
  cat("Stage 2 diagnostics did not flag major overlap concerns. ")
}

# The real dataset uses Scenario A parameters, so gate_A is the relevant gate
if (gate_A$decision == "GO") {
  cat("Stage 3 simulation returned **GO** under Scenario A (the DGP matching ",
      "our real-data assumptions), providing pre-registered evidence that the ",
      "TMLE pipeline produces adequate performance under these conditions. ",
      "The Stage 4 estimate therefore carries credibility beyond a post-hoc ",
      "TMLE analysis, because the analytic decisions were locked before ",
      "outcomes were accessed.\n\n")
} else {
  cat("Stage 3 simulation returned **STOP** under Scenario A, meaning the TMLE ",
      "pipeline did not meet pre-specified performance targets. In a real study, ",
      "this Stage 4 estimate would be reported with caveats, and the STOP finding ",
      "documented alongside the results. The team might also consider a protocol ",
      "amendment (e.g., richer learners, larger sample) before relying on these ",
      "results.\n\n")
}

cat("No analytic decisions — estimand, SL library, truncation bounds, covariate ",
    "set — were changed based on the observed outcomes.\n")
```


# Full Simulation Results {#full-sim}

```{r full-sim-load}
out_dir <- file.path("..", "results", "scenario1_v4")
has_full_sim <- dir.exists(out_dir)
if (has_full_sim) {
  load_mode <- function(mode) {
    d <- file.path(out_dir, mode)
    tryCatch(list(
      truth   = readRDS(file.path(d, "truth.rds")),
      results = readRDS(file.path(d, "replicate_results.rds")),
      metrics = read.csv(file.path(d, "summary_overall.csv")),
      gate    = readRDS(file.path(d, "gate_results.rds")),
      params  = readRDS(file.path(d, "params_used.rds"))
    ), error = function(e) NULL)
  }
  weak   <- load_mode("weak")
  strong <- load_mode("strong")
  gate_combined <- tryCatch(
    readRDS(file.path(out_dir, "gate_combined.rds")),
    error = function(e) NULL)
} else {
  weak <- strong <- gate_combined <- NULL
}
```

The compact simulation above demonstrates the staged workflow using a simple
enriched-GLM specification for both $g$ and $Q$. The production-scale
simulation uses richer estimators (TMLE-ML with SuperLearner, CV-TMLE,
PS-matched regression), additional sample sizes, and a mixed-overlap regime.
If those results are available, they are shown below as an extended reference.
Note that the production simulation uses a different DGP parameterization
than the tutorial example, so results are not directly comparable.

```{r full-sim-results, results='asis'}
if (!is.null(weak) && !is.null(strong)) {
  cat("### Full simulation results available\n\n")
  cat("These results come from the v4 simulation with N =", weak$params$n_main,
      "and", weak$params$reps, "replicates per mode.\n\n")

  point_methods <- c("Crude", "Adj. Regression (analytic)",
                     "IPTW (analytic)", "PS-Matched (pairs boot)",
                     "TMLE-GLM (IC)", "TMLE-ML (IC)", "CV-TMLE-ML (IC)")
  show_perf <- function(metrics, label) {
    sub <- metrics[metrics$subset == "all" &
                   metrics$method %in% point_methods, ]
    sub$mode <- label
    sub[, c("mode", "method", "bias", "rmse", "coverage", "mcse_coverage")]
  }
  combined <- rbind(show_perf(weak$metrics, "Weak W4"),
                    show_perf(strong$metrics, "Strong W4"))
  kable(combined, digits = 4, row.names = FALSE,
        caption = "Full simulation: estimator performance (all replicates)")

  if (!is.null(gate_combined)) {
    cat("\n### Stop/go gate results\n\n")
    kable(gate_combined, digits = 4, row.names = FALSE,
          caption = "Gate results from full simulation")
  }
} else {
  cat("Full simulation outputs not available. ",
      "Run: `Rscript scripts/sim_scenario1_v4_false_stop_inference.R`\n")
}
```


# Common Mistakes When Operationalizing TMLE in a Clean Room {#common-mistakes}

::: {.callout-warning}
### Pitfalls to avoid

1. **Fitting the real outcome regression $Q(A,W)$ before outcome access is
   permitted.** This violates the clean-room principle. All $Q$-fitting on
   real data belongs to Stage 4.

2. **Treating Stage 2 propensity score diagnostics as if they validate the
   full TMLE pipeline.** Stage 2 evaluates $g$ only. It says nothing about
   whether $Q$ estimation, the targeting step, or CI coverage will be adequate.

3. **Changing the Super Learner library, truncation rule, or estimand after
   seeing real outcomes without documenting a formal amendment.** Undocumented
   post-hoc changes undermine the entire clean-room rationale. All revisions
   must be recorded and justified.

4. **Interpreting acceptable overlap diagnostics as proof that TMLE inference
   will be valid.** Good overlap is neither necessary nor sufficient for valid
   TMLE. The outcome surface, sample size, and learner flexibility all matter.

5. **Assuming poor propensity score overlap automatically invalidates TMLE
   without checking whether the outcome model can compensate.** This ignores
   double robustness. Stage 3 simulation is the appropriate tool for resolving
   this question.

6. **Confusing a locked modeling strategy with a requirement to reuse the
   exact same fitted nuisance object.** What is locked is the specification
   (formula, learner library, truncation bounds). Re-running the same procedure
   on the same data is acceptable; changing the specification is not.
:::


# Key Takeaways {#takeaways}

```{r takeaways-text, results='asis'}
cat("1. **Stage 2 evaluates $g$ and overlap, not the entire estimator.** Propensity\n",
    "   score diagnostics are necessary but not sufficient for TMLE validation.\n",
    "   Acceptable overlap does not guarantee valid TMLE inference; flagged overlap\n",
    "   does not necessarily invalidate it.\n\n")

cat("2. **Stage 3 validates the full TMLE pipeline on synthetic data.** By testing\n",
    "   bias, variance, and coverage under a pre-specified DGP, the team makes an\n",
    "   evidence-based GO/STOP decision before seeing real outcomes.\n\n")

cat("3. **Stage 4 is where the real outcome regression and targeting step occur.**\n",
    "   The $g$-strategy is carried forward from Stage 2, but $Q$ is fit on real\n",
    "   data only at this final stage. No analytic decisions change after outcome\n",
    "   access.\n\n")

cat("4. **Outcome-blind simulation adds information that Stage 2 cannot provide.**\n")
if (gate_A$decision == "GO" && gate_B$decision == "STOP") {
  cat("   This tutorial demonstrated both cases: in Scenario A, Stage 3 showed\n",
      "   that TMLE performs adequately despite overlap stress (a false stop\n",
      "   averted). In Scenario B, Stage 3 confirmed that the same overlap pattern\n",
      "   genuinely degrades TMLE when the overlap-driving covariate also strongly\n",
      "   predicts the outcome (a true stop).\n\n")
} else if (gate_A$decision == "STOP" && gate_B$decision == "STOP") {
  cat("   In this tutorial run, Stage 3 revealed that the planned TMLE pipeline\n",
      "   did not meet performance targets under either scenario — information\n",
      "   that Stage 2 diagnostics alone could not have provided.\n\n")
} else {
  cat("   Stage 3 can distinguish cases where the TMLE pipeline works from cases\n",
      "   where it does not, based on the full estimator behavior rather than\n",
      "   just the propensity score.\n\n")
}

cat("5. **Whether overlap is a real problem depends on the outcome surface.**\n",
    "   The same propensity score distribution can be innocuous (when the\n",
    "   overlap driver weakly predicts the outcome) or fatal (when it strongly\n",
    "   predicts the outcome). IPTW, which relies entirely on $g$, is hurt by\n",
    "   poor overlap regardless. TMLE, through double robustness, can sometimes\n",
    "   compensate — but this is an asymptotic property, and Stage 3 simulation\n",
    "   is the tool for checking whether it holds in practice.\n")
```


# Practical Checklist for a Clean-Room TMLE Analysis {#checklist}

Use this checklist to ensure each stage is properly documented and archived.

### Stage 1: Pre-specification and Analysis Lock

- [ ] Causal question and estimand clearly stated
- [ ] Eligibility criteria, treatment definition, and outcome definition
- [ ] Covariate list with variable roles (confounders, effect modifiers)
- [ ] Super Learner libraries specified for both $g$ and $Q$
- [ ] PS truncation bounds
- [ ] Stage 2 diagnostic thresholds (extreme PS fraction, ESS, max weight, SMD)
- [ ] Stage 3 performance thresholds (max bias, min coverage)
- [ ] Versioned analysis lock file with timestamp and hash

### Stage 2: Outcome-Blind Design Diagnostics

- [ ] Fitted $g$-strategy (or locked specification to be re-run)
- [ ] Propensity score distribution summaries and overlap plots
- [ ] ESS and IPTW weight summaries
- [ ] Covariate balance diagnostics (love plot, weighted SMDs)
- [ ] List of triggered diagnostic flags
- [ ] Documented Stage 2 decision (PASS / FLAG — proceed to Stage 3)

### Stage 3: Outcome-Blind Simulation Validation

- [ ] Pre-specified DGP description (before real outcome access)
- [ ] Simulation seeds and settings (reps, sample size, scenarios)
- [ ] Code used for simulation runs (archived and versioned)
- [ ] Bias, RMSE, empirical SD, mean SE, and coverage for each estimator
- [ ] Comparison estimator results (crude, IPTW, etc.)
- [ ] Gate decision table: metric, threshold, observed value, GO/STOP
- [ ] Documented GO / FLAG / STOP decision with rationale
- [ ] Any plan amendment and its justification

### Stage 4: Unblinded TMLE Estimation

- [ ] Final $Q$ fit (first time on real outcomes)
- [ ] Final TMLE point estimate and 95% confidence interval
- [ ] Comparator estimates (crude, IPTW)
- [ ] Cross-reference to Stage 2 diagnostics and Stage 3 gate result
- [ ] Confirmation that no specifications were changed after outcome access
- [ ] Any deviations from the locked plan and their justification
- [ ] Final reproducibility metadata (session info, package versions, seed)


## Reproducing the Full Simulation

```bash
# Compact tutorial simulation: runs within this QMD (200 reps, enriched GLM)

# Production simulation with SuperLearner, 7 estimators, mixed overlap:
Rscript scripts/sim_scenario1_v4_false_stop_inference.R fast   # N=400, 120 reps
Rscript scripts/sim_scenario1_v4_false_stop_inference.R eval   # N=600, 300 reps
```


## References

- van der Laan MJ, Rose S. *Targeted Learning: Causal Inference for
  Observational and Experimental Data*. Springer, 2011.
- van der Laan MJ, Rose S. *Targeted Learning in Data Science*. Springer, 2018.
- Schuler MS, Rose S. Targeted maximum likelihood estimation for causal
  inference in observational studies. *American Journal of Epidemiology*, 2017.
- Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ. Diagnosing and
  responding to violations in the positivity assumption.
  *Statistical Methods in Medical Research*, 2012.
- Gruber S, van der Laan MJ. tmle: An R package for targeted maximum
  likelihood estimation. *Journal of Statistical Software*, 2012.
