---
title: "Operationalizing TMLE in a Clean-Room Staged Design"
subtitle: "A Pharmacoepidemiology Tutorial with Simulated HCV Safety Data"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

## Abstract {.unnumbered}

Targeted Maximum Likelihood Estimation (TMLE) is a doubly robust, semiparametric
efficient estimator increasingly used in pharmacoepidemiology and comparative
effectiveness research. However, TMLE's reliance on both a treatment model and
an outcome model creates an operational challenge: in a credible analysis, key
modeling decisions should be locked before outcome data are examined. This
tutorial introduces a **four-stage clean-room workflow** that resolves this
tension. In **Stage 1**, the analysis team pre-specifies and locks the estimand,
covariate set, learner libraries, truncation rules, and performance thresholds.
In **Stage 2**, the team accesses only treatment and covariate data to estimate
the propensity score, evaluate overlap, assess matching feasibility, and lock the
treatment-model strategy — without touching outcomes. In **Stage 3**, the team
runs the full TMLE pipeline on synthetic data generated under pre-specified,
outcome-blind data-generating processes with known truth, evaluating bias,
variance, and empirical coverage to make a GO/STOP gate decision. Only in
**Stage 4** are real outcomes accessed and the outcome regression fitted for the
first time. Using a simulated hepatitis C antiviral safety example, we
demonstrate the workflow with three simulation scenarios that illustrate distinct
clean-room decision states: acceptable overlap where TMLE achieves a precision
gain by using penalized regression (`glmnet`) to learn outcome-model structure
that design-only estimators cannot exploit, overlap stress where all estimators
— including TMLE — fail, and overlap stress where TMLE remains competitive
despite reduced matching feasibility because the outcome model compensates for
the stressed treatment model. The tutorial shows that propensity-score
diagnostics and full estimator validation are not interchangeable, that TMLE
can improve precision when the outcome surface contains learnable structure,
and that outcome-blind simulation is necessary to determine whether a planned
analysis should proceed.

```{r setup}
library(knitr)
library(ggplot2)
library(glmnet)
set.seed(2024)

# Utility functions
expit <- function(x) 1 / (1 + exp(-x))
logit <- function(p) log(p / (1 - p))
```

# What This Tutorial Is and Is Not

This is **not** a generic TMLE tutorial. For standalone introductions to TMLE,
see van der Laan and Rose (2011), the `tmle` R package vignette, or Schuler and
Rose (2017). This tutorial teaches how TMLE is **operationalized within a
clean-room staged workflow** — a governance structure that controls *when*
different data elements may be accessed and *what* analytic tasks are permitted
at each stage.

**Elevator pitch.** This tutorial shows how TMLE can be embedded in a clean-room
workflow where treatment-mechanism diagnostics occur before outcome access,
simulation is used to validate the full estimator pipeline under known truth,
and the real outcome regression and targeting step occur only after the earlier
stages are completed. The result is an analysis whose credibility does not rest
solely on post-hoc sensitivity checks, because the key analytic decisions were
locked before outcomes were seen.

::: {.callout-note}
### Tutorial vs. production clean-room implementation

In a real operational clean-room analysis, the workflow would also include
role separation (e.g., a blinded analyst team and an unblinded data
custodian), audit trails with timestamped artifacts, frozen data extracts,
and versioned analysis scripts archived at each stage transition. This
tutorial simplifies those operational components so the reader can focus on
the statistical logic — the sequence of analysis lock → outcome-blind
diagnostics → outcome-blind simulation validation → unblinded estimation.
The core sequencing principle is the same in both settings: analytic
decisions are locked before outcome data are accessed.
:::

### The running example

We use a simulated pharmacoepidemiology study loosely modeled on HCV antiviral
safety surveillance. The causal question: does a sofosbuvir (SOF)-containing
regimen increase the risk of acute kidney injury (AKI) compared to a non-SOF
regimen, among patients with chronic hepatitis C? The example is entirely
simulated — no real patient data are used — but the covariate structure,
treatment assignment mechanism, and outcome model reflect patterns seen in
real-world safety studies.

### The central tension

TMLE requires both a treatment model $g(W) = P(A=1|W)$ and an outcome model
$Q(A,W) = E[Y|A,W]$. In a clean-room workflow, these two components live in
different stages:

- **$g$ can be estimated early** (Stage 2), because it uses only treatment and
  covariates — no outcomes required.
- **$Q$ for the real data can only be estimated after outcome access**
  (Stage 4), because fitting an outcome regression requires $Y$.
- **The targeting step** combines $Q$ and $g$ and therefore also belongs to
  Stage 4 for the real analysis.
- **In simulation (Stage 3)**, the full pipeline ($g$, $Q$, targeting) runs on
  synthetic data with known truth, validating the estimator before real
  outcomes are unblinded.

### Workflow overview

```
┌─────────────────────────────────────────────┐
│  Stage 1. Pre-specification & Analysis Lock │
│  Lock estimand, covariates, SL libraries,   │
│  truncation rules, gate thresholds          │
└────────────────────┬────────────────────────┘
                     ↓
┌─────────────────────────────────────────────┐
│  Stage 2. Outcome-Blind Design Diagnostics  │
│  Access (W, A) only. Fit g(W), evaluate     │
│  overlap, balance, ESS. Lock g-strategy.    │
└────────────────────┬────────────────────────┘
                     ↓
┌─────────────────────────────────────────────┐
│  Stage 3. Outcome-Blind Simulation          │
│  Run full TMLE pipeline on synthetic data   │
│  with known truth. Evaluate bias, coverage. │
│  Make GO / STOP gate decision.              │
└────────────────────┬────────────────────────┘
                     ↓
┌─────────────────────────────────────────────┐
│  Stage 4. Unblinded TMLE Estimation         │
│  Access Y. Fit Q(A,W) for the first time.   │
│  Run targeting step. Report results.        │
└─────────────────────────────────────────────┘
```


## Quick Reference: Allowed vs Not Allowed Before Outcome Access

```{r permissions-table}
perms <- data.frame(
  Stage = c(
    "Stage 1: Pre-specification",
    "Stage 2: Outcome-Blind Diagnostics",
    "Stage 3: Simulation Validation",
    "Stage 4: Unblinded Estimation"
  ),
  `Data accessible` = c(
    "None (metadata, variable definitions only)",
    "W and A from the real dataset (not Y)",
    "Synthetic data with simulated Y",
    "Full dataset: W, A, Y"
  ),
  `Allowed TMLE-related tasks` = c(
    paste("Define causal question and estimand;",
          "specify covariates and treatment;",
          "select SL libraries for g and Q;",
          "set truncation rules;",
          "define diagnostic thresholds and gate criteria"),
    paste("Estimate g(W) on real data;",
          "PS overlap diagnostics;",
          "IPTW weight distributions and ESS;",
          "covariate balance (SMD)"),
    paste("Run the full TMLE pipeline on simulated data;",
          "evaluate bias, variance, and coverage;",
          "compare TMLE with simpler estimators;",
          "test SL specifications and truncation"),
    paste("Estimate Q(A,W) on real data;",
          "combine with locked g-strategy;",
          "run TMLE targeting step;",
          "compute estimates and CIs;",
          "report results")
  ),
  `Not allowed` = c(
    paste("Estimating g or Q on real data;",
          "any estimator involving real outcomes"),
    paste("Fitting Q(A,W) on real outcomes;",
          "computing treatment effects on real data;",
          "running TMLE on the real dataset"),
    paste("Using real outcomes;",
          "tuning estimator based on real Y"),
    paste("Changing estimand, SL library, or truncation",
          "without a documented protocol amendment")
  ),
  check.names = FALSE
)
kable(perms, caption = "Clean-room permissions by stage",
      col.names = c("Stage", "Data accessible", "Allowed TMLE-related tasks", "Not allowed"),
      align = c("l", "l", "l", "l")) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = TRUE, font_size = 13) |>
  kableExtra::column_spec(1, bold = TRUE, width = "12em") |>
  kableExtra::column_spec(2, width = "14em") |>
  kableExtra::column_spec(3, width = "20em") |>
  kableExtra::column_spec(4, width = "14em")
```


# Stage 1. Pre-specification and Analysis Lock {#stage-1}

## What happens at this stage

Before any real study data are accessed, the analysis team writes a statistical
analysis plan (SAP) that locks all major analytic decisions. This is analogous
to a clinical trial's SAP being finalized before database lock.

For our HCV example, the causal question is:

> Among adults with chronic HCV initiating antiviral therapy, what is the
> average causal effect of a SOF-containing regimen (vs. a non-SOF regimen)
> on the 90-day risk of acute kidney injury?

The estimand is the average treatment effect (ATE) on the risk difference scale:
$$\psi = E[Y^1] - E[Y^0]$$

where $Y^a$ denotes the potential AKI outcome under treatment $a$.

Identification assumptions: consistency ($Y = Y^A$), no unmeasured confounding
($Y^a \perp\!\!\!\perp A \mid W$), and positivity ($0 < P(A=1|W) < 1$).

### The analysis lock object

The pre-specified decisions are encoded as a versioned, timestamped artifact.
This object travels through every subsequent stage unchanged unless a formal
amendment is documented.

```{r analysis-lock}
create_analysis_lock <- function() {
  list(
    # --- Protocol metadata ---
    protocol_id   = "HCV-AKI-2024-001",
    version       = "1.0",
    locked_date   = Sys.Date(),
    locked_by     = "Analysis team (blinded)",
    blinded       = TRUE,  # TRUE until Stage 4 unblinding
    status        = "pre-outcome",

    # --- Causal question ---
    study       = "HCV antiviral safety: SOF vs non-SOF",
    estimand    = "ATE",
    effect_scale = "risk_difference",

    # --- Variable roles ---
    covariates  = c("age", "ckd", "cirrhosis", "diabetes", "nsaid", "renal_risk"),
    treatment   = "A",          # 1 = SOF-containing, 0 = non-SOF
    outcome     = "Y",          # 1 = AKI within 90 days

    # --- Treatment model (g) ---
    g_formula   = "A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk",
    g_SL_library = c("SL.mean", "SL.glm", "SL.glmnet"),

    # --- Outcome model (Q) ---
    Q_method    = "cv.glmnet (elastic net, alpha = 0.5)",
    Q_features  = paste("Main effects + age^2, age^3, renal_risk^2 +",
                        "treatment interactions (A:ckd, A:age, A:age^2,",
                        "A:renal_risk, A:cirrhosis, A:diabetes) +",
                        "covariate interactions (age:ckd, age:rr, ckd:rr)"),
    Q_SL_library = c("SL.mean", "SL.glm", "SL.glmnet", "SL.gam"),

    # --- PS truncation ---
    ps_truncation = c(lower = 0.025, upper = 0.975),

    # --- TMLE variant ---
    tmle_variant  = "TMLE-ML",

    # --- Stage 2 diagnostic thresholds ---
    diag_thresholds = list(
      extreme_ps_bound = 0.05,    # PS below this or above 1 - this
      max_extreme_frac = 0.15,    # max fraction of extreme PS
      min_ess_frac     = 0.50,    # minimum ESS / n
      max_iptw_weight  = 40,      # max single IPTW weight
      max_smd_weighted = 0.10     # max weighted SMD
    ),

    # --- Stage 3 simulation targets ---
    sim_targets = list(
      max_abs_bias = 0.01,
      min_coverage = 0.93,
      se_sd_low    = 0.80,
      se_sd_high   = 1.20,
      n_reps       = 200
    ),

    # --- Amendment rule ---
    amendment_rule = paste("If Stage 3 returns STOP, the team may file a",
                           "documented amendment to expand the learner library",
                           "or increase the sample size. The amended plan must",
                           "be re-validated through Stage 3 before proceeding",
                           "to Stage 4.")
  )
}

analysis_lock <- create_analysis_lock()

cat("Analysis lock (", analysis_lock$protocol_id, " v", analysis_lock$version, ")\n", sep = "")
cat("Study:", analysis_lock$study, "\n")
cat("Locked:", as.character(analysis_lock$locked_date), "\n")
cat("Blinded:", analysis_lock$blinded, " | Status:", analysis_lock$status, "\n")
cat("Estimand:", analysis_lock$estimand, "on", analysis_lock$effect_scale, "scale\n")
cat("Covariates:", paste(analysis_lock$covariates, collapse = ", "), "\n")
cat("PS truncation:", analysis_lock$ps_truncation["lower"], "-",
    analysis_lock$ps_truncation["upper"], "\n")
cat("Amendment rule:", analysis_lock$amendment_rule, "\n")
```

::: {.callout-note}
### What gets locked vs. what gets re-estimated

What is locked is the **modeling strategy**: the learner library, truncation
bounds, covariate set, and diagnostic thresholds. In practice, the team may
either carry the Stage 2 fitted $\hat{g}$ object forward into Stage 4, or
re-run the same locked procedure on the full $(W, A)$ data at Stage 4. Both
are valid — the key constraint is that the *specification* does not change
after outcome access. Locking a strategy is not the same as requiring reuse of
a specific fitted object.
:::

### Stage 1 summary

At this point, no data have been touched. The team has committed to an estimand
(ATE on the RD scale), identification assumptions, covariate list, SL
libraries, truncation rules, and simulation performance targets.


# Stage 2. Outcome-Blind Design Diagnostics {#stage-2}

## What happens at this stage

The analysis team receives the **treatment and covariate data** $(W, A)$ from
the real study. Outcomes $(Y)$ remain sealed. The purpose is to evaluate
whether the treatment mechanism and covariate structure are compatible with the
planned analysis.

Because $g(W) = P(A=1|W)$ does not involve outcomes, the entire treatment model
can be estimated here. This is the key operational insight: **the $g$-side of
TMLE is fully accessible during the outcome-blind stage.**

::: {.callout-note collapse="true"}
### Operational note: Stage 2 in practice
In a real clean-room implementation, the analyst would receive a blinded
analytic dataset or restricted data extract with outcomes masked or removed.
All Stage 2 diagnostics would be scripted, versioned, and archived as part of
the audit trail. Any changes to the $g$-strategy triggered by Stage 2
findings (e.g., adding a covariate interaction to improve balance) would
require documented justification and an updated analysis lock before
proceeding to Stage 3.
:::

### Generate the HCV tutorial dataset

We simulate a dataset that mimics an HCV antiviral safety study. We generate
the full dataset including $Y$, but the Stage 2 analyst sees only $(W, A)$.

```{r generate-data}
simulate_hcv_scenario <- function(n, seed,
                                   b_rr_g    = 0.7,    # g coeff for renal_risk
                                   beta_rr_Q = 0.35,   # Q coeff for renal_risk
                                   beta_int  = 0.00,   # A:renal_risk interaction
                                   beta_ckdA = 0.40,   # A:ckd interaction
                                   beta_ageA = 0.00,   # A:age interaction
                                   alpha_A   = -0.15,  # treatment intercept
                                   alpha_Y   = -2.80) {
  set.seed(seed)

  # --- Baseline covariates ---
  age_raw    <- rnorm(n, 55, 10)
  age        <- (age_raw - 55) / 10           # standardized
  ckd        <- rbinom(n, 1, 0.18)            # chronic kidney disease
  cirrhosis  <- rbinom(n, 1, 0.22)
  diabetes   <- rbinom(n, 1, 0.20)
  nsaid      <- rbinom(n, 1, 0.25)            # concurrent NSAID use
  renal_risk <- rnorm(n, 0, 1)               # composite renal-risk score

  # --- Treatment mechanism (g) ---
  # renal_risk drives overlap; b_rr_g varies by scenario
  lp_g <- alpha_A + 0.25*age + 0.45*ckd + 0.35*cirrhosis +
           0.20*diabetes - 0.15*nsaid + b_rr_g*renal_risk
  g_true <- pmin(pmax(expit(lp_g), 0.005), 0.995)
  A <- rbinom(n, 1, g_true)

  # --- Outcome model (Q): AKI ---
  # Includes nonlinear age (quadratic + cubic), renal_risk^2, and
  # treatment-effect heterogeneity (A:ckd, A:age, A:renal_risk)
  # so the DGP has learnable structure that a richer Q can exploit.
  lp_Q <- alpha_Y + 0.25*A + 0.20*age + 0.15*age^2 + 0.05*age^3 +
           0.90*ckd + 0.55*cirrhosis + 0.40*diabetes + 0.25*nsaid +
           beta_rr_Q*renal_risk + 0.08*renal_risk^2 +
           beta_ckdA*A*ckd + beta_ageA*A*age + beta_int*A*renal_risk
  Y <- rbinom(n, 1, expit(lp_Q))

  data.frame(age, ckd, cirrhosis, diabetes, nsaid, renal_risk, A, Y)
}

# Generate the "real" study data (n = 3000)
# Using Scenario 3 parameters: overlap stress (b_rr=2.4), weak renal_risk in Q
dat_full <- simulate_hcv_scenario(n = 3000, seed = 42,
                                   b_rr_g = 2.4, beta_rr_Q = 0.03,
                                   beta_int = 0.00, beta_ckdA = 0.25,
                                   beta_ageA = 0.10)

# Stage 2 analyst sees only W and A — Y is sealed
dat_stage2 <- dat_full[, c("age", "ckd", "cirrhosis", "diabetes",
                            "nsaid", "renal_risk", "A")]

cat("Stage 2 data dimensions:", nrow(dat_stage2), "x", ncol(dat_stage2), "\n")
cat("Treatment prevalence:", round(mean(dat_stage2$A), 3), "\n")
cat("Y column accessible:", "Y" %in% names(dat_stage2), "\n")
```

### Fit the treatment model

Using the locked g-formula from the analysis lock:

```{r fit-g}
fit_stage2_g <- function(dat, lock) {
  g_fit <- glm(as.formula(lock$g_formula), data = dat, family = "binomial")
  g_hat <- fitted(g_fit)
  g_trunc <- pmin(pmax(g_hat, lock$ps_truncation["lower"]),
                  lock$ps_truncation["upper"])
  list(fit = g_fit, g_hat = g_hat, g_trunc = g_trunc)
}

g_stage2 <- fit_stage2_g(dat_stage2, analysis_lock)

cat("PS range (raw):", round(range(g_stage2$g_hat), 4), "\n")
cat("PS range (truncated):", round(range(g_stage2$g_trunc), 4), "\n")
```

### Propensity score diagnostics

These diagnostics use only $(W, A)$ and are fully permitted at Stage 2.

```{r ps-diagnostics, fig.width=8, fig.height=4}
ps_df <- data.frame(
  ps = g_stage2$g_hat,
  Treatment = factor(dat_stage2$A,
                     labels = c("Non-SOF", "SOF-containing"))
)

ggplot(ps_df, aes(x = ps, fill = Treatment)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = c(analysis_lock$ps_truncation["lower"],
                             analysis_lock$ps_truncation["upper"]),
             linetype = "dashed", color = "red") +
  labs(title = "Stage 2: Propensity Score Distribution by Treatment",
       subtitle = "Red dashed lines = pre-specified truncation bounds",
       x = "Estimated P(SOF | W)", y = "Count") +
  theme_minimal(base_size = 12)
```

```{r overlap-metrics}
stage2_diagnostics <- function(g_hat, g_trunc, A, dat, lock) {
  thr <- lock$diag_thresholds
  n <- length(A)

  # Fraction with extreme PS
  frac_extreme <- mean(g_hat < thr$extreme_ps_bound |
                       g_hat > (1 - thr$extreme_ps_bound))

  # IPTW weights and ESS
  w_iptw <- A / g_trunc + (1 - A) / (1 - g_trunc)
  ess <- sum(w_iptw)^2 / sum(w_iptw^2)
  ess_frac <- ess / n
  max_w <- max(w_iptw)

  # Weighted SMDs
  compute_smd <- function(x, trt, weights = NULL) {
    if (is.null(weights)) weights <- rep(1, length(x))
    m1 <- weighted.mean(x[trt == 1], weights[trt == 1])
    m0 <- weighted.mean(x[trt == 0], weights[trt == 0])
    s <- sqrt((var(x[trt == 1]) + var(x[trt == 0])) / 2)
    if (s < 1e-10) return(0)
    (m1 - m0) / s
  }

  covars <- lock$covariates
  smd_unw <- sapply(covars, function(v) compute_smd(dat[[v]], A))
  smd_wtd <- sapply(covars, function(v) compute_smd(dat[[v]], A, w_iptw))

  diag_tab <- data.frame(
    Metric = c("Fraction extreme PS",
               "ESS fraction",
               "Max IPTW weight",
               "Max |SMD| (IPTW-weighted)"),
    Value = round(c(frac_extreme, ess_frac, max_w, max(abs(smd_wtd))), 3),
    Threshold = c(paste("<", thr$max_extreme_frac),
                  paste(">", thr$min_ess_frac),
                  paste("<", thr$max_iptw_weight),
                  paste("<", thr$max_smd_weighted)),
    Status = c(ifelse(frac_extreme < thr$max_extreme_frac, "PASS", "FLAG"),
               ifelse(ess_frac > thr$min_ess_frac, "PASS", "FLAG"),
               ifelse(max_w < thr$max_iptw_weight, "PASS", "FLAG"),
               ifelse(max(abs(smd_wtd)) < thr$max_smd_weighted, "PASS", "FLAG"))
  )

  list(diag_tab = diag_tab, smd_unw = smd_unw, smd_wtd = smd_wtd,
       w_iptw = w_iptw, any_flag = any(diag_tab$Status == "FLAG"))
}

s2_diag <- stage2_diagnostics(g_stage2$g_hat, g_stage2$g_trunc,
                               dat_stage2$A, dat_stage2, analysis_lock)
kable(s2_diag$diag_tab, caption = "Stage 2 overlap diagnostics")
```

```{r love-plot, fig.width=7, fig.height=4}
covars <- analysis_lock$covariates
smd_df <- data.frame(
  variable = rep(covars, 2),
  smd = c(s2_diag$smd_unw, s2_diag$smd_wtd),
  type = rep(c("Unweighted", "IPTW-Weighted"), each = length(covars))
)

ggplot(smd_df, aes(x = abs(smd), y = reorder(variable, abs(smd)), color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.10, linetype = "dashed") +
  labs(title = "Stage 2: Covariate Balance (Love Plot)",
       x = "|Standardized Mean Difference|", y = NULL, color = "Weighting") +
  theme_minimal(base_size = 12)
```

### Stage 2 decision

```{r stage2-decision}
stage2_status <- ifelse(s2_diag$any_flag,
                         "FLAG — proceed to Stage 3 simulation",
                         "PASS")
cat("Stage 2 status:", stage2_status, "\n")
if (s2_diag$any_flag) {
  flagged <- s2_diag$diag_tab$Metric[s2_diag$diag_tab$Status == "FLAG"]
  cat("Flagged metrics:", paste(flagged, collapse = "; "), "\n")
}
```

### Why Stage 2 is not enough for TMLE validation

Stage 2 evaluates the treatment mechanism $g$ and practical positivity. But
TMLE depends on *three* components: $g$, $Q$, and the targeting step. Stage 2
tells us nothing about whether the outcome model can be estimated well,
whether the targeting step is stable, whether the variance estimator is
reliable, or whether CI coverage is near nominal.

This matters in both directions:

- A study can have **flagged overlap but excellent TMLE performance** if the
  overlap-driving covariate is weakly related to the outcome — the outcome
  model compensates.
- A study can have **acceptable overlap but poor TMLE performance** if the
  outcome surface is difficult to learn at the given sample size.

Stage 2 alone cannot distinguish these cases. **Stage 3 fills the gap** by
testing the complete TMLE pipeline on synthetic data where the truth is known.

### Stage 2 summary

```{r stage2-summary, results='asis'}
cat("At this point the team has estimated $g(W)$, evaluated overlap and balance, ")
if (s2_diag$any_flag) {
  cat("and documented that **some diagnostic flags triggered**. ")
  cat("These flags indicate potential overlap concerns — but as discussed below, ")
  cat("overlap flags alone do not determine whether TMLE will perform adequately. ")
} else {
  cat("and found that **all diagnostic thresholds were met**. ")
  cat("However, passing Stage 2 does not guarantee that the full TMLE pipeline ")
  cat("will perform adequately — as Stage 3 will show. ")
}
cat("The outcome $Y$ has **not been touched**. The $g$-modeling strategy is now locked.")
```


# Why TMLE Can Survive Poor Overlap {#double-robustness}

Before proceeding to the simulation stage, it helps to understand *why* poor
propensity score diagnostics do not automatically invalidate TMLE. The answer
is **double robustness**.

TMLE relies on two nuisance components:

1. The **outcome regression** $Q(A,W) = E[Y|A,W]$: a model for the expected
   outcome given treatment and covariates.
2. The **treatment mechanism** $g(W) = P(A=1|W)$: a model for the probability
   of receiving treatment given covariates.

The key property is this: **TMLE is consistent if *either* the outcome model
or the treatment model is consistently estimated.** Both do not need to be
correct simultaneously for TMLE to converge to the right answer. If one model
is wrong but the other is right, TMLE uses the correct model to compensate for
the incorrect one.

When *both* models are consistently estimated, TMLE achieves an additional
benefit: it is **efficient**, meaning it reaches the smallest possible variance
among all regular estimators.

What does this mean practically?

- If the overlap-driving covariate barely affects the outcome (weak
  confounding through that variable), then the outcome model $Q$ can be
  estimated well even though $g$ is difficult. TMLE leans on $Q$ to get the
  right answer. (This is the mechanism behind Scenario 3 in Stage 3.)
- If the overlap-driving covariate strongly affects both treatment and outcome,
  then *both* $g$ and $Q$ are hard to estimate in the regions of poor overlap.
  TMLE has nothing to fall back on. (This is what happens in Scenario 2.)
- **IPTW** and **PS-matched regression**, by contrast, rely entirely on $g$.
  Poor overlap always hurts these methods, regardless of the outcome surface.

This is why overlap concerns — whether flagged at Stage 2 or not — do not
automatically determine whether TMLE will perform well or poorly. They mean
"the treatment model alone may be under stress — we need Stage 3 to check
whether the full TMLE pipeline, including the outcome model, can compensate."

::: {.callout-important}
Double robustness is an asymptotic property that holds under regularity
conditions, including adequate positivity and sufficiently good nuisance
estimation — it is not a finite-sample guarantee. In practice, extreme weights
from a poorly estimated $g$ can still destabilize the targeting step, inflate
variance, and degrade CI coverage even when the outcome model is well-specified.
This is exactly why simulation validation (Stage 3) is necessary: it checks
whether double robustness actually helps at the study's sample size and
covariate distribution, rather than relying on asymptotic theory alone.
:::


# Stage 3. Outcome-Blind Simulation Validation {#stage-3}

## What happens at this stage

Stage 2 told us about the treatment mechanism. Stage 3 answers the harder
question: **does the entire TMLE pipeline perform adequately under conditions
that resemble this study?**

The team runs a simulation using pre-specified data-generating processes (DGPs).
Each DGP is calibrated to match the observed $(W, A)$ distribution but generates
**synthetic outcomes** — not the real $Y$. Because the truth is known in the
simulation, the team can evaluate bias, variance, and CI coverage.

::: {.callout-note collapse="true"}
### Operational note: Stage 3 in practice
In a production clean-room analysis, Stage 3 simulations would be pre-run
from versioned scripts and their results archived as part of the audit trail.
The candidate DGPs are specified before outcome access, based on clinical
domain knowledge and the observed covariate/treatment distributions from
Stage 2. The analyst does not select a DGP post hoc to justify proceeding —
the scenarios are locked as part of the Stage 1 analysis plan or a documented
Stage 2 amendment. The purpose is to stress-test the locked estimator pipeline
under plausible candidate outcome surfaces before unblinding.
:::

We run **three scenarios** to demonstrate three distinct clean-room decision
states, using four estimators: crude RD, PS-matched regression, IPTW, and TMLE.

### What Stage 3 validates that Stage 2 cannot

| Question | Stage 2 | Stage 3 |
|----------|---------|---------|
| Is $\hat{g}$ well-estimated? | Yes | Yes |
| Are PS values extreme? | Yes | Yes |
| Can adequate matching be achieved on the observed data? | Yes | Partially |
| Does the matched estimator perform adequately under plausible DGPs? | **No** | Yes |
| Is $\hat{Q}$ adequate? | **No** | Yes (on synthetic data) |
| Does the targeting step work? | **No** | Yes |
| Is IC-based SE reliable? | **No** | Yes |
| Does CI coverage reach nominal? | **No** | Yes |
| Is TMLE better than alternatives? | **No** | Yes |


### Three scenarios for the HCV example

| Scenario | Overlap / matching | Outcome structure | Expected TMLE gate | Main lesson |
|----------|--------------------|-------------------|--------------------|-------------|
| **1** | Acceptable | Moderate `renal_risk` ($\beta_{rr}=0.35$), $A \times \text{ckd}$, $A \times \text{age}$ | GO | TMLE precision gain via richer Q |
| **2** | Overlap stress | Strong `renal_risk` ($\beta_{rr}=1.30$, $A \times \text{renal\_risk}$ $=0.60$) | STOP | Overlap stress + strong confounding defeats all |
| **3** | Overlap stress | Near-zero `renal_risk` ($\beta_{rr}=0.03$), $A \times \text{ckd}$, $A \times \text{age}$ | GO | TMLE competitive under overlap stress via Q-learning |

Scenarios 2 and 3 share the same treatment mechanism (`b_rr` = 2.4 in
$g$), so their Stage 2 overlap and matching diagnostics are similarly stressed.
The key difference is whether `renal_risk` matters for the outcome. All three
scenarios include nonlinear age effects (quadratic + cubic), quadratic
`renal_risk`, and treatment-effect heterogeneity ($A \times \text{ckd}$,
$A \times \text{age}$). TMLE's `glmnet`-based $Q$ model with a rich
feature matrix can learn this structure, while the PS-matched estimator uses
only a simple paired difference that cannot exploit outcome modeling.

### Define the simulation DGP

The DGP must be specified *before* real outcomes are seen. It should reflect
plausible outcome models given clinical domain knowledge.

```{r sim-dgp}
# Compute true ATE by Monte Carlo marginalization over the covariate distribution
compute_truth_hcv <- function(n_mc = 500000, seed = 1,
                               beta_rr_Q = 0.35, beta_int = 0.00,
                               beta_ckdA = 0.40, beta_ageA = 0.00,
                               alpha_Y = -2.80) {
  set.seed(seed)
  age        <- rnorm(n_mc, 0, 1)
  ckd        <- rbinom(n_mc, 1, 0.18)
  cirrhosis  <- rbinom(n_mc, 1, 0.22)
  diabetes   <- rbinom(n_mc, 1, 0.20)
  nsaid      <- rbinom(n_mc, 1, 0.25)
  renal_risk <- rnorm(n_mc, 0, 1)

  lp_base <- alpha_Y + 0.20*age + 0.15*age^2 + 0.05*age^3 +
             0.90*ckd + 0.55*cirrhosis + 0.40*diabetes + 0.25*nsaid +
             beta_rr_Q*renal_risk + 0.08*renal_risk^2
  EY1 <- mean(expit(lp_base + 0.25 + beta_ckdA*ckd +
                     beta_ageA*age + beta_int*renal_risk))
  EY0 <- mean(expit(lp_base))

  c(EY1 = EY1, EY0 = EY0, RD = EY1 - EY0)
}

# Scenario 1: Moderate overlap, moderate renal_risk, strong A:ckd heterogeneity
truth_1 <- compute_truth_hcv(beta_rr_Q = 0.35, beta_int = 0.00,
                              beta_ckdA = 0.40, beta_ageA = 0.15)
# Scenario 2: Overlap stress, strong renal_risk in Q + A:renal_risk interaction
truth_2 <- compute_truth_hcv(beta_rr_Q = 1.30, beta_int = 0.60,
                              beta_ckdA = 0.40, beta_ageA = 0.00)
# Scenario 3: Overlap stress, near-zero renal_risk, moderate heterogeneity
truth_3 <- compute_truth_hcv(beta_rr_Q = 0.03, beta_int = 0.00,
                              beta_ckdA = 0.25, beta_ageA = 0.10)

cat("True ATE by scenario:\n")
cat("  Scenario 1 (moderate renal_risk):", round(truth_1["RD"], 5), "\n")
cat("  Scenario 2 (strong renal_risk): ", round(truth_2["RD"], 5), "\n")
cat("  Scenario 3 (weak renal_risk):   ", round(truth_3["RD"], 5), "\n")
```

### Define the estimators

This function runs one TMLE analysis using the pre-specified pipeline. The $Q$
model uses `glmnet` (penalized logistic regression) with a rich design matrix
including polynomial terms, pairwise interactions, and treatment-by-covariate
interactions. This is a modest but meaningful upgrade over a plain GLM: `glmnet`
automatically selects among the candidate features via cross-validated
regularization, approximating what a small SuperLearner library would do but
at tutorial speed. The $g$ model remains a simple logistic GLM.

```{r tmle-function}
# Build the rich design matrix for Q estimation
build_Q_design <- function(dat) {
  A <- dat$A; age <- dat$age; ckd <- dat$ckd; cirrhosis <- dat$cirrhosis
  diabetes <- dat$diabetes; nsaid <- dat$nsaid; rr <- dat$renal_risk

  cbind(
    A         = A,
    age       = age,
    age2      = age^2,
    age3      = age^3,
    ckd       = ckd,
    cirrhosis = cirrhosis,
    diabetes  = diabetes,
    nsaid     = nsaid,
    rr        = rr,
    rr2       = rr^2,
    # Treatment-effect heterogeneity terms
    A_ckd     = A * ckd,
    A_age     = A * age,
    A_age2    = A * age^2,
    A_rr      = A * rr,
    A_cirr    = A * cirrhosis,
    A_dm      = A * diabetes,
    # Pairwise covariate interactions
    age_ckd   = age * ckd,
    age_rr    = age * rr,
    ckd_rr    = ckd * rr
  )
}

run_tmle_once <- function(dat, ps_clip = c(0.025, 0.975)) {
  n <- nrow(dat)

  # --- Fit g (treatment model) — simple logistic GLM ---
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])

  # --- Fit Q (outcome model) — glmnet with rich design matrix ---
  X_Q <- build_Q_design(dat)
  Q_fit <- cv.glmnet(X_Q, dat$Y, family = "binomial", alpha = 0.5,
                      nfolds = 5, type.measure = "deviance")
  Q_A <- as.numeric(predict(Q_fit, X_Q, s = "lambda.min", type = "response"))

  dat1 <- dat; dat1$A <- 1
  dat0 <- dat; dat0$A <- 0
  Q_1 <- as.numeric(predict(Q_fit, build_Q_design(dat1),
                             s = "lambda.min", type = "response"))
  Q_0 <- as.numeric(predict(Q_fit, build_Q_design(dat0),
                             s = "lambda.min", type = "response"))

  # Bound Q away from 0/1
  Q_A <- pmin(pmax(Q_A, 1e-5), 1 - 1e-5)
  Q_1 <- pmin(pmax(Q_1, 1e-5), 1 - 1e-5)
  Q_0 <- pmin(pmax(Q_0, 1e-5), 1 - 1e-5)

  # --- Targeting step ---
  H  <- dat$A / g_hat - (1 - dat$A) / (1 - g_hat)
  H1 <- 1 / g_hat
  H0 <- -1 / (1 - g_hat)

  eps_fit <- glm(dat$Y ~ -1 + H, offset = logit(Q_A), family = "binomial")
  eps <- coef(eps_fit)

  Q_1_star <- expit(logit(Q_1) + eps * H1)
  Q_0_star <- expit(logit(Q_0) + eps * H0)
  Q_A_star <- expit(logit(Q_A) + eps * H)

  # --- Point estimate ---
  psi <- mean(Q_1_star - Q_0_star)

  # --- IC-based inference ---
  eif <- H * (dat$Y - Q_A_star) + (Q_1_star - Q_0_star) - psi
  se  <- sqrt(var(eif) / n)
  ci  <- psi + c(-1.96, 1.96) * se

  list(estimate = psi, se = se, ci_low = ci[1], ci_high = ci[2])
}
```

We also define three comparators: crude RD, IPTW, and PS-matched regression.
These comparators use simpler estimation strategies that do not leverage
outcome modeling — reflecting realistic practice where matching and IPTW
rely primarily on design-stage adjustment.

```{r comparator-functions}
run_crude <- function(dat) {
  p1 <- mean(dat$Y[dat$A == 1]); p0 <- mean(dat$Y[dat$A == 0])
  rd <- p1 - p0
  n1 <- sum(dat$A == 1); n0 <- sum(dat$A == 0)
  se <- sqrt(p1*(1 - p1)/n1 + p0*(1 - p0)/n0)
  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}

run_iptw <- function(dat, ps_clip = c(0.025, 0.975)) {
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])
  A <- dat$A; Y <- dat$Y; n <- length(Y)
  w  <- A / g_hat + (1 - A) / (1 - g_hat)
  d1 <- sum(w * A); d0 <- sum(w * (1 - A))
  mu1 <- sum(w * A * Y) / d1
  mu0 <- sum(w * (1 - A) * Y) / d0
  rd <- mu1 - mu0
  # Sandwich SE
  phi1 <- w * A * (Y - mu1) / d1 * n
  phi0 <- w * (1 - A) * (Y - mu0) / d0 * n
  se <- sqrt(sum((phi1 - phi0)^2) / n^2)
  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}

run_ps_matched <- function(dat, ps_clip = c(0.025, 0.975)) {
  # Fit propensity score
  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = dat, family = "binomial")
  ps <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])
  lps <- logit(ps)

  # 1:1 nearest-neighbor matching without replacement, caliper = 0.2 * SD(logit PS)
  caliper <- 0.2 * sd(lps)
  tidx <- which(dat$A == 1)
  cidx <- which(dat$A == 0)

  # Greedy matching in original order
  matched_t <- matched_c <- integer(0)
  available <- rep(TRUE, length(cidx))

  for (ii in seq_along(tidx)) {
    i <- tidx[ii]
    if (!any(available)) break
    ac <- cidx[available]
    d_lps <- abs(lps[i] - lps[ac])
    j <- which.min(d_lps)
    if (d_lps[j] <= caliper) {
      matched_t <- c(matched_t, i)
      matched_c <- c(matched_c, ac[j])
      available[which(cidx == ac[j])] <- FALSE
    }
  }

  np <- length(matched_t)
  if (np < 20) {
    return(list(estimate = NA, se = NA, ci_low = NA, ci_high = NA,
                n_pairs = np, matched_frac = np / length(tidx), max_smd = NA))
  }

  # Paired difference estimator
  rd <- mean(dat$Y[matched_t] - dat$Y[matched_c])
  se <- sd(dat$Y[matched_t] - dat$Y[matched_c]) / sqrt(np)

  # Post-match balance
  covars <- c("age", "ckd", "cirrhosis", "diabetes", "nsaid", "renal_risk")
  smds <- sapply(covars, function(v) {
    xt <- dat[[v]][matched_t]; xc <- dat[[v]][matched_c]
    s <- sqrt((var(xt) + var(xc)) / 2)
    if (s < 1e-10) 0 else (mean(xt) - mean(xc)) / s
  })

  list(estimate = rd, se = se,
       ci_low = rd - 1.96*se, ci_high = rd + 1.96*se,
       n_pairs = np, matched_frac = np / length(tidx),
       max_smd = max(abs(smds)))
}
```

### Run the three-scenario simulation

```{r run-sim, cache=TRUE}
run_stage3_simulation <- function(reps, n, b_rr_g, beta_rr_Q, beta_int,
                                   beta_ckdA = 0.40, beta_ageA = 0.00,
                                   alpha_A = -0.15, alpha_Y = -2.80,
                                   true_rd, seed_offset = 0,
                                   ps_clip = c(0.025, 0.975)) {
  methods <- c("Crude", "IPTW", "PS-Matched", "TMLE")
  results <- vector("list", reps * length(methods))
  k <- 0

  for (r in seq_len(reps)) {
    d <- simulate_hcv_scenario(n, seed = seed_offset + r,
                                b_rr_g = b_rr_g, beta_rr_Q = beta_rr_Q,
                                beta_int = beta_int, beta_ckdA = beta_ckdA,
                                beta_ageA = beta_ageA, alpha_A = alpha_A,
                                alpha_Y = alpha_Y)

    for (meth in methods) {
      res <- tryCatch(
        switch(meth,
               Crude        = run_crude(d),
               IPTW         = run_iptw(d, ps_clip),
               "PS-Matched" = run_ps_matched(d, ps_clip),
               TMLE         = run_tmle_once(d, ps_clip)),
        error = function(e) list(estimate = NA, se = NA,
                                  ci_low = NA, ci_high = NA))
      k <- k + 1
      results[[k]] <- data.frame(
        rep = r, method = meth,
        est = res$estimate, se = res$se,
        ci_lo = res$ci_low, ci_hi = res$ci_high)
    }
  }

  results <- do.call(rbind, results)

  # Compute performance metrics per method
  metrics <- do.call(rbind, lapply(split(results, results$method), function(sub) {
    ok     <- !is.na(sub$est)
    has_ci <- ok & !is.na(sub$ci_lo)
    e <- sub$est[ok]
    data.frame(
      method   = sub$method[1],
      n_ok     = sum(ok),
      bias     = mean(e) - true_rd,
      rmse     = sqrt(mean((e - true_rd)^2)),
      emp_sd   = sd(e),
      mean_se  = mean(sub$se[ok], na.rm = TRUE),
      coverage = mean(
        (sub$ci_lo[has_ci] <= true_rd) & (true_rd <= sub$ci_hi[has_ci])),
      n_ci     = sum(has_ci))
  }))
  row.names(metrics) <- NULL

  list(results = results, metrics = metrics)
}

n_sim <- 3000   # sample size per replicate (larger for glmnet stability)
n_reps <- 200   # number of replicates

# --- Scenario 1: Moderate overlap, moderate renal_risk, A:ckd + A:age heterogeneity ---
sim_1 <- run_stage3_simulation(
  reps = n_reps, n = n_sim,
  b_rr_g = 0.7, beta_rr_Q = 0.35, beta_int = 0.00,
  beta_ckdA = 0.40, beta_ageA = 0.15,
  true_rd = truth_1["RD"], seed_offset = 1000,
  ps_clip = analysis_lock$ps_truncation)

# --- Scenario 2: Overlap stress, strong renal_risk in Q ---
sim_2 <- run_stage3_simulation(
  reps = n_reps, n = n_sim,
  b_rr_g = 2.4, beta_rr_Q = 1.30, beta_int = 0.60,
  beta_ckdA = 0.40, beta_ageA = 0.00,
  true_rd = truth_2["RD"], seed_offset = 2000,
  ps_clip = analysis_lock$ps_truncation)

# --- Scenario 3: Overlap stress, near-zero renal_risk, moderate heterogeneity ---
sim_3 <- run_stage3_simulation(
  reps = n_reps, n = n_sim,
  b_rr_g = 2.4, beta_rr_Q = 0.03, beta_int = 0.00,
  beta_ckdA = 0.25, beta_ageA = 0.10,
  true_rd = truth_3["RD"], seed_offset = 3000,
  ps_clip = analysis_lock$ps_truncation)
```

### Overlap and matching diagnostics by scenario

Before examining estimator performance, we compute Stage 2-style diagnostics
for each scenario using a single representative dataset.

```{r scenario-diagnostics}
compute_scenario_diagnostics <- function(n = 5000, seed = 99,
                                          b_rr_g, beta_rr_Q, beta_int,
                                          beta_ckdA = 0.40, beta_ageA = 0.00,
                                          alpha_A = -0.15, alpha_Y = -2.80,
                                          ps_clip = c(0.025, 0.975)) {
  d <- simulate_hcv_scenario(n, seed, b_rr_g, beta_rr_Q, beta_int,
                              beta_ckdA, beta_ageA, alpha_A, alpha_Y)

  g_fit <- glm(A ~ age + ckd + cirrhosis + diabetes + nsaid + renal_risk,
               data = d, family = "binomial")
  ps <- fitted(g_fit)
  ps_trunc <- pmin(pmax(ps, ps_clip[1]), ps_clip[2])

  frac_extreme <- mean(ps < 0.05 | ps > 0.95)
  w <- d$A / ps_trunc + (1 - d$A) / (1 - ps_trunc)
  ess <- sum(w)^2 / sum(w^2)
  ess_frac <- ess / n

  # Run matching on this dataset
  m <- run_ps_matched(d, ps_clip)

  data.frame(
    frac_extreme = round(frac_extreme, 3),
    ess_frac     = round(ess_frac, 3),
    max_weight   = round(max(w), 1),
    matched_frac = round(ifelse(is.null(m$matched_frac), NA, m$matched_frac), 3),
    max_smd_match = round(ifelse(is.null(m$max_smd), NA, m$max_smd), 3),
    trt_prev     = round(mean(d$A), 3),
    aki_prev     = round(mean(d$Y), 3)
  )
}

diag_1 <- compute_scenario_diagnostics(b_rr_g = 0.7,  beta_rr_Q = 0.35, beta_int = 0.00,
                                        beta_ckdA = 0.40, beta_ageA = 0.15)
diag_2 <- compute_scenario_diagnostics(b_rr_g = 2.4,  beta_rr_Q = 1.30, beta_int = 0.60,
                                        beta_ckdA = 0.40, beta_ageA = 0.00)
diag_3 <- compute_scenario_diagnostics(b_rr_g = 2.4,  beta_rr_Q = 0.03, beta_int = 0.00,
                                        beta_ckdA = 0.25, beta_ageA = 0.10)

diag_tab <- rbind(
  cbind(Scenario = "1: Moderate overlap", diag_1),
  cbind(Scenario = "2: Overlap stress, strong Q", diag_2),
  cbind(Scenario = "3: Overlap stress, weak Q", diag_3)
)
# Add qualitative label
diag_tab$quality <- ifelse(diag_tab$frac_extreme < 0.15 & diag_tab$matched_frac > 0.70,
                            "Acceptable", "Poor")
kable(diag_tab, row.names = FALSE,
      caption = "Stage 2-style diagnostics by scenario (n = 5000 representative datasets)")
```

### What the gate tests

The Stage 3 gate asks: does the TMLE pipeline meet pre-specified performance
targets?

- **GO** means TMLE bias is below the threshold, CI coverage is at or above
  the minimum, and SE calibration is acceptable — the pipeline is adequate.
- **FLAG** means bias and coverage are acceptable but SE calibration is
  borderline — proceed with caution.
- **STOP** means TMLE does not meet the primary targets — the pipeline is
  inadequate as specified.

The interpretation below follows the actual results from this tutorial run.

```{r gate-function}
gate_check <- function(metrics, scenario_name, targets) {
  tmle <- metrics[metrics$method == "TMLE", ]

  bias_ok <- abs(tmle$bias) < targets$max_abs_bias
  cov_ok  <- tmle$coverage >= targets$min_coverage
  se_sd_ratio <- tmle$mean_se / tmle$emp_sd
  se_cal  <- se_sd_ratio > targets$se_sd_low & se_sd_ratio < targets$se_sd_high

  gate_tab <- data.frame(
    Metric = c("TMLE |bias|", "TMLE coverage", "SE/SD ratio"),
    Target = c(paste("<", targets$max_abs_bias),
               paste(">=", targets$min_coverage),
               paste(targets$se_sd_low, "-", targets$se_sd_high)),
    Observed = round(c(abs(tmle$bias), tmle$coverage, se_sd_ratio), 4),
    Status = c(ifelse(bias_ok, "GO", "STOP"),
               ifelse(cov_ok, "GO", "STOP"),
               ifelse(se_cal, "GO", "FLAG"))
  )

  if (bias_ok & cov_ok & se_cal) {
    decision <- "GO"
  } else if (bias_ok & cov_ok) {
    decision <- "FLAG"
  } else {
    decision <- "STOP"
  }

  list(decision = decision, table = gate_tab, scenario = scenario_name)
}
```

### Scenario 1: Acceptable overlap, modest TMLE precision gain

```{r scenario1-results, results='asis'}
cat("**True RD:**", round(truth_1["RD"], 5), "\n\n")
kable(sim_1$metrics[, c("method", "n_ok", "bias", "rmse", "emp_sd", "mean_se", "coverage")],
      digits = 4, row.names = FALSE,
      caption = "Scenario 1: Moderate overlap (b_rr = 0.7), A:ckd + A:age heterogeneity")

gate_1 <- gate_check(sim_1$metrics, "Scenario 1", analysis_lock$sim_targets)
cat("\n")
kable(gate_1$table, row.names = FALSE, caption = "Scenario 1 gate assessment")
cat("\n**Gate decision:**", gate_1$decision, "\n\n")

# Decision summary
tmle_1 <- sim_1$metrics[sim_1$metrics$method == "TMLE", ]
psm_1  <- sim_1$metrics[sim_1$metrics$method == "PS-Matched", ]
cat("**Decision summary:** Matching was acceptable (retained ",
    round(100 * diag_1$matched_frac), "% of treated), ",
    "but TMLE achieved ", sep = "")
iptw_1 <- sim_1$metrics[sim_1$metrics$method == "IPTW", ]
if (!is.na(psm_1$emp_sd) && tmle_1$emp_sd < psm_1$emp_sd) {
  cat("lower empirical SD (",
      round(tmle_1$emp_sd, 4), " vs. ", round(psm_1$emp_sd, 4),
      " for matching, ", round(iptw_1$emp_sd, 4), " for IPTW",
      ") and lower RMSE (", round(tmle_1$rmse, 4), " vs. ", round(psm_1$rmse, 4),
      " for matching). ",
      "This precision gain reflects TMLE's glmnet-based Q model exploiting learnable ",
      "structure (nonlinear age effects, treatment-by-CKD and treatment-by-age interactions) ",
      "that design-only estimators cannot access.\n", sep = "")
} else {
  cat("comparable or better performance metrics, so TMLE would be preferred.\n")
}
```

### Scenario 2: Overlap stress, all estimators fail → STOP

```{r scenario2-results, results='asis'}
cat("**True RD:**", round(truth_2["RD"], 5), "\n\n")
kable(sim_2$metrics[, c("method", "n_ok", "bias", "rmse", "emp_sd", "mean_se", "coverage")],
      digits = 4, row.names = FALSE,
      caption = "Scenario 2: Overlap stress (b_rr = 2.4), strong renal_risk + A:renal_risk interaction")

gate_2 <- gate_check(sim_2$metrics, "Scenario 2", analysis_lock$sim_targets)
cat("\n")
kable(gate_2$table, row.names = FALSE, caption = "Scenario 2 gate assessment")
cat("\n**Gate decision:**", gate_2$decision, "\n\n")

# Decision summary
tmle_2 <- sim_2$metrics[sim_2$metrics$method == "TMLE", ]
cat("**Decision summary:** Matching feasibility was reduced (retained only ",
    round(100 * diag_2$matched_frac), "% of treated). ", sep = "")
if (gate_2$decision == "STOP") {
  cat("TMLE also failed the pre-specified Stage 3 gate ",
      "(|bias| = ", round(abs(tmle_2$bias), 4),
      ", coverage = ", round(tmle_2$coverage, 3),
      "), so the clean-room workflow would stop the analysis. ",
      "When `renal_risk` strongly predicts both treatment and outcome, ",
      "neither matching nor TMLE can adequately compensate for the positivity stress.\n", sep = "")
} else {
  cat("TMLE gate returned ", gate_2$decision, " with |bias| = ",
      round(abs(tmle_2$bias), 4), " and coverage = ", round(tmle_2$coverage, 3),
      ". Consult the gate criteria to determine whether this meets the pre-specified ",
      "performance targets.\n", sep = "")
}
```

### Scenario 3: Overlap stress, analysis pipeline adequate → GO

```{r scenario3-results, results='asis'}
cat("**True RD:**", round(truth_3["RD"], 5), "\n\n")
kable(sim_3$metrics[, c("method", "n_ok", "bias", "rmse", "emp_sd", "mean_se", "coverage")],
      digits = 4, row.names = FALSE,
      caption = "Scenario 3: Overlap stress (b_rr = 2.4), weak renal_risk, A:ckd + A:age heterogeneity")

gate_3 <- gate_check(sim_3$metrics, "Scenario 3", analysis_lock$sim_targets)
cat("\n")
kable(gate_3$table, row.names = FALSE, caption = "Scenario 3 gate assessment")
cat("\n**Gate decision:**", gate_3$decision, "\n\n")

# Decision summary
tmle_3 <- sim_3$metrics[sim_3$metrics$method == "TMLE", ]
psm_3 <- sim_3$metrics[sim_3$metrics$method == "PS-Matched", ]
iptw_3 <- sim_3$metrics[sim_3$metrics$method == "IPTW", ]
cat("**Decision summary:** Matching feasibility was reduced (retained only ",
    round(100 * diag_3$matched_frac), "% of treated). ", sep = "")
if (gate_3$decision == "GO" || gate_3$decision == "FLAG") {
  cat("However, the analysis pipeline performed adequately — TMLE met the Stage 3 gate ",
      "(|bias| = ", round(abs(tmle_3$bias), 4),
      ", coverage = ", round(tmle_3$coverage, 3),
      "), so the clean-room workflow would allow the analysis to proceed. ",
      "Because `renal_risk` barely affects the outcome, the outcome model compensates ",
      "for the stressed treatment model. ", sep = "")
  if (!is.na(psm_3$rmse) && tmle_3$rmse < psm_3$rmse) {
    cat("TMLE achieved lower RMSE than PS-matching (",
        round(tmle_3$rmse, 4), " vs. ", round(psm_3$rmse, 4),
        "), illustrating the benefit of outcome-model learning under overlap stress.\n",
        sep = "")
  } else {
    cat("\n")
  }
} else {
  cat("TMLE also failed the gate (|bias| = ", round(abs(tmle_3$bias), 4),
      ", coverage = ", round(tmle_3$coverage, 3),
      "). Even though `renal_risk` barely affects the outcome, finite-sample performance ",
      "did not meet the pre-specified targets at n = ", n_sim,
      ". The team would need to consider a larger sample or richer learners.\n", sep = "")
}
```

### Sampling distributions across scenarios

```{r sim-sampling-dist, fig.width=10, fig.height=9}
plot_dat <- rbind(
  cbind(scenario = "1: Moderate overlap", sim_1$results),
  cbind(scenario = "2: Overlap stress,\nstrong Q effect", sim_2$results),
  cbind(scenario = "3: Overlap stress,\nweak Q effect", sim_3$results)
)
plot_dat <- plot_dat[!is.na(plot_dat$est), ]
plot_dat$method <- factor(plot_dat$method,
                          levels = c("Crude", "PS-Matched", "IPTW", "TMLE"))

truth_lines <- data.frame(
  scenario = c("1: Moderate overlap",
               "2: Overlap stress,\nstrong Q effect",
               "3: Overlap stress,\nweak Q effect"),
  truth = c(truth_1["RD"], truth_2["RD"], truth_3["RD"])
)

ggplot(plot_dat, aes(x = est, fill = method)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(data = truth_lines, aes(xintercept = truth),
             linewidth = 0.8, linetype = "solid") +
  facet_grid(method ~ scenario, scales = "free") +
  labs(title = "Stage 3: Sampling Distributions Under Three DGP Scenarios",
       subtitle = "Black line = true ATE",
       x = "Estimated Risk Difference", y = "Count") +
  guides(fill = "none") +
  theme_minimal(base_size = 11)
```

### Stage 3 synthesis

```{r stage3-synthesis, results='asis'}
cat("The three scenarios tell a coherent story about the relationship between ",
    "propensity-score diagnostics and TMLE validation:\n\n")

cat("- **Scenario 1** (gate: ", gate_1$decision,
    ") shows that when overlap is acceptable, both matching and TMLE work well. ",
    "TMLE offers a precision gain because its glmnet-based outcome model can capture ",
    "nonlinear age effects and treatment-effect heterogeneity (A:ckd, A:age) that a ",
    "simple paired-difference estimator cannot exploit.\n\n", sep = "")

cat("- **Scenario 2** (gate: ", gate_2$decision,
    ") shows that when overlap is stressed *and* the overlap-driving covariate ",
    "strongly predicts the outcome, the full estimator pipeline fails. ",
    "Neither matching nor TMLE can rescue the analysis.\n\n", sep = "")

cat("- **Scenario 3** (gate: ", gate_3$decision,
    ") shows that reduced matching feasibility alone does not automatically ",
    "invalidate the analysis pipeline. When the overlap-driving covariate barely affects the ",
    "outcome, TMLE's glmnet-based outcome model can compensate for the stressed treatment ",
    "model by learning the remaining outcome structure (A:ckd, A:age interactions). ",
    "Only Stage 3 can reveal whether this compensation works at the study's sample size.\n\n",
    sep = "")

cat("Scenarios 2 and 3 have nearly identical Stage 2 overlap diagnostics ",
    "(both have `b_rr` = 2.4 in the treatment model). The difference is ",
    "entirely in the outcome surface, which Stage 2 cannot evaluate. ",
    "**Only Stage 3 distinguishes the two cases.**\n")
```

### What this tutorial run demonstrates

```{r tutorial-lesson, results='asis'}
all_go  <- (gate_1$decision %in% c("GO", "FLAG")) &&
           (gate_3$decision %in% c("GO", "FLAG"))
s2_stop <- (gate_2$decision == "STOP")

if (all_go && s2_stop) {
  cat("::: {.callout-tip}\n",
      "### Lesson from this run\n\n",
      "This tutorial run produced the intended three-way contrast:\n\n",
      "1. **Scenario 1 (GO):** Overlap was acceptable, and TMLE offered a precision gain by leveraging glmnet-based outcome-model learning.\n",
      "2. **Scenario 2 (STOP):** Overlap was stressed and all estimators failed — a genuine ",
      "problem that Stage 3 correctly identifies.\n",
      "3. **Scenario 3 (GO):** Overlap was stressed and matching feasibility was reduced, but the ",
      "analysis pipeline — including TMLE — performed adequately. Reduced matching feasibility ",
      "alone would have unnecessarily stopped a valid analysis.\n\n",
      "The core lesson: **propensity-score diagnostics and full estimator validation are not the ",
      "same thing.** Stage 2 evaluates overlap and matching feasibility on the observed data. ",
      "Stage 3 validates the full analysis pipeline under plausible DGPs. Clean-room outcome-blind ",
      "simulation is what resolves whether the analysis should GO or STOP.\n",
      ":::\n", sep = "")
} else {
  cat("::: {.callout-tip}\n",
      "### Lesson from this run\n\n",
      "The gate decisions (Scenario 1: ", gate_1$decision,
      ", Scenario 2: ", gate_2$decision,
      ", Scenario 3: ", gate_3$decision,
      ") show that Stage 3 provides information beyond Stage 2. ",
      "The three scenarios share different treatment mechanisms and outcome surfaces, ",
      "and Stage 3 can distinguish settings where TMLE works from those where it ",
      "does not.\n",
      ":::\n", sep = "")
}
```

::: {.callout-important}
### The core methodological point

Stage 2 evaluates $g$ alone — propensity scores, overlap, matching feasibility.
Stage 3 evaluates the entire TMLE pipeline ($g$, $Q$, targeting, inference).
Without Stage 3, the team can only assess whether the propensity score behaves
well — not whether the full estimator produces valid estimates. Outcome-blind
simulation is what turns a generic TMLE workflow into a credible clean-room
analysis.
:::

### When might TMLE be preferred?

TMLE is not uniformly superior to design-only estimators. When the outcome
surface is simple and overlap is adequate, matching and IPTW may perform
comparably. However, when the outcome model contains learnable structure —
nonlinear covariate effects, treatment-effect heterogeneity, or interactions
— TMLE can improve precision by combining information from both the treatment
and outcome models. In this tutorial, the `glmnet`-based $Q$ model with a
rich feature matrix can capture age nonlinearities and treatment-by-subgroup
interactions that a simple paired difference after matching cannot exploit.
This makes TMLE especially attractive in clean-room workflows: the richer
outcome-model specification (including the learner library and feature set)
is pre-committed at Stage 1, validated through simulation at Stage 3, and
applied to real data only at Stage 4, so any precision gain is achieved
without data-adaptive model selection on outcomes.


# A Brief Intuition for the Efficient Influence Function {#eif-intuition}

Before presenting the Stage 4 results and inference formulas, it helps to
have a brief intuition for the **efficient influence function (EIF)**, which
underlies TMLE's standard errors.

The EIF summarizes **how each observation contributes to the estimator**. For
the ATE, the EIF for observation $i$ is:

$$D_i = \frac{A_i}{\hat{g}(W_i)}\bigl(Y_i - \bar{Q}^*(1, W_i)\bigr)
      - \frac{1 - A_i}{1 - \hat{g}(W_i)}\bigl(Y_i - \bar{Q}^*(0, W_i)\bigr)
      + \bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i) - \hat\psi$$

Three key facts:

1. **TMLE's targeting step ensures that the sample mean of the EIF is
   approximately zero.** This is the "solved estimating equation" property —
   it means the estimator is locally efficient.

2. **The variance of the EIF gives the standard error.** Specifically,
   $\widehat{\text{SE}} = \sqrt{n^{-1}\widehat{\text{Var}}(D)}$.
   This is sometimes called the "IC-based" or "sandwich" variance estimator.

3. **The EIF combines information from both models.** The first two terms
   use $g$ through the weights $A/\hat{g}$ and $(1-A)/(1-\hat{g})$, and $Q$
   through the residuals $Y - \bar{Q}^*$. The third term uses $Q$ alone. If
   $Q$ is perfect, the residuals are zero and $g$ drops out. If $g$ is perfect,
   the weights are accurate and $Q$ errors are corrected. This is double
   robustness expressed through the influence function.

This is all you need to know to interpret the Stage 4 inference: the SE comes
from the variance of the EIF across observations, and the CI is
$\hat\psi \pm 1.96 \times \widehat{\text{SE}}$.


# Stage 4. Unblinded TMLE Estimation {#stage-4}

## What happens at this stage

The analyst now receives the full dataset including outcomes $Y$. This is the
**first point** at which the outcome regression $Q(A,W)$ is estimated on real
data. The treatment model strategy was locked in Stage 2; the simulation
validation was completed in Stage 3; only now does the targeting step run on
real data.

::: {.callout-note collapse="true"}
### Operational note: Stage 4 in practice
In a real clean-room analysis, outcome unblinding would occur only after
Stage 3 results have been reviewed and the GO/STOP decision documented.
The Stage 4 analysis script would be pre-specified (or at least minimally
modifiable) — typically the same code validated in Stage 3, applied to the
real dataset. Any deviations from the locked analysis plan triggered by
unexpected findings (e.g., convergence issues, data quality problems) would
need to be documented as formal amendments and reported alongside the primary
results.
:::

### Prerequisites

Stage 4 proceeds only after:

1. Stage 1 analysis lock is documented and versioned
2. Stage 2 diagnostics are reviewed and recorded
3. Stage 3 simulation gate is evaluated (if GO, proceed; if STOP, document
   the concern and any plan amendment)

### Run TMLE on the real data

```{r stage4-tmle}
run_stage4_tmle <- function(dat, lock) {
  cat("Stage 4: Outcome column now accessible.\n")
  cat("AKI prevalence:", round(mean(dat$Y), 3), "\n")
  cat("Treatment prevalence:", round(mean(dat$A), 3), "\n\n")

  result <- run_tmle_once(dat, ps_clip = lock$ps_truncation)

  cat("=== Stage 4 TMLE Result ===\n")
  cat("ATE estimate (RD):", round(result$estimate, 5), "\n")
  cat("Standard error:", round(result$se, 5), "\n")
  cat("95% CI: [", round(result$ci_low, 5), ",",
      round(result$ci_high, 5), "]\n")

  invisible(result)
}

stage4_result <- run_stage4_tmle(dat_full, analysis_lock)
```

### Comparator estimates

For transparency, we also run the pre-specified comparator estimators on the
real data:

```{r stage4-comparators}
stage4_crude   <- run_crude(dat_full)
stage4_iptw    <- run_iptw(dat_full, ps_clip = analysis_lock$ps_truncation)
stage4_matched <- run_ps_matched(dat_full, ps_clip = analysis_lock$ps_truncation)

est_tab <- data.frame(
  Estimator = c("Crude RD", "PS-Matched", "IPTW (Hajek)", "TMLE"),
  Estimate = round(c(stage4_crude$estimate, stage4_matched$estimate,
                     stage4_iptw$estimate, stage4_result$estimate), 5),
  SE = round(c(stage4_crude$se, stage4_matched$se,
               stage4_iptw$se, stage4_result$se), 5),
  CI_low = round(c(stage4_crude$ci_low, stage4_matched$ci_low,
                   stage4_iptw$ci_low, stage4_result$ci_low), 5),
  CI_high = round(c(stage4_crude$ci_high, stage4_matched$ci_high,
                    stage4_iptw$ci_high, stage4_result$ci_high), 5)
)
kable(est_tab, caption = "Stage 4: Effect estimates on the real (simulated) dataset")
```

### What just happened — step by step

1. **$Q$ was estimated on the real data** for the first time: a penalized
   logistic regression (`cv.glmnet`) with a rich feature set — polynomial
   terms, treatment-by-covariate interactions, and covariate interactions —
   was fit to $(W, A, Y)$ using 5-fold cross-validation to select the
   regularization parameter.

2. **$g$ was re-estimated** using the same locked strategy from Stage 2. (In
   a production analysis, one might load the Stage 2 $\hat{g}$ object instead.)

3. **The targeting step** constructed the clever covariate
   $H = A/\hat{g} - (1-A)/(1-\hat{g})$ and fit a logistic fluctuation
   $\text{logit}(\bar{Q}^*) = \text{logit}(\bar{Q}) + \epsilon H$.

4. **The ATE** was computed as
   $\hat\psi = n^{-1}\sum[\bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i)]$.

5. **Inference** used the EIF as described above.

### Interpreting the result in context

```{r stage4-context, results='asis'}
cat("This Stage 4 estimate should be interpreted alongside the evidence from ",
    "earlier stages.\n\n")

if (s2_diag$any_flag) {
  cat("Stage 2 flagged overlap concerns in the propensity score distribution. ")
} else {
  cat("Stage 2 diagnostics did not flag major overlap concerns. ")
}

# The real dataset uses Scenario 3 parameters (overlap stress, weak renal_risk in Q)
cat("The real study dataset was generated with the same DGP as Scenario 3 ",
    "(overlap stress, weak `renal_risk` effect on outcome). ",
    "In practice, the analyst would not know which candidate Stage 3 scenario ",
    "is closest to reality; interpretation should weigh results across all ",
    "plausible DGPs.\n\n")

if (gate_3$decision %in% c("GO", "FLAG")) {
  cat("Stage 3 simulation returned **", gate_3$decision,
      "** under Scenario 3, providing pre-registered evidence that the ",
      "analysis pipeline produces adequate finite-sample performance under these ",
      "conditions — even though matching feasibility was reduced. ",
      "The Stage 4 estimate therefore carries credibility beyond a post-hoc ",
      "TMLE analysis, because the analytic decisions were locked before ",
      "outcomes were accessed.\n\n", sep = "")
} else {
  cat("Stage 3 simulation returned **STOP** under Scenario 3, meaning the analysis ",
      "pipeline did not meet pre-specified performance targets. In a real study, ",
      "this Stage 4 estimate would be reported with caveats, and the STOP finding ",
      "documented alongside the results. The team might consider a protocol ",
      "amendment (e.g., richer learners, larger sample) before relying on these ",
      "results.\n\n")
}

cat("No analytic decisions — estimand, SL library, truncation bounds, covariate ",
    "set — were changed based on the observed outcomes.\n")
```


<details>
<summary><strong>Full Simulation Results (production-scale)</strong></summary>

# Full Simulation Results {#full-sim}

```{r full-sim-load}
out_dir <- file.path("..", "results", "scenario1_v4")
has_full_sim <- dir.exists(out_dir)
if (has_full_sim) {
  load_mode <- function(mode) {
    d <- file.path(out_dir, mode)
    tryCatch(list(
      truth   = readRDS(file.path(d, "truth.rds")),
      results = readRDS(file.path(d, "replicate_results.rds")),
      metrics = read.csv(file.path(d, "summary_overall.csv")),
      gate    = readRDS(file.path(d, "gate_results.rds")),
      params  = readRDS(file.path(d, "params_used.rds"))
    ), error = function(e) NULL)
  }
  weak   <- load_mode("weak")
  strong <- load_mode("strong")
  gate_combined <- tryCatch(
    readRDS(file.path(out_dir, "gate_combined.rds")),
    error = function(e) NULL)
} else {
  weak <- strong <- gate_combined <- NULL
}
```

The three-scenario simulation above demonstrates the staged workflow using
enriched-GLM specifications for both $g$ and $Q$. The production-scale
simulation uses richer estimators (TMLE-ML with SuperLearner, CV-TMLE,
PS-matched regression), additional sample sizes, and a mixed-overlap regime.
If those results are available, they are shown below as an extended reference.
Note that the production simulation uses a different DGP parameterization
than the tutorial scenarios, so results are not directly comparable.

```{r full-sim-results, results='asis'}
if (!is.null(weak) && !is.null(strong)) {
  cat("### Full simulation results available\n\n")
  cat("These results come from the v4 simulation with N =", weak$params$n_main,
      "and", weak$params$reps, "replicates per mode.\n\n")

  point_methods <- c("Crude", "Adj. Regression (analytic)",
                     "IPTW (analytic)", "PS-Matched (pairs boot)",
                     "TMLE-GLM (IC)", "TMLE-ML (IC)", "CV-TMLE-ML (IC)")
  show_perf <- function(metrics, label) {
    sub <- metrics[metrics$subset == "all" &
                   metrics$method %in% point_methods, ]
    sub$mode <- label
    sub[, c("mode", "method", "bias", "rmse", "coverage", "mcse_coverage")]
  }
  combined <- rbind(show_perf(weak$metrics, "Weak W4"),
                    show_perf(strong$metrics, "Strong W4"))
  kable(combined, digits = 4, row.names = FALSE,
        caption = "Full simulation: estimator performance (all replicates)")

  if (!is.null(gate_combined)) {
    cat("\n### Stop/go gate results\n\n")
    kable(gate_combined, digits = 4, row.names = FALSE,
          caption = "Gate results from full simulation")
  }
} else {
  cat("Full simulation outputs not available. ",
      "Run: `Rscript scripts/sim_scenario1_v4_false_stop_inference.R`\n")
}
```

</details>


# Common Mistakes When Operationalizing TMLE in a Clean Room {#common-mistakes}

::: {.callout-warning}
### Pitfalls to avoid

1. **Fitting the real outcome regression $Q(A,W)$ before outcome access is
   permitted.** This violates the clean-room principle. All $Q$-fitting on
   real data belongs to Stage 4.

2. **Treating Stage 2 propensity score diagnostics as if they validate the
   full TMLE pipeline.** Stage 2 evaluates $g$ only. It says nothing about
   whether $Q$ estimation, the targeting step, or CI coverage will be adequate.

3. **Changing the Super Learner library, truncation rule, or estimand after
   seeing real outcomes without documenting a formal amendment.** Undocumented
   post-hoc changes undermine the entire clean-room rationale. All revisions
   must be recorded and justified.

4. **Interpreting acceptable overlap diagnostics as proof that TMLE inference
   will be valid.** Good overlap is neither necessary nor sufficient for valid
   TMLE. The outcome surface, sample size, and learner flexibility all matter.

5. **Assuming poor propensity score overlap automatically invalidates TMLE
   without checking whether the outcome model can compensate.** This ignores
   double robustness. Stage 3 simulation is the appropriate tool for resolving
   this question.

6. **Confusing a locked modeling strategy with a requirement to reuse the
   exact same fitted nuisance object.** What is locked is the specification
   (formula, learner library, truncation bounds). Re-running the same procedure
   on the same data is acceptable; changing the specification is not.
:::


# Key Takeaways {#takeaways}

```{r takeaways-text, results='asis'}
cat("1. **Propensity-score diagnostics and full estimator validation are not the same thing.**\n",
    "   Stage 2 evaluates overlap and matching feasibility on the observed data.\n",
    "   Stage 3 validates the full analysis pipeline ($g$, $Q$, targeting, inference)\n",
    "   under plausible DGPs. Acceptable overlap does not guarantee valid inference;\n",
    "   overlap stress does not necessarily invalidate the pipeline.\n\n")

cat("2. **Stage 3 validates the full TMLE pipeline on synthetic data.** By testing\n",
    "   bias, variance, and empirical coverage under pre-specified DGPs, the team\n",
    "   makes an evidence-based GO/STOP decision before seeing real outcomes.\n\n")

cat("3. **TMLE may offer a precision gain when the outcome surface is learnable.**\n",
    "   Scenario 1 (gate: ", gate_1$decision,
    ") showed that both matching and TMLE worked well,\n",
    "   but TMLE achieved lower variance by using glmnet to exploit nonlinear age effects\n",
    "   and treatment-effect heterogeneity (A:ckd, A:age) that design-only estimators\n",
    "   cannot capture.\n\n", sep = "")

cat("4. **Overlap stress plus strong confounding through the overlap driver can\n",
    "   defeat all estimators.** Scenario 2 (gate: ", gate_2$decision,
    ") showed that when `renal_risk`\n",
    "   strongly predicts both treatment and outcome, neither matching nor TMLE\n",
    "   nor IPTW can rescue the analysis.\n\n", sep = "")

cat("5. **Reduced matching feasibility does not automatically invalidate the analysis pipeline.**\n",
    "   Scenario 3 (gate: ", gate_3$decision,
    ") showed that when the overlap-driving covariate\n",
    "   barely affects the outcome, TMLE can remain competitive or preferable by\n",
    "   learning the remaining outcome structure via a richer Q model. Stage 3\n",
    "   simulation is necessary to determine whether this compensation works at\n",
    "   the study's sample size and learner specification.\n\n", sep = "")

cat("6. **Stage 4 is where the real outcome regression and targeting step occur.**\n",
    "   The $g$-strategy is carried forward from Stage 2, but $Q$ is fit on real\n",
    "   data only at this final stage. No analytic decisions change after outcome\n",
    "   access.\n")
```


# Practical Checklist for a Clean-Room TMLE Analysis {#checklist}

Use this checklist to ensure each stage is properly documented and archived.

### Stage 1: Pre-specification and Analysis Lock

- [ ] Causal question and estimand clearly stated
- [ ] Eligibility criteria, treatment definition, and outcome definition
- [ ] Covariate list with variable roles (confounders, effect modifiers)
- [ ] Super Learner libraries specified for both $g$ and $Q$
- [ ] PS truncation bounds
- [ ] Stage 2 diagnostic thresholds (extreme PS fraction, ESS, max weight, SMD)
- [ ] Stage 3 performance thresholds (max bias, min coverage)
- [ ] Versioned analysis lock file with timestamp and hash

### Stage 2: Outcome-Blind Design Diagnostics

- [ ] Fitted $g$-strategy (or locked specification to be re-run)
- [ ] Propensity score distribution summaries and overlap plots
- [ ] ESS and IPTW weight summaries
- [ ] Covariate balance diagnostics (love plot, weighted SMDs)
- [ ] List of triggered diagnostic flags
- [ ] Documented Stage 2 decision (PASS / FLAG — proceed to Stage 3)

### Stage 3: Outcome-Blind Simulation Validation

- [ ] Pre-specified DGP description (before real outcome access)
- [ ] Simulation seeds and settings (reps, sample size, scenarios)
- [ ] Code used for simulation runs (archived and versioned)
- [ ] Bias, RMSE, empirical SD, mean SE, and coverage for each estimator
- [ ] Comparison estimator results (crude, IPTW, etc.)
- [ ] Gate decision table: metric, threshold, observed value, GO/STOP
- [ ] Documented GO / FLAG / STOP decision with rationale
- [ ] Any plan amendment and its justification

### Stage 4: Unblinded TMLE Estimation

- [ ] Final $Q$ fit (first time on real outcomes)
- [ ] Final TMLE point estimate and 95% confidence interval
- [ ] Comparator estimates (crude, IPTW)
- [ ] Cross-reference to Stage 2 diagnostics and Stage 3 gate result
- [ ] Confirmation that no specifications were changed after outcome access
- [ ] Any deviations from the locked plan and their justification
- [ ] Final reproducibility metadata (session info, package versions, seed)


## Reproducing the Full Simulation

```bash
# Compact tutorial simulation: runs within this QMD (200 reps, enriched GLM)

# Production simulation with SuperLearner, 7 estimators, mixed overlap:
Rscript scripts/sim_scenario1_v4_false_stop_inference.R fast   # N=400, 120 reps
Rscript scripts/sim_scenario1_v4_false_stop_inference.R eval   # N=600, 300 reps
```


## References

- van der Laan MJ, Rose S. *Targeted Learning: Causal Inference for
  Observational and Experimental Data*. Springer, 2011.
- van der Laan MJ, Rose S. *Targeted Learning in Data Science*. Springer, 2018.
- Schuler MS, Rose S. Targeted maximum likelihood estimation for causal
  inference in observational studies. *American Journal of Epidemiology*, 2017.
- Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ. Diagnosing and
  responding to violations in the positivity assumption.
  *Statistical Methods in Medical Research*, 2012.
- Gruber S, van der Laan MJ. tmle: An R package for targeted maximum
  likelihood estimation. *Journal of Statistical Software*, 2012.
