---
title: "Operationalizing TMLE in a Clean-Room Staged Design"
subtitle: "A Tutorial on Outcome-Blind Diagnostics, Simulation Validation, and Staged Estimation"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(ggplot2)

# Utility functions used throughout the tutorial
expit <- function(x) 1 / (1 + exp(-x))
logit <- function(p) log(p / (1 - p))

# Load full simulation outputs if available (for the Stage 3 section)
out_dir <- file.path("..", "results", "scenario1_v4")
has_full_sim <- dir.exists(out_dir)
if (has_full_sim) {
  load_mode <- function(mode) {
    d <- file.path(out_dir, mode)
    tryCatch(list(
      truth   = readRDS(file.path(d, "truth.rds")),
      results = readRDS(file.path(d, "replicate_results.rds")),
      metrics = read.csv(file.path(d, "summary_overall.csv")),
      gate    = readRDS(file.path(d, "gate_results.rds")),
      params  = readRDS(file.path(d, "params_used.rds"))
    ), error = function(e) NULL)
  }
  weak   <- load_mode("weak")
  strong <- load_mode("strong")
  gate_combined <- tryCatch(
    readRDS(file.path(out_dir, "gate_combined.rds")),
    error = function(e) NULL)
} else {
  weak <- strong <- gate_combined <- NULL
}
```

# What This Tutorial Is and Is Not

This is **not** a generic TMLE tutorial. It does not walk through TMLE as a
standalone statistical method. For that, see van der Laan and Rose (2011) or
the `tmle` R package vignette.

This tutorial teaches how TMLE is **operationalized within a clean-room staged
workflow** — a governance structure that controls *when* different data elements
may be accessed and *what* analytic tasks are permitted at each stage. The
clean-room design is borrowed from pharmaceutical manufacturing and adapted
for observational causal inference, where the investigator's access to outcome
data during model-building creates risk of bias.

The central tension: TMLE requires both a treatment model $g(W) = P(A=1|W)$
and an outcome model $Q(A,W) = E[Y|A,W]$. In a clean-room workflow, these
two components live in different stages:

- **$g$ can be estimated early** (Stage 2), because it uses only treatment and
  covariates — no outcomes required.
- **$Q$ for the real data can only be estimated after outcome access**
  (Stage 4), because fitting an outcome regression requires $Y$.
- **The targeting step** combines $Q$ and $g$ and therefore also belongs to
  Stage 4 for the real analysis.
- **In simulation (Stage 3)**, the full pipeline ($g$, $Q$, targeting) runs on
  synthetic data with known truth, validating the estimator before real
  outcomes are unblinded.

TMLE is not itself a clean-room design. The clean room governs access and
permissions; TMLE is one estimator that can be embedded in that governance
structure. The tutorial shows exactly how.

### Fitting $g$ early, carrying it forward

Because $g(W) = P(A=1|W)$ does not involve outcomes, the treatment model
strategy can be estimated and locked during the outcome-blind stage. The
locked $g$-strategy is then carried forward into the simulation validation
(Stage 3) and the real estimation (Stage 4). Revising $g$ after seeing real
outcomes would violate the clean-room logic unless the governance plan
explicitly allowed and documented such a revision.

Stage 3 simulation is the proper place to decide whether the pre-specified
TMLE pipeline — including the $g$-strategy — is adequate for the study at
hand.


## Quick Reference: Allowed vs Not Allowed Before Outcome Access

```{r permissions-table}
perms <- data.frame(
  Stage = c(
    "Stage 1:\nPre-specification",
    "Stage 2:\nOutcome-Blind Diagnostics",
    "Stage 3:\nSimulation Validation",
    "Stage 4:\nUnblinded Estimation"
  ),
  `Data accessible` = c(
    "None (metadata, variable definitions only)",
    "W and A from the real dataset (not Y)",
    "Synthetic data with simulated Y",
    "Full dataset: W, A, Y"
  ),
  `Allowed TMLE-related tasks` = c(
    paste("Define causal question and estimand;",
          "specify covariates and treatment;",
          "select SL libraries for g and Q;",
          "set truncation rules;",
          "define diagnostic thresholds and gate criteria"),
    paste("Estimate g(W) on real data;",
          "PS overlap diagnostics;",
          "IPTW weight distributions and ESS;",
          "covariate balance (SMD);",
          "test truncation thresholds"),
    paste("Run the full TMLE pipeline on simulated data;",
          "evaluate bias, variance, and coverage;",
          "compare TMLE with simpler estimators;",
          "test SL specifications and truncation"),
    paste("Estimate Q(A,W) on real data;",
          "combine with locked g-strategy;",
          "run TMLE targeting step;",
          "compute estimates and CIs;",
          "report results and sensitivity analyses")
  ),
  `Not allowed` = c(
    paste("Estimating g or Q on real data;",
          "any estimator involving real outcomes"),
    paste("Fitting Q(A,W) on real outcomes;",
          "computing treatment effects on real data;",
          "running TMLE on the real dataset"),
    paste("Using real outcomes;",
          "tuning estimator based on real Y"),
    paste("Changing estimand without documented justification;",
          "switching estimator pipeline without a protocol amendment")
  ),
  check.names = FALSE
)
kable(perms, caption = "Clean-room permissions by stage")
```

Many steps of a TMLE analysis — particularly estimation and diagnostics of
the treatment mechanism — can occur during outcome-blind stages. However, the
outcome regression and targeting step that produce the final causal estimate
occur only after outcome access. Outcome-blind simulations provide a mechanism
for validating the full TMLE pipeline before this final step.


# Stage 1. Pre-specification and Analysis Lock {#stage-1}

## What happens at this stage

Before any real study data are accessed, the analysis team writes a
statistical analysis plan (SAP) that locks all major analytic decisions. This
is analogous to a clinical trial's SAP being finalized before database lock.

The SAP for a TMLE clean-room analysis must specify:

1. **Causal question and estimand.** We target the average treatment effect
   (ATE) on the risk difference scale:
   $$\psi = E[Y^1] - E[Y^0]$$

2. **Identification assumptions.** Consistency ($Y = Y^A$), no unmeasured
   confounding ($Y^a \perp\!\!\!\perp A \mid W$), and positivity
   ($0 < P(A=1|W) < 1$).

3. **Variable roles.** Which variables are $W$ (confounders), $A$ (treatment),
   and $Y$ (outcome).

4. **Modeling strategies.** The SuperLearner library for $g$ and $Q$, PS
   truncation bounds, and the TMLE variant (standard vs cross-validated).

5. **Stage 2 diagnostic thresholds.** What constitutes acceptable overlap
   and balance.

6. **Stage 3 simulation targets.** Performance criteria (bias, coverage) that
   the estimator must meet before proceeding.

7. **Conditions for revision.** Under what circumstances the plan must be
   amended.

### The analysis lock object

The pre-specified decisions are encoded as an operational artifact — a named
list that is versioned and timestamped. This object travels through every
subsequent stage unchanged (unless a formal amendment is documented).

```{r analysis-lock}
analysis_lock <- list(
  # --- Causal question ---
  estimand     = "ATE",
  effect_scale = "risk_difference",

  # --- Variable roles ---
  covariates   = c("W1", "W2", "W3", "W4"),
  treatment    = "A",
  outcome      = "Y",

  # --- Treatment model (g) ---
  g_formula    = "A ~ W1 + W2 + W3 + W4",
  g_SL_library = c("SL.mean", "SL.glmnet", "SL.xgboost"),

  # --- Outcome model (Q) ---
  Q_formula    = "Y ~ A + W1 + W2 + W3 + W4",
  Q_SL_library = c("SL.mean", "SL.glmnet", "SL.xgboost"),

  # --- PS truncation ---
  ps_truncation = c(lower = 0.025, upper = 0.975),

  # --- TMLE variant ---
  tmle_variant = "TMLE-ML",
  cv_folds     = 3,

  # --- Stage 2 diagnostic thresholds ---
  diag_thresholds = list(
    extreme_ps_bound  = 0.05,   # PS values below this or above 1-this
    max_extreme_frac  = 0.40,   # max fraction of extreme PS
    min_ess_frac      = 0.20,   # minimum ESS / n
    max_iptw_weight   = 60,     # max single IPTW weight
    max_smd_weighted  = 0.10    # max weighted SMD
  ),

  # --- Stage 3 simulation targets ---
  sim_targets = list(
    max_abs_bias = 0.01,
    min_coverage = 0.93
  ),

  # --- Metadata ---
  version    = "1.0",
  locked_date = Sys.Date(),
  locked_by   = "Analysis team"
)

cat("Analysis lock object (version", analysis_lock$version, ")\n")
cat("Locked:", as.character(analysis_lock$locked_date), "\n")
cat("Estimand:", analysis_lock$estimand, "on", analysis_lock$effect_scale, "scale\n")
cat("g library:", paste(analysis_lock$g_SL_library, collapse = ", "), "\n")
cat("Q library:", paste(analysis_lock$Q_SL_library, collapse = ", "), "\n")
cat("PS truncation:", analysis_lock$ps_truncation["lower"], "-",
    analysis_lock$ps_truncation["upper"], "\n")
```

::: {.callout-note}
In a real study, this object would be saved to a version-controlled file
(e.g., `analysis_lock.rds` or `analysis_lock.yml`) and timestamped before
any data access. The hash of this file becomes part of the audit trail.
:::

### Stage 1 summary

At this point, no data have been touched. The team has committed to:

- An estimand ($\psi$, ATE on RD scale)
- Identification assumptions
- Specific SL libraries for both $g$ and $Q$
- Truncation rules and diagnostic thresholds
- Simulation performance targets for the go/no-go gate


# Stage 2. Outcome-Blind Design Diagnostics {#stage-2}

## What happens at this stage

The analysis team now receives the **treatment and covariate data** $(W, A)$
from the real study. Outcomes $(Y)$ remain sealed. The purpose of this stage
is to evaluate whether the study's treatment mechanism and covariate structure
are compatible with the planned TMLE analysis.

Because $g(W) = P(A=1|W)$ does not involve outcomes, the entire treatment
model can be estimated and evaluated here. This is the key operational insight:
**the $g$-side of TMLE is fully accessible during the outcome-blind stage.**

### Generate tutorial data

For this tutorial, we simulate a dataset that mimics a pharmacoepi study with
four confounders and a binary treatment. We generate the full dataset (including
$Y$) but the Stage 2 analyst only has access to $(W, A)$.

```{r generate-data}
set.seed(42)
n <- 400

# Covariates
W1 <- rnorm(n)
W2 <- rbinom(n, 1, 0.5)
W3 <- rnorm(n)
W4 <- rnorm(n)  # overlap driver

# Treatment mechanism (g) — W4 has large coefficient
s_g <- 1.5  # overlap severity
lp_g_raw <- s_g * (1.5 * W1 + 0.8 * W2 - 0.5 * W3 + 3.0 * W4)
alpha0_g <- -mean(lp_g_raw) - 0.5  # shift P(A=1) ~ 0.35
g_true <- expit(alpha0_g + lp_g_raw)
g_true <- pmin(pmax(g_true, 0.005), 0.995)
A <- rbinom(n, 1, g_true)

# Outcome (generated but NOT accessible until Stage 4)
lp_Q <- -0.5 + 0.40 * A +
  0.50 * W1 + 0.25 * W1^2 + 0.70 * W2 + 0.35 * abs(W3) + 0.04 * W4 +
  0.15 * A * (W1 > 0) + 0.05 * A * W2
Y <- rbinom(n, 1, expit(lp_Q))

# The full dataset
dat <- data.frame(W1 = W1, W2 = W2, W3 = W3, W4 = W4, A = A, Y = Y)

# Stage 2 analyst sees only this:
dat_stage2 <- dat[, c("W1", "W2", "W3", "W4", "A")]
cat("Stage 2 data dimensions:", nrow(dat_stage2), "x", ncol(dat_stage2), "\n")
cat("Treatment prevalence:", round(mean(A), 3), "\n")
cat("Y column accessible:", "Y" %in% names(dat_stage2), "\n")
```

### Fit the treatment model

Using the pre-specified formula from the analysis lock:

```{r fit-g}
g_fit <- glm(as.formula(analysis_lock$g_formula),
             data = dat_stage2, family = "binomial")
g_hat <- fitted(g_fit)

# Apply pre-specified truncation
g_trunc <- pmin(pmax(g_hat, analysis_lock$ps_truncation["lower"]),
                analysis_lock$ps_truncation["upper"])
cat("PS range (raw):", round(range(g_hat), 4), "\n")
cat("PS range (truncated):", round(range(g_trunc), 4), "\n")
```

### Propensity score diagnostics

These diagnostics evaluate the treatment mechanism and practical positivity.
They are permitted at this stage because they use only $(W, A)$.

```{r ps-diagnostics, fig.width=8, fig.height=4}
# PS distribution by treatment group
ps_df <- data.frame(ps = g_hat, Treatment = factor(A, labels = c("Control", "Treated")))

ggplot(ps_df, aes(x = ps, fill = Treatment)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = c(analysis_lock$ps_truncation["lower"],
                             analysis_lock$ps_truncation["upper"]),
             linetype = "dashed", color = "red") +
  labs(title = "Stage 2: Propensity Score Distribution by Treatment",
       subtitle = "Red dashed lines = pre-specified truncation bounds",
       x = "Estimated P(A=1|W)", y = "Count") +
  theme_minimal(base_size = 12)
```

```{r overlap-metrics}
# Compute overlap metrics using the analysis lock thresholds
thr <- analysis_lock$diag_thresholds

frac_extreme <- mean(g_hat < thr$extreme_ps_bound |
                     g_hat > (1 - thr$extreme_ps_bound))

w_iptw <- A / g_trunc + (1 - A) / (1 - g_trunc)
ess <- sum(w_iptw)^2 / sum(w_iptw^2)
ess_frac <- ess / n
max_w <- max(w_iptw)

# Weighted SMDs
compute_smd <- function(x, trt, weights = NULL) {
  if (is.null(weights)) weights <- rep(1, length(x))
  m1 <- weighted.mean(x[trt == 1], weights[trt == 1])
  m0 <- weighted.mean(x[trt == 0], weights[trt == 0])
  s <- sqrt((var(x[trt == 1]) + var(x[trt == 0])) / 2)
  if (s < 1e-10) return(0)
  (m1 - m0) / s
}

covars <- c("W1", "W2", "W3", "W4")
smd_unw <- sapply(covars, function(v) compute_smd(dat_stage2[[v]], A))
smd_wtd <- sapply(covars, function(v) compute_smd(dat_stage2[[v]], A, w_iptw))

# Results
diag_results <- data.frame(
  Metric = c("Fraction extreme PS",
             "ESS fraction",
             "Max IPTW weight",
             "Max |SMD| (weighted)"),
  Value = round(c(frac_extreme, ess_frac, max_w, max(abs(smd_wtd))), 3),
  Threshold = c(
    paste("<", thr$max_extreme_frac),
    paste(">", thr$min_ess_frac),
    paste("<", thr$max_iptw_weight),
    paste("<", thr$max_smd_weighted)
  ),
  Status = c(
    ifelse(frac_extreme < thr$max_extreme_frac, "PASS", "FLAG"),
    ifelse(ess_frac > thr$min_ess_frac, "PASS", "FLAG"),
    ifelse(max_w < thr$max_iptw_weight, "PASS", "FLAG"),
    ifelse(max(abs(smd_wtd)) < thr$max_smd_weighted, "PASS", "FLAG")
  )
)
kable(diag_results, caption = "Stage 2 overlap diagnostics")
```

```{r love-plot, fig.width=7, fig.height=4}
smd_df <- data.frame(
  variable = rep(covars, 2),
  smd = c(smd_unw, smd_wtd),
  type = rep(c("Unweighted", "IPTW-Weighted"), each = length(covars))
)

ggplot(smd_df, aes(x = abs(smd), y = reorder(variable, abs(smd)), color = type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.10, linetype = "dashed") +
  labs(title = "Stage 2: Covariate Balance (SMD Love Plot)",
       x = "|Standardized Mean Difference|", y = NULL, color = "Weighting") +
  theme_minimal(base_size = 12)
```

### Stage 2 decision

```{r stage2-decision}
any_flag <- any(diag_results$Status == "FLAG")
stage2_status <- ifelse(any_flag, "FLAG — proceed to Stage 3 simulation", "PASS")
cat("Stage 2 status:", stage2_status, "\n")
if (any_flag) {
  flagged <- diag_results$Metric[diag_results$Status == "FLAG"]
  cat("Flagged metrics:", paste(flagged, collapse = "; "), "\n")
}
```

### Why Stage 2 is not enough for TMLE validation

Stage 2 diagnostics evaluate the treatment mechanism $g$ and practical
positivity. They answer: "Is the propensity score distribution well-behaved?"

But TMLE depends on *three* components: $g$, $Q$, and the targeting step.
Stage 2 tells us nothing about:

- Whether the SuperLearner can estimate $Q$ well enough under these conditions
- Whether the targeting step (which uses $g$ through the clever covariate
  $H = A/g - (1-A)/(1-g)$) produces stable updates
- Whether the IC-based variance estimator is reliable at this sample size
- Whether the CI coverage is near nominal

A study can have poor overlap diagnostics but excellent TMLE performance (a
**false stop**) if the overlap-driving covariate is weakly related to the
outcome. Conversely, acceptable overlap does not guarantee that TMLE will
perform well if the outcome surface is difficult to learn.

**Stage 3 fills this gap** by testing the complete TMLE pipeline on synthetic
data where the truth is known.

### Stage 2 summary

At this point the team has:

- Estimated $g(W)$ using the pre-specified model
- Evaluated overlap, ESS, weight distributions, and balance
- Documented whether any diagnostic flags triggered
- **Not touched** the outcome $Y$

The $g$-modeling strategy is now locked. If diagnostics are flagged, Stage 3
simulation determines whether the full TMLE pipeline still works.


# Stage 3. Outcome-Blind Simulation Validation {#stage-3}

## What happens at this stage

Stage 2 told us about the treatment mechanism. Stage 3 answers the harder
question: **does the entire TMLE pipeline perform adequately under conditions
that resemble this study?**

The team runs a simulation study using a pre-specified data-generating process
(DGP). The DGP is calibrated to match the observed $(W, A)$ distribution but
generates **synthetic outcomes** — not the real $Y$. Because the truth is known
in the simulation, the team can evaluate bias, variance, and CI coverage.

This stage is outcome-blind with respect to the *real* data: the simulated
outcomes come from a parametric model specified before real outcomes are
accessed.

### What Stage 3 validates that Stage 2 cannot

| Question | Stage 2 | Stage 3 |
|----------|---------|---------|
| Is $\hat{g}$ well-estimated? | Yes | Yes |
| Are PS extreme? | Yes | Yes |
| Is $\hat{Q}$ adequate? | **No** | Yes (on simulated data) |
| Does the targeting step work? | **No** | Yes |
| Is IC-based SE reliable? | **No** | Yes |
| Does CI coverage reach nominal? | **No** | Yes |
| Is the estimator better than alternatives? | **No** | Yes |

### Define the simulation DGP

The DGP must be specified *before* real outcomes are seen. It should reflect
plausible outcome models given domain knowledge. Here we define two scenarios:
one where the overlap-driving covariate barely affects $Y$ (weak W4), and one
where it matters substantially (strong W4).

```{r sim-dgp}
# Shared DGP function: generates data with known truth
sim_data <- function(n, seed, s_g, beta4_Q = 0.04, tau_int3 = 0.0) {
  set.seed(seed)
  W1 <- rnorm(n); W2 <- rbinom(n, 1, 0.5)
  W3 <- rnorm(n); W4 <- rnorm(n)

  # Treatment model
  lp_g_raw <- s_g * (1.5 * W1 + 0.8 * W2 - 0.5 * W3 + 3.0 * W4)
  alpha0_g <- -mean(lp_g_raw) - 0.5
  g_true <- expit(alpha0_g + lp_g_raw)
  g_true <- pmin(pmax(g_true, 0.005), 0.995)
  A <- rbinom(n, 1, g_true)

  # Outcome model (nonlinear — misspecified by GLM)
  lp_Q <- -0.5 + 0.40 * A +
    0.50 * W1 + 0.25 * W1^2 + 0.70 * W2 + 0.35 * abs(W3) + beta4_Q * W4 +
    0.15 * A * (W1 > 0) + 0.05 * A * W2 + tau_int3 * A * W4
  Y <- rbinom(n, 1, expit(lp_Q))

  data.frame(W1, W2, W3, W4, A, Y)
}

# Compute true ATE by marginalizing Q over W
compute_truth <- function(n_mc = 100000, seed = 1, beta4_Q = 0.04, tau_int3 = 0.0) {
  set.seed(seed)
  W1 <- rnorm(n_mc); W2 <- rbinom(n_mc, 1, 0.5)
  W3 <- rnorm(n_mc); W4 <- rnorm(n_mc)

  lp_base <- -0.5 + 0.50 * W1 + 0.25 * W1^2 + 0.70 * W2 +
             0.35 * abs(W3) + beta4_Q * W4
  lp_Q1 <- lp_base + 0.40 + 0.15 * (W1 > 0) + 0.05 * W2 + tau_int3 * W4
  lp_Q0 <- lp_base

  c(EY1 = mean(expit(lp_Q1)), EY0 = mean(expit(lp_Q0)),
    RD_true = mean(expit(lp_Q1)) - mean(expit(lp_Q0)))
}

truth_weak   <- compute_truth(beta4_Q = 0.04, tau_int3 = 0.0)
truth_strong <- compute_truth(beta4_Q = 0.35, tau_int3 = 0.2)

cat("True ATE (weak W4):", round(truth_weak["RD_true"], 5), "\n")
cat("True ATE (strong W4):", round(truth_strong["RD_true"], 5), "\n")
```

### Define the TMLE estimator

This function runs one TMLE analysis using the pre-specified pipeline from
the analysis lock. It fits $Q$ (here via GLM for speed), fits $g$, performs
the targeting step, and returns the estimate with IC-based inference.

```{r tmle-function}
run_tmle <- function(dat, ps_clip = c(0.025, 0.975)) {
  W <- dat[, c("W1", "W2", "W3", "W4")]
  A <- dat$A; Y <- dat$Y; n <- nrow(dat)

  # Fit g (treatment model)
  g_fit <- glm(A ~ W1 + W2 + W3 + W4, data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])

  # Fit Q (outcome model — using enriched GLM for tutorial speed)
  Q_fit <- glm(Y ~ A * (W1 + W2 + W3 + W4) + I(W1^2) + I(abs(W3)),
               data = dat, family = "binomial")
  Q_A  <- predict(Q_fit, type = "response")
  Q_1  <- predict(Q_fit, newdata = transform(dat, A = 1), type = "response")
  Q_0  <- predict(Q_fit, newdata = transform(dat, A = 0), type = "response")

  # Bound Q away from 0/1
  Q_A <- pmin(pmax(Q_A, 1e-5), 1 - 1e-5)
  Q_1 <- pmin(pmax(Q_1, 1e-5), 1 - 1e-5)
  Q_0 <- pmin(pmax(Q_0, 1e-5), 1 - 1e-5)

  # Targeting step
  H  <- A / g_hat - (1 - A) / (1 - g_hat)
  H1 <- 1 / g_hat
  H0 <- -1 / (1 - g_hat)

  eps_fit <- glm(Y ~ -1 + H, offset = logit(Q_A), family = "binomial")
  eps <- coef(eps_fit)

  Q_1_star <- expit(logit(Q_1) + eps * H1)
  Q_0_star <- expit(logit(Q_0) + eps * H0)
  Q_A_star <- expit(logit(Q_A) + eps * H)

  # Estimate
  psi <- mean(Q_1_star - Q_0_star)

  # IC-based inference
  eif <- H * (Y - Q_A_star) + (Q_1_star - Q_0_star) - psi
  se <- sqrt(var(eif) / n)
  ci <- psi + c(-1.96, 1.96) * se

  list(estimate = psi, se = se, ci_low = ci[1], ci_high = ci[2])
}

# Simpler comparators for Stage 3
run_crude <- function(dat) {
  p1 <- mean(dat$Y[dat$A == 1]); p0 <- mean(dat$Y[dat$A == 0])
  rd <- p1 - p0
  n1 <- sum(dat$A == 1); n0 <- sum(dat$A == 0)
  se <- sqrt(p1*(1-p1)/n1 + p0*(1-p0)/n0)
  list(estimate = rd, se = se, ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}

run_iptw <- function(dat, ps_clip = c(0.025, 0.975)) {
  g_fit <- glm(A ~ W1 + W2 + W3 + W4, data = dat, family = "binomial")
  g_hat <- pmin(pmax(fitted(g_fit), ps_clip[1]), ps_clip[2])
  A <- dat$A; Y <- dat$Y; n <- length(Y)
  w <- A / g_hat + (1 - A) / (1 - g_hat)
  d1 <- sum(w * A); d0 <- sum(w * (1 - A))
  mu1 <- sum(w * A * Y) / d1; mu0 <- sum(w * (1 - A) * Y) / d0
  rd <- mu1 - mu0
  phi1 <- w * A * (Y - mu1) / d1 * n
  phi0 <- w * (1 - A) * (Y - mu0) / d0 * n
  se <- sqrt(sum((phi1 - phi0)^2) / n^2)
  list(estimate = rd, se = se, ci_low = rd - 1.96*se, ci_high = rd + 1.96*se)
}
```

### Run the simulation

We run a compact simulation: 60 replicates per scenario, fast enough for a
tutorial but sufficient to assess performance.

```{r run-sim, cache=TRUE}
run_sim <- function(reps, n, s_g, beta4_Q, tau_int3, true_rd, seed_offset = 0) {
  results <- data.frame(
    rep = integer(0), method = character(0),
    est = numeric(0), se = numeric(0),
    ci_lo = numeric(0), ci_hi = numeric(0)
  )

  for (r in seq_len(reps)) {
    d <- sim_data(n, seed = seed_offset + r, s_g = s_g,
                  beta4_Q = beta4_Q, tau_int3 = tau_int3)

    for (meth in c("Crude", "IPTW", "TMLE")) {
      res <- tryCatch(
        switch(meth,
               Crude = run_crude(d),
               IPTW  = run_iptw(d),
               TMLE  = run_tmle(d)),
        error = function(e) list(estimate = NA, se = NA,
                                  ci_low = NA, ci_high = NA))

      results <- rbind(results, data.frame(
        rep = r, method = meth,
        est = res$estimate, se = res$se,
        ci_lo = res$ci_low, ci_hi = res$ci_high
      ))
    }
  }

  # Compute metrics
  metrics <- do.call(rbind, lapply(split(results, results$method), function(sub) {
    ok <- !is.na(sub$est)
    has_ci <- ok & !is.na(sub$ci_lo)
    e <- sub$est[ok]
    data.frame(
      method = sub$method[1],
      n_ok = sum(ok),
      bias = round(mean(e) - true_rd, 5),
      rmse = round(sqrt(mean((e - true_rd)^2)), 5),
      emp_sd = round(sd(e), 5),
      mean_se = round(mean(sub$se[ok], na.rm = TRUE), 5),
      coverage = round(mean(
        (sub$ci_lo[has_ci] <= true_rd) & (true_rd <= sub$ci_hi[has_ci])), 3),
      n_ci = sum(has_ci)
    )
  }))
  row.names(metrics) <- NULL

  list(results = results, metrics = metrics)
}

# --- Scenario A: Bad overlap, weak W4 (candidate false stop) ---
sim_weak <- run_sim(reps = 60, n = 400, s_g = 1.5,
                    beta4_Q = 0.04, tau_int3 = 0.0,
                    true_rd = truth_weak["RD_true"],
                    seed_offset = 1000)

# --- Scenario B: Bad overlap, strong W4 (candidate true stop) ---
sim_strong <- run_sim(reps = 60, n = 400, s_g = 1.5,
                      beta4_Q = 0.35, tau_int3 = 0.2,
                      true_rd = truth_strong["RD_true"],
                      seed_offset = 2000)
```

### Simulation results

```{r sim-results}
cat("--- Scenario A: Weak W4 (candidate false stop) ---\n")
cat("True RD:", round(truth_weak["RD_true"], 5), "\n\n")
kable(sim_weak$metrics, digits = 4,
      caption = "Scenario A: Weak W4, bad overlap (s_g = 1.5)")

cat("\n--- Scenario B: Strong W4 (candidate true stop) ---\n")
cat("True RD:", round(truth_strong["RD_true"], 5), "\n\n")
kable(sim_strong$metrics, digits = 4,
      caption = "Scenario B: Strong W4, bad overlap (s_g = 1.5)")
```

```{r sim-sampling-dist, fig.width=9, fig.height=6}
# Combine for plotting
plot_dat <- rbind(
  cbind(scenario = "A: Weak W4 (false stop?)", sim_weak$results),
  cbind(scenario = "B: Strong W4 (true stop?)", sim_strong$results)
)
plot_dat <- plot_dat[!is.na(plot_dat$est), ]
plot_dat$method <- factor(plot_dat$method, levels = c("Crude", "IPTW", "TMLE"))

truth_lines <- data.frame(
  scenario = c("A: Weak W4 (false stop?)", "B: Strong W4 (true stop?)"),
  truth = c(truth_weak["RD_true"], truth_strong["RD_true"])
)

ggplot(plot_dat, aes(x = est, fill = method)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  geom_vline(data = truth_lines, aes(xintercept = truth),
             linewidth = 0.8, linetype = "solid") +
  facet_grid(method ~ scenario, scales = "free") +
  labs(title = "Stage 3: Sampling Distributions Under Two DGP Scenarios",
       subtitle = "Black line = true ATE",
       x = "Estimated RD", y = "Count") +
  guides(fill = "none") +
  theme_minimal(base_size = 11)
```

### Interpret the simulation gate

```{r sim-gate}
gate_check <- function(metrics, scenario_name, targets) {
  tmle <- metrics[metrics$method == "TMLE", ]
  bias_ok <- abs(tmle$bias) < targets$max_abs_bias
  cov_ok  <- tmle$coverage >= targets$min_coverage

  cat(scenario_name, "\n")
  cat("  TMLE bias:", tmle$bias, " (threshold: <", targets$max_abs_bias, ")",
      ifelse(bias_ok, "PASS", "FLAG"), "\n")
  cat("  TMLE coverage:", tmle$coverage, " (threshold: >=", targets$min_coverage, ")",
      ifelse(cov_ok, "PASS", "FLAG"), "\n")
  cat("  Gate decision:", ifelse(bias_ok & cov_ok, "GO", "STOP"), "\n\n")
  invisible(bias_ok & cov_ok)
}

gate_A <- gate_check(sim_weak$metrics, "Scenario A (Weak W4)",
                     analysis_lock$sim_targets)
gate_B <- gate_check(sim_strong$metrics, "Scenario B (Strong W4)",
                     analysis_lock$sim_targets)
```

### Stage 3 interpretation

The simulation reveals something that Stage 2 cannot:

- **Scenario A (weak W4)**: Overlap diagnostics flag the PS distribution, but
  TMLE still performs well because W4 barely affects the outcome. The overlap
  flag was a **false alarm**. The simulation-based gate says GO.

- **Scenario B (strong W4)**: The same overlap problem now matters because W4
  enters both $g$ and $Q$ strongly. TMLE cannot compensate. The gate says
  STOP.

Both scenarios have identical treatment mechanisms and identical Stage 2
diagnostics. The difference is visible only through Stage 3 simulation, which
tests the full estimator pipeline — not just the propensity score.

::: {.callout-important}
This is the core methodological point: **Stage 2 evaluates $g$ alone. Stage 3
evaluates the entire TMLE pipeline ($g$, $Q$, targeting, inference).** A GO
decision from Stage 3 provides evidence that the pre-specified TMLE analysis
will produce valid results despite overlap concerns.
:::

### Stage 3 summary

At this point the team has:

- Run the full TMLE pipeline on synthetic data under pre-specified DGPs
- Evaluated bias, variance, and CI coverage
- Compared TMLE against simpler estimators (crude, IPTW)
- Made a gate decision: GO for the weak-W4 scenario, STOP for strong-W4
- **Still not touched** the real outcome $Y$

If this were a real study and the simulation gate said GO, the team would
proceed to Stage 4. If STOP, they would document the concern and consider
alternatives (e.g., different estimand, different PS truncation, or reporting
the analysis as inconclusive).


# Stage 4. Unblinded TMLE Estimation {#stage-4}

## What happens at this stage

The analyst now receives the full dataset including outcomes $Y$. This is the
first point at which the **outcome regression $Q(A,W)$** is estimated on real
data. The treatment model strategy was locked in Stage 2; the simulation
validation was completed in Stage 3; only now does the targeting step run on
real data.

### Prerequisites

Stage 4 proceeds only after:

1. Stage 1 analysis lock is documented and versioned
2. Stage 2 diagnostics are reviewed and recorded
3. Stage 3 simulation gate says GO (or, if STOP, the team documents why they
   proceed or what alternative they adopt)

### Run TMLE on the real data

The `dat` object created earlier now has its $Y$ column accessed for the first
time:

```{r stage4-tmle}
cat("Stage 4: Outcome column now accessible.\n")
cat("Y prevalence:", round(mean(dat$Y), 3), "\n\n")

# Run the pre-specified TMLE analysis
stage4_result <- run_tmle(dat, ps_clip = analysis_lock$ps_truncation)

cat("=== Stage 4 TMLE Result ===\n")
cat("ATE estimate (RD):", round(stage4_result$estimate, 5), "\n")
cat("Standard error:", round(stage4_result$se, 5), "\n")
cat("95% CI: [", round(stage4_result$ci_low, 5), ",",
    round(stage4_result$ci_high, 5), "]\n")
```

### Comparator estimates

For transparency, we also run the pre-specified comparator estimators:

```{r stage4-comparators}
stage4_crude <- run_crude(dat)
stage4_iptw  <- run_iptw(dat, ps_clip = analysis_lock$ps_truncation)

est_tab <- data.frame(
  Estimator = c("Crude RD", "IPTW (Hajek)", "TMLE"),
  Estimate = round(c(stage4_crude$estimate, stage4_iptw$estimate,
                     stage4_result$estimate), 5),
  SE = round(c(stage4_crude$se, stage4_iptw$se, stage4_result$se), 5),
  CI_low = round(c(stage4_crude$ci_low, stage4_iptw$ci_low,
                   stage4_result$ci_low), 5),
  CI_high = round(c(stage4_crude$ci_high, stage4_iptw$ci_high,
                    stage4_result$ci_high), 5)
)
kable(est_tab, caption = "Stage 4: Effect estimates from pre-specified estimators")
```

### What just happened — step by step

1. **$Q$ was estimated on the real data** for the first time: an enriched GLM
   (with $W_1^2$, $|W_3|$, and treatment interactions) was fit to $(W, A, Y)$.

2. **$g$ was re-estimated** using the same locked strategy from Stage 2. In a
   production implementation, one might literally load the Stage 2 $g$-fit
   rather than re-estimating.

3. **The targeting step** constructed the clever covariate
   $H = A/\hat{g} - (1-A)/(1-\hat{g})$ and ran a logistic fluctuation
   $\text{logit}(\bar{Q}^*) = \text{logit}(\bar{Q}) + \epsilon H$ to solve
   the efficient influence function estimating equation.

4. **The ATE** was computed as $\hat\psi = n^{-1}\sum (\bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i))$.

5. **Inference** used the efficient influence function:
   $D_i = H_i(Y_i - \bar{Q}^*_i) + \bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i) - \hat\psi$
   with $\hat{\text{SE}} = \sqrt{n^{-1}\hat{\text{Var}}(D)}$.

### Stage 4 summary

- The real outcome regression was fit for the first time
- It was combined with the locked $g$-strategy
- The TMLE targeting step produced the final estimate
- The estimate was reported alongside the Stage 2 diagnostics and Stage 3
  simulation evidence
- No analytic decisions were changed based on the observed outcomes


# Full Simulation Results {#full-sim}

The compact simulation above demonstrates the staged workflow. For the
production-scale simulation with richer estimators (TMLE-ML with SuperLearner,
CV-TMLE, PS-matched regression) and larger sample sizes, see the outputs
from `sim_scenario1_v4_false_stop_inference.R`.

```{r full-sim-results, results='asis'}
if (!is.null(weak) && !is.null(strong)) {
  cat("### Full simulation results available\n\n")
  cat("These results come from the v4 simulation with N =", weak$params$n_main,
      "and", weak$params$reps, "replicates per mode.\n\n")

  # Performance table
  point_methods <- c("Crude", "Adj. Regression (analytic)",
                     "IPTW (analytic)", "PS-Matched (pairs boot)",
                     "TMLE-GLM (IC)", "TMLE-ML (IC)", "CV-TMLE-ML (IC)")
  show_perf <- function(metrics, label) {
    sub <- metrics[metrics$subset == "all" &
                   metrics$method %in% point_methods, ]
    sub$mode <- label
    sub[, c("mode", "method", "bias", "rmse", "coverage", "mcse_coverage")]
  }
  combined <- rbind(show_perf(weak$metrics, "Weak W4"),
                    show_perf(strong$metrics, "Strong W4"))
  kable(combined, digits = 4, row.names = FALSE,
        caption = "Full simulation: estimator performance (all replicates)")

  # Gate results
  if (!is.null(gate_combined)) {
    cat("\n### Stop/go gate results\n\n")
    kable(gate_combined, digits = 4, row.names = FALSE,
          caption = "Gate results from full simulation")
  }
} else {
  cat("Full simulation outputs not available.\n",
      "Run: `Rscript scripts/sim_scenario1_v4_false_stop_inference.R`\n")
}
```

```{r full-sim-coverage, fig.width=10, fig.height=7}
if (!is.null(weak) && !is.null(strong)) {
  cov_all <- rbind(
    cbind(mode = "Weak W4", weak$metrics[weak$metrics$subset == "all" &
                                          !is.na(weak$metrics$coverage), ]),
    cbind(mode = "Strong W4", strong$metrics[strong$metrics$subset == "all" &
                                              !is.na(strong$metrics$coverage), ])
  )
  if (nrow(cov_all) > 0) {
    cov_all$ci_lo <- pmax(cov_all$coverage - 1.96 * cov_all$mcse_coverage, 0)
    cov_all$ci_hi <- pmin(cov_all$coverage + 1.96 * cov_all$mcse_coverage, 1)
    ggplot(cov_all, aes(x = method, y = coverage, fill = mode)) +
      geom_col(position = "dodge", alpha = 0.8, width = 0.7) +
      geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                    position = position_dodge(0.7), width = 0.2) +
      geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
      coord_flip() + ylim(0, 1) +
      scale_fill_manual(values = c("Weak W4" = "steelblue",
                                    "Strong W4" = "firebrick")) +
      labs(title = "Full Simulation: 95% CI Coverage by Inference Method",
           x = NULL, y = "Coverage", fill = "Mode") +
      theme_minimal(base_size = 11)
  }
}
```


# Key Takeaways for Operationalizing TMLE in a Clean Room

1. **Stage 2 evaluates $g$ and overlap, not the entire estimator.** Propensity
   score diagnostics are necessary but not sufficient for TMLE validation.
   Acceptable overlap does not guarantee valid TMLE inference; poor overlap
   does not necessarily invalidate it.

2. **Stage 3 validates the full TMLE pipeline on synthetic data.** This is the
   step that turns a generic TMLE analysis into a credible clean-room workflow.
   By testing bias, variance, and coverage under a pre-specified DGP, the team
   can make an evidence-based GO/STOP decision before seeing real outcomes.

3. **Stage 4 is where the real outcome regression and targeting step occur.**
   The $g$-strategy is carried forward from Stage 2, but $Q$ is fit on real
   data only at this final stage. No analytic decisions should change based on
   what $Y$ shows.

4. **Outcome-blind simulation is what turns a generic TMLE workflow into a
   more credible clean-room analysis.** Without Stage 3, the team can only
   say "overlap looks okay" or "overlap looks concerning." With Stage 3, they
   can say "under plausible DGPs, our pre-specified TMLE pipeline produces
   valid estimates with near-nominal coverage, even when overlap is flagged."

5. **The false-stop / true-stop distinction is estimator-specific.** The same
   overlap environment can be a false stop for TMLE (which has $Q$ to fall
   back on) and a true stop for IPTW (which depends entirely on $g$). This is
   TMLE's double robustness in action.


## Reproducing the Full Simulation

```bash
# Compact tutorial simulation (runs within this QMD)
# — defined inline above

# Production simulation with SuperLearner, 7 estimators, mixed overlap:
Rscript scripts/sim_scenario1_v4_false_stop_inference.R fast   # N=400, 120 reps
Rscript scripts/sim_scenario1_v4_false_stop_inference.R eval   # N=600, 300 reps
```

## References

- van der Laan MJ, Rose S. *Targeted Learning: Causal Inference for
  Observational and Experimental Data*. Springer, 2011.
- van der Laan MJ, Rose S. *Targeted Learning in Data Science*. Springer, 2018.
- Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ. Diagnosing and
  responding to violations in the positivity assumption.
  *Statistical Methods in Medical Research*, 2012.
- Gruber S, van der Laan MJ. tmle: An R package for targeted maximum
  likelihood estimation. *Journal of Statistical Software*, 2012.
