---
title: "Clean-Room TMLE for Pharmacoepidemiology: A Tutorial"
subtitle: "Staged Design, Outcome-Blind Gating, and Simulation-Based Estimator Selection"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(ggplot2)
library(knitr)

# Load simulation outputs if available
out_dir <- file.path("..", "results", "scenario1_v4")
has_sim <- dir.exists(out_dir)

if (has_sim) {
  load_mode <- function(mode) {
    d <- file.path(out_dir, mode)
    list(
      truth   = readRDS(file.path(d, "truth.rds")),
      results = readRDS(file.path(d, "replicate_results.rds")),
      metrics = read.csv(file.path(d, "summary_overall.csv")),
      gate    = readRDS(file.path(d, "gate_results.rds")),
      params  = readRDS(file.path(d, "params_used.rds"))
    )
  }
  weak   <- tryCatch(load_mode("weak"),   error = function(e) NULL)
  strong <- tryCatch(load_mode("strong"), error = function(e) NULL)
  gate_combined <- tryCatch(
    readRDS(file.path(out_dir, "gate_combined.rds")),
    error = function(e) {
      if (!is.null(weak) && !is.null(strong))
        rbind(cbind(mode = "weak", weak$gate), cbind(mode = "strong", strong$gate))
      else NULL
    })
} else {
  weak <- strong <- gate_combined <- NULL
  message("No simulation outputs found. Narrative sections will render without data.\n",
          "To generate outputs: Rscript scripts/sim_scenario1_v4_false_stop_inference.R")
}
```

# Part I: The Clean-Room Staged Design

## 1. Why a Staged Design?

Observational pharmacoepidemiology studies face a fundamental credibility
challenge: the investigator can see the data while making analytic decisions.
This creates risk of conscious or unconscious bias — choosing a model that
"works" for the outcome, dropping inconvenient subgroups, or selecting among
estimators post hoc. Regulatory agencies and journal reviewers rightly worry
about these researcher degrees of freedom.

The **clean-room staged design** addresses this by separating the analysis
into phases with explicit rules about what information the analyst can access
at each stage. The key principle is:

> All decisions about the treatment model, estimand, estimator, and
> feasibility criteria are locked *before any outcome data are examined*.

This does not mean the analyst works blind. They have full access to covariates
$(W)$ and treatment assignment $(A)$ — only the outcome $(Y)$ is withheld
until Stage 3. The separation provides a verifiable audit trail: the analysis
plan, propensity score model, and stop/go decisions can all be documented and
locked before outcomes are unblinded.

## 2. The Four Stages

```{r stages-diagram, fig.height=3, fig.width=9}
# Visual overview of staged design
stage_df <- data.frame(
  stage = factor(c("Stage 1", "Stage 2", "Stage 2b", "Stage 3"),
                 levels = c("Stage 1", "Stage 2", "Stage 2b", "Stage 3")),
  label = c("Pre-specification", "Outcome-Blind\nDiagnostics",
            "Outcome-Blind\nSimulation OC", "Estimation"),
  data_access = c("Protocol only", "W, A only", "W, A only\n(simulated Y)",
                   "W, A, Y"),
  x = 1:4
)

ggplot(stage_df, aes(x = x, y = 0)) +
  geom_tile(aes(fill = stage), width = 0.85, height = 0.5, alpha = 0.7) +
  geom_text(aes(label = label), size = 3.5, fontface = "bold") +
  geom_text(aes(y = -0.35, label = data_access), size = 2.8, color = "grey30") +
  scale_fill_manual(values = c("Stage 1" = "#4DAF4A", "Stage 2" = "#377EB8",
                                 "Stage 2b" = "#984EA3", "Stage 3" = "#E41A1C")) +
  geom_segment(aes(x = x + 0.45, xend = x + 0.55, y = 0, yend = 0),
               data = stage_df[1:3, ],
               arrow = arrow(length = unit(0.15, "cm")), color = "grey40") +
  coord_cartesian(ylim = c(-0.55, 0.35)) +
  theme_void() + guides(fill = "none")
```

### Stage 1: Pre-specification

Before touching any data, the investigator specifies:

- **Estimand**: What causal quantity is targeted? (e.g., ATE on the risk
  difference scale: $\psi = E[Y^1] - E[Y^0]$)
- **Estimators**: Which statistical methods will be applied? (e.g., TMLE-ML,
  IPTW, G-computation)
- **Treatment model**: What variables enter the propensity score? What
  modeling strategy? (e.g., SuperLearner with specific library)
- **Overlap diagnostics**: What metrics and thresholds define "adequate" vs
  "inadequate" overlap?
- **Decision rules**: Under what conditions does the analysis proceed vs stop?

This pre-specification is recorded in a protocol or statistical analysis plan
(SAP) that is timestamped and locked — analogous to a clinical trial's SAP
being finalized before database lock.

### Stage 2: Outcome-Blind Diagnostics

The analyst now has access to covariates $(W)$ and treatment assignment $(A)$,
but **not outcomes** $(Y)$. At this stage:

1. **Fit the propensity score model** $\hat{g}(W) = \hat{P}(A=1|W)$
2. **Evaluate overlap** using pre-specified metrics:

| Diagnostic | What it measures | Typical threshold |
|-----------|-----------------|-------------------|
| Fraction of extreme PS | $P(\hat{g} < \epsilon$ or $\hat{g} > 1-\epsilon)$ | $< 30$--$40\%$ of sample |
| ESS fraction | $\text{ESS}/n$ where $\text{ESS} = (\sum w_i)^2 / \sum w_i^2$ | $> 20$--$30\%$ |
| Max IPTW weight | $\max_i |w_i|$ | $< 40$--$60$ |
| SMD after weighting | Standardized mean difference | $< 0.1$ |

3. **Assess covariate balance** in the weighted/matched sample via love plots
4. **Make the stop/go decision**: Do we proceed to estimation, or do we flag a
   concern?

::: {.callout-important}
**No outcome data are examined at this stage.** The propensity score, overlap
diagnostics, and stop/go decision are all functions of $(W, A)$ only. This is
what makes the design "clean room" — decisions about feasibility cannot be
influenced by knowledge of $Y$.
:::

### Stage 2b: Outcome-Blind Simulation-Based Operating Characteristics

This is the novel stage that connects the clean-room design to the simulation
study presented in this tutorial.

**The problem**: Stage 2 diagnostics tell us that overlap is poor, but they
don't tell us whether a specific estimator (e.g., TMLE-ML) can still produce
valid inference despite the poor overlap. Standard overlap metrics are
conservative — they flag any PS near the boundary, regardless of whether
the boundary matters for the specific outcome model.

**The solution**: Run a **simulation study** that mimics the observed covariate
and treatment distributions but generates synthetic outcomes under a known
data-generating process (DGP). Because we control the DGP, we know the true
causal effect and can evaluate:

- **Bias**: Is the estimator's point estimate close to the truth?
- **Coverage**: Do the 95% confidence intervals contain the truth at the
  nominal rate?
- **RMSE**: How variable is the estimator?

This simulation uses only $(W, A)$ from the real data (or distributions
calibrated to them) — it does not require the actual outcomes. The synthetic
outcomes are generated from a parametric model that the analyst specifies
*before* seeing real outcomes, making this step outcome-blind.

### Stage 3: Estimation

Only after Stages 1--2b are completed and documented does the analyst gain
access to outcomes $(Y)$. At this point:

- Apply the pre-specified estimators to the real data
- Compute confidence intervals using the pre-specified inference method
- Report results alongside the Stage 2 diagnostics
- If Stage 2b flagged a concern, report it transparently

The analyst does **not** switch estimators, change the propensity score model,
or modify the analysis based on what the outcomes show. Any such changes
require a documented protocol amendment with justification.


# Part II: The Gating Process

## 3. What Is a Stop/Go Gate?

A stop/go gate is a pre-specified decision rule applied at Stage 2 (or 2b)
that determines whether the analysis should proceed with the planned estimator
or whether the overlap environment is too hostile for valid inference.

The gate converts continuous diagnostics into a binary decision:

```
IF (|bias| < threshold_bias AND coverage >= threshold_coverage):
    → GO: proceed with estimation
ELSE:
    → STOP: flag concern, consider alternatives
```

### 3.1 Why Not Just Trim or Clip?

A common response to poor overlap is to truncate the propensity score (clip
extreme values to some bound, e.g., [0.025, 0.975]). This is necessary but
insufficient:

- **Trimming changes the estimand**: Excluding subjects with extreme PS changes
  the population for which we estimate the effect
- **Clipping introduces bias**: Bounding $\hat{g}$ away from 0/1 prevents
  infinite weights but introduces systematic bias
- **The degree of truncation is ad hoc**: There's no principled way to choose
  the clipping bounds without knowing something about the outcome model

The gate approach is different: it asks whether the *entire estimation
pipeline* (including any trimming/clipping) produces valid results under
realistic conditions, rather than just whether individual weights are
well-behaved.

### 3.2 False Stops and True Stops

A **false stop** occurs when the overlap flag triggers but the estimator
would have performed well — the gate is too conservative. A **true stop**
occurs when the flag correctly identifies a genuine barrier to valid inference.

Whether a stop is true or false depends on the *estimator*, not just the
overlap environment. This is the key insight:

> The same overlap violation can be a false stop for TMLE-ML (which adapts
> through both $Q$ and $g$ models) but a true stop for IPTW (which relies
> entirely on $g$).

This is where TMLE's **double robustness** matters. If the overlap-driving
covariate is nearly irrelevant to the outcome (weak in $Q$), TMLE can
effectively "route around" the positivity violation by leaning on a good
outcome model. IPTW has no such fallback.

### 3.3 Gate Criteria Used in This Study

The gate evaluates TMLE-ML in the simulation and requires:

- **|Bias| < 0.01** on the RD scale (point estimation validity)
- **95% CI coverage ≥ 0.93** (inferential validity)

The coverage threshold of 0.93 (rather than 0.95) accounts for the
finite-sample simulation uncertainty (MCSE) and the inherent conservatism of
requiring exact nominal coverage. Both criteria must hold in the flagged
overlap stratum for the gate to pass.


# Part III: Where TMLE Fits In

## 4. TMLE as the Estimator of Choice

### 4.1 What TMLE Does

Targeted Minimum Loss-Based Estimation (TMLE) is a two-step procedure:

1. **Initial estimation**: Fit models for $Q(A,W) = E[Y|A,W]$ (outcome model)
   and $g(W) = P(A=1|W)$ (treatment model) using flexible machine learning
   (SuperLearner)
2. **Targeting step**: Update the initial $Q$ estimate by "fluctuating" it in
   the direction of the efficient influence function (EIF), using $g$ to
   construct the clever covariate $H = A/g - (1-A)/(1-g)$

The result is an estimator that:

- Is **doubly robust**: consistent if either $Q$ or $g$ is correctly specified
- Is **efficient**: achieves the semiparametric efficiency bound when both
  models are well-estimated
- Respects the **parameter space**: point estimates stay within natural bounds
  (e.g., probabilities between 0 and 1)

### 4.2 Why TMLE, Not IPTW or G-Computation?

| Property | TMLE | IPTW | G-computation |
|----------|------|------|---------------|
| Relies on $g$ only | No | **Yes** | No |
| Relies on $Q$ only | No | No | **Yes** |
| Double robustness | Yes | No | No |
| Respects bounds | Yes | No | Yes |
| Efficiency | Optimal | No | No |
| Handles near-positivity | Better (via $Q$) | Poorly | Ignores $g$ |

Under near-positivity violations:

- **IPTW** suffers directly because extreme weights $1/\hat{g}$ inflate
  variance and can drive bias
- **G-computation** ignores $g$ entirely, so overlap doesn't directly hurt it,
  but it relies completely on correct specification of $Q$
- **TMLE** has a middle path: the targeting step uses $g$ through the clever
  covariate, but if $g$ is extreme in regions where $Q$ is well-estimated,
  the update is small and the estimator remains stable

### 4.3 TMLE Variants in This Study

| Variant | Q model | g model | Cross-fitting | Use case |
|---------|---------|---------|---------------|----------|
| TMLE-GLM | `glm(Y ~ A+W)` | `glm(A ~ W)` | No | Misspecification benchmark |
| TMLE-ML | SuperLearner | SuperLearner | No | Main estimator |
| CV-TMLE-ML | Ranger/GLM per fold | Ranger/GLM per fold | Yes (V-fold) | Theoretical gold standard |

**CV-TMLE** (cross-validated TMLE) uses V-fold cross-fitting: the data are
split into $V$ folds, nuisance models are trained on $V-1$ folds and
predictions are made on the held-out fold. This ensures that the nuisance
estimates used in the targeting step are independent of the data used to
evaluate the EIF, relaxing conditions needed for asymptotic linearity.

### 4.4 Inference for TMLE

TMLE provides multiple inference approaches:

**IC-based (influence curve)**:
The efficient influence function $D_i = H_i(Y_i - \bar{Q}^*(A_i, W_i)) + \bar{Q}^*(1, W_i) - \bar{Q}^*(0, W_i) - \hat\psi$ provides an asymptotically linear representation:
$$\hat\psi - \psi_0 \approx \frac{1}{n}\sum_{i=1}^n D_i$$
The variance is estimated as $\hat\sigma^2 = n^{-1}\sum D_i^2$, giving
Wald-type CIs. This is fast but can under-cover at moderate $n$ because it
treats the nuisance estimates as fixed.

**Full nonparametric bootstrap**:
Resample the data with replacement and refit the *entire* TMLE pipeline
(nuisance models + targeting) on each resample. The bootstrap distribution
of $\hat\psi^*$ captures nuisance estimation uncertainty. This is the most
honest inference but computationally expensive.

**Multiplier bootstrap (Rademacher)**:
Perturb the EIF with random signs:
$\psi^*_b = \hat\psi + n^{-1}\sum\xi_i D_i$ where $\xi_i \sim \{-1, +1\}$.
This is fast (no refitting) but inherits the same limitation as IC-based
inference — it doesn't account for nuisance estimation uncertainty.


# Part IV: Outcome-Blind Simulation for Estimator Selection

## 5. The Core Idea

In a real analysis, after Stage 2 diagnostics flag an overlap concern, the
analyst faces a decision: should I stop the analysis, switch to a different
estimand (e.g., ATT), or proceed with a robust estimator? The
**outcome-blind simulation** provides evidence to support this decision.

The simulation works as follows:

1. **Calibrate a DGP** to match the observed $(W, A)$ distribution. The
   covariate distributions and propensity score model are taken directly from
   the real data.

2. **Specify a range of outcome models** $Q_0(A, W)$ that span plausible
   scenarios. These are pre-specified parametric models — the analyst does
   *not* see the real outcomes when choosing them.

3. **For each scenario**, generate $R$ replicate datasets with known truth
   $\psi_0$, run all candidate estimators, and evaluate bias, coverage, RMSE.

4. **Select the working estimator** based on which method achieves acceptable
   bias and coverage across the plausible scenarios, especially in the
   flagged overlap stratum.

::: {.callout-tip}
**Key insight**: Even when PS overlap is poor, TMLE-ML may still work well
if the overlap-driving covariate has a weak relationship with the outcome.
The simulation can demonstrate this by comparing "weak W4" (false stop)
vs "strong W4" (true stop) scenarios.
:::

## 6. Simulation Design

### 6.1 Data-Generating Process

The DGP used in this study generates data with four covariates:

- $W_1 \sim N(0,1)$ — continuous confounder
- $W_2 \sim \text{Bernoulli}(0.5)$ — binary confounder
- $W_3 \sim N(0,1)$ — continuous confounder
- $W_4 \sim N(0,1)$ — the **overlap driver**: large coefficient in $g$,
  variable coefficient in $Q$

**Treatment model** (propensity score):
$$
\text{logit}(g_0(W)) = \alpha_0 + s_g \cdot \left(
  1.5 W_1 + 0.8 W_2 - 0.5 W_3 + 3.0 W_4
\right)
$$

where $s_g$ is a treatment-strength multiplier that varies by overlap regime:

- $s_g = s_{ok}$ (e.g., 0.5) for **ok-overlap** replicates
- $s_g = s_{bad}$ (e.g., 1.5--2.0) for **bad-overlap** replicates
- Each replicate is randomly assigned to ok or bad with equal probability

The intercept $\alpha_0$ is adjusted so that $P(A=1) \approx 0.35$, ensuring
more controls than treated (important for PS matching to have enough donor
units).

W4 has the largest coefficient (3.0) and is the primary driver of overlap
problems. When $s_g$ is large, subjects with extreme W4 values have propensity
scores very close to 0 or 1.

**Outcome model** (intentionally nonlinear to create misspecification):
$$
\text{logit}(Q_0(A,W)) = \beta_0 + \beta_A \cdot A +
  \underbrace{0.50 W_1 + 0.25 W_1^2}_{h_1} +
  \underbrace{0.70 W_2}_{h_2} +
  \underbrace{0.35 |W_3|}_{h_3} +
  \underbrace{\beta_{4,Q} W_4}_{h_4}
$$
$$
  + \tau_1 A \cdot I(W_1 > 0) + \tau_2 A \cdot W_2 + \tau_3 A \cdot W_4
$$

The nonlinear terms ($W_1^2$, $|W_3|$, and the treatment-effect interactions)
mean that a simple main-effects GLM is misspecified for both $Q$ and $g$.
This is realistic — in real pharmacoepi studies, the true functional forms
are never linear.

### 6.2 The Two Modes

The simulation runs two scenarios that differ in how strongly W4 affects
the outcome:

```{r mode-comparison, results='asis'}
if (!is.null(weak)) {
  mode_tab <- data.frame(
    Parameter = c("s_bad (overlap severity)",
                  "beta4_Q (W4 in outcome)",
                  "tau_int1 (A × I(W1>0))",
                  "tau_int2 (A × W2)",
                  "tau_int3 (A × W4)"),
    `Weak W4` = c(weak$params$s_bad, weak$params$beta4_Q,
                  weak$params$tau_int1, weak$params$tau_int2,
                  weak$params$tau_int3),
    `Strong W4` = c(strong$params$s_bad, strong$params$beta4_Q,
                    strong$params$tau_int1, strong$params$tau_int2,
                    strong$params$tau_int3),
    check.names = FALSE
  )
  kable(mode_tab, digits = 3,
        caption = "Mode-specific DGP parameters: Weak vs Strong W4")
} else {
  cat("\n| Parameter | Weak W4 | Strong W4 |")
  cat("\n|-----------|---------|-----------|")
  cat("\n| s_bad (overlap severity) | 1.50 | 2.00 |")
  cat("\n| beta4_Q (W4 in outcome) | 0.04 | 0.35 |")
  cat("\n| tau_int1 (A × I(W1>0)) | 0.15 | 0.25 |")
  cat("\n| tau_int2 (A × W2) | 0.05 | 0.10 |")
  cat("\n| tau_int3 (A × W4) | 0.00 | 0.20 |\n")
}
```

**Weak W4 mode (candidate false stop):**

- W4 has a large coefficient in $g$ (drives overlap) but barely enters $Q$
  ($\beta_{4,Q} = 0.04$, no $A \times W4$ interaction)
- Treatment-effect heterogeneity is modest ($\tau_1 = 0.15$, $\tau_2 = 0.05$)
- The overlap flag fires (PS diagnostics see the extreme W4 effect in $g$),
  but TMLE can compensate by fitting $Q$ well — W4 barely matters for $Y$
- **Expected result**: TMLE-ML maintains valid inference → **false stop**

**Strong W4 mode (candidate true stop):**

- W4 is important in both $g$ and $Q$ ($\beta_{4,Q} = 0.35$, $\tau_3 = 0.20$)
- Treatment-effect heterogeneity is larger ($\tau_1 = 0.25$, $\tau_2 = 0.10$)
- The overlap flag fires, *and* TMLE cannot compensate because getting $Q$
  right in the extreme-W4 region matters for the ATE
- **Expected result**: TMLE-ML shows bias and/or under-coverage → **true stop**

### 6.3 Mixed Overlap Regimes

Within each mode, replicates are randomly assigned to an ok-overlap or
bad-overlap regime (50/50 by default). This creates a within-simulation
comparison:

```
+---------------------+------------+------------+
|                     | Weak W4    | Strong W4  |
+---------------------+------------+------------+
| OK  overlap flag    | baseline   | baseline   |
| BAD overlap flag    | FALSE stop | TRUE stop  |
+---------------------+------------+------------+
```

The ok-overlap stratum serves as a **positive control** — all estimators
should perform well there. The bad-overlap stratum is where the interesting
comparison happens.

### 6.4 Truth Computation

The true ATE is computed by marginalizing over the covariate distribution:
$$
\psi_0 = E_W\left[\text{expit}(Q_0(1,W))\right] -
         E_W\left[\text{expit}(Q_0(0,W))\right]
$$

This is approximated via plug-in on a large Monte Carlo sample ($n = 100{,}000$).
Critically, **the truth depends only on the outcome model $Q_0$ and the
marginal distribution of $W$** — not on the treatment model $g_0$. The
treatment mechanism affects the overlap environment but not the estimand.

## 7. Estimator Comparison

The simulation evaluates seven estimators, spanning the spectrum from simple
to sophisticated:

### 7.1 Estimators

**1. Crude RD**: $\hat{P}(Y=1|A=1) - \hat{P}(Y=1|A=0)$ — no adjustment for
confounding. Included as a "worst case" baseline.

**2. Adjusted Regression (G-computation)**: Fit a misspecified GLM
$\text{logit}(E[Y|A,W]) = \beta_0 + \beta_A A + \beta_1 W_1 + \ldots$ and
predict $\hat{E}[Y|A=1,W] - \hat{E}[Y|A=0,W]$, averaged over the sample.
Inference via GLM bootstrap (B=50). G-computation relies entirely on $Q$;
misspecification of $Q$ directly biases the estimate.

**3. IPTW (Hajek)**: Inverse probability of treatment weighting with
stabilized weights. Relies entirely on $g$; near-positivity violations
directly inflate variance.

**4. PS-Matched Regression**: Nearest-neighbor matching on the logit of the
propensity score, followed by G-computation on the matched sample. Combines
elements of both $g$ (matching) and $Q$ (outcome regression). Inference via
pairs bootstrap.

**5. TMLE-GLM**: TMLE with misspecified GLM for both $Q$ and $g$. Included
to show that double robustness requires *at least one* model to be correct —
misspecifying both can still bias TMLE.

**6. TMLE-ML**: TMLE with SuperLearner (SL.mean, SL.glmnet, SL.xgboost)
for both $Q$ and $g$. The main estimator of interest. IC-based inference.

**7. CV-TMLE-ML**: Cross-validated TMLE with V-fold sample splitting.
Theoretical gold standard for coverage properties. IC-based inference.

### 7.2 What the Simulation Reveals About Each Estimator

```{r estimator-table, results='asis'}
if (!is.null(weak)) {
  point_methods <- c("Crude", "Adj. Regression (analytic)",
                     "IPTW (analytic)", "PS-Matched (pairs boot)",
                     "TMLE-GLM (IC)", "TMLE-ML (IC)", "CV-TMLE-ML (IC)")

  show_point <- function(metrics, label) {
    sub <- metrics[metrics$subset == "all" &
                   metrics$method %in% point_methods, ]
    sub$mode <- label
    sub[, c("mode", "method", "n_success", "bias", "rmse", "emp_sd",
            "mean_se", "coverage", "mcse_coverage")]
  }

  combined <- rbind(show_point(weak$metrics, "Weak"),
                    show_point(strong$metrics, "Strong"))
  kable(combined, digits = 4, row.names = FALSE,
        caption = "Estimator performance: all replicates")
} else {
  cat("*Simulation outputs not available. Run the simulation to see results.*\n")
}
```

### 7.3 Performance by Overlap Stratum

This is the key table for the false-stop / true-stop demonstration. Splitting
results by overlap flag reveals whether TMLE-ML maintains performance when
the PS diagnostics are concerning.

```{r overlap-strat, results='asis'}
if (!is.null(weak)) {
  show_strat <- function(metrics, label) {
    tmle_rows <- metrics[grepl("TMLE-ML \\(IC\\)|CV-TMLE-ML \\(IC\\)", metrics$method) &
                         metrics$subset %in% c("overlap_ok", "overlap_bad"), ]
    if (nrow(tmle_rows) > 0) {
      tmle_rows$mode <- label
      tmle_rows[, c("mode", "subset", "method", "n_ci",
                     "bias", "coverage", "mcse_coverage")]
    }
  }
  strat_tab <- rbind(show_strat(weak$metrics, "Weak"),
                     show_strat(strong$metrics, "Strong"))
  if (!is.null(strat_tab) && nrow(strat_tab) > 0) {
    kable(strat_tab, digits = 4, row.names = FALSE,
          caption = "TMLE coverage by overlap stratum and mode")
  }
}
```


## 8. Selecting a Working Estimator Despite Poor PS Overlap

The simulation provides a principled basis for estimator selection when
overlap is flagged. The decision tree is:

### Step 1: Check baseline performance (ok-overlap stratum)

All estimators should perform well in the ok-overlap stratum. If they don't,
the problem is model misspecification or sample size, not overlap.

### Step 2: Check TMLE-ML in the flagged stratum

If TMLE-ML maintains |bias| < threshold and coverage ≥ threshold in the
bad-overlap stratum, the overlap flag is a **false stop**. The analyst can
document this evidence and proceed with TMLE-ML.

### Step 3: Check if the problem is overlap-specific

Compare TMLE-ML performance in ok vs bad strata. If coverage drops
substantially in the bad stratum, the overlap violation is genuine.

### Step 4: Consider alternatives

If TMLE-ML fails the gate in the flagged stratum:

- **Trim to the overlap region** (changes the estimand to ATT or ATO)
- **Switch to a more robust estimand** (ATT may be identifiable even when
  ATE is not, because it only requires $g > 0$, not $g < 1$)
- **Report the analysis as inconclusive** for the flagged scenario and
  describe what additional data or design features would be needed

### Decision Flow

```
Stage 2 overlap flag fires
         │
         ▼
   Run Stage 2b simulation
         │
    ┌────┴────┐
    │         │
   PASS     FAIL
    │         │
    ▼         ▼
Proceed    Consider:
with       • Trim to overlap region
TMLE-ML    • Switch estimand (ATT)
           • Flag as inconclusive
```

::: {.callout-note}
The simulation-based gate is *not* a test of the null hypothesis. It is an
operating-characteristics assessment: "Under a plausible DGP, does this
estimator produce valid inference in this overlap environment?" The DGP is
pre-specified and the evaluation criteria are pre-specified. The analyst
commits to the decision rule before seeing outcomes.
:::


## 9. Stop/Go Gate Results

```{r gate-display, results='asis'}
if (!is.null(gate_combined)) {
  kable(gate_combined, digits = 4, row.names = FALSE,
        caption = "Stop/go gate results by mode and overlap stratum")
} else {
  cat("*No gate results available. Run the simulation first.*\n")
}
```

```{r gate-interpretation, results='asis'}
if (!is.null(gate_combined) && !is.null(weak) && !is.null(strong)) {
  cat("\n**Interpretation:**\n\n")

  # Weak mode, bad overlap
  wb <- gate_combined[gate_combined$mode == "weak" &
                      gate_combined$overlap_flag == "bad", ]
  if (nrow(wb) > 0 && !is.na(wb$acceptable[1])) {
    if (wb$acceptable[1]) {
      cat("- **Weak W4, bad overlap: PASS (false stop confirmed).**",
          sprintf("TMLE-ML bias = %.4f, coverage = %.3f.",
                  wb$tmle_ml_bias[1],
                  if (!is.na(wb$tmle_ml_cov_np[1])) wb$tmle_ml_cov_np[1]
                  else wb$tmle_ml_cov_ic[1]),
          "The overlap flag was conservative; TMLE-ML inference remained valid.\n")
    } else {
      cov_val <- if (!is.na(wb$tmle_ml_cov_np[1])) wb$tmle_ml_cov_np[1]
                 else wb$tmle_ml_cov_ic[1]
      cat("- **Weak W4, bad overlap: FAIL.**",
          sprintf("TMLE-ML bias = %.4f, coverage = %.3f.",
                  wb$tmle_ml_bias[1], cov_val),
          "The gate did not pass; further tuning of the DGP or",
          "larger N may be needed to demonstrate a false stop.\n")
    }
  }

  # Strong mode, bad overlap
  sb <- gate_combined[gate_combined$mode == "strong" &
                      gate_combined$overlap_flag == "bad", ]
  if (nrow(sb) > 0 && !is.na(sb$acceptable[1])) {
    if (!sb$acceptable[1]) {
      cov_val <- if (!is.na(sb$tmle_ml_cov_np[1])) sb$tmle_ml_cov_np[1]
                 else sb$tmle_ml_cov_ic[1]
      cat("- **Strong W4, bad overlap: FAIL (true stop confirmed).**",
          sprintf("TMLE-ML bias = %.4f, coverage = %.3f.",
                  sb$tmle_ml_bias[1], cov_val),
          "The overlap violation genuinely degraded inference.\n")
    } else {
      cat("- **Strong W4, bad overlap: PASS.** The true-stop was not",
          "demonstrated; the DGP parameters may need adjustment.\n")
    }
  }
}
```


# Part V: Simulation Details

## 10. Figures

::: {.panel-tabset}

### Sampling Distributions

```{r fig-samp, fig.width=9, fig.height=14}
if (!is.null(weak)) {
  methods <- c("Crude", "Adj.Reg", "IPTW", "PS-Match",
               "TMLE-GLM", "TMLE-ML", "CV-TMLE")
  ests <- list(weak$results$crude_est, weak$results$regadj_est,
               weak$results$iptw_est, weak$results$psmatch_est,
               weak$results$tmle_glm_est, weak$results$tmle_ml_est,
               weak$results$cvtmle_est)
  est_long <- do.call(rbind, lapply(seq_along(methods), function(i) {
    data.frame(method = methods[i], estimate = ests[[i]],
               overlap_flag = weak$results$overlap_flag,
               stringsAsFactors = FALSE)
  }))
  est_long <- est_long[!is.na(est_long$estimate), ]
  est_long$method <- factor(est_long$method, levels = methods)

  ggplot(est_long, aes(x = estimate, fill = overlap_flag)) +
    geom_histogram(bins = 35, alpha = 0.6, position = "identity") +
    geom_vline(xintercept = weak$truth$RD_true, linewidth = 0.8) +
    facet_wrap(~ method, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = c(ok = "steelblue", bad = "firebrick")) +
    labs(title = "Sampling distributions (Weak W4 mode)",
         subtitle = paste0("True RD = ", round(weak$truth$RD_true, 4),
                           " | Vertical line = truth"),
         x = "Estimated RD", y = "Count", fill = "Overlap") +
    theme_minimal(base_size = 11)
}
```

### Coverage

```{r fig-cov, fig.width=10, fig.height=7}
if (!is.null(weak) && !is.null(strong)) {
  cov_all <- rbind(
    cbind(mode = "Weak W4", weak$metrics[weak$metrics$subset == "all" &
                                          !is.na(weak$metrics$coverage), ]),
    cbind(mode = "Strong W4", strong$metrics[strong$metrics$subset == "all" &
                                              !is.na(strong$metrics$coverage), ])
  )

  if (nrow(cov_all) > 0) {
    cov_all$ci_lo <- pmax(cov_all$coverage - 1.96 * cov_all$mcse_coverage, 0)
    cov_all$ci_hi <- pmin(cov_all$coverage + 1.96 * cov_all$mcse_coverage, 1)

    ggplot(cov_all, aes(x = method, y = coverage, fill = mode)) +
      geom_col(position = "dodge", alpha = 0.8, width = 0.7) +
      geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                    position = position_dodge(0.7), width = 0.2) +
      geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
      coord_flip() + ylim(0, 1) +
      scale_fill_manual(values = c("Weak W4" = "steelblue",
                                    "Strong W4" = "firebrick")) +
      labs(title = "95% CI Coverage by Inference Method and Mode",
           subtitle = "Error bars: ±1.96 × MCSE | Dashed line: nominal 95%",
           x = NULL, y = "Coverage", fill = "Mode") +
      theme_minimal(base_size = 11)
  }
}
```

### Overlap Flags

```{r fig-flags, fig.width=7, fig.height=4}
if (!is.null(weak) && !is.null(strong)) {
  flag_tab <- data.frame(
    Mode = c("Weak", "Weak", "Strong", "Strong"),
    Flag = factor(rep(c("ok", "bad"), 2), levels = c("ok", "bad")),
    N = c(sum(weak$results$overlap_flag == "ok"),
          sum(weak$results$overlap_flag == "bad"),
          sum(strong$results$overlap_flag == "ok"),
          sum(strong$results$overlap_flag == "bad"))
  )
  flag_tab$Pct <- round(100 * flag_tab$N / c(
    rep(nrow(weak$results), 2), rep(nrow(strong$results), 2)), 1)

  ggplot(flag_tab, aes(x = Mode, y = N, fill = Flag)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_text(aes(label = paste0(Pct, "%")),
              position = position_dodge(0.9), vjust = -0.3, size = 3.5) +
    scale_fill_manual(values = c(ok = "steelblue", bad = "firebrick")) +
    labs(title = "Overlap Flag Distribution by Mode",
         x = NULL, y = "Number of Replicates", fill = "Flag") +
    theme_minimal(base_size = 12)
}
```

:::


## 11. Technical Parameters

```{r tech-params, results='asis'}
if (!is.null(weak)) {
  p <- weak$params
  tech <- data.frame(
    Parameter = c(
      "N (per replicate)", "Replicates", "Preset",
      "Truth sample size",
      "NP bootstrap B", "Multiplier bootstrap B",
      "GLM bootstrap B (adj reg)", "Pairs bootstrap B (matching)",
      "CV-TMLE folds",
      "PS clipping (estimation)", "PS clipping (truth)",
      "g intercept shift", "s_ok", "p_bad",
      "Q SL library", "g SL library",
      "Overlap: extreme_prop threshold", "Overlap: ess_frac threshold",
      "Overlap: max_w threshold"
    ),
    Value = c(
      p$n_main, p$reps, p$preset,
      format(p$n_truth, big.mark = ","),
      p$B_npboot_tmle, p$B_multboot,
      p$B_regboot, p$B_matchboot,
      p$V_cf,
      paste(p$ps_clip_hat, collapse = ", "),
      paste(p$ps_clip_true, collapse = ", "),
      p$g_intercept_shift, p$s_ok, p$p_bad,
      paste(p$Q_SL_library_ml, collapse = ", "),
      paste(p$g_SL_library_ml, collapse = ", "),
      p$overlap_flag_thresholds$extreme_prop,
      p$overlap_flag_thresholds$ess_frac,
      p$overlap_flag_thresholds$max_w
    )
  )
  kable(tech, caption = "Simulation parameters (from params_used.rds)")
} else {
  cat("*Parameters will be displayed after running the simulation.*\n")
}
```


## 12. Limitations and Caveats

1. **This is a controlled simulation, not a real analysis.** The DGP is known
   and the true ATE is computable. In a real study, the analyst must pre-specify
   the DGP for Stage 2b simulations without seeing outcomes, which requires
   domain knowledge and sensitivity analysis across plausible DGPs.

2. **Coverage estimates have Monte Carlo uncertainty.** With $R$ replicates,
   the MCSE of a coverage estimate $\hat{c}$ is $\sqrt{\hat{c}(1-\hat{c})/R}$.
   For $R = 120$ and $\hat{c} = 0.95$, MCSE $\approx 0.020$.

3. **The SuperLearner library is deliberately small** (SL.mean, SL.glmnet,
   SL.xgboost) for computational feasibility. Richer libraries (GAMs, neural
   nets, deeper forests) may improve or worsen finite-sample performance.

4. **IC-based CIs can under-cover at moderate N.** The influence-curve variance
   estimator treats nuisance models as known. At $N = 400$, this approximation
   may be poor, especially under near-positivity violations where the clever
   covariate $H$ has extreme values. The NP bootstrap addresses this but is
   computationally expensive.

5. **Single DGP family.** Results are specific to this covariate structure
   and outcome model. Generalization requires running the simulation under
   alternative DGPs (different covariate distributions, different functional
   forms, time-to-event outcomes, etc.).

6. **The "clean room" property depends on execution discipline.** The staged
   design is only credible if the analyst genuinely does not peek at outcomes
   before Stage 3. In practice, this requires institutional safeguards
   (separate data extracts, audit logs, or a second analyst who controls
   outcome data access).

## 13. Reproducing the Simulation

To run the simulation from scratch:

```bash
# Fast preset (N=400, 120 reps, ~minutes per mode)
Rscript scripts/sim_scenario1_v4_false_stop_inference.R fast

# Evaluation preset (N=600, 300 reps, longer runtime)
Rscript scripts/sim_scenario1_v4_false_stop_inference.R eval
```

Outputs are saved to `results/scenario1_v4/{weak,strong}/`:

- `truth.rds` — true ATE and potential outcome means
- `replicate_results.rds` — per-replicate estimates, diagnostics, flags
- `summary_overall.csv` — bias, RMSE, coverage by estimator and stratum
- `gate_results.rds` — stop/go gate decision
- `params_used.rds` — exact parameters used for reproducibility
- `figures/` — diagnostic plots

The helper functions are in `R/sim_helpers_v4.R` and the driver script is
`scripts/sim_scenario1_v4_false_stop_inference.R`.

## References

- van der Laan MJ, Rose S. *Targeted Learning: Causal Inference for
  Observational and Experimental Data*. Springer, 2011.
- van der Laan MJ, Rose S. *Targeted Learning in Data Science*. Springer, 2018.
- Petersen ML, Porter KE, Gruber S, Wang Y, van der Laan MJ. Diagnosing and
  responding to violations in the positivity assumption.
  *Statistical Methods in Medical Research*, 2012.
- Gruber S, van der Laan MJ. tmle: An R package for targeted maximum
  likelihood estimation. *Journal of Statistical Software*, 2012.
