---
title: "Simulation Scenario 1 v3: Mixed Overlap, Nonlinear Q, CV-TMLE"
subtitle: "False stops and true stops under a clean-room feasibility gate"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 8
    fig-height: 5
  pdf:
    toc: true
    toc-depth: 3
    fig-width: 7
    fig-height: 4.5
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| code-fold: false

library(ggplot2)
theme_set(theme_minimal(base_size = 13))

base_dir <- "../results/scenario1_v3"

load_mode <- function(mode) {
  d <- file.path(base_dir, mode)
  rds <- c(params  = file.path(d, "params_used.rds"),
           truth   = file.path(d, "truth.rds"),
           results = file.path(d, "replicate_results.rds"),
           gate    = file.path(d, "gate_results.rds"))
  csv <- file.path(d, "summary_overall.csv")
  missing <- names(rds)[!file.exists(rds)]
  if (length(missing) > 0 || !file.exists(csv))
    stop("Outputs not found for mode '", mode, "' in ", d,
         "\nRun: Rscript scripts/sim_scenario1_v3_mixed_overlap_nonlinear_cvTMLE.R",
         call. = FALSE)
  list(params  = readRDS(rds["params"]),
       truth   = readRDS(rds["truth"]),
       res     = readRDS(rds["results"]),
       gate    = readRDS(rds["gate"]),
       summary = read.csv(csv, stringsAsFactors = FALSE))
}

weak   <- load_mode("weak")
strong <- load_mode("strong")

gate_file <- file.path(base_dir, "gate_combined.csv")
gate_combined <- if (file.exists(gate_file))
  read.csv(gate_file, stringsAsFactors = FALSE) else NULL
```

## Overview

This simulation extends Scenario 1 to a richer design that demonstrates both
**false stops** and **true stops** under a clean-room overlap feasibility gate.
The key innovations are:

1. **Mixed overlap regimes**: each replicate is randomly assigned OK or BAD
   overlap by varying the treatment-model strength multiplier, producing a
   non-degenerate stop/go table.
2. **Nonlinear outcome model**: $Q_0(A,W)$ includes quadratic and absolute-
   value main effects and treatment-covariate interactions, so that
   intentionally misspecified GLM estimators (crude, adjusted regression,
   TMLE-GLM) exhibit nontrivial bias.
3. **Two W4 outcome-dependence modes**: "weak" ($\beta_4 \approx 0$, no
   $A \times W_4$) and "strong" ($\beta_4 = 0.35$, $A \times W_4$ present),
   creating the false-stop vs true-stop contrast.
4. **Seven estimators**: crude, misspecified adjusted regression, IPTW,
   PS-matched regression, TMLE-GLM, TMLE-ML, and cross-fitted TMLE-ML.
5. **Inference**: IC-based Wald CIs for TMLE variants plus Rademacher
   multiplier bootstrap CIs; nonparametric bootstrap for regression and
   IPTW; pairs bootstrap for PS-matching.

### Point estimation vs inference

This report explicitly separates two properties:

- **Point estimation stability**: whether an estimator's sampling distribution
  is centered near the truth (low bias, low RMSE).
- **Inferential validity**: whether 95% confidence intervals achieve nominal
  coverage, as measured by the proportion of replicates where the CI contains
  the true RD. Coverage is computed only over replicates where both the
  estimate and CI were successfully produced (`n_ci` is reported alongside
  `n_success`).

## Estimand

$$\text{RD} = E[Y^1] - E[Y^0]$$

The estimand depends only on the outcome model $Q_0$ and the marginal
covariate distribution, not on the treatment mechanism $g_0$. Therefore
the truth is the same regardless of overlap regime.

## Data-Generating Process

### Scenario grid

| Dimension | Levels | Purpose |
|-----------|--------|---------|
| Overlap regime $Z_r$ | OK ($s_g = `r weak$params$s_ok`$) vs BAD ($s_g = `r weak$params$s_bad`$) | Controls PS extremity per replicate |
| W4 outcome dependence | Weak ($\beta_4 = `r weak$params$beta4_Q`$) vs Strong ($\beta_4 = `r strong$params$beta4_Q`$) | Controls whether overlap driver matters for Y |

### Covariates

| Variable | Distribution | Role |
|----------|-------------|------|
| $W_1$ | $N(0, 1)$ | Confounder with nonlinear effect on Y ($0.6 W_1 + 0.4 W_1^2$) |
| $W_2$ | Bernoulli(0.5) | Binary confounder |
| $W_3$ | $N(0, 1)$ | Confounder with non-smooth effect on Y ($0.5 |W_3|$) |
| $W_4$ | $N(0, 1)$ | Overlap driver: $\alpha_4 = `r weak$params$alpha4_g`$ in g; $\beta_4$ mode-dependent in Q |

### Treatment mechanism

$$g_0(W) = \text{expit}\!\bigl(\alpha_0 + s_g \cdot (
  `r weak$params$alpha1_g` W_1 + `r weak$params$alpha2_g` W_2
  `r weak$params$alpha3_g` W_3 + `r weak$params$alpha4_g` W_4)\bigr)$$

where $s_g \in \{`r weak$params$s_ok`, `r weak$params$s_bad`\}$ and
$\alpha_0$ centers $P(A=1) \approx 0.5$.

### Outcome model (nonlinear + interactions)

$$Q_0(A,W) = \text{expit}\!\bigl(
  \beta_0 + \beta_A A +
  \underbrace{0.6 W_1 + 0.4 W_1^2}_{h_1} +
  \underbrace{0.7 W_2}_{h_2} +
  \underbrace{0.5 |W_3|}_{h_3} +
  \beta_4 W_4 +
  `r weak$params$tau_int1` \cdot A \cdot \mathbb{1}(W_1 > 0) +
  `r weak$params$tau_int2` \cdot A \cdot W_2 +
  \tau_3 \cdot A \cdot W_4
\bigr)$$

The misspecified GLM (Y ~ A + W1 + W2 + W3 + W4) cannot represent the
$W_1^2$, $|W_3|$, or interaction terms.

### Ground truth

```{r truth}
#| label: truth
#| code-fold: false

cat(sprintf("Weak W4 mode:   RD_true = %.6f  (EY1=%.4f, EY0=%.4f)\n",
            weak$truth$RD_true, weak$truth$EY1, weak$truth$EY0))
cat(sprintf("Strong W4 mode: RD_true = %.6f  (EY1=%.4f, EY0=%.4f)\n",
            strong$truth$RD_true, strong$truth$EY1, strong$truth$EY0))
```

## Simulation Design

```{r design}
#| label: design

p <- weak$params
cat(sprintf("Replicates per mode:   %d\n", p$reps))
cat(sprintf("Sample size (N):       %d\n", p$n_main))
cat(sprintf("P(bad overlap):        %.2f\n", p$p_bad))
cat(sprintf("PS clip (true):        [%.3f, %.3f]\n", p$ps_clip_true[1], p$ps_clip_true[2]))
cat(sprintf("PS clip (estimated):   [%.2f, %.2f]\n", p$ps_clip_hat[1], p$ps_clip_hat[2]))
cat(sprintf("V-fold CV-TMLE:        %d folds\n", p$V_cf))
cat(sprintf("Multiplier boot B:     %d\n", p$B_multboot))
cat(sprintf("Regression boot B:     %d\n", p$B_regboot))
cat(sprintf("IPTW boot B:           %d\n", p$B_iptw))
cat(sprintf("Pairs boot B:          %d\n", p$B_matchboot))
```

### Estimators

| Estimator | Q model | g model | Inference |
|-----------|---------|---------|-----------|
| **Crude RD** | — | — | Wald CI (binomial SE) |
| **Adj. Regression** | Misspecified GLM | — | Nonparametric bootstrap (B=`r p$B_regboot`) |
| **IPTW** | — | GLM, bounded | Nonparametric bootstrap (B=`r p$B_iptw`) |
| **PS-Matched** | Misspecified GLM on matched data | GLM (logit PS + caliper) | Pairs bootstrap (B=`r p$B_matchboot`) |
| **TMLE-GLM** | Misspecified GLM via SL | GLM via SL | IC Wald + Rademacher multiplier boot |
| **TMLE-ML** | Rich SL library | Rich SL library | IC Wald + Rademacher multiplier boot |
| **CV-TMLE-ML** | V-fold cross-fitted ML | V-fold cross-fitted ML | IC Wald + Rademacher multiplier boot |

**Adjusted regression** fits Y ~ A + W1 + W2 + W3 + W4 (no quadratic terms,
no $|W_3|$, no interactions) and computes the ATE by g-computation. This
intentional misspecification ensures nontrivial bias for comparison. The CI
is computed via nonparametric bootstrap to avoid misleadingly precise
delta-method SEs.

**PS-matched regression** performs nearest-neighbor matching on logit($\hat{g}$)
with caliper $0.2 \times \text{sd}(\text{logit}(\hat{g}))$. Under bad overlap,
matching often discards many observations. Match diagnostics (SMD before/after,
number matched) are recorded. Inference uses a pairs bootstrap that resamples
matched pairs with replacement.

**Cross-fitted TMLE-ML** implements V-fold cross-fitting: Q and g models are
fit on training folds and predicted on validation folds; cross-fitted
predictions are combined for a single targeting step. This avoids
Donsker-class conditions required by standard TMLE and is appropriate
when using highly adaptive learners.

**Multiplier bootstrap** (Rademacher): for each TMLE variant, draws
$\xi_i \stackrel{iid}{\sim} \text{Rademacher}(\pm 1)$ and computes
$\hat\psi^{(b)} = \hat\psi + n^{-1} \sum_i \xi_i \hat{D}_i$
where $\hat{D}_i$ are the estimated EIF values. The resulting SE and
percentile CIs provide a complementary check on variance estimation.
Nuisance fits are held fixed across bootstrap draws.

## Stage 2: Outcome-Blind Overlap Diagnostics

```{r overlap-summary}
#| label: overlap-summary

for (mode_name in c("weak", "strong")) {
  m <- get(mode_name)
  cat(sprintf("\n=== %s W4 mode ===\n", toupper(mode_name)))
  cat(sprintf("  Replicates flagged 'bad': %d/%d (%.0f%%)\n",
              sum(m$res$overlap_flag == "bad"), nrow(m$res),
              100 * mean(m$res$overlap_flag == "bad")))
  cat(sprintf("  Mean fraction extreme PS: %.3f\n",
              mean(m$res$fraction_extreme)))
  cat(sprintf("  Mean ESS/N:              %.3f\n",
              mean(m$res$ess_frac)))
  cat(sprintf("  Mean max weight:         %.1f\n",
              mean(m$res$max_w)))
}
```

```{r ps-hist}
#| label: ps-hist
#| fig-height: 5
#| fig-cap: "Distribution of overlap severity (fraction extreme PS) across replicates, stratified by overlap regime actually drawn. Dashed line shows the flagging threshold."

thr <- weak$params$overlap_flag_thresholds
comb <- rbind(
  data.frame(mode = "Weak W4", fraction_extreme = weak$res$fraction_extreme,
             overlap_flag = weak$res$overlap_flag, stringsAsFactors = FALSE),
  data.frame(mode = "Strong W4", fraction_extreme = strong$res$fraction_extreme,
             overlap_flag = strong$res$overlap_flag, stringsAsFactors = FALSE)
)
ggplot(comb, aes(x = fraction_extreme, fill = overlap_flag)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = thr$extreme_prop, linetype = "dashed", color = "red") +
  facet_wrap(~ mode) +
  scale_fill_manual(values = c(ok = "steelblue", bad = "firebrick")) +
  labs(title = "Overlap severity across replicates",
       x = "Fraction with extreme PS", y = "Count", fill = "Flag")
```

## Stage 2b: Stop/Go Gate

The gate criterion for TMLE-ML with multiplier bootstrap CI is:
$|\text{bias}| < 0.01$ AND coverage $\geq 0.93$ AND RMSE acceptable.

```{r gate-table}
#| label: gate-table
#| code-fold: false

if (!is.null(gate_combined)) {
  knitr::kable(gate_combined, digits = 4,
               caption = "Stop/go gate by W4 mode and overlap flag. 'Acceptable' = |bias| < 0.01 and multiplier bootstrap coverage >= 0.93.")
} else {
  cat("Gate combined file not found.\n")
}
```

```{r gate-interpret}
#| label: gate-interpret

cat("Interpretation:\n\n")

# Weak mode, bad overlap
wb <- weak$gate[weak$gate$overlap_flag == "bad", ]
if (!is.na(wb$acceptable) && wb$acceptable) {
  cat("WEAK W4, BAD overlap: TMLE-ML meets OC criteria despite overlap flag.\n")
  cat("This is a FALSE STOP scenario — the PS-only gate would unnecessarily\n")
  cat("terminate the analysis when TMLE can succeed because the overlap driver\n")
  cat("(W4) barely affects the outcome.\n\n")
} else if (!is.na(wb$acceptable)) {
  cat("WEAK W4, BAD overlap: TMLE-ML does NOT meet OC criteria.\n\n")
}

# Strong mode, bad overlap
sb <- strong$gate[strong$gate$overlap_flag == "bad", ]
if (!is.na(sb$acceptable) && !sb$acceptable) {
  cat("STRONG W4, BAD overlap: TMLE-ML fails OC criteria under bad overlap.\n")
  cat("This is a TRUE STOP scenario — the overlap driver matters for the outcome,\n")
  cat("so near-positivity genuinely degrades both point estimation and inference.\n")
} else if (!is.na(sb$acceptable)) {
  cat("STRONG W4, BAD overlap: TMLE-ML meets OC criteria (possibly with larger N\n")
  cat("or richer learners). The true-stop effect may require more severe parameters.\n")
}
```

## Stage 3: Estimator Performance

### Weak W4 Mode (False-Stop Demonstration)

```{r weak-table}
#| label: weak-table
#| code-fold: false

knitr::kable(weak$summary, digits = 4,
             caption = "Weak W4: estimator performance by subset (all / overlap ok / overlap bad). Coverage computed over n_ci replicates with non-NA CIs.")
```

### Strong W4 Mode (True-Stop Demonstration)

```{r strong-table}
#| label: strong-table
#| code-fold: false

knitr::kable(strong$summary, digits = 4,
             caption = "Strong W4: estimator performance by subset. Under bad overlap, estimators relying on the overlap-driver W4 show degraded performance.")
```

### Sampling Distributions

```{r sampling-dist-weak}
#| label: sampling-dist-weak
#| fig-height: 10
#| fig-cap: "Weak W4 mode: sampling distributions of RD estimators. Colored by overlap flag. Misspecified estimators (crude, adj. regression, TMLE-GLM) show bias from model misspecification. TMLE-ML and CV-TMLE are better centered."

plot_sampling <- function(res, truth_rd, mode_label) {
  methods <- c("Crude", "Adj.Reg", "IPTW", "PS-Match",
               "TMLE-GLM", "TMLE-ML", "CV-TMLE")
  ests <- list(res$crude_est, res$regadj_est, res$iptw_est,
               res$psmatch_est, res$tmle_glm_est, res$tmle_ml_est,
               res$cvtmle_est)
  el <- do.call(rbind, lapply(seq_along(methods), function(i) {
    data.frame(method = methods[i], estimate = ests[[i]],
               overlap_flag = res$overlap_flag, stringsAsFactors = FALSE)
  }))
  el <- el[!is.na(el$estimate), ]
  el$method <- factor(el$method, levels = methods)

  ggplot(el, aes(x = estimate, fill = overlap_flag)) +
    geom_histogram(bins = 35, alpha = 0.6, position = "identity") +
    geom_vline(xintercept = truth_rd, linewidth = 0.8) +
    facet_wrap(~ method, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = c(ok = "steelblue", bad = "firebrick")) +
    labs(title = paste0("Sampling distributions (", mode_label, ")"),
         subtitle = paste0("True RD = ", round(truth_rd, 4)),
         x = "Estimated RD", y = "Count", fill = "Overlap") +
    theme_minimal(base_size = 11)
}

plot_sampling(weak$res, weak$truth$RD_true, "weak W4")
```

```{r sampling-dist-strong}
#| label: sampling-dist-strong
#| fig-height: 10
#| fig-cap: "Strong W4 mode: when W4 drives both treatment and outcome, bad overlap degrades all estimators. TMLE-ML may still improve over GLM-based methods but with wider spread."

plot_sampling(strong$res, strong$truth$RD_true, "strong W4")
```

### Bias Comparison Across Modes

```{r bias-comparison}
#| label: bias-comparison
#| fig-height: 6
#| fig-cap: "Bias by estimator, overlap flag, and W4 mode. Under weak W4, TMLE-ML and CV-TMLE maintain low bias even under bad overlap (false stop). Under strong W4, bad overlap degrades all estimators (true stop)."

make_bias_df <- function(summ, mode_label) {
  d <- summ[summ$subset %in% c("overlap_ok", "overlap_bad") &
            !grepl("mult\\.boot", summ$method), ]
  d$mode <- mode_label
  d
}
bd <- rbind(make_bias_df(weak$summary, "Weak W4"),
            make_bias_df(strong$summary, "Strong W4"))
if (nrow(bd) > 0) {
  bd$method <- factor(bd$method)
  ggplot(bd, aes(x = method, y = bias, fill = subset)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    coord_flip() +
    facet_wrap(~ mode) +
    scale_fill_manual(values = c(overlap_ok = "steelblue",
                                 overlap_bad = "firebrick")) +
    labs(title = "Bias by estimator and overlap regime",
         x = NULL, y = "Bias", fill = "Overlap") +
    theme_minimal(base_size = 11)
}
```

### Coverage: IC-Based vs Multiplier Bootstrap

```{r coverage-comparison}
#| label: coverage-comparison
#| fig-height: 6
#| fig-cap: "Coverage of 95% CIs (all replicates). IC-based Wald CIs and Rademacher multiplier bootstrap CIs shown side by side for each TMLE variant. Coverage computed only over replicates where CI was successfully produced (n_ci)."

make_cov_df <- function(summ, mode_label) {
  d <- summ[summ$subset == "all" & !is.na(summ$coverage), ]
  d$mode <- mode_label
  d
}
cd <- rbind(make_cov_df(weak$summary, "Weak W4"),
            make_cov_df(strong$summary, "Strong W4"))
if (nrow(cd) > 0) {
  cd$method <- factor(cd$method, levels = rev(unique(cd$method)))
  cd$ci_lo <- pmax(cd$coverage - 1.96 * cd$mcse_coverage, 0)
  cd$ci_hi <- pmin(cd$coverage + 1.96 * cd$mcse_coverage, 1)
  ggplot(cd, aes(x = method, y = coverage)) +
    geom_col(aes(fill = mode), position = "dodge", alpha = 0.8) +
    geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
    coord_flip() + ylim(0, 1) +
    scale_fill_manual(values = c("Weak W4" = "steelblue",
                                 "Strong W4" = "firebrick")) +
    labs(title = "95% CI coverage by estimator and W4 mode",
         x = NULL, y = "Coverage", fill = "Mode") +
    theme_minimal(base_size = 11)
}
```

### SE Calibration: TMLE-ML

```{r se-calibration}
#| label: se-calibration

for (mode_name in c("weak", "strong")) {
  m <- get(mode_name)
  sm <- m$summary[m$summary$subset == "all", ]
  ic_row <- sm[sm$method == "TMLE-ML (IC)", ]
  mb_row <- sm[sm$method == "TMLE-ML (mult.boot)", ]

  cat(sprintf("\n=== %s W4 mode: TMLE-ML SE calibration ===\n", toupper(mode_name)))
  if (nrow(ic_row) > 0 && !is.na(ic_row$emp_sd) && !is.na(ic_row$mean_se)) {
    cat(sprintf("  Empirical SD:     %.4f\n", ic_row$emp_sd))
    cat(sprintf("  Mean IC SE:       %.4f  (ratio: %.3f)\n",
                ic_row$mean_se, ic_row$mean_se / ic_row$emp_sd))
  }
  if (nrow(mb_row) > 0 && !is.na(mb_row$mean_se) && !is.na(ic_row$emp_sd)) {
    cat(sprintf("  Mean mult.boot SE: %.4f  (ratio: %.3f)\n",
                mb_row$mean_se, mb_row$mean_se / ic_row$emp_sd))
  }
  if (nrow(ic_row) > 0 && !is.na(ic_row$coverage)) {
    cat(sprintf("  Coverage (IC):     %.1f%%\n", ic_row$coverage * 100))
  }
  if (nrow(mb_row) > 0 && !is.na(mb_row$coverage)) {
    cat(sprintf("  Coverage (m.boot): %.1f%%\n", mb_row$coverage * 100))
  }
}
```

### PS Matching Diagnostics

```{r match-diag}
#| label: match-diag

for (mode_name in c("weak", "strong")) {
  m <- get(mode_name)
  r <- m$res
  matched <- r[!is.na(r$psmatch_n_matched), ]
  cat(sprintf("\n=== %s W4: PS matching ===\n", toupper(mode_name)))
  cat(sprintf("  Successful matches: %d/%d\n", nrow(matched), nrow(r)))
  if (nrow(matched) > 0) {
    cat(sprintf("  Median matched N:  %d / %d (%.0f%%)\n",
                median(matched$psmatch_n_matched),
                m$params$n_main,
                100 * median(matched$psmatch_n_matched) / m$params$n_main))
    # By overlap flag
    for (fl in c("ok", "bad")) {
      sub <- matched[matched$overlap_flag == fl, ]
      if (nrow(sub) > 0)
        cat(sprintf("    [%s] median matched N: %d\n", fl,
                    median(sub$psmatch_n_matched)))
    }
  }
}
```

## Discussion

### False stop (weak W4 mode)

When $W_4$ has negligible influence on the outcome ($\beta_4 = `r weak$params$beta4_Q`$,
no $A \times W_4$ interaction), the overlap flag triggered by extreme propensity
scores is a **false alarm** for doubly robust estimators. TMLE-ML and CV-TMLE-ML
maintain low bias and adequate coverage even in bad-overlap replicates because
the outcome surface is smooth in the direction of PS extremity. The multiplier
bootstrap provides a more reliable variance estimate than IC-based SEs when the
clever covariate is occasionally large.

Under a propensity-score-only feasibility gate, these replicates would have
triggered redesign or termination. The outcome-blind simulation-based OC check
(Stage 2b) demonstrates that proceeding is defensible: TMLE's operating
characteristics remain acceptable because the $Q$ model is learnable.

### True stop (strong W4 mode)

When $W_4$ meaningfully affects the outcome ($\beta_4 = `r strong$params$beta4_Q`$)
and interacts with treatment ($\tau_3 = `r strong$params$tau_int3`$), the overlap
flag is a **genuine warning**. Now the outcome surface is no longer flat in the
direction of PS extremity: subjects with extreme propensity scores also have
different treatment effects, and the $Q$ model must extrapolate into a region
where data is sparse. Even TMLE-ML and CV-TMLE show increased bias and
degraded coverage under bad overlap.

### Misspecification contrast

The nonlinear DGP produces a clear ordering:

- **Crude RD**: heavily biased (confounding).
- **Adjusted regression** and **TMLE-GLM**: biased due to model misspecification
  ($W_1^2$, $|W_3|$, and interactions absent from the linear predictor).
- **TMLE-ML** and **CV-TMLE-ML**: reduced bias because flexible learners
  (ranger or enriched GLM) can capture the nonlinearities.
- **IPTW** and **PS-matching**: performance depends on overlap quality, not
  outcome model quality. Under bad overlap, both degrade regardless of mode.

### Inference methods

For TMLE variants, we report both IC-based Wald CIs and Rademacher multiplier
bootstrap CIs. The multiplier bootstrap uses
$\hat\psi^{(b)} = \hat\psi + n^{-1} \sum_i \xi_i \hat{D}_i$
with $\xi_i \in \{-1, +1\}$ and nuisance fits held fixed. This is
computationally light and avoids the Donsker-class conditions needed for
IC-based inference to be valid with data-adaptive learners.

For regression-only and PS-matching estimators, nonparametric and pairs
bootstraps provide CIs that account for the full estimation uncertainty
including model fitting.

### Limitations

Because the DGP is stylized and the scenario grid is constructed to
demonstrate specific principles (false stop vs true stop), these results
illustrate mechanisms rather than provide universal guarantees. In
particular:

- The 2×2 grid (overlap regime × W4 dependence) captures two extremes;
  real data may fall on a continuum.
- With $N = `r weak$params$n_main`$ and $`r weak$params$reps`$ replicates,
  Monte Carlo standard errors for coverage are nontrivial; scale-up is
  needed for precise coverage estimation.
- The cross-fitted TMLE implementation uses ranger (if available) or enriched
  GLM as fallback; with richer learner libraries, performance may improve
  further.

## Appendix: Replicate-Level Sample

```{r replicate-table}
#| label: replicate-table

cols <- c("rep", "overlap_regime_true", "overlap_flag", "event_rate",
          "fraction_extreme", "ess_frac",
          "tmle_ml_est", "cvtmle_est", "regadj_est", "crude_est")
cols <- intersect(cols, names(weak$res))
knitr::kable(head(weak$res[, cols], 10), digits = 4,
             caption = "Weak W4 mode: first 10 replicates")
```

### Session Info

```{r session-info}
#| label: session-info

si_file <- file.path(base_dir, "weak", "sessionInfo.txt")
if (file.exists(si_file)) {
  cat(readLines(si_file), sep = "\n")
} else {
  sessionInfo()
}
```
