---
title: "Simulation Scenario 1: Bad Overlap, TMLE OK"
subtitle: "Double robustness under near-positivity violations"
author: "Andrew N. Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| code-fold: false

library(ggplot2)
theme_set(theme_minimal(base_size = 13))

# Load precomputed simulation outputs
out_dir <- "../outputs/sim_scenario1"
rds_files <- c(
  params  = file.path(out_dir, "params_used.rds"),
  truth   = file.path(out_dir, "truth.rds"),
  results = file.path(out_dir, "replicate_results.rds")
)
csv_file <- file.path(out_dir, "summary_metrics.csv")

missing <- names(rds_files)[!file.exists(rds_files)]
if (length(missing) > 0 || !file.exists(csv_file)) {
  stop("Simulation outputs not found in ", out_dir,
       "\n\nRun the simulation first:\n",
       "  Rscript scripts/sim_scenario1_bad_overlap_tmle_ok.R\n",
       call. = FALSE)
}

params  <- readRDS(rds_files["params"])
truth   <- readRDS(rds_files["truth"])
res     <- readRDS(rds_files["results"])
summary <- read.csv(csv_file, stringsAsFactors = FALSE)
```

## Motivation

Propensity score--based estimators require *positivity*: every subject must
have a nonzero probability of receiving each treatment level. When this
assumption is nearly violated --- many propensity scores crowd near 0 or 1 ---
inverse probability of treatment weighted (IPTW) estimators become unstable
because a few observations receive extreme weights that dominate the estimate.
Unadjusted comparisons are biased by confounding regardless of overlap quality.

Targeted minimum loss-based estimation (TMLE), however, is *doubly robust*:
it is consistent if either the outcome model $Q(A,W)$ or the treatment model
$g(W)$ is correctly specified. When $Q$ is well-learned, TMLE can produce
valid inference even when the propensity score model struggles. This simulation
demonstrates that property.

## Estimand

The target estimand is the average treatment effect on the risk difference
scale:

$$\text{RD} = E[Y^1] - E[Y^0]$$

where $Y^a$ is the counterfactual binary outcome under intervention $A = a$.

## Data-Generating Process

### Covariates

Four baseline covariates are generated independently:

| Variable | Distribution | Role |
|----------|-------------|------|
| $W_1$ | $N(0, 1)$ | Shared confounder (moderate in both $g$ and $Q$) |
| $W_2$ | Bernoulli(0.5) | Binary confounder |
| $W_3$ | $N(0, 1)$ | Continuous confounder |
| $W_4$ | $N(0, 1)$ | **Overlap driver**: strong in $g$, weak in $Q$ |

### Treatment mechanism (poor overlap by design)

The propensity score is generated as:

$$g(W) = \text{expit}\!\bigl(\alpha_g + s \cdot (
  `r params$coef_W1_g` \cdot W_1 +
  `r params$coef_W2_g` \cdot W_2 +
  (`r params$coef_W3_g`) \cdot W_3 +
  `r params$coef_W4_g` \cdot W_4
)\bigr)$$

where $s = `r params$strength_overlap`$ is the overlap strength multiplier and
$\alpha_g$ is chosen so that $P(A=1) \approx 0.5$. True propensity scores are
clipped to $[`r params$eps_clip`,\; `r 1 - params$eps_clip`]$.

The key structural feature: **$W_4$ has coefficient `r params$coef_W4_g` in the
treatment model but only `r params$coef_W4_Q` in the outcome model.** This means
the PS is extreme primarily in regions of the covariate space where the outcome
is not changing much, so $Q(A,W)$ remains smooth and learnable even where $g$ is
near 0 or 1.

### Outcome model (learnable)

$$Q(A,W) = \text{expit}\!\bigl(
  `r params$alpha_Q` +
  `r params$effect_A` \cdot A +
  `r params$coef_W1_Q` \cdot W_1 +
  `r params$coef_W2_Q` \cdot W_2 +
  (`r params$coef_W3_Q`) \cdot W_3 +
  `r params$coef_W4_Q` \cdot W_4
  `r if (params$nonlinear_Q) paste0("+ 0.3 \\cdot W_1^2") else ""`
\bigr)$$

The outcome model includes all four covariates with a correctly specified
GLM as one of the learners in the SuperLearner library, ensuring $Q$ is
well-estimated.

### Ground truth

```{r truth-display}
#| label: truth-display
#| code-fold: false

cat(sprintf("True RD:    %.6f\n", truth$RD_true))
cat(sprintf("E[Y^1]:     %.4f\n", truth$EY1))
cat(sprintf("E[Y^0]:     %.4f\n", truth$EY0))
cat(sprintf("Truth N:    %s\n", format(truth$truth_N, big.mark = ",")))
cat(sprintf("Method:     %s\n", truth$note))
```

## Simulation Design

```{r sim-design}
#| label: sim-design

cat(sprintf("Replicates:          %d\n", params$reps))
cat(sprintf("Sample size (N):     %d\n", params$N))
cat(sprintf("Overlap strength:    %.1f\n", params$strength_overlap))
cat(sprintf("PS clipping:         [%.3f, %.3f]\n", params$eps_clip, 1 - params$eps_clip))
cat(sprintf("g truncation:        [%.2f, %.2f]\n", params$g_trunc[1], params$g_trunc[2]))
cat(sprintf("IPTW bootstrap B:    %d\n", params$iptw_boot_B))
cat(sprintf("Master seed:         %d\n", params$master_seed))
```

### Estimators

| Estimator | Description | Inference |
|-----------|-------------|-----------|
| **TMLE** (primary) | Doubly robust; outcome model via SuperLearner (`r paste(params$Q_SL_library, collapse=", ")`), PS via SuperLearner (`r paste(params$g_SL_library, collapse=", ")`) | IC-based Wald CI |
| **Crude RD** | Unadjusted difference in sample proportions | Wald CI (binomial SE) |
| **IPTW RD** | Hajek estimator with truncated inverse probability weights | Percentile bootstrap CI (B=`r params$iptw_boot_B`) |

## Stage 2: Outcome-Blind Overlap Diagnostics

Before examining any treatment--outcome associations, we assess propensity
score overlap using only treatment assignment and covariates. This mirrors the
clean-room protocol where design adequacy is evaluated before unblinding
outcomes.

```{r overlap-diagnostics}
#| label: overlap-diagnostics

cat(sprintf("Fraction of replicates flagged 'bad' overlap: %.1f%%\n",
            mean(res$overlap_flag == "bad") * 100))
cat(sprintf("\nAcross %d replicates:\n", nrow(res)))
cat(sprintf("  Mean fraction g_hat < 0.05 or > 0.95:  %.3f\n",
            mean(res$fraction_extreme)))
cat(sprintf("  Mean ESS / N:                           %.3f\n",
            mean(res$ess_frac)))
cat(sprintf("  Mean max IPTW weight:                   %.1f\n",
            mean(res$max_w)))
cat(sprintf("  Median max IPTW weight:                 %.1f\n",
            median(res$max_w)))
cat(sprintf("  Mean min g_hat:                         %.4f\n",
            mean(res$min_g)))
cat(sprintf("  Mean max g_hat:                         %.4f\n",
            mean(res$max_g)))
```

### Distribution of Propensity Score Extremity

```{r ps-extreme-hist}
#| label: ps-extreme-hist
#| fig-height: 4

ggplot(res, aes(x = fraction_extreme)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = params$diag_extreme_threshold,
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = params$diag_extreme_threshold + 0.02,
           y = Inf, vjust = 2, hjust = 0, color = "red", size = 3.5,
           label = paste0("threshold = ", params$diag_extreme_threshold)) +
  labs(title = "Fraction of observations with extreme PS per replicate",
       x = "Fraction with g_hat < 0.05 or g_hat > 0.95",
       y = "Count")
```

### Effective Sample Size Under IPTW

```{r ess-hist}
#| label: ess-hist
#| fig-height: 4

ggplot(res, aes(x = ess_frac)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  geom_vline(xintercept = params$diag_ess_threshold,
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = params$diag_ess_threshold + 0.01,
           y = Inf, vjust = 2, hjust = 0, color = "red", size = 3.5,
           label = paste0("threshold = ", params$diag_ess_threshold)) +
  labs(title = "Effective sample size fraction (ESS / N) per replicate",
       x = "ESS / N", y = "Count")
```

### Maximum IPTW Weight

```{r max-weight-hist}
#| label: max-weight-hist
#| fig-height: 4

ggplot(res, aes(x = max_w)) +
  geom_histogram(bins = 30, fill = "firebrick", alpha = 0.7) +
  labs(title = "Maximum IPTW weight per replicate",
       x = "Max weight", y = "Count")
```

The diagnostics confirm that this DGP produces a *practical near-positivity
violation* in most replicates: many estimated propensity scores fall near the
boundaries, the effective sample size is substantially reduced by extreme
weights, and maximum weights are large. In a real analysis, Stage 2 would flag
these issues before any outcome data are examined.

## Stage 3: Estimator Performance

### Summary Table

```{r summary-table}
#| label: summary-table
#| code-fold: false

knitr::kable(summary, digits = 4,
             caption = "Simulation performance metrics by estimator and subset")
```

### Interpretation of Metrics

```{r metrics-interpret}
#| label: metrics-interpret

sm <- summary[summary$subset == "all", ]
tmle_row  <- sm[sm$method == "TMLE", ]
crude_row <- sm[sm$method == "Crude", ]
iptw_row  <- sm[sm$method == "IPTW", ]

cat("=== TMLE (primary) ===\n")
cat(sprintf("  Bias:      %+.4f\n", tmle_row$bias))
cat(sprintf("  RMSE:      %.4f\n", tmle_row$rmse))
cat(sprintf("  Emp SD:    %.4f\n", tmle_row$emp_sd))
cat(sprintf("  Mean SE:   %.4f\n", tmle_row$mean_se))
cat(sprintf("  Coverage:  %.1f%%\n", tmle_row$coverage * 100))

cat("\n=== Crude RD ===\n")
cat(sprintf("  Bias:      %+.4f\n", crude_row$bias))
cat(sprintf("  RMSE:      %.4f\n", crude_row$rmse))
cat(sprintf("  Coverage:  %.1f%%\n", crude_row$coverage * 100))

cat("\n=== IPTW ===\n")
cat(sprintf("  Bias:      %+.4f\n", iptw_row$bias))
cat(sprintf("  RMSE:      %.4f\n", iptw_row$rmse))
cat(sprintf("  Emp SD:    %.4f\n", iptw_row$emp_sd))
cat(sprintf("  Mean SE:   %.4f\n", iptw_row$mean_se))
cat(sprintf("  Coverage:  %.1f%%\n", iptw_row$coverage * 100))
```

### Sampling Distribution of Estimators

```{r sampling-dist}
#| label: sampling-dist
#| fig-height: 6

est_long <- data.frame(
  method = rep(c("TMLE", "Crude", "IPTW"), each = nrow(res)),
  estimate = c(res$tmle_est, res$crude_est, res$iptw_est),
  stringsAsFactors = FALSE
)
est_long <- est_long[!is.na(est_long$estimate), ]
est_long$method <- factor(est_long$method, levels = c("TMLE", "IPTW", "Crude"))

ggplot(est_long, aes(x = estimate, fill = method)) +
  geom_histogram(bins = 40, alpha = 0.7) +
  geom_vline(xintercept = truth$RD_true, linetype = "solid",
             color = "black", linewidth = 1) +
  annotate("text", x = truth$RD_true, y = Inf, vjust = 2, hjust = -0.1,
           label = paste0("True RD = ", round(truth$RD_true, 4)),
           fontface = "bold", size = 3.5) +
  facet_wrap(~ method, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(TMLE = "steelblue", IPTW = "darkorange",
                                Crude = "firebrick")) +
  labs(title = "Sampling distributions of RD estimators",
       x = "Estimated RD", y = "Count") +
  guides(fill = "none")
```

The sampling distributions reveal the core finding:

- **Crude RD** is centered far from the true RD (`r round(truth$RD_true, 4)`),
  with the entire distribution shifted by confounding bias. Its narrow width
  reflects the precision of the wrong answer.
- **IPTW** is approximately centered on the truth but has a wide, heavy-tailed
  distribution. The extreme weights amplify variability, and occasional
  replicates produce outlying estimates.
- **TMLE** is tightly centered on the truth. Despite the poor overlap, the
  well-learned outcome model allows the targeting step to correct the plug-in
  estimate without relying heavily on the unstable PS-based weights.

### Bias Comparison

```{r bias-plot}
#| label: bias-plot
#| fig-height: 4

bias_df <- data.frame(
  method = c("TMLE", "IPTW", "Crude"),
  bias = c(tmle_row$bias, iptw_row$bias, crude_row$bias)
)
bias_df$method <- factor(bias_df$method, levels = c("TMLE", "IPTW", "Crude"))

ggplot(bias_df, aes(x = method, y = bias, fill = method)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c(TMLE = "steelblue", IPTW = "darkorange",
                                Crude = "firebrick")) +
  labs(title = "Bias by estimator",
       x = NULL, y = "Bias (estimate - truth)") +
  guides(fill = "none")
```

### Coverage Comparison

```{r coverage-plot}
#| label: coverage-plot
#| fig-height: 4

cov_df <- data.frame(
  method = c("TMLE", "IPTW", "Crude"),
  coverage = c(tmle_row$coverage, iptw_row$coverage, crude_row$coverage),
  n_reps = c(tmle_row$n_reps, iptw_row$n_reps, crude_row$n_reps)
)
cov_df$method <- factor(cov_df$method, levels = c("TMLE", "IPTW", "Crude"))
# Wilson CI for coverage proportion
cov_df$ci_lo <- cov_df$coverage - 1.96 * sqrt(cov_df$coverage * (1 - cov_df$coverage) / cov_df$n_reps)
cov_df$ci_hi <- cov_df$coverage + 1.96 * sqrt(cov_df$coverage * (1 - cov_df$coverage) / cov_df$n_reps)

ggplot(cov_df, aes(x = method, y = coverage, fill = method)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0.2) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.5, y = 0.95, label = "95% nominal",
           hjust = -0.1, vjust = -0.5, color = "grey40", size = 3.5) +
  scale_fill_manual(values = c(TMLE = "steelblue", IPTW = "darkorange",
                                Crude = "firebrick")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  labs(title = "95% CI coverage by estimator",
       subtitle = "Error bars: Monte Carlo uncertainty in coverage estimate",
       x = NULL, y = "Coverage") +
  guides(fill = "none")
```

### SE Calibration: TMLE

A well-calibrated SE estimator should produce $\text{mean}(\widehat{\text{SE}})
\approx \text{empirical SD}$ of the point estimates. If the IC-based SE
systematically underestimates variability, coverage will fall below nominal.

```{r se-calibration}
#| label: se-calibration

if (!is.na(tmle_row$mean_se) && !is.na(tmle_row$emp_sd)) {
  ratio <- tmle_row$mean_se / tmle_row$emp_sd
  cat(sprintf("TMLE SE calibration:\n"))
  cat(sprintf("  Empirical SD of estimates:  %.4f\n", tmle_row$emp_sd))
  cat(sprintf("  Mean IC-based SE:           %.4f\n", tmle_row$mean_se))
  cat(sprintf("  Ratio (mean SE / emp SD):   %.3f\n", ratio))
  if (ratio > 0.90 && ratio < 1.10) {
    cat("  Interpretation: well-calibrated (ratio near 1.0)\n")
  } else if (ratio < 0.90) {
    cat("  Interpretation: SE underestimates variability; coverage may be below nominal\n")
  } else {
    cat("  Interpretation: SE overestimates variability; CIs are conservative\n")
  }
}
```

```{r se-scatter}
#| label: se-scatter
#| fig-height: 4

tmle_ok <- res[res$tmle_ok == TRUE, ]
if (nrow(tmle_ok) > 0) {
  ggplot(tmle_ok, aes(x = tmle_se, y = abs(tmle_est - truth$RD_true))) +
    geom_point(alpha = 0.4, color = "steelblue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    labs(title = "TMLE: estimated SE vs absolute error",
         subtitle = "Points near the diagonal indicate well-calibrated SEs",
         x = "IC-based SE", y = "|Estimate - Truth|")
}
```

### Overlap Severity and Estimator Error

```{r overlap-vs-error}
#| label: overlap-vs-error
#| fig-height: 5

if (nrow(tmle_ok) > 0) {
  err_df <- data.frame(
    fraction_extreme = rep(tmle_ok$fraction_extreme, 3),
    abs_error = c(
      abs(tmle_ok$tmle_est - truth$RD_true),
      abs(tmle_ok$iptw_est - truth$RD_true),
      abs(tmle_ok$crude_est - truth$RD_true)
    ),
    method = rep(c("TMLE", "IPTW", "Crude"), each = nrow(tmle_ok))
  )
  err_df$method <- factor(err_df$method, levels = c("TMLE", "IPTW", "Crude"))

  ggplot(err_df, aes(x = fraction_extreme, y = abs_error, color = method)) +
    geom_point(alpha = 0.3, size = 1.5) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    scale_color_manual(values = c(TMLE = "steelblue", IPTW = "darkorange",
                                   Crude = "firebrick")) +
    labs(title = "Absolute error vs overlap severity",
         x = "Fraction of observations with extreme PS",
         y = "|Estimate - Truth|",
         color = "Estimator")
}
```

This plot shows how each estimator's error relates to the severity of the
overlap problem within each replicate. Crude error is large regardless of
overlap (confounding bias dominates). IPTW error increases with overlap
severity (extreme weights amplify noise). TMLE error stays low even as
overlap worsens, demonstrating its robustness through outcome model learning.

## Discussion

### Why TMLE succeeds here

TMLE's double robustness means it is consistent if *either* $Q$ or $g$ is
correctly specified. In this DGP:

1. The outcome model $Q(A,W)$ is a logistic regression with a mild
   nonlinearity ($0.3 W_1^2$), which is well-captured by GLM or GAM
   learners in the SuperLearner ensemble.

2. The variable driving poor overlap ($W_4$) has a strong coefficient in
   the treatment model ($`r params$coef_W4_g`$) but a near-zero coefficient
   in the outcome model ($`r params$coef_W4_Q`$). This means the outcome
   surface is *flat* in the direction of PS extremity. The targeting step
   in TMLE, which updates the initial $Q$ estimate using the clever
   covariate $H = A/g - (1-A)/(1-g)$, can make small, well-behaved
   corrections even when $g$ is near the boundary.

3. The IC-based standard error accounts for the double robustness and
   remains well-calibrated because the influence curve is not inflated in
   the extreme-PS region (where $Q$ is smooth).

### Why IPTW struggles

IPTW relies entirely on the treatment model $g(W)$. When propensity scores
are near 0 or 1:

- Inverse probability weights become very large for a small number of
  observations, inflating variance.
- The effective sample size shrinks, reducing the amount of information
  available for estimation.
- Bootstrap CIs may under-cover because the bootstrap distribution of the
  Hajek estimator can be irregular with extreme weights.

### Why crude RD fails

The unadjusted estimator ignores confounding entirely. Covariates $W_1$
through $W_4$ affect both treatment assignment and (to varying degrees) the
outcome, inducing selection bias. The confounding bias dominates the sampling
variability, producing CIs that never contain the truth.

### Implications for practice

This scenario illustrates a realistic situation in observational studies:

1. **Stage 2 diagnostics would flag a problem.** A clean-room analysis would
   detect the poor overlap before any outcomes are examined, alerting analysts
   to the positivity concern.

2. **TMLE provides a path forward** when the outcome model is well-specified
   and the positivity violation is *practical* (no structural zeros). The
   analyst should invest in a rich, correctly-specified $Q$ library.

3. **IPTW alone is insufficient** under near-positivity violations. Truncation
   helps but does not fully address the instability.

4. **Unadjusted analyses are not a fallback.** Confounding bias makes crude
   comparisons unreliable regardless of overlap quality.

## Appendix: Replicate-Level Data

```{r replicate-table}
#| label: replicate-table

# Show first 10 replicates as a sample
display_cols <- c("rep", "event_rate", "fraction_extreme", "ess_frac",
                  "max_w", "overlap_flag", "tmle_est", "crude_est", "iptw_est")
knitr::kable(head(res[, display_cols], 10), digits = 4,
             caption = "First 10 replicates (sample)")
```

### Session Info

```{r session-info}
#| label: session-info

si_file <- file.path(out_dir, "sessionInfo.txt")
if (file.exists(si_file)) {
  cat(readLines(si_file), sep = "\n")
} else {
  sessionInfo()
}
```
