---
title: "Does plasmode-based candidate selection improve TMLE estimator performance? A simulation study"
subtitle: "DRAFT — placeholders for full-study results"
author:
  - name: "Andrew Mertens"
    affiliation: "University of California, Berkeley"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    self-contained: true
    code-fold: true
  docx:
    toc: true
    toc-depth: 3
    number-sections: true
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

# Abstract {.unnumbered}

**Background.** Targeted maximum likelihood estimation (TMLE) requires the analyst to prespecify a SuperLearner library, a propensity-score truncation level, and a targeting routine. The traditional choice is to fix these *a priori*. The outcome-blind plasmode candidate-selection approach in cleanTMLE selects them from a small candidate grid using a synthetic outcome generated from the observed covariate and treatment structure. Whether plasmode selection actually improves operating characteristics, or simply adds compute, is an open empirical question.

**Methods.** We simulate from five data-generating processes (DGPs) that share a common baseline-covariate distribution but vary in how a linear outcome model approximates the truth: a linear DGP, a smooth nonlinear DGP, a DGP with strong two-way interactions, a sparse DGP in which only one covariate matters, and a high-dimensional-noise DGP with 20 nuisance covariates added to the adjustment set. At each simulated dataset we run two fixed-library TMLE estimators (a parametric library and a rich SuperLearner library, both with truncation 0.01) and a plasmode-selected TMLE that picks a candidate from a 6-element grid spanning library tier and PS truncation. We compare bias, RMSE, 95% CI coverage, and the candidate-selection frequencies across DGPs.

**Results.** At a pilot budget of 10 outer replicates per DGP, n = 500, and 10 inner plasmode replicates, the four workflows (fixed-parametric, fixed-rich, plasmode min-RMSE, plasmode FIORD-two-stage) produce RD estimates that differ by less than 0.005 in RMSE on every DGP. All workflows are biased toward zero by 0.018–0.026 on the three mis-specified DGPs (smooth nonlinear, interactions, sparse) and are essentially unbiased on the linear and high-dim-noise DGPs. The plasmode min-RMSE selector does respond to DGP structure — it selects screener-augmented candidates on 5 of 10 high-dim-noise replicates and 5 of 10 interactions replicates, and selects parametric-plus-glmnet on 4 of 10 linear replicates — but at this budget the selection does not translate into a detectably different point estimate from the fixed-rich workflow. The FIORD two-stage selector reduces to fixed-parametric at this inner_reps budget (it selects `param_t01` on 7–8 of 10 replicates for every DGP except smooth-nonlinear).

**Conclusions.** The pilot supports a guarded message: *at modest sample sizes (n ≈ 500) and modest inner-plasmode budgets (≤ 10 inner reps), library choice is operationally indifferent and plasmode candidate selection does not change the resulting RD estimate*. Selection machinery does not equal selection benefit. Before drawing inferential conclusions a full-budget run at larger n_per_rep and inner_reps is required; we delay that run pending package-side changes (a richer Q0 surface for the inner plasmode generator and a coverage-stability fix to the FIORD selector) and a fresh pilot.

---

# Introduction

In observational comparative-effectiveness research, the analyst running a TMLE pipeline faces three small but consequential choices before the comparative effect is computed: which SuperLearner library to use for the outcome (Q) and propensity-score (g) regressions, where to truncate g to avoid practical positivity violations, and whether to cross-fit. Phillips and colleagues [@phillips2023sl] and Gruber and colleagues [@gruber2022sap] argue that these choices should be **prespecified** but offer little guidance on *how* to prespecify them when the analyst does not yet know the truth.

The cleanTMLE package [@cleantmle2026] operationalises an alternative: rather than pick a single library and truncation a priori, the analyst declares a *grid of candidate specifications*, evaluates the grid on an outcome-blind plasmode simulation derived from the realised covariate and treatment structure, and locks the best-performing candidate before unblinding. This procedure is FIORD-inspired [@nance2026fiord] and respects outcome-blindness in the sense that the inner plasmode never inspects the observed treatment-outcome association.

The procedure is intuitive and the cost is modest, but the empirical question is open: **does plasmode-based candidate selection actually improve TMLE operating characteristics relative to a fixed-library default**, and if so, under which DGPs?

This paper answers that question with a five-DGP simulation study. We hold the propensity-score model fixed across DGPs and vary only the outcome-generation mechanism so that any differences in operating characteristics can be attributed to the candidate-selection step.

# Methods

```{r setup}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
# Load pilot results if the full study has not yet been run.
sim_path_full  <- "../results/sim_full.rds"
sim_path_pilot <- "../results/sim_pilot.rds"
sim_path <- if (file.exists(sim_path_full)) sim_path_full else sim_path_pilot
has_sim  <- file.exists(sim_path)
sim_label <- if (identical(sim_path, sim_path_full)) "full study"
              else "pilot run"
if (has_sim) sim <- readRDS(sim_path)
true_RD <- -0.05
```

## Data-generating processes

The covariate distribution is shared across DGPs and consists of six baseline variables: age (mean 60, SD 10), sex (Bernoulli 0.5), a continuous biomarker (standard normal), a binary comorbidity flag (prevalence 0.30), a continuous BMI z-score (standard normal), and a binary smoking indicator (prevalence 0.25). The propensity score is logit-linear in these six variables with coefficients chosen so the marginal treatment prevalence is approximately 0.5.

The five DGPs differ only in the outcome model:

1. **Linear.** $\log\frac{p_{Y=1}}{1-p_{Y=1}}$ is linear in the six baseline variables. The fixed parametric library is correctly specified.
2. **Smooth nonlinear.** The biomarker and BMI enter the outcome with quadratic and sinusoidal terms respectively. A linear GLM is mis-specified but glmnet with a basis expansion would not be.
3. **Interactions.** Two strong two-way interactions enter the outcome (biomarker × comorbidity and BMI × smoking). Main-effects-only GLMs miss them; tree-based learners pick them up.
4. **Sparse.** Only the biomarker matters; the other five baseline variables are noise. Lasso-style learners shrink them away; an unregularised GLM does not.
5. **High-dimensional noise.** Same outcome model as Linear but with 20 additional standard-normal noise covariates added to the adjustment set. The fixed parametric library now over-fits.

The true marginal RD is fixed at −0.05 across all five DGPs.

## Workflows compared

We compare three workflows. Each takes a simulated dataset and returns a point estimate of the marginal RD, its IF-based SE, and the 95% CI.

- **Fixed-parametric TMLE.** SuperLearner library = `(SL.glm, SL.mean)`, truncation = 0.01. Represents the traditional reflex of an applied analyst who fits a logistic GLM by default.
- **Fixed-rich TMLE.** SuperLearner library = `(SL.glm, SL.glmnet, SL.gam, SL.ranger, SL.mean)`, truncation = 0.01. Represents an analyst who has read the workshop guidance and uses a rich library a priori without further tuning.
- **Plasmode-selected TMLE.** The cleanTMLE candidate grid (`build_candidate_grid()`) contains six candidates: a parametric tier, a parametric+glmnet tier, a rich tier at two truncation levels, and a rich+corP-screener tier at two truncation levels. For each simulated outer dataset we run a 50-replicate inner plasmode at effect size −0.05, pick the candidate with the smallest mean RMSE on the inner plasmode (the *min-RMSE* rule), and fit that candidate on the outer real outcome. We also report a second selector — the FIORD two-stage rule that screens candidates on oracle coverage before picking the smallest mean SE — for sensitivity.

Each workflow uses its own library and truncation throughout — so the fixed-parametric workflow fits both the propensity score and the outcome model with `(SL.glm, SL.mean)`, the fixed-rich workflow fits both with the rich library, and the plasmode workflow fits both nuisance functions with the library prescribed by the candidate it selected on the inner plasmode loop. Truncation likewise follows the locked candidate. The inner plasmode loop itself uses a baseline GLM Q0 (covariates only, outcome-blind) regardless of the candidate grid; only the inner candidate fits use the candidate-specific libraries.

## Performance metrics

For each workflow and DGP we report bias, RMSE, 95% CI coverage (with Monte Carlo standard error), the empirical SD, the mean reported SE, and the ratio of mean SE to empirical SD as a calibration diagnostic. For the plasmode-selected workflow we additionally report the empirical frequency with which each candidate was selected (per DGP), which is the diagnostic that tells us whether plasmode selection is responding to the DGP at all.

## Replicate budget

The full study uses `n_per_rep = 1000` and `n_reps = 500` outer replicates per DGP, plus 50-replicate inner plasmode loops at each outer replicate. The Monte Carlo standard error of the 95% coverage estimate is approximately 0.010. The full study should be run on a workstation or cluster (~6–24 hours wall-clock depending on parallelism).

# Results

```{r results-prep, eval = has_sim}
sim <- readRDS(sim_path)
sim$bias <- sim$estimate - true_RD
sim$covers <- (sim$ci_lower <= true_RD) & (sim$ci_upper >= true_RD)
sim_label_caption <- paste0(
  "(Source: ", basename(sim_path), "; ",
  length(unique(sim$rep)), " replicates per DGP × workflow; ",
  "results are pilot-quality and **not yet inferentially defensible**.)"
)
```

::: {.callout-warning}
The numbers below are populated from a **pilot run** unless the full-study results file `results/sim_full.rds` is present. The pilot uses 10 outer replicates per DGP at n = 500 (with 10 inner plasmode replicates) and is intended only to validate the simulation pipeline and serve as a placeholder. Re-render after running `scripts/run_full.R` for the paper-quality version.
:::

## Headline operating characteristics

```{r headline-table, eval = has_sim, results = "asis"}
summary_tbl <- sim |>
  group_by(dgp, workflow) |>
  summarise(
    n_valid  = sum(!is.na(estimate)),
    bias     = mean(bias, na.rm = TRUE),
    abs_bias = mean(abs(bias), na.rm = TRUE),
    emp_sd   = sd(estimate, na.rm = TRUE),
    mean_se  = mean(se, na.rm = TRUE),
    se_sd    = mean(se, na.rm = TRUE) / sd(estimate, na.rm = TRUE),
    rmse     = sqrt(mean(bias^2, na.rm = TRUE)),
    coverage = mean(covers, na.rm = TRUE),
    mc_se_cov = sqrt(mean(covers, na.rm = TRUE) *
                      (1 - mean(covers, na.rm = TRUE)) /
                      sum(!is.na(estimate))),
    .groups = "drop"
  ) |>
  arrange(dgp, workflow)
knitr::kable(
  summary_tbl,
  digits = c(NA, NA, 0, 4, 4, 4, 4, 3, 4, 3, 4),
  caption = paste0("Bias, RMSE, and 95% CI coverage by DGP and workflow. ",
                    sim_label_caption)
)
```

[INTERPRETATION TO BE WRITTEN AFTER FULL RUN. Anticipated patterns, to be confirmed:

- **Linear DGP.** All three workflows are unbiased and well-calibrated. Plasmode adds compute but not statistical advantage.
- **Smooth nonlinear and Interactions DGPs.** Fixed-parametric is biased and undercovers; fixed-rich and plasmode-selected both recover the truth, with plasmode-selected slightly cheaper in the sense that it does not always engage the heaviest learners.
- **Sparse DGP.** Fixed-parametric pays the price of unregularised regression; plasmode-selected and fixed-rich agree, with both routinely picking glmnet-augmented candidates.
- **High-dimensional noise DGP.** Fixed-parametric over-fits and undercovers; plasmode-selected picks the screener-augmented candidate. The fixed-rich library, without a screener, also struggles.]

## Candidate-selection frequencies

```{r selection-table, eval = has_sim, results = "asis"}
all_cands <- c("param_t01", "paramplus_t01",
               "rich_t01", "rich_t05",
               "rich_screener_t01", "rich_screener_t05")
pick <- sim |>
  filter(workflow == "plasmode_min_rmse") |>
  # strip the *_FALLBACK suffix so fallbacks are counted as the
  # underlying candidate (with a separate column for transparency)
  mutate(
    fallback = grepl("_FALLBACK$", selected_candidate),
    sel      = factor(sub("_FALLBACK$", "", selected_candidate),
                      levels = all_cands)
  ) |>
  group_by(dgp, sel) |>
  summarise(n = n(), .groups = "drop") |>
  tidyr::pivot_wider(names_from = sel, values_from = n,
                     values_fill = 0L, names_expand = TRUE)
knitr::kable(
  pick,
  caption = paste0("Empirical candidate-selection frequencies by DGP ",
                    "under the min-RMSE plasmode rule. Rows are DGPs; ",
                    "columns are the six candidates in the grid. ",
                    sim_label_caption)
)
```

[The candidate-selection cross-tab is the diagnostic that tells us whether plasmode selection is responding to the DGP. If the selected candidate is identical across DGPs, plasmode is not adapting; if it shifts in the expected direction (glmnet under nonlinear / sparse, screener under high-dim-noise), the selection is doing real work.]

## Forest plot of replicate-level RD estimates

```{r forest-plot, eval = has_sim, fig.height = 8, fig.cap = "Replicate-level RD estimates by workflow and DGP. Dashed red line marks the true RD (-0.05). Workflows are arranged top-to-bottom: fixed-parametric, fixed-rich, plasmode (min-RMSE)."}
fp <- sim |>
  filter(workflow %in% c("fixed_parametric", "fixed_rich",
                          "plasmode_min_rmse"))
ggplot(fp, aes(x = estimate, y = workflow)) +
  geom_vline(xintercept = true_RD, linetype = "dashed",
              colour = "red") +
  geom_jitter(alpha = 0.4, size = 1, height = 0.15) +
  facet_wrap(~ dgp, ncol = 1, scales = "free_y") +
  labs(x = "RD estimate", y = NULL) +
  theme_minimal(base_size = 10)
```

## FIORD two-stage selector as sensitivity

```{r fiord-table, eval = has_sim, results = "asis"}
fiord_compare <- sim |>
  filter(workflow %in% c("plasmode_min_rmse",
                          "plasmode_fiord_two_stage")) |>
  group_by(dgp, workflow) |>
  summarise(
    abs_bias = mean(abs(bias), na.rm = TRUE),
    rmse     = sqrt(mean(bias^2, na.rm = TRUE)),
    coverage = mean(covers, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(fiord_compare, digits = 4,
  caption = paste0("Plasmode-selected workflow under two selection ",
                    "rules. ", sim_label_caption))
```

In the pilot the FIORD two-stage rule collapses onto the parametric candidate (`param_t01`) on 7--8 of 10 replicates for every DGP except smooth-nonlinear. The mechanism is mechanical: the coverage estimate per candidate has Monte Carlo SE roughly $\sqrt{0.5 \cdot 0.5 / r_{\text{inner}}}$, so at $r_{\text{inner}} = 10$ the MC SE is 0.158 — wider than the $\pm 0.02$ tolerance the FIORD screen uses. With coverage unable to discriminate, the second stage picks the candidate with the smallest mean SE, which is almost always the parametric library. We recommend FIORD-style selection only with $r_{\text{inner}} \geq 30$.

# Discussion

The pilot's central finding is that *at modest sample sizes the choice of TMLE library does not matter very much*. Fixed-parametric, fixed-rich, and plasmode-selected TMLE produce point estimates that agree to within Monte Carlo noise, and all four workflows inherit the same bias on the three DGPs where the outcome model is mis-specified. Plasmode candidate selection therefore does not pay for itself here in operating characteristics, though the selection cross-tab does demonstrate that the selector is responding to DGP structure — a necessary, but not sufficient, condition for selection to add value.

Three caveats temper this conclusion before a full study is launched.

First, the pilot's n = 500 is small relative to what flexible learners need. Random forests and generalized additive models need sample sizes in the low thousands per learner before their bias advantage exceeds their variance penalty against an unregularised GLM. A full run at $n \geq 1000$ may separate the workflows.

Second, the inner plasmode loop in cleanTMLE fits the baseline outcome model Q0 as a logistic GLM by default, which biases the synthetic outcomes — and therefore candidate selection — toward learners that handle linear-in-logit structure well. Since cleanTMLE 0.1.3 a `q0_library` argument lets the user pass a richer SuperLearner library; this paper's full run will use it.

Third, the SE-to-empirical-SD calibration ratio drops below 1 on the smooth-nonlinear DGP (0.69--0.74 across all four workflows). Influence-function variance estimates may therefore *under-state* uncertainty in the regime where outcome-model mis-specification is real. The full study should report this calibration ratio routinely; a robust or bootstrap variance option is the natural sensitivity.

What this pilot says about study design (a Statistical Analysis Plan question): if the planned analytic sample is below roughly 1000 and the analyst has any reason to suspect mis-specification, neither a fixed-rich library nor a plasmode-selected library will rescue an underpowered estimator. The honest move is to acknowledge that library choice is operationally indifferent at small n and to pre-register a single, defensible library rather than a costly grid search whose result does not move the point estimate.

# Limitations

- The pilot uses only 10 outer replicates per DGP, n = 500, and 10 inner plasmode replicates. The Monte Carlo SE of coverage at this budget is roughly 0.16, so coverage differences are not yet inferentially defensible. The full study (`scripts/run_full.R`) is deferred pending the package-side changes flagged in the Discussion.
- The FIORD two-stage rule collapses to a parametric pick at inner_reps = 10; we recommend it only with inner_reps $\geq$ 30.
- Plasmode candidate selection is driven by the outcome model's RMSE on the inner loop, but a candidate's PS library and truncation are bundled with the Q library. A separable selection rule for g vs Q is a natural extension.
- The DGPs span outcome-model mis-specification but not positivity strain. A positivity-strained sixth DGP is a natural extension; we excluded it to keep the focus.
- The inner plasmode generator in cleanTMLE uses a logistic-GLM Q0 by default, biasing candidate selection toward linear-in-logit learners. Since cleanTMLE 0.1.3 a `q0_library` argument accepts a richer SuperLearner library; the full study will use it (the pilot did not, so its selection inherits the linear-in-logit bias).
- Variance is computed with IF-based SEs throughout. The se_sd ratio < 1 finding on smooth-nonlinear motivates a robust or bootstrap variance sensitivity in the full study.

# References

::: {#refs}
:::
